### 2018-07-16 mscthesis.tex
init

\documentclass[m,times]{cgMA}

\usepackage{listings}
\usepackage[hyphens]{url} 
\usepackage{hyperref}
\usepackage{breakurl}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{color}
\usepackage{subfigure}
\usepackage{wrapfig}
\lstdefinelanguage{GLSL}
{
sensitive=true,
morekeywords=[1]{
attribute, const, uniform, varying,
layout, centroid, flat, smooth,
noperspective, break, continue, do,
for, while, switch, case, default, if,
else, in, out, inout, float, int, void,
bool, true, false, invariant, discard,
return, mat2, mat3, mat4, mat2x2, mat2x3,
mat2x4, mat3x2, mat3x3, mat3x4, mat4x2,
mat4x3, mat4x4, vec2, vec3, vec4, ivec2,
ivec3, ivec4, bvec2, bvec3, bvec4, uint,
uvec2, uvec3, uvec4, lowp, mediump, highp,
precision, sampler1D, sampler2D, sampler3D,
samplerCube, sampler1DShadow,
sampler2DShadow, samplerCubeShadow,
sampler1DArray, sampler2DArray,
sampler1DArrayShadow, sampler2DArrayShadow,
isampler1D, isampler2D, isampler3D,
isamplerCube, isampler1DArray,
isampler2DArray, usampler1D, usampler2D,
usampler3D, usamplerCube, usampler1DArray,
usampler2DArray, sampler2DRect,
sampler2DRectShadow, isampler2DRect,
usampler2DRect, samplerBuffer,
isamplerBuffer, usamplerBuffer, sampler2DMS,
isampler2DMS, usampler2DMS,
sampler2DMSArray, isampler2DMSArray,
usampler2DMSArray, struct},
morekeywords=[2]{
radians,degrees,sin,cos,tan,asin,acos,atan,
atan,sinh,cosh,tanh,asinh,acosh,atanh,pow,
exp,log,exp2,log2,sqrt,inversesqrt,abs,sign,
floor,trunc,round,roundEven,ceil,fract,mod,modf,
min,max,clamp,mix,step,smoothstep,isnan,isinf,
floatBitsToInt,floatBitsToUint,intBitsToFloat,
uintBitsToFloat,length,distance,dot,cross,
normalize,faceforward,reflect,refract,
matrixCompMult,outerProduct,transpose,
determinant,inverse,lessThan,lessThanEqual,
greaterThan,greaterThanEqual,equal,notEqual,
any,all,not,textureSize,texture,textureProj,
textureLod,textureOffset,texelFetch,
texelFetchOffset,textureProjOffset,
textureLodOffset,textureProjLod,
textureProjLodOffset,textureGrad,
textureGradOffset,textureProjGrad,
textureProjGradOffset,texture1D,texture1DProj,
texture1DProjLod,texture2D,texture2DProj,
texture2DLod,texture2DProjLod,texture3D,
texture3DProj,texture3DLod,texture3DProjLod,
textureCube,textureCubeLod,shadow1D,shadow2D,
shadow1DProj,shadow2DProj,shadow1DLod,
shadow2DLod,shadow1DProjLod,shadow2DProjLod,
dFdx,dFdy,fwidth,noise1,noise2,noise3,noise4,
EmitVertex,EndPrimitive},
morekeywords=[3]{
gl_VertexID,gl_InstanceID,gl_Position,
gl_PointSize,gl_ClipDistance,gl_PerVertex,
gl_Layer,gl_ClipVertex,gl_FragCoord,
gl_FrontFacing,gl_ClipDistance,gl_FragColor,
gl_FragData,gl_MaxDrawBuffers,gl_FragDepth,
gl_PointCoord,gl_PrimitiveID,
gl_MaxVertexAttribs,gl_MaxVertexUniformComponents,
gl_MaxVaryingFloats,gl_MaxVaryingComponents,
gl_MaxVertexOutputComponents,
gl_MaxGeometryInputComponents,
gl_MaxGeometryOutputComponents,
gl_MaxFragmentInputComponents,
gl_MaxVertexTextureImageUnits,
gl_MaxCombinedTextureImageUnits,
gl_MaxTextureImageUnits,
gl_MaxFragmentUniformComponents,
gl_MaxDrawBuffers,gl_MaxClipDistances,
gl_MaxGeometryTextureImageUnits,
gl_MaxGeometryOutputVertices,
gl_MaxGeometryOutputVertices,
gl_MaxGeometryTotalOutputComponents,
gl_MaxGeometryUniformComponents,
gl_MaxGeometryVaryingComponents,gl_DepthRange},
morecomment=[l]{//},
morecomment=[s]{/*}{*/},
morecomment=[l][keywordstyle4]{\#},
}
\lstset{
tabsize=2,
rulecolor=,
basicstyle=\scriptsize,
upquote=true,
aboveskip={0.5\baselineskip},
belowskip={1.5\baselineskip},
columns=fixed,
showstringspaces=false,
extendedchars=true,
breaklines=true,
prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
frame=single,
showtabs=false,
showspaces=false,
showstringspaces=false,
identifierstyle=\ttfamily,
keywordstyle=\color[rgb]{1.0,0,0},
keywordstyle=[1]\color[rgb]{0,0,0.75},
keywordstyle=[2]\color[rgb]{0.5,0.0,0.0},
keywordstyle=[3]\color[rgb]{0.127,0.427,0.514},
keywordstyle=[4]\color[rgb]{0.4,0.4,0.4},
commentstyle=\color[rgb]{0.133,0.545,0.133},
stringstyle=\color[rgb]{0.639,0.082,0.082},
numbers=left,
}

\renewcommand{\lstlistingname}{Source code}

\renewcommand{\lstlistlistingname}{List of source codes}

\begin{document}
\selectlanguage{english}
\bibliographystyle{alphadin}
\author{Fabian Meyer}
\title{GPU-Beschleunigung der Material Point Method}
\zweitgutachter{Bastian Krayer, M.Sc.}
\zweitgutachterInfo{(Institut f{\"u}r Computervisualistik, AG Computergraphik)}
% Umschalten der Sprache (fuer englische Rubrikbezeichnungen etc.)

\pagenumbering{roman}
\maketitle

\clearpage

\subsubsection*{Abstract}

\noindent Physic simulations allow the creation of dynamic scenes on the computer. Computer generated images become lively and find use in movies, games and engineering applications. GPGPU techniques make use of the graphics card to simulate physics. The simulation of dynamic snow is still little researched. The Material Point Method is the first technique which is capable of showing the dynamics and characteristics of snow.\\
The hybrid use of Lagrangian particles and a regular cartesian grid enables solving of partial differential equations. Therefore particles are transformed to the grid. The grid  velocities can be updated with the calculation of gradients in an FEM-manner (finite element method). Finally grid node velocities are weight back to the particles to move them across the scene. This method is coupled with a constitutive model to cover the dynamic nature of snow. This includes collisions and breaking.\\
This bachelor thesis connects the recent developments in GPGPU techniques of OpenGL with the Material Point Method to efficiently simulate visually compelling, dynamic snow scenes.

%\vfill~
%ggf. noch'n \vfill~

\clearpage
\tableofcontents

%\setcounter{page}{3}
\clearpage         % oder \cleardoublepage bei zweiseitigem Druck
% \listoffigures   % fuer ein eventuelles Abbildungsverzeichnis
% \clearpage
\pagenumbering{arabic}

\section{Introduction}\label{intro}
bib-test \cite{Stomakhin:2013:MPM:2461912.2461948}
$$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
\section{Problems of a full GPU approach}
\section{SoA vs AoS}
\section{Sparse Datastructures}
\section{Validation of conservation equations}
\listoffigures
\clearpage
\listoftables
\lstlistoflistings
\clearpage
\bibliography{mscthesis}
\end{document}

### 2018-09-26 mscthesis.tex
lit add, bibtex -> biblatex

@@
 \usepackage{listings}
-\usepackage[hyphens]{url} 
+\usepackage{xurl}
+\urlstyle{sf}
 \usepackage{hyperref}
-\usepackage{breakurl}
 \usepackage{amsmath}
@@
 \usepackage{wrapfig}
+\usepackage[english,ngerman]{babel}
+\usepackage[backend=biber,style=alphabetic]{biblatex}
+\addbibresource{mscthesis.bib}
 \lstdefinelanguage{GLSL}
@@
 \begin{document}
-\selectlanguage{english}
-\bibliographystyle{alphadin}
+
 \author{Fabian Meyer}
@@
 \clearpage
-
+\selectlanguage{english}
 \subsubsection*{Abstract}
 
-\noindent Physic simulations allow the creation of dynamic scenes on the computer.
-Computer generated images become lively and find use in movies, games and engineering applications.
-GPGPU techniques make use of the graphics card to simulate physics.
-The simulation of dynamic snow is still little researched.
-The Material Point Method is the first technique which is capable of showing the dynamics and characteristics of snow.\\
-The hybrid use of Lagrangian particles and a regular cartesian grid enables solving of partial differential equations.
-Therefore particles are transformed to the grid.
-The grid  velocities can be updated with the calculation of gradients in an FEM-manner (finite element method).
-Finally grid node velocities are weight back to the particles to move them across the scene.
-This method is coupled with a constitutive model to cover the dynamic nature of snow.
-This includes collisions and breaking.\\
-This bachelor thesis connects the recent developments in GPGPU techniques of OpenGL with the Material Point Method to efficiently simulate visually compelling, dynamic snow scenes.
+\noindent Physic
 
@@
 
-\section{Introduction}\label{intro}
-bib-test \cite{Stomakhin:2013:MPM:2461912.2461948}
+\section{Introduction} \label{intro}
+bib-test \cite{MPM:SNOW}
 $$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
-\section{Problems of a full GPU approach}
-\section{SoA vs AoS}
-\section{Sparse Datastructures}
-\section{Validation of conservation equations}
+\subsection{Motivation}
+\subsection{Related Work}
+\cite{PIC:GPU}
+\cite{MPM:GPU}
+%look again for MPM/PIC gpu impl
+\section{Basics}
+\subsection{Continuum Mechanics}
+\cite{MPM:COURSE}
+\subsection{Constitutive Models}
+\cite{MPM:DRUCKER} %Sand Plasticity
+\cite{MPM:MULTI} %Sand+Water
+\cite{MPM:INVERT} %Neo-Hookean
+\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
+\cite{MPM:SHELLS}
+\subsubsection{Fixed Corotated Hyperelasticity}
+\cite{MPM:INVERT}
+\subsection{Singular Value Decomposition}
+\cite{SVD:3x3}
+\subsection{Material Point Method}
+\subsubsection{Conservation Equations}
+\cite{MPM:APIC}
+\subsubsection{CFL Condition}
+\subsection{GPU memory}
+\subsubsection{Optimization opportunities}
+\cite{NVIDIA:BEST:PRACTICE}
+\cite{AMD:GPU_OPEN}
+\begin{itemize}
+  \item Minimize CPU-GPU transfers
+  \item Coalesced memory access
+  \item Bank conflicts
+\end{itemize}
+\subsubsection{SoA vs. AoS}
+\section{Implementation}
+\subsection{Parallel Reduction}
+\cite{NVIDIA:PARALLEL_REDUCTION}
+\subsection{Verifying Simulation}
+\subsubsection{Validation of Conservation Equations}
+\section{Evaluation}
+\cite{NVIDIA:PEAK_PERFORMANCE}
+\cite{KHRONOS:TIMER_QUERY}
+\section{Conclusion}
+\subsection{Review}
+\subsection{Future Work}
 \listoffigures
@@
 \clearpage
-\bibliography{mscthesis}
+\selectlanguage{english}
+\printbibliography
 \end{document}

### 2018-09-26 mscthesis.tex
pseudo code

@@
 \addbibresource{mscthesis.bib}
+
 \lstdefinelanguage{GLSL}
@@
 }
-\lstset{
+\lstdefinestyle{GL}{
   tabsize=2,
@@
 }
+
+\lstdefinelanguage{alg}{
+  keywordstyle=[1]{\keywordstyle},
+  keywordstyle=[2]{\operatorstyle},
+  keywordstyle=[3]{\typestyle},
+  keywordstyle=[4]{\functionstyle},
+  identifierstyle={\identifierstyle},
+  keywords=[1]{%
+    struct,
+    begin,end,%
+    program,procedure,function,subroutine,%
+    while,do,for,to,next,repeat,until,loop,continue,endwhile,endfor,endloop,%
+    if,then,else,endif%
+  return},
+  keywords=[2]{mod},
+  keywords=[3]{Particle,int},
+  keywords=[4]{Euclidean},
+  literate={-}{$-$}1 {^}{$^\wedge$}1
+  {>}{{$>$\ }}1 {<}{{$<$\ }}1
+  {>=}{{$\geqslant$\ }}1 {<=}{{$\leqslant$\ }}1
+  {:=}{{$\gets$\ }}1 {!=}{{$\ne$\ }}1 {<>}{{$\ne$\ }}1
+  {->}{{$\;\to\;$}}1
+  {&&}{{\keywordstyle and\ }}4 {{||}}{{\keywordstyle or\ }}3
+  {;}{\hspace{0.2em};}2 {,}{\hspace{0.2em},}2,
+}
+
+\lstdefinestyle{Pseudo}{
+  language={alg},
+  columns=fullflexible,
+  numbers=left,
+  numberstyle=\scriptsize,
+}
+\newcommand\keywordstyle{\rmfamily\bfseries\upshape}
+\newcommand\operatorstyle{\rmfamily\mdseries\upshape}
+\newcommand\typestyle{\rmfamily\mdseries\upshape}
+\newcommand\functionstyle{\rmfamily\mdseries\scshape}
+\newcommand\identifierstyle{\rmfamily\mdseries\itshape}
 
@@
 \subsection{Motivation}
-\subsection{Related Work}
+\section{Related Work}
 \cite{PIC:GPU}
@@
 %look again for MPM/PIC gpu impl
+%SPH, FEM methods
 \section{Basics}
@@
 \cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
-\cite{MPM:SHELLS}
+\cite{MPM:SHELLS} %Thin Shells
+\cite{MPM:PHASE_CHANGE} %Heat-Model
+\cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule
 \subsubsection{Fixed Corotated Hyperelasticity}
@@
 \subsubsection{SoA vs. AoS}
+
+\begin{lstlisting}[style = Pseudo]
+struct Particle{
+  int val;
+};
+\end{lstlisting}
 \section{Implementation}
+
+\begin{lstlisting}[style =Pseudo]
+procedure Euclidean(a,b)
+  r := a mod b
+  while r != 0 do
+    a := b
+    b := r
+    r := a mod b
+  end while
+  return b
+end procedure
+\end{lstlisting}
 \subsection{Parallel Reduction}
 \cite{NVIDIA:PARALLEL_REDUCTION}
+\subsection{Affine particle-in-cell}
+\cite{MPM:APIC}
+\subsection{Iterative solvers for linear systems}
+%GMRES, Conjugate Gradients
 \subsection{Verifying Simulation}
-\subsubsection{Validation of Conservation Equations}
+\begin{itemize}
+  \item Conservation equations
+  \item Singular Value Decomposition
+  \item Parallel Reduction
+\end{itemize}
 \section{Evaluation}

### 2018-09-30 mscthesis.tex
or lets just use algorithmpseudocode
binning+scan

--- 2018-09-26 mscthesis.tex
@@
 \usepackage[backend=biber,style=alphabetic]{biblatex}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
 \addbibresource{mscthesis.bib}
+
 \lstdefinelanguage{GLSL}
@@
 }
-
 \renewcommand{\lstlistingname}{Source code}
@@
 \subsection{Motivation}
-\subsection{Related Work}
-\cite{PIC:GPU}
+\section{Related Work}
+\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}
+benefits over radix sort /thrust sort-by-key -> counting sort
+
 \cite{MPM:GPU}
 %look again for MPM/PIC gpu impl
+%SPH, FEM methods
 \section{Basics}
@@
 \section{Implementation}
+\newcommand{\var}[1]{{\ttfamily#1}}% variable
+\begin{algorithm}[t]
+  \caption{Euclid’s algorithm}\label{euclid}
+  \begin{algorithmic}[1]
+    \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
+    \State $r\gets a\bmod b$ and \var{foobar}\label{foobar}
+    \While{$r\not=0$}\Comment{We have the answer if r is 0}
+    \State $a\gets b$
+    \State $b\gets r$
+    \State $r\gets a\bmod b$
+    \EndWhile\label{euclidendwhile}
+    \State \textbf{return} $b$\Comment{The gcd is b}
+    \EndProcedure
+  \end{algorithmic}
+\end{algorithm}
 \subsection{Parallel Reduction}
 \cite{NVIDIA:PARALLEL_REDUCTION}
+\subsection{Binning}
+\cite{NVIDIA:BINNING}
+\subsection{Scanning}
+\cite{NVIDIA:SCAN}
+\subsection{Affine particle-in-cell}
+\cite{MPM:APIC}
+\subsection{Iterative solvers for linear systems}
+%GMRES, Conjugate Gradients
 \subsection{Verifying Simulation}
-\subsubsection{Validation of Conservation Equations}
+\begin{itemize}
+  \item Conservation equations
+  \item Singular Value Decomposition
+  \item Parallel Reduction
+\end{itemize}
 \section{Evaluation}
@@
 \subsection{Future Work}
+Particle Activation: \cite{MPM:GPU}
+
+Sparse: \cite{OPENVDB} \cite{NVIDIA:GVDB_VOXELS}
+
 \listoffigures

### 2018-11-08 mscthesis.tex
hyperelasticity theory

@@
 benefits over radix sort /thrust sort-by-key -> counting sort
+\cite{NVIDIA:SHUFFLE}
 
@@
 %SPH, FEM methods
+\section{Notation}
+Einstein-Notation
 \section{Basics}
@@
 \cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule
+
+The first aim will be to find the strain energy density $\psi(\mathbf{\epsilon})$ of the strain $\bm{\epsilon}$.
+For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\psi$ as:
+The most general linear stress-strain relationship is given by Hooke's Law in three dimensions
+\begin{equation}\label{eq:hook}
+  \sigma_{ij} = C_{ijkl} \epsilon_{kl},
+\end{equation}
+where $\mathbf{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.
+Assuming following symmetries reduces the tensors to 6 and 21 unique elements respectively:
+\begin{enumerate}
+  \item Conservation of angular momentum: $\sigma_{ij} = \sigma_{ji} \Rightarrow C_{ijkl} = C_{jikl} $
+  \item No-net-rotation: $\epsilon_{kl} = \epsilon_{lk} \Rightarrow C_{ijkl} = C_{ijlk}$
+  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\psi$:
+    \begin{equation}
+      C _ { i j k l } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },
+    \end{equation}
+\end{enumerate}
+where item \ref{it:mixed_partials} holds for the strain energy density functional of an (hyper-)elastic material. The stress may then also be calculated by
+\begin{equation}\label{eq:partial_energy}
+  \sigma _ { i j } = \sigma _ { i j } ( \mathbf{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },
+\end{equation}
+if such a $\psi$ is given.
+An isotropic (direction-independent) linear elastic material further only has three unique elements $C _ {i j k l}$.
+Using Voigt-Notation, which collapses indices $i=j$ and $k=l$, equation \ref{eq:hook} can be rewritten as:
+\begin{equation} \label{eq:voigt}
+  \left[ \begin{array} {c}
+      {\sigma_{11}}\\
+      {\sigma_{22}}\\
+      {\sigma_{33}}\\
+      {\sigma_{23}}\\
+      {\sigma_{13}}\\
+      {\sigma_{12}}\\
+  \end{array} \right]
+  = \left[ \begin{array} { c c c c c c }
+      { C _ { 11 }   } & { C _ { 12 } } & { C _ { 12 } } & 0                & 0                & 0 \\
+      {}               & { C _ { 11 } } & { C _ { 12 } } & 0                & 0                & 0 \\
+      {}               & {}             & { C _ { 11 } } & 0                & 0                & 0 \\
+      {}               & {}             & {}             & { C _ { 22 } }   & 0                & 0 \\
+      {}               & {}             & {}             & {}               & { C _ { 22 } }   & 0 \\
+      {sym}            & {}             & {}             & {}               & {}               & { C _ { 22 } } \\
+  \end{array} \right]
+  \left[ \begin{array} {c}
+      {\epsilon_{11}}\\
+      {\epsilon_{22}}\\
+      {\epsilon_{33}}\\
+      {\gamma_{23}}\\
+      {\gamma_{13}}\\
+      {\gamma_{12}}\\
+  \end{array} \right]
+\end{equation}
+Experimental results of Hooke's law commonly give
+\begin{equation}
+  \boldsymbol{\epsilon} =
+  \epsilon _ { i j } = \frac { 1 } { E } \left[ ( 1 + \nu ) \sigma _ { i j } - \nu \sigma _ { k k } \delta _ { i j } \right]
+  = \mathbf{C}^{-1} \bm{\sigma}
+\end{equation}
+using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\mathbf{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:
+\begin{equation} \label{eq:stress_strain}
+  \sigma_{ij} = 2\mu \epsilon_{ij} + \lambda \text{tr}(\boldsymbol{\epsilon}) \delta_{ij}.
+\end{equation}
+Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$ and $C_{22} = \mu$.
+$\gamma$ is also referred to as the engineering strain.
+Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:
+\begin{equation}\label{eq:linear_elas}
+  \psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).
+\end{equation}
+\cite{MIT:LINEAR_ELASTICITY}
 \subsubsection{Fixed Corotated Hyperelasticity}
+The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
+  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \mathbf { F } + \mathbf { F } ^ { T } \right) - \mathbf { I }
+\end{equation}
+The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
+to $\mathbf{F}$: $\psi(\mathbf{R}_{0} \mathbf{F}) \neq \psi(\mathbf{F})$.
+The deformation caused by stress however does not affect rigid body motions.
+Given the polar decomposition $\mathbf{F} = \mathbf{RS}$, an alternate strain measure may be defined as:
+\begin{equation}
+  \hat { \boldsymbol { \epsilon } } ( \mathbf { F } ) = \boldsymbol { \epsilon } \left( \mathbf { R } ^ { T } \mathbf { F } \right) = \frac { 1 } { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } + \left( \mathbf { R } ^ { T } \mathbf { F } \right) ^ { T } \right) - \mathbf { I } = \mathbf { S } - \mathbf { I }
+\end{equation}
+Substituting $\hat{\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity using rotation-invariance of Frobenius norm and trace:
+\begin{equation}
+  \Psi = \mu \| \mathbf { F } - \mathbf { R } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } - \mathbf { I } \right)
+\end{equation}
+
+\begin{equation}
+  \Psi(\boldsymbol{\Sigma}) = \mu \| \boldsymbol { \Sigma } - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma } - \mathbf { I } )
+\end{equation}
+\cite{ADAMS:ELASTICITY}
 \cite{MPM:INVERT}
+
 \subsection{Singular Value Decomposition}
@@
     \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-    \State $r\gets a\bmod b$ and \var{foobar}\label{foobar}
+    \State $r\gets a\mathbfod b$ and \var{foobar}\label{foobar}
     \While{$r\not=0$}\Comment{We have the answer if r is 0}
@@
     \State $b\gets r$
-    \State $r\gets a\bmod b$
+    \State $r\gets a\mathbfod b$
     \EndWhile\label{euclidendwhile}
@@
 \end{algorithm}
+\subsection{Framework}
+decorator pattern
 \subsection{Parallel Reduction}
@@
 \cite{NVIDIA:SCAN}
+\cite{NVIDIA:SCAN_MODERN}
 \subsection{Affine particle-in-cell}

### 2018-11-09 mscthesis.tex
singular value descr

@@
 \usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{mathtools}
+\usepackage{subdepth}
 \addbibresource{mscthesis.bib}
@@
 \cite{MPM:COURSE}
+\subsection{Singular Value Decomposition}
+\cite{SVD:3x3}
 \subsection{Constitutive Models}
@@
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
-  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \mathbf { F } + \mathbf { F } ^ { T } \right) - \mathbf { I }
+  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \mathbf { F }_E + \mathbf { F }_E ^ { T } \right) - \mathbf { I }
 \end{equation}
 The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
-to $\mathbf{F}$: $\psi(\mathbf{R}_{0} \mathbf{F}) \neq \psi(\mathbf{F})$.
+to $\mathbf{F}_E$: $\psi(\mathbf{R}_{0} \mathbf{F}_E) \neq \psi(\mathbf{F}_E)$.
 The deformation caused by stress however does not affect rigid body motions.
-Given the polar decomposition $\mathbf{F} = \mathbf{RS}$, an alternate strain measure may be defined as:
+Given the polar decomposition $\mathbf{F}_E = \mathbf{R}_E\mathbf{S}_E$, an alternate strain measure may be defined as:
 \begin{equation}
-  \hat { \boldsymbol { \epsilon } } ( \mathbf { F } ) = \boldsymbol { \epsilon } \left( \mathbf { R } ^ { T } \mathbf { F } \right) = \frac { 1 } { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } + \left( \mathbf { R } ^ { T } \mathbf { F } \right) ^ { T } \right) - \mathbf { I } = \mathbf { S } - \mathbf { I }
+  \hat { \boldsymbol { \epsilon } } ( \mathbf { F }_E ) = \boldsymbol { \epsilon } \left( \mathbf { R }_E ^ { T } \mathbf { F }_E \right) = \frac { 1 } { 2 } \left( \mathbf { R }_E ^ { T } \mathbf { F }_E + \left( \mathbf { R }_E ^ { T } \mathbf { F }_E \right) ^ { T } \right) - \mathbf { I } = \mathbf { S }_E - \mathbf { I }
 \end{equation}
-Substituting $\hat{\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity using rotation-invariance of Frobenius norm and trace:
+Substituting $\hat{\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity:
+\begin{equation}\label{eq:corot_S}
+  \Psi = \mu \| \mathbf{ S }_E - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \mathbf { S }_E - \mathbf { I } )
+\end{equation}
+Using rotation-invariance of the Frobenius norm:
 \begin{equation}
-  \Psi = \mu \| \mathbf { F } - \mathbf { R } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } - \mathbf { I } \right)
+  \Psi = \mu \| \mathbf {F}_E  - \mathbf {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \mathbf { R }_E ^ { T } \mathbf { F }_E - \mathbf { I } \right)
 \end{equation}
-
-\begin{equation}
-  \Psi(\boldsymbol{\Sigma}) = \mu \| \boldsymbol { \Sigma } - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma } - \mathbf { I } )
+Even more insight yields the relationship to their singular values $\sigma _ i$, also called the principal stretches:
+$$\operatorname {tr}\left(\mathbf{S}\right) = \sum _ { i = 1 } \sigma _ { i } = \operatorname{tr}\left(\mathbf{\Sigma}_E\right)$$
+$$\| \mathbf{S}_E \| _ { F }^2 = \sum _ { i = 1 } \sigma _ { i } ^ { 2 } = \| \mathbf{\Sigma}_E\|_{F}^2$$
+$$\Rightarrow \| \mathbf{S}_E\mathbf{-I} \| _ { F }^2 = \|\mathbf{S}_E \| _ { F }^2 - 2 \operatorname{tr}(\mathbf{S}_E)  + \|\mathbf{I}\|_{F}^2 = \| \boldsymbol { \Sigma }_E - \mathbf { I } \| _ { F } ^ { 2 }$$
+\begin{equation} \label{eq:corot_sing}
+  \Psi(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \mathbf { I } )
 \end{equation}
+Equation \ref{eq:corot_sing} uses the diagonal matrix $\mathbf{\Sigma}_E$ containing the principal stretches typically acquired by the singular value decomposition $\mathbf{F}_E \mathbf{= U \Sigma}_E\mathbf{ V ^T }.$ Since the energy density in \ref{eq:corot_sing} is a function of only three (singular) values, which describe stretch/compression of the material, isotropy of the material is underlined.
 \cite{ADAMS:ELASTICITY}
+Numerical Stepping algorithms
+-> Newton Raphson scheme
+-> Problem: Bad Primary contour
+Benefit: Easier description in terms of Invariants
 \cite{MPM:INVERT}
 
-\subsection{Singular Value Decomposition}
-\cite{SVD:3x3}
 \subsection{Material Point Method}

### 2018-11-10 mscthesis.tex
einstein notation, intro mpm

@@
 \section{Notation}
-Einstein-Notation
+Multi-component types are generally printed in bold letters $\mathbf{A}$, $\mathbf{a}$. Vectors furthermore use lower-case letters $\mathbf{v}$, while Matrices use upper-case letters $\mathbf{M}$.
+\subsection{Einstein-Notation}
+Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors.
+Let $a$ and $b$ be vectors of dimension $n$.
+Repeated indices that are also defined on the variable imply component-wise operations.
+$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $a$ and $b$. Vector and tensor/matrix addition thus become:
+\begin{equation} \label{EINSTEIN:ADD}
+  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad c _ { i j } = a _ { i j } + b _ { i j }
+\end{equation}
+Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
+\begin{equation}
+  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
+\end{equation}
+The Frobenius inner product between two second-order tensors $\mathbf{A,B}$ $\in \mathbb { R } ^ { m \times n }$.
+\begin{equation} \label{EINSTEIN:FROBENIUS1}
+  \mathbf { A } : \mathbf { B } \equiv a _ { i j } b _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } a _ { ij } b _ { ij }
+\end{equation}
+The Frobenius inner product between a fourth-order tensor $C \in \mathbb { R } ^ {r \times s \times m \times n }$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
+\begin{equation}
+  \mathbf { A } = a_ {ij} =  \mathbf {C} : \mathbf { B } = c _ {k l} b _ { k l  } = \sum _ { k = 1 } ^ { n } \sum _ { l = 1 } ^ { m } c _ { ijkl } b _ { ijkl }
+\end{equation}
+The definition of the Kronecker Delta is as follows:
+\begin{equation}
+  \delta _ { i j } = \left\{
+    \begin{array} { l l l }
+      { 1 } & {\text{if}} &  {i = j}  \\
+      { 0 } & {\text{if}} &  { i \neq j }
+  \end{array} \right.
+\end{equation}
+The Kronecker Delta as an operator is most efficiently described as a substitution:
+\begin{equation}
+  a _ { i } \delta _ { i j } = a _ { j }
+\end{equation}
+Or i.e. this component is only evaluated if $i = j$.
+\cite{MCGINTY:CONTINUUM}
+\subsection{Lagrange-Euler Notation}
 \section{Basics}
-\subsection{Continuum Mechanics}
+\subsection{Material Point Method}
+\subsubsection{Continuum Mechanics}
+Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
+Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t.
+a microscopic view does not need to be adopted.
+A particular body is composed of a set of particles and can adapt different configurations due to changes in shape.
+These changes are cause by external or internal effects (forces etc.) on it and deform the body over time.
+\cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}
+
+When modelling solids changes of quantities from one configuration (reference) $\mathbf{X}$ to another configuration (current) $\mathbf{x}$ need to be measured.
+In the material point method the reference configuration $\mathbf{X}$ is just the initial configuration of the body (at time $t=0$).
+This is similiar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
+Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ { \mathrm { 3 } }$ be the set of (material) points in the initial and current configuration respectively.
+Then one may define a function or mapping $\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the current configuration to the initial configuration.
+If we assume for simplicity that $\mathbf{X}$ and $\mathbf{x}$ describe the position of the particle in their respective configurations, this mapping can be thought of as the deformation of the body from the reference configuration $\mathbf{X}$:
+\begin{equation}
+  \mathbf{x} =  \phi ( \mathbf { X } , \mathbf { t } )
+\end{equation}
+It is helpful to abstract away from the reference configuration and define it to be in a different fixed material space.
+Physically this has the impact of moving with the particle.
+Commonly known as the Lagrangian formulation.
+In contrast
+
+\cite{MIT:CONTINUUM_MECHANICS}
 \cite{MPM:COURSE}
+\subsubsection{Conservation Equations}
+\cite{MPM:APIC}
+\subsubsection{CFL Condition}
 \subsection{Singular Value Decomposition}
@@
 
-\subsection{Material Point Method}
-\subsubsection{Conservation Equations}
-\cite{MPM:APIC}
-\subsubsection{CFL Condition}
 \subsection{GPU memory}

### 2018-11-11 mscthesis.tex
push-forward,pull-back, eulerian+lagrangian, notations

--- 2018-11-08 mscthesis.tex
@@
 \section{Notation}
-Einstein-Notation
+Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$.
+Vectors furthermore use lower-case letters $\boldsymbol{v}$, while Matrices use upper-case letters $\boldsymbol{M}$.
+A variable in the Lagrangian form has a subscript 0 $ _0{\boldsymbol{x}}$, the Eulerian description will be written with an 'overline' $_t\boldsymbol{x}$ if distinction is needed.
+Occasionally notation may be omitted where it is apparent from context.
+\subsection{Einstein-Notation}
+Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors.
+Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors.
+Repeated indices that are also defined on the variable imply component-wise operations.
+$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
+\begin{equation} \label{EINSTEIN:ADD}
+  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }
+\end{equation}
+Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
+\begin{equation}
+  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
+\end{equation}
+Following this notation the Frobenius inner product between two second-order tensors is:
+\begin{equation} \label{EINSTEIN:FROBENIUS1}
+  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }
+\end{equation}
+The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
+\begin{equation}
+  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
+\end{equation}
+The definition of the Kronecker Delta is as follows:
+\begin{equation}
+  \delta _ { i j } = \left\{
+    \begin{array} { l l l }
+      { 1 } & {\text{if}} &  {i = j}  \\
+      { 0 } & {\text{if}} &  { i \neq j }
+  \end{array} \right.
+\end{equation}
+The Kronecker Delta as an operator is most efficiently described as a substitution:
+\begin{equation}
+  a _ { i } \delta _ { i j } = a _ { j }
+\end{equation}
+I.e. this component is only evaluated if $i = j$.
+\cite{MCGINTY:CONTINUUM}
 \section{Basics}
-\subsection{Continuum Mechanics}
+\subsection{Material Point Method}
+\subsubsection{Continuum Mechanics}
+Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
+Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t.
+a microscopic view does not need to be adopted.
+A particular body is composed of a set of particles and can adapt different configurations due to changes in shape.
+These changes are cause by external or internal effects (forces etc.) on it and deform the body over time.
+\cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}
+
+When modelling solids changes of quantities from one configuration (reference) $_0\boldsymbol{x}$ to another configuration (current) $_t\boldsymbol{x}$ need to be measured.
+In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
+This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
+Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the initial and current configuration respectively.
+Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the initial configuration to the current configuration.
+If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
+\begin{equation}
+  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
+\end{equation}
+If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
+\begin{equation}
+  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
+\end{equation}
+The velocity of a material point in $_0\boldsymbol{x}$ can be defined using this mapping
+\begin{equation}\label{eq:velocity}
+  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial {^t_0}\phi}{\partial t}(_0\boldsymbol{x},t)
+\end{equation}
+and similarly the acceleration is defined
+\begin{equation}
+  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 {^t_0}\phi}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
+\end{equation}
+It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space.
+Physically this has the impact of moving with the particle in world space.
+Commonly known as the Lagrangian form.
+It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed.
+The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
+These descriptions are different but they will yield the same measurements when related correctly. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
+\begin{equation}
+  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
+\end{equation}
+and the (Eulerian) push forward
+\begin{equation}\label{eq:push_forward}
+  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
+\end{equation}
+with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.
+To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t.
+an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.
+
+The difficulty in the eulerian formulation becomes apparent when differentiating (due to the chain rule):
+\begin{equation}
+  \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
+  \frac { \partial  {_tv_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tv_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t )
+\end{equation}
+Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative:
+\begin{equation}
+  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } )
+\end{equation}
+The Jacobian of the deformation map $\phi$ is the deformation gradient and is one of the key components to measure strain:
+\begin{equation}
+  {^t_0 F_{ij}} ( _0\boldsymbol {x} , \boldsymbol { t } ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , \boldsymbol { t } ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , \boldsymbol { t } )
+\end{equation}
+Intuitively it measures the local changes in length from the reference to the current configuration.
+\cite{MIT:CONTINUUM_MECHANICS}
 \cite{MPM:COURSE}
+\subsubsection{Conservation Equations}
+\cite{MPM:APIC}
+\subsubsection{CFL Condition}
+\subsection{Singular Value Decomposition}
+\cite{SVD:3x3}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%                         Constitutive Models                         %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Constitutive Models}
@@
 
-The first aim will be to find the strain energy density $\psi(\mathbf{\epsilon})$ of the strain $\bm{\epsilon}$.
+The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
 For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\psi$ as:
@@
 \end{equation}
-where $\mathbf{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.
+where $\boldsymbol{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.
 Assuming following symmetries reduces the tensors to 6 and 21 unique elements respectively:
@@
 \begin{equation}\label{eq:partial_energy}
-  \sigma _ { i j } = \sigma _ { i j } ( \mathbf{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },
+  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },
 \end{equation}
@@
   \epsilon _ { i j } = \frac { 1 } { E } \left[ ( 1 + \nu ) \sigma _ { i j } - \nu \sigma _ { k k } \delta _ { i j } \right]
-  = \mathbf{C}^{-1} \bm{\sigma}
+  = \boldsymbol{C}^{-1} \bm{\sigma}
 \end{equation}
-using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\mathbf{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:
+using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\boldsymbol{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:
 \begin{equation} \label{eq:stress_strain}
@@
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
-  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \mathbf { F } + \mathbf { F } ^ { T } \right) - \mathbf { I }
+  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \boldsymbol { F }_E + \boldsymbol { F }_E ^ { T } \right) - \boldsymbol { I }
 \end{equation}
 The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
-to $\mathbf{F}$: $\psi(\mathbf{R}_{0} \mathbf{F}) \neq \psi(\mathbf{F})$.
+to $\boldsymbol{F}_E$: $\psi(\boldsymbol{R}_{0} \boldsymbol{F}_E) \neq \psi(\boldsymbol{F}_E)$.
 The deformation caused by stress however does not affect rigid body motions.
-Given the polar decomposition $\mathbf{F} = \mathbf{RS}$, an alternate strain measure may be defined as:
+Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:
 \begin{equation}
-  \hat { \boldsymbol { \epsilon } } ( \mathbf { F } ) = \boldsymbol { \epsilon } \left( \mathbf { R } ^ { T } \mathbf { F } \right) = \frac { 1 } { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } + \left( \mathbf { R } ^ { T } \mathbf { F } \right) ^ { T } \right) - \mathbf { I } = \mathbf { S } - \mathbf { I }
+  { \boldsymbol { \epsilon } } ( \boldsymbol { F }_E ) = \boldsymbol { \epsilon } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) = \frac { 1 } { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E + \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) ^ { T } \right) - \boldsymbol { I } = \boldsymbol { S }_E - \boldsymbol { I }
 \end{equation}
-Substituting $\hat{\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity using rotation-invariance of Frobenius norm and trace:
+Substituting $ {\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity:
+\begin{equation}\label{eq:corot_S}
+  \Psi = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )
+\end{equation}
+Using rotation-invariance of the Frobenius norm:
 \begin{equation}
-  \Psi = \mu \| \mathbf { F } - \mathbf { R } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \mathbf { R } ^ { T } \mathbf { F } - \mathbf { I } \right)
+  \Psi = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)
 \end{equation}
-
-\begin{equation}
-  \Psi(\boldsymbol{\Sigma}) = \mu \| \boldsymbol { \Sigma } - \mathbf { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma } - \mathbf { I } )
+Even more insight yields the relationship to their singular values $\sigma _ i$, also called the principal stretches:
+$$\operatorname {tr}\left(\boldsymbol{S}\right) = \sum _ { i = 1 } \sigma _ { i } = \operatorname{tr}\left(\boldsymbol{\Sigma}_E\right)$$
+$$\| \boldsymbol{S}_E \| _ { F }^2 = \sum _ { i = 1 } \sigma _ { i } ^ { 2 } = \| \boldsymbol{\Sigma}_E\|_{F}^2$$
+$$\Rightarrow \| \boldsymbol{S}_E\boldsymbol{-I} \| _ { F }^2 = \|\boldsymbol{S}_E \| _ { F }^2 - 2 \operatorname{tr}(\boldsymbol{S}_E)  + \|\boldsymbol{I}\|_{F}^2 = \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 }$$
+\begin{equation} \label{eq:corot_sing}
+  \Psi(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )
 \end{equation}
+Equation \ref{eq:corot_sing} uses the diagonal matrix $\boldsymbol{\Sigma}_E$ containing the principal stretches typically acquired by the singular value decomposition $\boldsymbol{F}_E \boldsymbol{= U \Sigma}_E\boldsymbol{ V ^T }.$ Since the energy density in \ref{eq:corot_sing} is a function of only three (singular) values, which describe stretch/compression of the material, isotropy of the material is underlined.
 \cite{ADAMS:ELASTICITY}
+Numerical Stepping algorithms
+-> Newton Raphson scheme
+-> Problem: Bad Primary contour
+Benefit: Easier description in terms of Invariants
 \cite{MPM:INVERT}
 
-\subsection{Singular Value Decomposition}
-\cite{SVD:3x3}
-\subsection{Material Point Method}
-\subsubsection{Conservation Equations}
-\cite{MPM:APIC}
-\subsubsection{CFL Condition}
 \subsection{GPU memory}
@@
     \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-    \State $r\gets a\mathbfod b$ and \var{foobar}\label{foobar}
+    \State $r\gets a\boldsymbolod b$ and \var{foobar}\label{foobar}
     \While{$r\not=0$}\Comment{We have the answer if r is 0}
@@
     \State $b\gets r$
-    \State $r\gets a\mathbfod b$
+    \State $r\gets a\boldsymbolod b$
     \EndWhile\label{euclidendwhile}

### 2018-11-12 mscthesis.tex
finished integral discussions for governing equations

@@
 \end{equation}
-Following this notation the Frobenius inner product between two second-order tensors is:
+Following this notation the Frobenius inner product between two second-order tensors is (also called Frobenius scalar product):
 \begin{equation} \label{EINSTEIN:FROBENIUS1}
@@
   \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
+\end{equation}
+Matrix-Vector and Matrix-Matrix multiplication can be expressed this way as:
+\begin{equation}
+  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }
 \end{equation}
@@
 \end{equation}
-I.e. this component is only evaluated if $i = j$.
+I.e. this component is only evaluated if $i = j$. In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$.
 \cite{MCGINTY:CONTINUUM}
 \section{Basics}
-\subsection{Material Point Method}
-\subsubsection{Continuum Mechanics}
+\subsection{Reference and current configuration}
 Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
@@
 \end{equation}
-The Jacobian of the deformation map $\phi$ is the deformation gradient and is one of the key components to measure strain:
+The Jacobian of the deformation map $\phi$ is the deformation gradient $\boldsymbol{F}$ and is one of the key components to measure strain:
 \begin{equation}
-  {^t_0 F_{ij}} ( _0\boldsymbol {x} , \boldsymbol { t } ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , \boldsymbol { t } ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , \boldsymbol { t } )
+  {^t_0 F_{ij}} ( _0\boldsymbol {x} ,  t  ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , t ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , t )
 \end{equation}
-Intuitively it measures the local changes in length from the reference to the current configuration.
+Intuitively it measures the local deformation of all particles in a small neighborhood of $_0B$ to $_tB$.
+Topology specifies the neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$.
+This allows to describe infinitesimal changes in position from the reference to the current configuration
+\begin{equation}\label{eq:def_grad_pos}
+  d {_tx_i} = {^t_0}F{_{ij}} d{_0x_j}.
+\end{equation}
+With this quantity in place volume and area changes are calculable.
+In a typical analytical fashion a coordinate system change $_0x \rightarrow {_tx}$ is done using the Jacobian (determinant) matrix.
+The determinant is given a separate name $^t_0J =\text{det}(^t_0\boldsymbol{F})$.
+The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ is $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
+\begin{equation}
+  \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} )  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}} ) {^t_0 J}d {_0\boldsymbol{x}}.
+\end{equation}
+This can also be achieved by the cross product.
+A cube spanned by vectors $_0\boldsymbol{x}_i$ ($i=1,2,3$) becomes a parallelepiped in the deformed configuration $d_tV =|{^t_0\boldsymbol{F}}d_0\boldsymbol{x}_0 \cdot({^t_0\boldsymbol{F}}d_0\boldsymbol{x}_1 \times {^t_0\boldsymbol{F}}d_0\boldsymbol{x}_2)| {d_0V}$:
+\begin{equation}\label{eq:j}
+  d_tV = {^t_0J} d_0V.
+\end{equation}
+The area change is given by Nanson's Formula. Where $\boldsymbol{A}$ points in the direction of the normal of the area:
+$$ d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}, \quad d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}} $$
+$$ \xRightarrow{\ref{eq:def_grad_pos},\ref{eq:j}} {^t_0J} d_0\boldsymbol{A} \cdot d_0\boldsymbol{l} = d{_t\boldsymbol{A}} \cdot ({^t_0\boldsymbol{F}}d{_0\boldsymbol{l}})$$
+\begin{equation}
+  \Rightarrow d_t\boldsymbol{A} = {^t_0\boldsymbol{F}}^{-T} {^t_0}J d_0\boldsymbol{A} = {^0_t\boldsymbol{F}}^{T} {^t_0}J d_0\boldsymbol{A}.
+\end{equation}
+A surface intergral may then be transformed to current configuration by
+\begin{equation}
+  \int _ { \partial B ^ { \mathfrak { t } } } \boldsymbol{h} ( {_t\boldsymbol{x}} , t ) \cdot d \boldsymbol{A} ( {_t\boldsymbol{x}} ) =  \int _ { \partial B ^ { \mathfrak { 0 } } } \boldsymbol{h} ( {_0\boldsymbol{x}} , t ) \cdot \boldsymbol{F}^{-T}J d \boldsymbol{A} ( {_0\boldsymbol{x}} )
+\end{equation}
+where $_0h = h(_0\boldsymbol{x},\cdot)$ is the pull back of $_th = h(_t\boldsymbol{x},\cdot)$.
 \cite{MIT:CONTINUUM_MECHANICS}
 \cite{MPM:COURSE}
-\subsubsection{Conservation Equations}
-\cite{MPM:APIC}
-\subsubsection{CFL Condition}
-\subsection{Singular Value Decomposition}
+\subsection{Polar- and Singular Value Decomposition}
 \cite{SVD:3x3}
@@
 
+\subsubsection{Linear Elasticity}
 The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
@@
 \cite{MIT:LINEAR_ELASTICITY}
-\subsubsection{Fixed Corotated Hyperelasticity}
+\subsubsection{Corotated Hyperelasticity}
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
@@
 The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
-to $\boldsymbol{F}_E$: $\psi(\boldsymbol{R}_{0} \boldsymbol{F}_E) \neq \psi(\boldsymbol{F}_E)$.
-The deformation caused by stress however does not affect rigid body motions.
+to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$.
+Rigid body motions however don't result in deformation and consequently don't need to be recovered from.
+So energy should not change.
 Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:
@@
 \cite{ADAMS:ELASTICITY}
-Numerical Stepping algorithms
--> Newton Raphson scheme
--> Problem: Bad Primary contour
-Benefit: Easier description in terms of Invariants
+\subsubsection{Fixed Corotated Hyperelasticity}
+Numerical Stepping algorithms\\
+-> Newton Raphson scheme\\
+-> Problem: Bad Primary contour\\
+Benefit: Easier description in terms of Invariants\\
 \cite{MPM:INVERT}
+\subsubsection{Conservation Equations}
+\cite{MPM:APIC}
+\subsubsection{CFL Condition}
 

### 2018-11-12 mscthesis.tex
polar svd

@@
 \end{equation}
+A transpose operator swaps the indices:
+\begin{equation}
+  D^T _ { i j }  =D_{ji}
+\end{equation}
 Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
@@
 \end{equation}
-I.e. this component is only evaluated if $i = j$. In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$.
+I.e.
+this component is only evaluated if $i = j$.
+In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$.
+The Kronecker Delta comes up in differentiating a variable by itself with different indexing:
+\begin{equation}
+  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}
+\end{equation}
 \cite{MCGINTY:CONTINUUM}
@@
 If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
-\begin{equation}
+\begin{equation}\label{eq:rigid}
   _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
@@
 \end{equation}
-The area change is given by Nanson's Formula. Where $\boldsymbol{A}$ points in the direction of the normal of the area:
+The area change is given by Nanson's Formula. Where $d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area:
 $$ d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}, \quad d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}} $$
@@
 \end{equation}
-A surface intergral may then be transformed to current configuration by
+A surface integral may then be transformed to reference configuration by
 \begin{equation}
@@
 \end{equation}
-where $_0h = h(_0\boldsymbol{x},\cdot)$ is the pull back of $_th = h(_t\boldsymbol{x},\cdot)$.
+where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$.
 \cite{MIT:CONTINUUM_MECHANICS}
@@
 \subsection{Polar- and Singular Value Decomposition}
+The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$.
+In equation \ref{eq:rigid} a rigid body movement was introduced.
+A problem arises when calculating the deformation gradient of this equation.
+Let $b_i(t) = v_it$ be more generally a translation:
+\begin{equation}
+  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t)
+\end{equation}
+As can be seen the deformation gradient contains a rigid rotation.
+For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation.
+I.e.
+the deformation gradient has two components a constant rotation and the actual distortion or strain.
+There is two ways to deal with this:
+\begin{enumerate}
+  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-strain with quadratic components:
+    \begin{equation}\label{eq:green}
+      E _ { i j } = \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right).
+    \end{equation}
+  \item \label{it:polar} Polar Decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric) distortion $\boldsymbol{U}$ part $\boldsymbol{F}=\boldsymbol{U}\boldsymbol{R}$.
+\end{enumerate}
+That equation \ref{eq:green} cancels out the rotational part can be shown by item \ref{it:polar}
+$$ \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {m i}R_{k m} R _ { k n} U_{n j} - \delta _ { i j } \right)$$
+$$= \frac { 1 } { 2 } \left( U _ {m i} \delta_{mn} U_{n j} - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {i m} U_{m j} - \delta _ { i j } \right)$$
+$$= \frac{1}{2} \left(\boldsymbol{U}^2 - I \right).$$
+
 \cite{SVD:3x3}
@@
 to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$.
-Rigid body motions however don't result in deformation and consequently don't need to be recovered from.
+Rigid body motions however don't result in strain and consequently don't need to be recovered from.
 So energy should not change.

### 2018-11-14 mscthesis.tex
singular value decompositon, todo: polar svd

@@
 \usepackage{subdepth}
+\usepackage{enumitem}
 \addbibresource{mscthesis.bib}
@@
     \end{equation}
-  \item \label{it:polar} Polar Decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric) distortion $\boldsymbol{U}$ part $\boldsymbol{F}=\boldsymbol{U}\boldsymbol{R}$.
+  \item \label{it:polar} Polar decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric positive definite) distortion $\boldsymbol{S}$ part $\boldsymbol{F}=\boldsymbol{R}\boldsymbol{S}$.
 \end{enumerate}
-That equation \ref{eq:green} cancels out the rotational part can be shown by item \ref{it:polar}
-$$ \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {m i}R_{k m} R _ { k n} U_{n j} - \delta _ { i j } \right)$$
-$$= \frac { 1 } { 2 } \left( U _ {m i} \delta_{mn} U_{n j} - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {i m} U_{m j} - \delta _ { i j } \right)$$
-$$= \frac{1}{2} \left(\boldsymbol{U}^2 - I \right).$$
+That equation \ref{eq:green} cancels out the rotational part can be shown by item \ref{it:polar}:
+$$ \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( S _ {m i}R_{k m} R _ { k n} S_{n j} - \delta _ { i j } \right)$$
+$$= \frac { 1 } { 2 } \left( S _ {m i} \delta_{mn} S_{n j} - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( S _ {i m} S_{m j} - \delta _ { i j } \right)$$
+$$= \frac{1}{2} \left(\boldsymbol{S}^2 - I \right).$$
+Regarding item \ref{it:polar}: Assuming a singular value decomposition of
+\begin{equation}\label{eq:svd}
+  \boldsymbol{F} = \boldsymbol{U\Sigma V}^T
+\end{equation}
+is already computed, where $\boldsymbol{U}$,$\boldsymbol{V}$ are orthogonal matrices and $\boldsymbol{\Sigma}$ is a diagonal matrix containing the singular values $\sigma_1 \geq \sigma_2 \geq ...
+\geq \sigma_r$ of $\boldsymbol{F}$.
+$(r-n)$ forms the dimension of the null space.
+The polar decomposition is computable as:
+\begin{equation}
+  \boldsymbol{R} = \boldsymbol{V} \boldsymbol{W} ^ { T }, \quad \boldsymbol{S} = \boldsymbol{W} \boldsymbol{\Sigma} \boldsymbol{W} ^ { T }.
+\end{equation}
+Since singular values are positive, it is straightforward to see that the properties for $\boldsymbol{R}$ and $\boldsymbol{S}$ hold.
 
-\cite{SVD:3x3}
+The components of the singular value decomposition are important to gain an intuition: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors, respectively \cite{MIT:SVD}.
+For illustrating purposes imagine the manipulation of $\boldsymbol{v}_1$ due to $\boldsymbol{U\Sigma V}^T$:
+\begin{enumerate}
+  \item Transform from the right singular vector space to standard basis space: $\boldsymbol{V}^T\boldsymbol{v}_1 = \boldsymbol{e}_1$
+  \item Scale by singular values to transform to principal stretch space: $\boldsymbol{\Sigma} \boldsymbol{e}_1 = \sigma_1 \boldsymbol{e}_1$
+  \item \label{it:tr_sing_vect} Transform to left singular vector space: $\boldsymbol{U}\sigma_1\boldsymbol{e}_1 = \sigma_1 \boldsymbol{u}_{1}$
+\end{enumerate}
+For a positive definite matrix the singular value decomposition becomes even easier as $\boldsymbol{U}=\boldsymbol{V}$. Item \ref{it:tr_sing_vect} then effectively just becomes a transform 'back'.
+Following is a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3}.
+\begin{enumerate}
+  \item \label{it:eig} \textbf{Symmetric eigenanalysis:} A Jacobi eigenvalue algorithm begins with the symmetric positive definite matrix $\boldsymbol { S } ^{(0)}= \boldsymbol { A } ^ { T } \boldsymbol { A } = \boldsymbol { V } \boldsymbol { \Sigma } ^ { 2 } \boldsymbol { V } ^ { T }$.
+    \begin{enumerate}[label*=\arabic*.]
+      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$, where $\boldsymbol{Q}$ is a Givens rotation aiming to eliminate off $s_{12}$.
+Store $\boldsymbol{V}^{(k+1)} = \boldsymbol{V}^{(k)}  \boldsymbol{Q}^{(k)}$.
+      \item \label{item:sii} Do \ref{item:sk} again for the other off-diagonal entries $s_{13},s_{23}$.
+      \item \label{item:redo_sii} Redo \ref{item:sk} - \ref{item:sii} a fixed amount of steps $m$.
+    \end{enumerate}
+  \item \textbf{Sorting singular values:} Compute $\boldsymbol{B} := \boldsymbol{AV}$, where $\boldsymbol{V} = \boldsymbol{V}^{(3m)}$.
+Acquire $\boldsymbol{\Sigma}$ by $\|\boldsymbol{b}_i \|_2 = \|\boldsymbol{u}_i\sigma_i\|_2 = |\sigma_i|$, since also $\boldsymbol{B=U\Sigma}$ holds.
+Permute the singular values by sorting them in decreasing order.
+Apply the same permutation to the columns of $\boldsymbol{B}$ and $\boldsymbol{V}$, where switches in $\boldsymbol{V}$ also cause a sign change.
+  \item \label{item:qr} \textbf{QR-Factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-Factorization with Givens Rotations, where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
+The $\boldsymbol{QR}$-Factorization is done once in the same fashion as in item \ref{item:sk} - \ref{item:sii}.
+\end{enumerate}
+Due to inherent normalization, fast multiplication and storage efficiency quaternions are preferred over actual rotation matrices.
+In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy for near-zero singular values.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-11-14 mscthesis.tex
convention added

@@
 Since singular values are positive, it is straightforward to see that the properties for $\boldsymbol{R}$ and $\boldsymbol{S}$ hold.
-
-The components of the singular value decomposition are important to gain an intuition: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors, respectively \cite{MIT:SVD}.
+The components of the singular value decomposition are important to gain an intuition: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors $\boldsymbol{u}_i,\boldsymbol{v}_i$, respectively \cite{MIT:SVD}.
 For illustrating purposes imagine the manipulation of $\boldsymbol{v}_1$ due to $\boldsymbol{U\Sigma V}^T$:
@@
 For a positive definite matrix the singular value decomposition becomes even easier as $\boldsymbol{U}=\boldsymbol{V}$. Item \ref{it:tr_sing_vect} then effectively just becomes a transform 'back'.
-Following is a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3}.
+Following is a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3}. The proposed singular value decomposition is also called 'Polar SVD' and follows a specific convention.
+\begin{itemize}\label{ref:itemize_conv}
+  \item $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e. both $\text{det}(U), \text{det}(V) = 1$.
+  \item As a result if $\text{det}(\boldsymbol{F}) = -1$ the negative sign needs to move on to $\boldsymbol{\Sigma}$: The lowest singular value in magnitude will get a negative sign attached.
+\end{itemize}
+This convention does not change the existence or uniqueness of the singular value or polar decomposition, although strictly speaking $\boldsymbol{S}$ in the Polar decomposition is not positive definite anymore.
 \begin{enumerate}
@@
     \begin{enumerate}[label*=\arabic*.]
-      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$, where $\boldsymbol{Q}$ is a Givens rotation aiming to eliminate off $s_{12}$.
+      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$, where $\boldsymbol{Q}$ is a Givens rotation aiming to eliminate off $S_{12}$.
 Store $\boldsymbol{V}^{(k+1)} = \boldsymbol{V}^{(k)}  \boldsymbol{Q}^{(k)}$.
-      \item \label{item:sii} Do \ref{item:sk} again for the other off-diagonal entries $s_{13},s_{23}$.
+      \item \label{item:sii} Do \ref{item:sk} again for the other off-diagonal entries $S_{13},S_{23}$.
       \item \label{item:redo_sii} Redo \ref{item:sk} - \ref{item:sii} a fixed amount of steps $m$.
@@
   \item \textbf{Sorting singular values:} Compute $\boldsymbol{B} := \boldsymbol{AV}$, where $\boldsymbol{V} = \boldsymbol{V}^{(3m)}$.
-Acquire $\boldsymbol{\Sigma}$ by $\|\boldsymbol{b}_i \|_2 = \|\boldsymbol{u}_i\sigma_i\|_2 = |\sigma_i|$, since also $\boldsymbol{B=U\Sigma}$ holds.
+Acquire $\boldsymbol{\Sigma}$ by $\|\boldsymbol{b}_i \|_2 = \|\boldsymbol{u}_i\sigma_i\|_2 = |\sigma_i|$, where also $\boldsymbol{B=U\Sigma}$ holds.
 Permute the singular values by sorting them in decreasing order.
 Apply the same permutation to the columns of $\boldsymbol{B}$ and $\boldsymbol{V}$, where switches in $\boldsymbol{V}$ also cause a sign change.
+Enforce the convention mentioned above.
   \item \label{item:qr} \textbf{QR-Factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-Factorization with Givens Rotations, where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
@@
 Due to inherent normalization, fast multiplication and storage efficiency quaternions are preferred over actual rotation matrices.
-In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy for near-zero singular values.
+In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-11-15 mscthesis.tex
conservational

@@
 
-When modelling solids changes of quantities from one configuration (reference) $_0\boldsymbol{x}$ to another configuration (current) $_t\boldsymbol{x}$ need to be measured.
+When modelling solids changes of quantities from a initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
 In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
 This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
-Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the initial and current configuration respectively.
-Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the initial configuration to the current configuration.
+Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.
+Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the reference configuration to the current configuration.
 If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
@@
 The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
-These descriptions are different but they will yield the same measurements when related correctly. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
+These descriptions are different but they will yield the same measurements when related correctly.
+I.e.
+both configuration refer variables defined on them do the deformed state, but the position where we 'look up' that value is different.
+In the reference configuration we look up at the initial position of the particle.
+In the current configuration we look up at the particle's 'world position'.
+These relations for some particle quantity $f$ are called the (Lagrangian) pull back
 \begin{equation}
@@
 The difficulty in the eulerian formulation becomes apparent when differentiating (due to the chain rule):
-\begin{equation}
+\begin{equation} \label{eq:eulerian_general}
   \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
-  \frac { \partial  {_tv_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tv_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t )
+  \frac { \partial  {_tf_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tf_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t )
 \end{equation}
 Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative:
-\begin{equation}
+\begin{equation} \label{eq:material_derivative}
   \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } )
@@
 \end{equation}
-Intuitively it measures the local deformation of all particles in a small neighborhood of $_0B$ to $_tB$.
+Intuitively it measures the local deformation of all particles in a small neighborhood of $_0B_\epsilon$ to $_tB_\epsilon$.
 Topology specifies the neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$.
+$_0B_\epsilon$ becomes the pre-image of $_tB_\epsilon$ under $_0^t\phi$.
 This allows to describe infinitesimal changes in position from the reference to the current configuration
@@
 The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ is $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
-\begin{equation}
-  \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} )  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}} ) {^t_0 J}d {_0\boldsymbol{x}}.
+\begin{equation} \label{eq:pull_back_vol}
+  \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} ,t)  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}},t) {^t_0 J}d {_0\boldsymbol{x}}.
 \end{equation}
@@
 A surface integral may then be transformed to reference configuration by
-\begin{equation}
-  \int _ { \partial B ^ { \mathfrak { t } } } \boldsymbol{h} ( {_t\boldsymbol{x}} , t ) \cdot d \boldsymbol{A} ( {_t\boldsymbol{x}} ) =  \int _ { \partial B ^ { \mathfrak { 0 } } } \boldsymbol{h} ( {_0\boldsymbol{x}} , t ) \cdot \boldsymbol{F}^{-T}J d \boldsymbol{A} ( {_0\boldsymbol{x}} )
+\begin{equation}\label{eq:surface_integral}
+  \int _ { \partial {_0B}  } \boldsymbol{h} ( {_t\boldsymbol{x}} , t ) \cdot d \boldsymbol{A} ( {_t\boldsymbol{x}} ) =  \int _ { \partial {_tB} } \boldsymbol{h} ( {_0\boldsymbol{x}} , t ) \cdot \boldsymbol{F}^{-T}J d \boldsymbol{A} ( {_0\boldsymbol{x}} )
 \end{equation}
 where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$.
+$d_0\boldsymbol{A}$,$d_t\boldsymbol{A}$ point in the direction of the surface normal of $\partial {_0B}(_0\boldsymbol{x})$, $\partial {_tB}(_t\boldsymbol{x})$, respectively.
 \cite{MIT:CONTINUUM_MECHANICS}
@@
 In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
+\cite{MPM:COURSE}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule
-
+$$\boldsymbol{F}_E \boldsymbol{F}_P$$ \cite{ochsner2014elasto}
 \subsubsection{Linear Elasticity}
@@
 \cite{MPM:INVERT}
-\subsubsection{Conservation Equations}
-\cite{MPM:APIC}
+\subsection{Governing equations}
+deriving the weak form
+\subsubsection{Conservation of mass}
+Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$ and its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
+After \ref{eq:pull_back_vol} they are related as:
+\begin{equation} \label{eq:density_pull_back}
+  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}
+\end{equation}
+An open ball $ _ { 0 }B _ { \epsilon } $ in the reference configuration will have the same mass as its respective open ball $B_{\epsilon}^t$ in the current configuration.
+Keep in mind that both refer to a deformed state.
+
+Conservation of mass dictates dictates that mass does not depend on motion or time. Only the space occupied by this mass may be more or less.
+\begin{equation}
+  \frac{d}{dt} \int _ {_tB _ { \epsilon }} _t\rho(_t\boldsymbol{x},t) d_t\boldsymbol{x} = 0
+\end{equation}
+Equivalently this can be formulated with the constant undeformed initial mass in Lagrangian view:
+\begin{equation}
+  \left(\int _ {_tB _ { \epsilon }} _t\rho(_t\boldsymbol{x},t) d_t\boldsymbol{x} \stackrel{\text{\ref{eq:density_pull_back}}}{=} \right)
+  \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}=
+  \int _ {_0B _ { \epsilon }} _0\rho(_0\boldsymbol{x},0)d_0\boldsymbol{x}.
+\end{equation}
+In Eulerian view the conservation of mass is more difficult to develop and starts with the Lagrangian view. Since the integrals do account for arbitrary volumes, they are left out in the following:
+\begin{equation}\label{eq:euler_density_evol}
+  \frac{\partial}{\partial t}(_0\rho^t_0J) = \frac{\partial _0\rho}{\partial t}{^t_0J} + \frac{\partial{^t_0J}}{\partial t} {_0\rho} = 0.
+\end{equation}
+The left side could be immediately pushed forward, the right side is harder:
+\begin{equation}\label{eq:evol_jacobian}
+  \frac{\partial{J}}{\partial t}
+  = \frac{\partial{J}}{\partial {F_{ij}}} \frac{\partial {F_{ij}}}{\partial t}
+  \stackrel{\text{\ref{eq:det_deriv},\ref{eq:evol_def_grad}}}{=} {J}{F}^{-1}_{ji}\frac{\partial v_i}{\partial x_k}F_{kj}
+  = J \delta_{ik}\frac{\partial v_i}{\partial x_k} =  J \frac{\partial v_i}{\partial x_i}.
+\end{equation}
+The determinant differentiation rule can be shown by expressing the determinant with Laplace's expansion and applying the derivative on it:
+\begin{equation}\label{eq:det_deriv}
+  \frac{\partial J}{\partial F_{ij}} = \frac{\partial (F_{ik}\text{adj}(F)_{ki})}{\partial F_{ij}} =  \text{adj}(F)_{ji} = JF^{-1}_{ji}.
+\end{equation}
+The time-evolution of the deformation gradient is:
+\begin{equation}\label{eq:evol_def_grad}
+  \frac{\partial {{^t_0F_{ij}}}}{\partial t} =
+  \frac{\partial}{\partial t}\frac{\partial {^t_0\phi_i}}{\partial _0x_j}(_0\boldsymbol{x},t) = \frac{\partial {_0v_i}}{\partial _0x_j}(_0\boldsymbol{x},t) =
+  \frac{\partial {_tv_i}}{\partial _tx_k}(^t_0\phi(_0\boldsymbol{x},t),t)\frac{\partial {_tx_k}}{\partial {_0x_j}}(_0\boldsymbol{x},t).
+\end{equation}
+Pushing forward \ref{eq:euler_density_evol} with the result of \ref{eq:evol_jacobian} using the material derivative formulation \ref{eq:material_derivative}:
+\begin{equation}
+  \frac{D}{Dt}\rho(_t\boldsymbol{x},t) + \rho(_t\boldsymbol{x},t) \vec{\nabla} \cdot \boldsymbol{v}(_t\boldsymbol{x},t)=0
+\end{equation}
+Commonly used is the Nabla operator: $\vec { \nabla } = \left( \frac { \partial } { \partial x _ { 1 } } , \dots , \frac { \partial } { \partial x _ { n } } \right)$.
+\cite{MPM:COURSE}
+\subsubsection{Conservation of momentum}
+Continuum forces are divided up into body and surface forces.
+A surface force acts upon the surface of the material $\partial _tB_\epsilon$.
+While a body force scales upon the volume of the material $_tB_\epsilon$.
+Conservation of momentum may then be expressed in a similar way to the conservation of mass as:
+\begin{equation}
+  \frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial t_B _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
+\end{equation}
+Beginning with a mix out of Lagrangian and Eulerian view
+$$\left(\frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} \stackrel{\text{\ref{eq:eulerian_general}}}{=} \right)
+\frac { d } { d t } \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{v} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}$$
+\begin{equation}
+  \stackrel{\text{\ref{eq:euler_density_evol}}}{=}
+  \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{a} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}
+  = \int _ { \partial t_B _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x},
+\end{equation}
+where conservation of mass (equation \ref{eq:euler_density_evol}) was applied.
+The Eulerian push-forward of the left side combined with the divergence theorem becomes:
+\begin{equation}
+  \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} =
+  \int _ { \partial t_B _ {\epsilon } } \vec{\nabla} \cdot \boldsymbol{\sigma} d_t\boldsymbol{x}  + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
+\end{equation}
+The acceleration $_t\boldsymbol{a}$ is again defined due to the material derivative \ref{eq:material_derivative}. Thus the Eulerian momentum balance equation becomes:
+\begin{equation}
+  _t\rho \frac{D_t\boldsymbol{v}}{Dt} = \vec{\nabla} \cdot \boldsymbol{\sigma} +\boldsymbol{f} ^ {\text{body} }.
+\end{equation}
+There is a quantity left to be defined for the Lagrangian view.
+The Cauchy stress $\boldsymbol{\sigma}$ is defined in the current configuration.
+Pulling back the Cauchy stress leads to a stress measure named the first Piola Kirchhoff stress, denoted in the literature as $\boldsymbol{P} = \boldsymbol{\sigma}\boldsymbol{F}^{-T}J$:
+\begin{equation}
+  \int _ { \partial t_B _ {\epsilon  } } _t\boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} )
+  \stackrel{\text{\ref{eq:surface_integral}}}{=}
+  \int _ { \partial 0_B _ {\epsilon  } }  _0\boldsymbol{\sigma}\boldsymbol{F}^{-T} Jd_0\boldsymbol{A} (_0\boldsymbol{x} )=
+  \int _ { \partial 0_B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} )
+\end{equation}
+Summarized the Lagrangian view of the momentum equation with an initial momentum is:
+\begin{equation}
+  _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
+= \vec{\nabla}} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
+\end{equation}
 \subsubsection{CFL Condition}
-
 \subsection{GPU memory}
@@
   \item Conservation equations
+    \cite{MPM:APIC}
   \item Singular Value Decomposition

### 2018-11-16 mscthesis.tex
angular mom

@@
 \end{equation}
+The alternating tensor or Levi-Civita symbol
+\begin{equation}
+  \varepsilon _ { i j k } = \left\{ \begin{array} { r l } { + 1 } & { \text { if } ( i , j , k ) \text { is } ( 1,2,3 ) , ( 2,3,1 ) , \text { or } ( 3,1,2 ) } \\ { - 1 } & { \text { if } ( i , j , k ) \text { is } ( 3,2,1 ) , ( 1,3,2 ) , \text { or } ( 2,1,3 ) } \\ { 0 } & { \text { if } i = j , \text { or } j = k , \text { or } k = i } \end{array} \right.
+\end{equation}
+is used to express cross-products $\boldsymbol{c}=\boldsymbol{a} \times \boldsymbol{b}$:
+\begin{equation}
+  c_i = \varepsilon_{ijk} a_j b_k \left(= \sum _ { j = 1 } ^ { n } \sum _ { k = 1 } ^ { n } \varepsilon_{ijk} a_j b_k\right).
+\end{equation}
+A useful relation to the Kronecker-Delta is:
+\begin{equation}\label{eq:compact_levi}
+  \varepsilon _ { i j k } \varepsilon _ { i m n } = \delta _ { j m } \delta _ { k n } - \delta _ { j n } \delta _ { k m }.
+\end{equation}
 \cite{MCGINTY:CONTINUUM}
@@
 $$\boldsymbol{F}_E \boldsymbol{F}_P$$ \cite{ochsner2014elasto}
-\subsubsection{Linear Elasticity}
+\subsubsection{Linear Elasticity}\label{sec:linear_elasticty}
 The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
@@
 \begin{equation}
-  \frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial t_B _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
+  \frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
 \end{equation}
+Where the assumption due to angular momentum conservation $\boldsymbol{\sigma}^T  = \boldsymbol{\sigma} $ out of the second part of this section has already been made for completion.
 Beginning with a mix out of Lagrangian and Eulerian view
@@
 The Eulerian push-forward of the left side combined with the divergence theorem becomes:
-\begin{equation}
+\begin{equation}\label{eq:eul_transport}
   \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} =
@@
 \begin{equation}
-  \int _ { \partial t_B _ {\epsilon  } } _t\boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} )
+  \int _ { \partial _tB _ {\epsilon  } } _t\boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} )
   \stackrel{\text{\ref{eq:surface_integral}}}{=}
-  \int _ { \partial 0_B _ {\epsilon  } }  _0\boldsymbol{\sigma}\boldsymbol{F}^{-T} Jd_0\boldsymbol{A} (_0\boldsymbol{x} )=
-  \int _ { \partial 0_B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} )
+  \int _ { \partial _0B _ {\epsilon  } }  _0\boldsymbol{\sigma}\boldsymbol{F}^{-T} Jd_0\boldsymbol{A} (_0\boldsymbol{x} )=
+  \int _ { \partial _0B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} )
 \end{equation}
-Summarized the Lagrangian view of the momentum equation with an initial momentum is:
+Summarized the Lagrangian view of the momentum equation with an initial momentum is ($^0_0J=1$):
 \begin{equation}
@@
 \end{equation}
+For the stress strain relationship $\boldsymbol{C}$ in chapter \ref{sec:linear_elasticty} as well as the start of this chapter conservation of angular momentum caused $\sigma_{ij} = \sigma_{ji}$.
+This can be shown as follows.
+The description of angular momentum follows that of linear momentum where product rule generally applies $\left(\frac{D_t\boldsymbol{x}}{Dt} \times {_t\boldsymbol{v}} = {_t\boldsymbol{v}} \times {_t\boldsymbol{v}} = 0\right)$ :
+\begin{equation}
+  \int _ {_tB _ { \epsilon}  } _t\boldsymbol{x} \times \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } }_t\boldsymbol{x} \times  \boldsymbol{\sigma}^T d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon }_t\boldsymbol{x} \times  \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
+\end{equation}
+In component form this becomes:
+\begin{equation}\label{eq:gen_ang_mom}
+  \int _ {_tB _ { \epsilon}  } \varepsilon_{ijk} \rho x_j a_k d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } } \varepsilon_{ijk} x_j \sigma_{mk} dA_m(_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \varepsilon_{ijk} x_j f_k ^ {\text{body} } d _t\boldsymbol{x}.
+\end{equation}
+The divergence theorem is again applied to the surface forces $\left(\vec{\nabla} = \frac{\partial}{\partial x_m}\right)$:
+$$
+\int _ { \partial _tB _ {\epsilon  } } \varepsilon_{ijk} x_j \sigma_{mk} dA_m (_t\boldsymbol{x} ) =
+\int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \frac{\partial(x_j \sigma_{mk})}{\partial x_m} d_t\boldsymbol{x}
+$$
+\begin{equation}\label{eq:surface_ang_mom}
+  = \int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \left(\delta_{jm} \sigma_{mk}+ x_j\frac{ \partial \sigma_{mk}}{\partial x_m}\right) d_t\boldsymbol{x}.
+\end{equation}
+The conservation of momentum (eq. \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} into \ref{eq:gen_ang_mom} leaving only:
+\begin{equation}
+  \int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \sigma_{jk} d_t\boldsymbol{x} = 0.
+\end{equation}
+Leaving out the integral and multiplying by ${\varepsilon_irs}$ enables equation \label{eq:compact_levi}. The Cauchy stress is constrained to
+\begin{equation}
+  \sigma_{ij} =\sigma_{ji}
+\end{equation}
+as assumed before. Note that $\boldsymbol{P}$ however is not constrained to be symmetric.
 \subsubsection{CFL Condition}

### 2018-11-17 mscthesis.tex
weak form

@@
 A useful relation to the Kronecker-Delta is:
-\begin{equation}\label{eq:compact_levi}
+\begin{equation} \label{eq:compact_levi}
   \varepsilon _ { i j k } \varepsilon _ { i m n } = \delta _ { j m } \delta _ { k n } - \delta _ { j n } \delta _ { k m }.
@@
 The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ is $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
-\begin{equation} \label{eq:pull_back_vol}
+\begin{equation} \label{eq:volume_integral}
   \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} ,t)  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}},t) {^t_0 J}d {_0\boldsymbol{x}}.
@@
   \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{a} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}
-  = \int _ { \partial t_B _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x},
+  = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x},
 \end{equation}
@@
   \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} =
-  \int _ { \partial t_B _ {\epsilon } } \vec{\nabla} \cdot \boldsymbol{\sigma} d_t\boldsymbol{x}  + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
+  \int _ { \partial _tB _ {\epsilon } } \vec{\nabla} \cdot \boldsymbol{\sigma} d_t\boldsymbol{x}  + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
 \end{equation}
@@
 The Cauchy stress $\boldsymbol{\sigma}$ is defined in the current configuration.
-Pulling back the Cauchy stress leads to a stress measure named the first Piola Kirchhoff stress, denoted in the literature as $\boldsymbol{P} = \boldsymbol{\sigma}\boldsymbol{F}^{-T}J$:
+Pulling back the Cauchy stress leads to a stress measure named the first Piola Kirchhoff stress
 \begin{equation}
@@
   \int _ { \partial _0B _ {\epsilon  } }  _0\boldsymbol{\sigma}\boldsymbol{F}^{-T} Jd_0\boldsymbol{A} (_0\boldsymbol{x} )=
-  \int _ { \partial _0B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} )
+  \int _ { \partial _0B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} ),
+\end{equation}
+denoted in the literature as:
+\begin{equation}\label{eq:piola}
+  \boldsymbol{P} = \boldsymbol{\sigma}\boldsymbol{F}^{-T}J
 \end{equation}
 Summarized the Lagrangian view of the momentum equation with an initial momentum is ($^0_0J=1$):
-\begin{equation}
+\begin{equation}\label{eq:lagr_mom}
   _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
@@
 \end{equation}
-Leaving out the integral and multiplying by ${\varepsilon_irs}$ enables equation \label{eq:compact_levi}. The Cauchy stress is constrained to
+Leaving out the integral and multiplying by ${\varepsilon_{irs}}$ enables equation \ref{eq:compact_levi}. The Cauchy stress is constrained to
 \begin{equation}
@@
 as assumed before. Note that $\boldsymbol{P}$ however is not constrained to be symmetric.
-\subsubsection{CFL Condition}
+\subsubsection{Weak formulation}
+Before deriving the weak form an explanation of what it achieves is in order.
+The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain (and is as such influenced by the whole domain in general).
+Such a solution can be found for simplified models analytically.
+These act as a ground truth to numerical methods like the material point method.
+Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.
+
+The complete mathematical description of the weak formulation is beyond this thesis.
+Numeric books that discuess finite element methods will provide one (\cite{bathe2006finite}, \cite{dahmen2008numerik}).
+For simplicity: The weak formulation combined with Galerkin discretization restricts the globalility of the strong method due to so called 'test functions': $_0\boldsymbol{q}_h$ .
+A 'test function' generally only has influence ($ _0\boldsymbol{q}_h\neq 0$) on a very small subset of the whole domain $\Omega^0$.
+This is mostly used to gather information on a local neighborhood.
+
+The weak formulation requires an observation of the dot product. Consider the conservation of momentum in the strong form as formulated before (\ref{eq:lagr_mom}) as
+\begin{equation}\label{eq:strong}
+  _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
+  = \vec{\nabla} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
+\end{equation}
+Multiply both sides with the dot product of an arbitrary function $_0\boldsymbol{q}(\cdot,t):\Omega^0 \rightarrow \mathbb{R}^d$ and integrate over $\Omega^0$.
+If a solution solves the balance of \ref{eq:strong} then it also solves:
+$$
+\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \left(_0\rho (_0\boldsymbol{x} , 0 ) _0a _{ i } (_0\boldsymbol{x} , t ) - {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}\right) d _0\boldsymbol{x}
+$$
+\begin{equation}
+  =\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \frac{\partial P _ {ij}}{\partial x_j} ( _0\boldsymbol{x} , t ) d_0\boldsymbol{x}.
+\end{equation}
+With the help of partial integration in multiple dimension the derivative moves over to the test-function.
+Reducing the order of derivatives by moving a derivative to the test function is the main motive of the weak form:
+$$
+\int _ { \Omega ^ { 0 } } \frac{\partial (_0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t ))}{\partial _0x_j}
+- \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
+d_0\boldsymbol{x}
+$$
+The divergence theorem allows to convert the first term to a boundary integral.
+\begin{equation}
+  \int _ { \partial \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t)
+  d_0\boldsymbol{A}(_0\boldsymbol{x})
+  -\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
+  d_0\boldsymbol{x}
+\end{equation}
+Mathematically the boundary integral serves as a boundary condition which is set by the specific problem (e.g.
+context of the simulation).
+Finally putting together the previous results gives the weak form of force balance in the Lagrangian view:
+$$
+\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \left(_0\rho (_0\boldsymbol{x} , 0 ) _0a _{ i } (_0\boldsymbol{x} , t ) - {_0f_i} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}\right) d _0\boldsymbol{x}
+$$
+\begin{equation}
+  = \int _ { \partial \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t)
+  d_0\boldsymbol{A}(_0\boldsymbol{x})
+  -\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
+  d_0\boldsymbol{x}
+\end{equation}
+In the material point method stress computations are more naturally done in the current configuration.
+Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
+$$
+\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij} d_0\boldsymbol{x}
+\stackrel{\text{\ref{eq:piola}, \ref{eq:volume_integral}}}{=}
+\int _ { \Omega ^ { 0 } }  \left({^t_0}F_{kj}\frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\right) \left({_t^0}F_{kj} \sigma _ {ik} {_0^tJ}\right) {_t^0J} d_t\boldsymbol{x}
+$$
+\begin{equation}
+  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}
+\end{equation}
+This completes the Eulerian view to be:
+$$
+\int _ { \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \left(_t\rho (_t\boldsymbol{x} , t) _ta _{ i } (_t\boldsymbol{x} , t ) - {_tf_i} ^ {\text{body}}(_t\boldsymbol{x},t)\right) d _t\boldsymbol{x}
+$$
+\begin{equation}
+  = \int _ { \partial \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \sigma _ {ij} d_tA_t\boldsymbol{x} -  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
+\end{equation}
+\cite{MPM:COURSE}
+\cite{strang2007computational}
+\cite{bathe2006finite}
+\subsubsection{CFL condition}
 \subsection{GPU memory}

### 2018-11-17 mscthesis.tex
apic paper starts now

@@
   -\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
-  d_0\boldsymbol{x}
+  d_0\boldsymbol{x}.
 \end{equation}
@@
 \begin{equation}
-  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}
+  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
 \end{equation}
@@
 \begin{equation}
-  = \int _ { \partial \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \sigma _ {ij} d_tA_t\boldsymbol{x} -  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
+  = \int _ { \partial \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \sigma _ {ij} d_tA_j(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
 \end{equation}
@@
 \cite{bathe2006finite}
+\subsection{Material Particles}
+\begin{equation}
+  w( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array} \right.
+\end{equation}
+\begin{equation}
+  \nabla w_{\boldsymbol{i}}(\boldsymbol{x})
+  =\frac { 1 } {  { h } }
+  \left(
+    \begin{array} {l}
+      w^{\prime} (\frac { 1 } {  { h } } ( x  - x _ { i } ))  { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
+      w(\frac{1} {  { h } } ( x  - x _ { i } ))               { w }^{\prime} ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
+      w(\frac{1} {  { h } } ( x  - x _ { i } ))               { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }^{\prime}( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
+    \end{array}
+  \right)
+\end{equation}
 \subsubsection{CFL condition}

### 2018-11-19 mscthesis.tex
transfers

@@
 \usepackage{enumitem}
+\usepackage[toc,page]{appendix}
+\usepackage{minted}
 \addbibresource{mscthesis.bib}
-
+\makeatletter
+\def\ext@algorithm{lol}% algorithm captions will be written to the .lol file
+% share the list making commands and redefine the heading
+\AtBeginDocument{%
+  \let\l@algorithm\l@lstlisting
+  \let\c@algorithm\c@lstlisting
+  \let\thealgorithm\thelstlisting
+  \renewcommand{\lstlistlistingname}{Algorithms and source code}%
+}
+\makeatother
 \lstdefinelanguage{GLSL}
@@
 }
-\renewcommand{\lstlistingname}{Source code}
 
-\renewcommand{\lstlistlistingname}{List of source codes}
-
+\newenvironment{code}{\captionsetup{type=algorithm}}{}
 \begin{document}
@@
 %SPH, FEM methods
-\section{Notation}
+\section{Notation}\label{sec:notation}
 Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$.
@@
 Occasionally notation may be omitted where it is apparent from context.
+Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have subscript $p$ (e.g.
+$\boldsymbol{x}_p$).
+Grid cells will be assigned a subscript $i$ denoting the grid index (e.g.
+$\boldsymbol{x}_i$).
+This is not to be confused with the following Einstein-Notation.
+Therefore beginning with chapter \ref{sec:mat_point} Einstein-Notation will use greek letters ($\alpha,\beta$).
+A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n} t^i$ will have a right superscript: $\boldsymbol{x}^n$.
 \subsection{Einstein-Notation}
@@
 
-\subsection{Constitutive Models}
+\subsection{Constitutive Models}\label{seq:constitutive_models}
 \cite{MPM:DRUCKER} %Sand Plasticity
@@
 Equivalently this can be formulated with the constant undeformed initial mass in Lagrangian view:
-\begin{equation}
+\begin{equation}\label{eq:lagr_mass}
   \left(\int _ {_tB _ { \epsilon }} _t\rho(_t\boldsymbol{x},t) d_t\boldsymbol{x} \stackrel{\text{\ref{eq:density_pull_back}}}{=} \right)
@@
 \cite{bathe2006finite}
-\subsection{Material Particles}
+\subsection{Material Point Method}\label{sec:mat_point}
+The key idea of the material point method is to use (Lagrangian) particles as a consistent storage of material properties.
+All stress based forces are computed on a Eulerian grid however.
+This grid does not store any material properties and is therefore often referred to as a scratch pad.
+
+As a corollary there needs to be a way to transfer information from a particle to the neighboring grid cells.
+This also induces switching from Lagrangian to Eulerian view.
+After stresses are computed there also needs to be a way to get back the relevant information from the grid to the particles.
+After the information is transferred back the particles are moved.
+This advection is typically hard to do in Eulerian/FEM-like methods and cause of a lot of problems down the development pipeline.
+In a Lagrangian view particle advection is trivial.
+
+Its is very important that the two transfers as well as the grid solver are in compliance with the governing equations.
+While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that also conserve the properties defined in them.
+\cite{MPM:COURSE}
+\subsubsection{Interpolation weights}
+The choice of interpolation weights is flexible. Nevertheless a kernel $w(\boldsymbol{x})$ requires some important properties to be qualified for MPM:
+\begin{enumerate}
+
+  \item Partition of unity:
+    \begin{equation}\label{eq:partition_unity}
+      \sum _ { i } w \left(  { \boldsymbol{x} } -  { \boldsymbol{x} } _ { i } ^ { n } \right) & = 1.
+    \end{equation}
+  \item Identity relation:
     \begin{equation}
-  w( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array} \right.
+      \sum _ { i }  { \boldsymbol{x} } _ { i } ^ { n } w \left(  { \boldsymbol{x} } -  {\boldsymbol{ x} } _ { i } ^ { n } \right) & =  { \boldsymbol{x} }.
+    \end{equation}
+  \item $C^1$-continuity s.t. $\nabla w_i$ is continuous.
+\end{enumerate}
+As a reminder of chapter \ref{sec:notation}: $\boldsymbol{x}_p$ refers to a particle's position. $\boldsymbol{x}_i$ to a grid cell's position.
+\\
+Often dyadic products of one-dimensional interpolation functions suffice
+\begin{equation}
+  w(\boldsymbol{x}-\boldsymbol{x}_i^n) = w_i^n(\boldsymbol{x}) = w\left(\frac{1}{h}\left(x-x_i^n\right)\right)w\left(\frac{1}{h}\left(y-y_i^n\right)\right)w\left(\frac{1}{h}\left(z-z_i^n\right)\right)
 \end{equation}
 \begin{equation}
-  \nabla w_{\boldsymbol{i}}(\boldsymbol{x})
+  \nabla w_{i}^n(\boldsymbol{x})
   =\frac { 1 } {  { h } }
@@
     \begin{array} {l}
-      w^{\prime} (\frac { 1 } {  { h } } ( x  - x _ { i } ))  { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
-      w(\frac{1} {  { h } } ( x  - x _ { i } ))               { w }^{\prime} ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
-      w(\frac{1} {  { h } } ( x  - x _ { i } ))               { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}  ) )   { w }^{\prime}( \frac { 1 } {  { h } } ( z  - z _ { i } ) \\
+      w^{\prime} (\frac { 1 } {  { h } } ( x  - x _ { i }^n ))  { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
+      w(\frac{1} {  { h } } ( x  - x _ { i }^n ))               { w }^{\prime} ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
+      w(\frac{1} {  { h } } ( x  - x _ { i }^n ))               { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }^{\prime}( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
     \end{array}
-  \right)
+  \right),
 \end{equation}
+where $h$ is the grid spacing (of a uniform grid). An interpolation function often employed is a cubic B-spline ($C^2$):
+\begin{equation}
+  w(x) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array} \right.
+.
+\end{equation}
+Since these function are used to weight (or filter) particles a shortening in notation may be employed as $ w_{ip}^n = w(\boldsymbol{x}_p^n-\boldsymbol{x}_i^n)$.
+\cite{MPM:APIC}\cite{MPM:COURSE}
+\subsubsection{Mass transfer}
+Each material point will be assigned an initial volume $_0V_p$ as well as a initial mass $_0m_p$.
+The volume the material point occupies may change in time due to eq.
+\ref{eq:j}: ${^t_0}J _0V_p = {_tV_p}$.
+But due to conservation of mass \ref{eq:lagr_mass} it will have a constant, initial mass associated with it.
+I.e.
+there will only be a transfer to the grid and no transfer back.
+A transfer of mass to the grid may then be expressed as:
+\begin{equation}\label{eq:mpm:mass}
+  m_i = \sum_p w_{ip}m_p.
+\end{equation}
+$\sum_i m_i =\sum_p m_p$ is a complete prove this fulfills the conservation of mass.
+Remember there is no transfer back.
+In doing so no information can be lost on the particles.
+Proving that the mass transfer to the grid is conserving is enough.
+The stress based-solver may manipulate this information further (mass-lumping strategies).
+\begin{equation}
+  \sum_i m_i
+  \stackrel{\text{\ref{eq:mpm:mass}}}{=}
+  \sum _i \sum_p w_{ip} m_p =
+  \sum _p m_p \sum_i w_{ip}
+  \stackrel{\text{\ref{eq:partition_unity}}}{=}
+  \sum_p m_p
+\end{equation}
+\cite{MPM:APIC}\cite{MPM:COURSE}
+\subsubsection{APIC momentum Transfers}
+The momentum transfer round trip could be defined as:
+\begin{enumerate}
+  \item Particle to grid momentum transfer:
+    \begin{equation}\label{eq:lin_p2g}
+      (m\boldsymbol{v})_i^n = \sum_p w_{ip}m_p\boldsymbol{v}^n_p.
+    \end{equation}
+  \item Factoring out mass:
+    \begin{equation}
+      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
+    \end{equation}
+  \item Coupled with either ($\alpha \in \{0,1\}$) or a combination ($\alpha \in ]0;1[$) of:
+    \begin{equation}
+      \boldsymbol{v}_p^{n+1} = \alpha\boldsymbol{v}_{p,{PIC}}^{n+1} + (1-\alpha)\boldsymbol{v}_{p,{FLIP}}^{n+1}
+    \end{equation}
+    \begin{equation}\label{eq:PIC}
+      \boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}.
+    \end{equation}
+    \begin{equation}\label{eq:FLIP}
+      \boldsymbol{v}_{p,{FLIP}}^{n+1} = \boldsymbol{v}_p^{n} + \sum_i w_{ip}(\boldsymbol{v}_i^{n+1}-\boldsymbol{v}_i^{n+1}).
+    \end{equation}
+\end{enumerate}
+While $PIC$-Transfers are very stable they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
+This causes a heavy loss in angular momentum and velocity modes.
+$FLIP$-Transfers avoid dissipation and loss of angular momentum by only updating the velocity with a difference \ref{eq:FLIP}.
+However some velocity modes are also not recognized on the grid and may cause unpredictable and unstable behavior in following steps.
+Therefore often a combination of both is taken.
+
+$APIC$ builds on top of the very stable $PIC$-transfers and effectively only adds an extra term of the Taylor series to increase accuracy.
+This extra term $\boldsymbol{C}_p$ may be in short just referred to as the velocity derivative.
+The local velocity field around a particle may then be characterized by the affine function $\boldsymbol{v}(\boldsymbol{x}) = \boldsymbol{C}_p(\boldsymbol{x}-\boldsymbol{x}_p)$.
+\cite{MPM:APIC}\cite{MPM:OLD_APIC}
 \subsubsection{CFL condition}
@@
 Sparse: \cite{OPENVDB} \cite{NVIDIA:GVDB_VOXELS}
+\begin{appendices}
+  \section{Inertia Tensor for Cubic Splines}
+  \begin{code}
+\caption{$\boldsymbol{D}_p$ proof}
+  \begin{minted}{python}
+import numpy as np
+from sympy import *
 
+def round_expr(expr, num_digits):
+  return expr.xreplace({n : round(n, num_digits) for n in expr.atoms(Number)})
+
+# Limit a,b,c,x to interval [0,1] for simplyfing
+a, b, c = symbols('a b c', nonnegative=true)
+a = a/(1+a)
+b = b/(1+b)
+c = c/(1+c)
+
+x = symbols('x', nonnegative=true)
+x = x/(1+x)
+
+# Define cubic interpolation function
+def N1(x):
+  return 0.5*pow(abs(x),3)-pow(x,2)+2/3
+def N2(x):
+  return -1/6*pow(abs(x),3)+pow(x,2)-2*abs(x)+4/3
+def wip(i,x):
+  if(i==1 or i==2):
+    return N1(x)
+  else:
+    return N2(x)
+
+# Define the parametrized position in the grid
+def grid_points(x):
+  return np.array([-x-1,-x,1-x,2-x])
+
+alphas = grid_points(a)
+betas = grid_points(b)
+gammas = grid_points(c)
+
+D_temp = np.array([[0,0,0],[0,0,0],[0,0,0]])
+
+# Loop over all grid nodes in the vicinity
+for i,ai in enumerate(alphas):
+  for j,bj in enumerate(betas):
+    for k,ck in enumerate(gammas):
+      # xi_xp is the distance from parametrized position to grid node [i,j,k]
+      xi_xp = np.array([ai,bj,ck])
+      # each outer product weighted by interpolation functions
+      this_outer = wip(i,ai)*wip(j,bj)*wip(k,ck)*np.outer(xi_xp,xi_xp)
+      # summed up over all grid nodes
+      D_temp = np.add(D_temp,this_outer)
+
+D = np.array([[0.0,0.0,0.0],[0.0,0.0,0.0],[0.0,0.0,0.0]])
+
+for i,D_row in enumerate(D_temp):
+  for j,D_ij in enumerate(D_row):
+    #simplify to cancel polynoms, round_expr because numerical cancellation
+    D[i][j] = round_expr(simplify(D_ij),14)
+    print(D)
+\end{minted}
+\end{code}
+  Prints out:
+  $$
+  \left[
+    \begin{array}{ccc}
+      0.33333333 &0.         &0.        \\
+      0.         &0.33333333 &0.	\\
+      0.         &0.         &0.33333333\\
+  \end{array}\right]=\boldsymbol{D}_p
+  $$
+\end{appendices}
+\clearpage
 \listoffigures
-\clearpage
-\listoftables
 \lstlistoflistings

### 2018-11-20 mscthesis.tex
apic,cfl,start discretize

@@
 For simplicity: The weak formulation combined with Galerkin discretization restricts the globalility of the strong method due to so called 'test functions': $_0\boldsymbol{q}_h$ .
-A 'test function' generally only has influence ($ _0\boldsymbol{q}_h\neq 0$) on a very small subset of the whole domain $\Omega^0$.
+A 'test function' generally only has limited support ($ _0\boldsymbol{q}_h\neq 0$ on a very small subset of the whole domain $\Omega^0$).
 This is mostly used to gather information on a local neighborhood.
@@
 $$
-\begin{equation}
+\begin{equation} \label{eq:lagr_force_bal}
   = \int _ { \partial \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t)
@@
 $$
-\begin{equation}
-  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
+\begin{equation}  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
 \end{equation}
@@
 $$
-\begin{equation}
+\begin{equation}\label{eq:eul_force_bal}
   = \int _ { \partial \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \sigma _ {ij} d_tA_j(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
@@
 
-Its is very important that the two transfers as well as the grid solver are in compliance with the governing equations.
+Its is very important that the two transfers as well as the grid solver are in compliance with all governing equations.
 While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that also conserve the properties defined in them.
@@
 where $h$ is the grid spacing (of a uniform grid). An interpolation function often employed is a cubic B-spline ($C^2$):
-\begin{equation}
+\begin{equation}\label{eq:cubic_weights}
   w(x) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array} \right.
@@
 The local velocity field around a particle may then be characterized by the affine function $\boldsymbol{v}(\boldsymbol{x}) = \boldsymbol{C}_p(\boldsymbol{x}-\boldsymbol{x}_p)$.
+
+Motivated by the theory of angular momentum and moment of inertia one can define a quantity
+\begin{equation}
+\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T },
+\end{equation}
+which is similar to the classically known inertia tensor:
+\begin{equation}
+  \boldsymbol{I} _ { p } = - \sum _ { i } m _ { i } [\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}][\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}]^T
+\end{equation}
+$$
+=
+\sum _ { i } m _ { i }((\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})\boldsymbol{I} -(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T)
+$$
+Bearing in mind that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion, where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
+In classical mechanics the angular velocity $\boldsymbol{\omega_p}$ can be then described using the inertia tensor $\boldsymbol{I}_p$ with the help of the angular momentum $\boldsymbol{L}_p$:
+\begin{equation}
+  \boldsymbol{\omega}_p = \boldsymbol{I}_p^{-1} \boldsymbol{L}_p.
+\end{equation}
+This motivates the velocity derivative to be equally defined by a similar relationship, where $\boldsymbol{B}^n_p$ holds momentum information.
+\begin{equation}
+  \boldsymbol{C}^n_p = (\boldsymbol{D}_p^n)^{-1} \boldsymbol{B}^n_p.
+\end{equation}
+The transfers of the $APIC$-scheme are then summarized:
+\begin{enumerate}
+\item Particle to grid:
+  \begin{equation}\label{eq:apic_mom_p2g}
+    (m\boldsymbol{v})^n_i = \sum_p w^n_{ip}m_p (\boldsymbol { v } _ { p } ^ { n } + \boldsymbol { B } _ { p } ^ { n } ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) )
+\end{equation}
+  \item Factoring out mass.
+    \begin{equation}
+      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
+    \end{equation}
+
+  \item Grid to Particle Transfer (in a $PIC$-manner):
+$$
+    \boldsymbol { B } _ { p } ^ { n + 1 } = \frac { 1 } { 2 } \sum _ { i } w _ { i p } ^ { n } (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } +  { \boldsymbol { x } } _ { i } ^ { n + 1 } - \boldsymbol { x } _ { p } ^ { n + 1 } ) ^ { T }
+    $$
+    \begin{equation}\label{eq:apic_mom_g2p}
++ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } )
+    \end{equation}
+    \begin{equation}
+      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}.
+    \end{equation}
+\end{enumerate}
+For a full proof that these transfers preserve linear and angular momentum consult \cite{MPM:APIC}.
+For the choice of dyadic products of cubic b-splines (\ref{eq:cubic_weights}) $\boldsymbol{D}_p$ takes on the simple form:
+\begin{equation}
+\boldsymbol{D}_p^n = \frac { 1 } { 3 } h ^ { 2 } \boldsymbol { I }.
+\end{equation}
+A simple proof (Appendices: \ref{app:dp_proof}) cancelling out the numerous polynomials can be done for instance using \textbf{SymPy} (\cite{Sympy}).
 \cite{MPM:APIC}\cite{MPM:OLD_APIC}
 \subsubsection{CFL condition}
+The CFL condition is prominent for FEM-like methods: For a stable integration a particle should not travel father than the grid spacing $h$ in a discrete time-step $\Delta t$.
+$\Delta t$ is thus limited by:
+\begin{equation}
+  \Delta t \leq \frac{h}{\|\boldsymbol{v}^n_i\|_2}.
+\end{equation}
+Assuming $\|\boldsymbol{x}_i^n -\boldsymbol{x}_p^n\| \leq \kappa h$, where $\kappa$ is determined by the interpolation stencil support (cubic 3D: $\kappa = 2\sqrt{3}$) and additionally assuming $\boldsymbol{D}_p^n = k\boldsymbol{I} \Rightarrow (\boldsymbol{D}_p^n)^{-1} = \frac{1}{k}\boldsymbol{I}$ (cubic 3D: $k=\frac{1}{3}h^2$).
+$\|\boldsymbol{v}^n_i\|_2$ can be estimated on the particles. Typically the number of particles is lower. Given Eq. \ref{eq:apic_mom_p2g}:
+$$
+\|\boldsymbol{v}^n_i\|_2 \leq \frac{1}{m_i^n}\left(\sum _ { p } w _ { i p } ^ { n } m _ { p } \| \boldsymbol { v } _ { p } ^ { n } \|_2 + \sum _ { p } w _ { i p } ^ { n } m _ { p } \| \boldsymbol { B } _ { p } ^ { n } \| _ { F } \| ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) \|_2\right)
+$$
+\begin{equation}
+\leq \max _ { p } ( \| \boldsymbol { v } _ { p } ^ { n } \|_2 + \frac { \kappa } { k } \Delta x \| \boldsymbol { B } _ { p } ^ { n } \| _ { F } )
+\end{equation}
+\cite{MPM:APIC}
+\subsection{Discretization}
+The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following for MPM preferable description:
+\begin{equation}
+  \int _ { \Omega ^ { 0 } } (_0q _ \alpha) (_0\rho_0) (_0a _\alpha)  d _0\boldsymbol{x}= \int _ { \partial \Omega ^ { t^n } } _tq _ \alpha \sigma _ {\alpha\beta} d_tA_\beta(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t^n} } \frac{\partial _tq _ \alpha}{\partial _tx_\beta} \sigma _ {\alpha\beta} d_t\boldsymbol{x}.
+\end{equation}
+The boundary integral is mostly due to collisions and will be ignored for now.
+\cite{MPM:OPTIMI_INTEGR} discusses level set collisions due to constraint collisions, object penalty collision and penalty self-collisions.
+An handling of collision would need to take them into account when solving the equation.
+A simple though less accurate method is to process collisions separately in a typical computer graphics manner, which is assumed for now.
+\subsubsection{Discretize time}
+Any conservative integrator could be used to discretize time. The implicit midpoint rule is one of such ($\lambda =\frac{1}{2}$)
+\begin{equation}\label{eq:midpoint}
+  \frac{y ^ { n + 1 } - y ^ { n }}{ \Delta t} =  f \left(t^n+\frac{\Delta t}{2}, \frac { y ^ { n } + y^{n+1}}{2}\right)
+\end{equation}
+for a differential equation of order one:
+\begin{equation}
+  \frac{\partial y}{\partial t}( t ) = f (t,y(t)) , \quad y (0) = y _ { 0 }.
+\end{equation}
+Replacing the Lagrangian acceleration $_0a_\alpha$ with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
+\begin{equation}
+  \frac{1}{\Delta t} \int _{ \Omega^{t^n}}(_tq_\alpha)(_t\rho)({_tv_\alpha^{n+1}}-{_tv_\alpha^{n}}) d_t\boldsymbol{x} =
+  -  \int _ { \Omega ^ { t^n} } \frac{\partial _tq _ \alpha}{\partial _tx_\beta} \sigma _ {\alpha\beta}^{n+\lambda} d_t\boldsymbol{x}.
+\end{equation}
 \subsection{GPU memory}
@@
 \begin{appendices}
-  \section{Inertia Tensor for Cubic Splines}
+  \section{"Inertia Tensor" for Cubic Splines}\label{app:dp_proof}
   \begin{code}
@@
 def round_expr(expr, num_digits):
-  return expr.xreplace({n : round(n, num_digits) for n in expr.atoms(Number)})
+  return expr.xreplace(
+  {n : round(n, num_digits) for n in expr.atoms(Number)}
+  )
 
@@
     for k,ck in enumerate(gammas):
-      # xi_xp is the distance from parametrized position to grid node [i,j,k]
+      # xi_xp is the distance from parametrized
+      # position to grid node [i,j,k]
       xi_xp = np.array([ai,bj,ck])
       # each outer product weighted by interpolation functions
-      this_outer = wip(i,ai)*wip(j,bj)*wip(k,ck)*np.outer(xi_xp,xi_xp)
+      this_outer = wip(i,ai)*wip(j,bj)*wip(k,ck)*
+		    np.outer(xi_xp,xi_xp)
       # summed up over all grid nodes
@@
   for j,D_ij in enumerate(D_row):
-    #simplify to cancel polynoms, round_expr because numerical cancellation
+    # simplify to cancel polynoms
+    # round_expr because of numerical cancellation
     D[i][j] = round_expr(simplify(D_ij),14)
@@
       0.         &0.         &0.33333333\\
-  \end{array}\right]=\boldsymbol{D}_p
+  \end{array}\right]= \frac{\boldsymbol{D}_p}{h^2}
   $$

### 2018-11-22 mscthesis.tex
discretization space

@@
   _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
-= \vec{\nabla}} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
+= \vec{\nabla} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
 \end{equation}
@@
 as assumed before. Note that $\boldsymbol{P}$ however is not constrained to be symmetric.
-\subsubsection{Weak formulation}
+\subsubsection{Weak formulation}\label{sec:weak}
 Before deriving the weak form an explanation of what it achieves is in order.
@@
 \subsubsection{Discretize time}
-Any conservative integrator could be used to discretize time. The implicit midpoint rule is one of such ($\lambda =\frac{1}{2}$)
+Any conservative integrator could be used to discretize time. The implicit midpoint rule ($\lambda =\frac{1}{2}$) is one of such
 \begin{equation}\label{eq:midpoint}
-  \frac{y ^ { n + 1 } - y ^ { n }}{ \Delta t} =  f \left(t^n+\frac{\Delta t}{2}, \frac { y ^ { n } + y^{n+1}}{2}\right)
+  \frac{y ^ { n + 1 } - y ^ { n }}{ \Delta t} = f^{n+\lambda} = f \left(t^n+\lambda\Delta t, (1-\lambda) y ^ { n } +\lambda y^{n+1}\right)
 \end{equation}
@@
 Replacing the Lagrangian acceleration $_0a_\alpha$ with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
-\begin{equation}
+\begin{equation}\label{eq:time_disc}
   \frac{1}{\Delta t} \int _{ \Omega^{t^n}}(_tq_\alpha)(_t\rho)({_tv_\alpha^{n+1}}-{_tv_\alpha^{n}}) d_t\boldsymbol{x} =
@@
 \end{equation}
+\subsubsection{Discretize space}
+A Galerkin discretization brings all spatial terms of equation \ref{eq:time_disc} to a finite-dimensional space: $\boldsymbol{q} \rightarrow \boldsymbol{q}_h$.
+To not further clutter up the notation, the $h$ will be omitted.
+This will replace $q_\alpha, v^n_\alpha , v_\alpha^{n+1}$ with their finite-dimensional grid-based interpolants:
+\begin{equation}
+  {_t q } _ { \alpha } ^ {  { n } } =  ({_t q } _ {  { i } \alpha } ^ {  { n } })  ({_t w } _ { i }) , \quad _tv _ { \alpha } ^ {  { n } } = ({_tv _ {  { j\alpha } } ^ {  { n } }} )( {_t w } _ {  { j } }) , \quad {_tv _ { \alpha } ^ {  { n } + 1 }} = ({_tv _ {  { j } \alpha } ^ {  { n } + 1 }}) ({_t w } _ {  { j } }).
+\end{equation}
+Further chapter \ref{sec:weak} mentions that $\boldsymbol{q}$ can be chosen arbitrarily.
+The Galerkin discretization of a $d$-dimensional space with $m$ grid nodes therefore uses the standard basis functions $e_1,e_2, ...
+,e_{d \times m}$.
+Due to the scalar-product $d\times m$ equations would need to be solved:
+\begin{equation}\label{eq:galerkin}
+  \frac{1}{\Delta t} \int _{ \Omega^{t^n}} ({ _tw } _ {i  })(_t\rho)({_t w } _ { j  })({_tv_{j\alpha}^{n+1}}-{_tv_{j\alpha}^{n}}) d_t\boldsymbol{x} =
+ - \int _ { \Omega ^ {  t^ n } }   \frac{\partial{_t w } _ { {i  }}}{\partial_tx_\beta }  \sigma _ {{ \alpha } \beta }  { d } \boldsymbol { x }.
+\end{equation}
+A mass matrix can be factored out as:
+\begin{equation}
+  m_{ij}^n = \int _{ \Omega^{t^n}} {_t w } _ {i  }(_t\rho){_t w } _ { j  } d_t\boldsymbol{x}.
+\end{equation}
+The Lagrangian pull-back relates this to the initial density in the Lagrangian view and discretizing the integral with the initial time-invariant particle mass $m_p \approx V_p^0\rho(_0x_p,0)$:
+\begin{equation}
+m_{ij}^n = \int _{ \Omega^{t^0}} ({_tw_i})(_0\rho_0)({_tw} _ { j  }) d_0 \boldsymbol{x} \approx \sum _ {  { p } }  { m } _ {  { p } }  { w } _ {  { i } } (\boldsymbol { x } _ {  { p } } )  { w } _ {  { j } } ( \boldsymbol { x } _ {  { p } } ).
+\end{equation}
+This matrix is symmetric positive semi-definite (since mass is positive).
+Numerically this matrix is mostly not used as is due to possibility of it being singular.
+This is solved commonly due to a mass lumping strategy.
+Replace $m^n_{ii}$ with the $i$-th row sum to get:
+\begin{equation}\label{eq:mass_discr}
+\sum _ {  { p } }  { m } _ {  { p } }  { w } _ {ip}   w_{jp}
+\stackrel{\text{\ref{eq:partition_unity}}}{\approx}
+\sum_p m_p w_{ip}
+\stackrel{\text{\ref{eq:mpm:mass}}}{=}
+m_i,
+\end{equation}
+where partition of unity $\sum_j w_{jp} = 1$ is used. This is exactly the mass transfer as in eq. \ref{eq:mpm:mass}, s.t. no further assembling of a mass matrix is needed.
+Discretizing the right side of \ref{eq:galerkin} with an estimated per particle stress $\boldsymbol{\sigma}_p^{n+\lambda}$:
+\begin{equation}\label{eq:stress_discr}
+\int _ { \Omega ^ {  t^ n } }   \frac{\partial{_t w } _ { {i  }}}{\partial_tx_\beta }  \sigma _ {{ \alpha } \beta }  { d } \boldsymbol { x } \approx
+\sum _ { p } (\sigma _ { p }^{n+\lambda}) _ { \alpha \beta } \frac{{ \partial w } _ { {ip}}^n}{ \partial x_\beta }  V _ { p } ^ { n }.
+\end{equation}
+Setting in equation \ref{eq:mass_discr} and \ref{eq:stress_discr} into \ref{eq:galerkin} summarizes the space discretization as:
+\begin{equation}\label{eq:space_discr}
+  \frac{1}{\Delta t} ({m\boldsymbol{v}_i^{n+1}}-{m\boldsymbol{v}_i^{n}}) =
+  -\sum _ { p } \boldsymbol{\sigma} _ { p }^{n+\lambda} \nabla w_{ip}^n V _ { p } ^ { n }
+  = \boldsymbol{f}_i^{n+\lambda}.
+\end{equation}
+The momentum change of the left side is by construction equal to a (grid node) force.\\
+Given that the material point method keeps track of the deformation by a deformation gradient.
+Each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
+Based on this one may also gain a volume change measure around the particle as $J^n_p = \text{det}(\boldsymbol{F}^n_p)$.
+Starting with an initial volume of a particle $V_p^0$ the volume may be tracked in time by:
+\begin{equation}\label{eq:volume_discr}
+  V^n_p \stackrel{\text{\ref{eq:j}}}{\approx} V_p^0J^n_p.
+\end{equation}
+In eq. \ref{eq:piola} an alternate measure for the stress by the first Piola Kirchhoff stress. The results of \ref{eq:space_discr} may therefore equally expressed by it:
+$$
+  f_{i\alpha} \stackrel{\text{\ref{eq:piola},\ref{eq:volume_discr}}}{=}
+  - \sum_p \frac{1}{J^n_p}P^n_{p\alpha\gamma}F^n_{p\beta\gamma}\frac{\partial w_{ip}}{\partial x_\beta}V^0_pJ^n_p
+$$
+\begin{equation}
+  = - \sum_p P^n_{p\alpha\gamma}F^n_{p\beta\gamma}\frac{\partial w_{ip}}{\partial x_\beta}V^0_p
+\end{equation}
+\cite{MPM:COURSE} \cite{MPM:APIC} \cite{bathe2006finite}
+\subsubsection{Deformation gradient evolution}
 \subsection{GPU memory}

### 2018-11-23 mscthesis.tex
implicit midpoint rule, little problematic

@@
 This is not to be confused with the following Einstein-Notation.
-Therefore beginning with chapter \ref{sec:mat_point} Einstein-Notation will use greek letters ($\alpha,\beta$).
+Therefore beginning with chapter \ref{sec:mat_point} Einstein-Notation for components will use greek letters ($\alpha,\beta$).
+Occasionally summation over grid index $i$ and particles $p$ may be implied.
 A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n} t^i$ will have a right superscript: $\boldsymbol{x}^n$.
@@
 \stackrel{\text{\ref{eq:mpm:mass}}}{=}
-m_i,
+m_i^n,
 \end{equation}
@@
 \begin{equation}\label{eq:space_discr}
-  \frac{1}{\Delta t} ({m\boldsymbol{v}_i^{n+1}}-{m\boldsymbol{v}_i^{n}}) =
+  \frac{1}{\Delta t} (({m^n\boldsymbol{v}^{n+1}})_i-({m^n\boldsymbol{v}^n})_i) =
   -\sum _ { p } \boldsymbol{\sigma} _ { p }^{n+\lambda} \nabla w_{ip}^n V _ { p } ^ { n }
@@
 $$
-  f_{i\alpha} \stackrel{\text{\ref{eq:piola},\ref{eq:volume_discr}}}{=}
-  - \sum_p \frac{1}{J^n_p}P^n_{p\alpha\gamma}F^n_{p\beta\gamma}\frac{\partial w_{ip}}{\partial x_\beta}V^0_pJ^n_p
-$$
-\begin{equation}
-  = - \sum_p P^n_{p\alpha\gamma}F^n_{p\beta\gamma}\frac{\partial w_{ip}}{\partial x_\beta}V^0_p
+\boldsymbol{f}^{n+\lambda}_i \stackrel{\text{\ref{eq:piola},\ref{eq:volume_discr}}}{=}
+- \sum_p \frac{1}{J^n_p}\boldsymbol{P}^{n+\lambda}_{p}(\boldsymbol{F}^{n}_{p})^T \nabla w_{ip}V^0_pJ^n_p
+$$
+\begin{equation}
+  = - \sum_p \boldsymbol{P}^{n+\lambda}_{p}(\boldsymbol{F}^{n}_{p})^T \nabla w_{ip}V^0_p.
 \end{equation}
@@
 \subsubsection{Deformation gradient evolution}
+In eq. \ref{eq:evol_def_grad} the evolution of the deformation gradient is shown to be:
+\begin{equation}
+  \frac{\partial ^t_0\boldsymbol{F}}{\partial t} =  \nabla{_0\boldsymbol{v}}(_0\boldsymbol{x},t).
+\end{equation}
+Discretizing the Lagrangian deformation gradient in time with eq. \ref{eq:midpoint} results in:
+\begin{equation}
+  \frac{\boldsymbol{F}^{n+1}_p + \boldsymbol{F}^{n}_p}{\Delta t} =  \nabla_0\boldsymbol{v}^{n+\lambda}(_0\boldsymbol{x}).
+\end{equation}
+Pushing the right side forward to Eulerian view
+\begin{equation}
+  \frac{\boldsymbol{F}^{n+1}_p + \boldsymbol{F}^{n}_p}{\Delta t} =  \nabla_t\boldsymbol{v}^{n+\lambda}(_t\boldsymbol{x})^t_0\boldsymbol{F} = \nabla_t\boldsymbol{v}^{n+\lambda}(_t\boldsymbol{x})\boldsymbol{F}^n_p
+\end{equation}
+and further applying the Galerkin discretization
+\begin{equation}
+  (_tv^{n+\lambda})_\alpha = (v_{i}^{n+\lambda})_\alpha w_i \Rightarrow \frac{\partial (_tv^{n+\lambda})_\alpha}{\partial x_\beta} =(v_i^{n+\lambda})_\alpha\frac{\partial w_i}{\partial x_\beta}
+\end{equation}
+leads to the final update rule for the deformation gradient:
+\begin{equation}
+\boldsymbol{F}^{n+1}_p = \left( \boldsymbol{I} + \Delta t\sum_i \boldsymbol{v}_i^{n+\lambda}({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
+\end{equation}
+\subsubsection{Grid nodal forces}
+Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint}:
+$$
+\frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda} =  (1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}
+  $$
+\begin{equation}
+  \Leftrightarrow \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \left((1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}\right).
+\end{equation}
+Due to \ref{eq:space_discr} a velocity update can be put together as
+\begin{equation}
+  \frac{1}{\Delta t} (({m^n\boldsymbol{\hat{v}}^{n+1}})_i-({m^n\boldsymbol{v}^n})_i)
+  = \boldsymbol{f}_i^{n+\lambda}
+ \Leftrightarrow
+ \boldsymbol{\hat{v}}^{n+1}_i
+ = \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\lambda},
+\end{equation}
+where the force can be also described in terms of the in-between point:
+\begin{equation}
+ \boldsymbol{f}_i^{n+\lambda} = \boldsymbol{\sigma}^{n+\lambda}_p(\boldsymbol{F}_p^{n+\lambda}(\boldsymbol{\hat{x}}_i^{n+1})).
+\end{equation}
 \subsection{GPU memory}

### 2018-11-24 mscthesis.tex
grid nodal forces

@@
 $$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
-\subsection{Motivation}
 \section{Related Work}
@@
 \cite{NVIDIA:SHUFFLE}
-
+\cite{MPM:POLYPIC}
 \cite{MPM:GPU}
@@
 \subsubsection{Linear Elasticity}\label{sec:linear_elasticty}
-The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
-For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\psi$ as:
+The first aim will be to find the strain energy density $\Psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
+For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\Psi$ as:
 The most general linear stress-strain relationship is given by Hooke's Law in three dimensions
@@
   \item No-net-rotation: $\epsilon_{kl} = \epsilon_{lk} \Rightarrow C_{ijkl} = C_{ijlk}$
-  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\psi$:
-    \begin{equation}
-      C _ { i j k l } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },
+  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\Psi$:
+    \begin{equation}
+      C _ { i j k l } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },
     \end{equation}
@@
 \begin{equation}\label{eq:partial_energy}
-  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },
-\end{equation}
-if such a $\psi$ is given.
+  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \Psi } } { \partial \epsilon _ { i j } },
+\end{equation}
+if such a $\Psi$ is given.
 An isotropic (direction-independent) linear elastic material further only has three unique elements $C _ {i j k l}$.
@@
 $\gamma$ is also referred to as the engineering strain.
-Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:
+Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\Psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:
 \begin{equation}\label{eq:linear_elas}
-  \psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).
+  \Psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).
 \end{equation}
 \cite{MIT:LINEAR_ELASTICITY}
-\subsubsection{Corotated Hyperelasticity}
+\subsubsection{Corotated Hyperelasticity}\label{sec:cor_hyper}
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
@@
 The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
-to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$.
+to $\boldsymbol{F}_E$: $\Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$.
 Rigid body motions however don't result in strain and consequently don't need to be recovered from.
@@
 \begin{equation}\label{eq:corot_S}
-  \Psi = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )
+  \Psi_{CH} = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )
 \end{equation}
@@
 \begin{equation}
-  \Psi = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)
+  \Psi_{CH} = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)
 \end{equation}
@@
 \begin{equation} \label{eq:corot_sing}
-  \Psi(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )
+  \Psi_{CH}(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )
 \end{equation}
@@
 \begin{equation}\label{eq:piola}
-  \boldsymbol{P} = \boldsymbol{\sigma}\boldsymbol{F}^{-T}J
+  \boldsymbol{P} =
+  \frac{\partial \Psi}{\partial \boldsymbol{F}}=
+  \boldsymbol{\sigma}\boldsymbol{F}^{-T}J.
 \end{equation}
@@
 \cite{MPM:APIC}\cite{MPM:COURSE}
-\subsubsection{APIC momentum Transfers}
+\subsubsection{APIC transfers}
 The momentum transfer round trip could be defined as:
@@
 \end{equation}
-  \item Factoring out mass.
+  \item Factoring out mass:
     \begin{equation}
@@
 
-  \item Grid to Particle Transfer (in a $PIC$-manner):
+  \item Grid to particle transfer (in a $PIC$-manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
+    \begin{equation}
+      \boldsymbol{x}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{x}_i^{n+1}
+    \end{equation}
+
+    \begin{equation}
+      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}
+    \end{equation}
 $$
@@
     \begin{equation}\label{eq:apic_mom_g2p}
-+ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } )
-    \end{equation}
-    \begin{equation}
-      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}.
++ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } ).
     \end{equation}
@@
 The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following for MPM preferable description:
-\begin{equation}
+\begin{equation} \label{eq:weak_mpm}
   \int _ { \Omega ^ { 0 } } (_0q _ \alpha) (_0\rho_0) (_0a _\alpha)  d _0\boldsymbol{x}= \int _ { \partial \Omega ^ { t^n } } _tq _ \alpha \sigma _ {\alpha\beta} d_tA_\beta(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t^n} } \frac{\partial _tq _ \alpha}{\partial _tx_\beta} \sigma _ {\alpha\beta} d_t\boldsymbol{x}.
@@
 \subsubsection{Discretize time}
-Any conservative integrator could be used to discretize time. The implicit midpoint rule ($\lambda =\frac{1}{2}$) is one of such
-\begin{equation}\label{eq:midpoint}
+Any integrator conserving linear and angular momentum could be used to discretize time. The class of time integrators used here are characterized by \begin{equation}\label{eq:midpoint}
   \frac{y ^ { n + 1 } - y ^ { n }}{ \Delta t} = f^{n+\lambda} = f \left(t^n+\lambda\Delta t, (1-\lambda) y ^ { n } +\lambda y^{n+1}\right)
@@
 \end{equation}
-Replacing the Lagrangian acceleration $_0a_\alpha$ with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
+A prominent member of this class is the implicit midpoint rule ($\lambda =\frac{1}{2}$).
+Replacing the Lagrangian acceleration $_0a_\alpha$ in eq.
+\ref{eq:weak_mpm} with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
 \begin{equation}\label{eq:time_disc}
@@
 $$
-\begin{equation}
+\begin{equation}\label{eq:force_disc}
   = - \sum_p \boldsymbol{P}^{n+\lambda}_{p}(\boldsymbol{F}^{n}_{p})^T \nabla w_{ip}V^0_p.
@@
 leads to the final update rule for the deformation gradient:
-\begin{equation}
+\begin{equation}\label{eq:evol_def_grad_disc}
 \boldsymbol{F}^{n+1}_p = \left( \boldsymbol{I} + \Delta t\sum_i \boldsymbol{v}_i^{n+\lambda}({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
 \end{equation}
+The discretization of the position will also be of need to advance the particles and weight them back:
+\begin{equation}  \frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1}
+  - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda}.
+\end{equation}
+The grid position $\boldsymbol{\hat{x}}_i^{n+1}$ does not correspond to an actual deformation.
+The grid never actually gets deformed (unlike in FEM-methods).
+Therefore the discretized evolution of the deformation gradient  \ref{eq:evol_def_grad_disc} is directly a function of $\boldsymbol{\hat{x}}$.
+For the point $\boldsymbol{\hat{x}}_i^{n+1}$ this becomes:
+\begin{equation}\label{eq:evol_def_grad_disc_x}
+  \boldsymbol{\hat{F}}^{n+1}_p(\boldsymbol{\hat{x}}_i^{n+1}) =  \left( \boldsymbol{I} + \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
+\end{equation}
+As part of the class of time integrators in use (\ref{eq:midpoint}) a function of $\boldsymbol{x}_i$ gets evaluated at an in-between point given by:
+\begin{equation}
+  \boldsymbol{x}_i^{n+\lambda} = \lambda \boldsymbol{x}_i^{n+1} + (1-\lambda)\boldsymbol{x}_i^n.
+\end{equation}
+Plugging this point into \ref{eq:evol_def_grad_disc_x} leads to the following generalization:
+$$\boldsymbol{\hat{F}}^{n+\lambda}_p(\boldsymbol{x}_i^{n+\lambda}) = \left( \boldsymbol{I} + \lambda \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p
+  $$
+  \begin{equation} \label{eq:def_grad_lambda}
+  = (1-\lambda)\boldsymbol{F}^n_p + \lambda \boldsymbol{F}^{n+1}_p
+\end{equation}
+This
+\cite{MPM:COURSE}
 \subsubsection{Grid nodal forces}
-Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint}:
-$$
-\frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda} =  (1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}
-  $$
-\begin{equation}
-  \Leftrightarrow \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \left((1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}\right).
-\end{equation}
-Due to \ref{eq:space_discr} a velocity update can be put together as
-\begin{equation}
-  \frac{1}{\Delta t} (({m^n\boldsymbol{\hat{v}}^{n+1}})_i-({m^n\boldsymbol{v}^n})_i)
-  = \boldsymbol{f}_i^{n+\lambda}
- \Leftrightarrow
+The notion of a total elastic potential energy function $\Psi$ was introduced in chapter \ref{sec:cor_hyper}. The MPM approximation to this function can be defined by:
+\begin{equation}
+  e(\hat{\boldsymbol{x}}) = \sum_pV^0_p \Psi(\hat{\boldsymbol{F}}_{Ep}(\hat{\boldsymbol{x}})).
+\end{equation}
+The evolution of the deformation gradient \ref{eq:evol_def_grad_disc_x} for a general $\boldsymbol{\hat{x}}_i$  is necessary
+$$\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
+= \delta_{\omega \alpha}
+\frac{\partial w_{ip}}{\partial x_\gamma}F_{\gamma\beta} = \delta_{\omega \alpha}F_{\gamma\beta} \frac{\partial{w_{ip}}}{\partial x_\gamma}$$
+for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of eq. \ref{eq:force_disc}:
+$$
+  \frac{\partial e}{\partial \hat{x}_{i\alpha}}(\hat{\boldsymbol{x}})
+  = \sum_p V^0_p \frac{\partial \Psi}{\partial \hat{F}_{\omega\beta}}(\hat{\boldsymbol{F}}_{Ep})\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
+  $$
+\begin{equation}
+  = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}
+\end{equation}
+where the force can be also described in terms of the Cauchy stress:
+\begin{equation}
+  \boldsymbol{\hat{f}}_i = - \sum_p V^n_p \boldsymbol{\sigma}_p(\boldsymbol{\hat{F}}_{Ep})\nabla{w^n_{ip}}.
+ \end{equation}
+Due to eq. \ref{eq:def_grad_lambda} we can for the class of functions summarize the stress computation as:
+\begin{equation}
+  \boldsymbol{P}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{P}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}),
+  \quad
+  \boldsymbol{\sigma}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{\sigma}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}).
+\end{equation}
+\cite{MPM:COURSE}\cite{MPM:APIC}
+\subsubsection{Symplectic midpoint scheme}
+Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
+\begin{equation}
+  \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \boldsymbol{v}_i^{n+\frac{1}{2}}.
+\end{equation}
+Due to \ref{eq:space_discr} a velocity update can be put together as:
+\begin{equation}
 \boldsymbol{\hat{v}}^{n+1}_i
- = \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\lambda},
-\end{equation}
-where the force can be also described in terms of the in-between point:
-\begin{equation}
- \boldsymbol{f}_i^{n+\lambda} = \boldsymbol{\sigma}^{n+\lambda}_p(\boldsymbol{F}_p^{n+\lambda}(\boldsymbol{\hat{x}}_i^{n+1})).
-\end{equation}
+ = \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\frac{1}{2}}.
+\end{equation}
+\begin{equation}
+\end{equation}
+\cite{MPM:APIC}
 \subsection{GPU memory}

### 2018-11-25 mscthesis.tex
conj. gradients, midpoint

@@
     \end{equation}
-  \item $C^1$-continuity s.t. $\nabla w_i$ is continuous.
+  \item $C^1$-continuity s.t. $\nabla w$ is continuous.
 \end{enumerate}
@@
 \begin{equation}
-  = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}
-\end{equation}
-where the force can be also described in terms of the Cauchy stress:
+  = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}.
+\end{equation}
+The nodal force can be also described in terms of the Cauchy stress:
 \begin{equation}
@@
  \end{equation}
-Due to eq. \ref{eq:def_grad_lambda} we can for the class of functions summarize the stress computation as:
+Due to eq. \ref{eq:def_grad_lambda} the stress computation is summarized for the class of functions in use as:
 \begin{equation}
@@
 Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
-\begin{equation}
+\begin{equation}\label{eq:vel_midpoint}
   \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \boldsymbol{v}_i^{n+\frac{1}{2}}.
@@
 Due to \ref{eq:space_discr} a velocity update can be put together as:
-\begin{equation}
+\begin{equation}\label{eq:force_midpoint}
 \boldsymbol{\hat{v}}^{n+1}_i
@@
 \end{equation}
-\begin{equation}
+The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the Material Point Method differs from this by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
+\begin{equation}\label{eq:trapezoidal}
+  \boldsymbol{v}_i^{n+\frac{1}{2}} \approx \frac{\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n}}{2}.
+\end{equation}
+The trapezoidal rule has the same order of error $O(\Delta t^2)$ as the implicit midpoint scheme.
+This modification allows for a more direct one-step scheme as shown in the following.
+Plugging in eq.
+\ref{eq:vel_midpoint} into \ref{eq:force_midpoint}:
+\begin{equation}
+  m_i^n \frac{\boldsymbol{\hat{v}}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} = \boldsymbol{f}_i \left( \frac{\boldsymbol{x_i}^n+\boldsymbol{x_i}^n + \Delta t \left( \frac{\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n}}{2}
+\right)
+    }{2}\right)
 \end{equation}
@@
 \begin{algorithm}[t]
-  \caption{Euclid’s algorithm}\label{euclid}
+  \caption{Conjugate gradient}\label{alg:conj_grad}
   \begin{algorithmic}[1]
-    \Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-    \State $r\gets a\boldsymbolod b$ and \var{foobar}\label{foobar}
-    \While{$r\not=0$}\Comment{We have the answer if r is 0}
-    \State $a\gets b$
-    \State $b\gets r$
-    \State $r\gets a\boldsymbolod b$
-    \EndWhile\label{euclidendwhile}
-    \State \textbf{return} $b$\Comment{The gcd is b}
+    \Procedure{CONJUGATE-GRADIENT}{$A,x,b$}
+    \State $\boldsymbol{r}\gets \boldsymbol{b} - \boldsymbol{Ax}$
+    \State $\boldsymbol{p}\gets \boldsymbol{r}$
+    \State $\gamma \gets \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
+    \Repeat
+      \State $\boldsymbol { s } \gets \boldsymbol { A } \boldsymbol { p }$
+      \State $\alpha \gets \frac { \gamma } { \langle \boldsymbol { p } , \boldsymbol { s } \rangle }$
+      \State $\boldsymbol { x } \gets \boldsymbol { x } + \alpha \boldsymbol { p }$
+      \State $\boldsymbol { r } \leftarrow \boldsymbol { r } - \alpha \boldsymbol { s }$
+      \State $\kappa \leftarrow \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
+      \If{$ \kappa < \epsilon$} \Comment{alt. fixed amount of steps}
+        \State \Return \label{alg:conj_grad:exit}
+      \EndIf
+      \State $\beta \gets \frac{\kappa}{\gamma}$
+      \State $\boldsymbol { p } \leftarrow \boldsymbol { r } + \beta \boldsymbol { p }$
+      \State $\gamma \gets \kappa$
+      \Until{false} \Comment{exit at \ref{alg:conj_grad:exit}}
     \EndProcedure
@@
 Sparse: \cite{OPENVDB} \cite{NVIDIA:GVDB_VOXELS}
+\clearpage
 \begin{appendices}

### 2018-11-26 mscthesis.tex
objective

@@
 \addbibresource{mscthesis.bib}
+
+\DeclareMathOperator*{\argmin}{argmin}
 \makeatletter
@@
 As part of the class of time integrators in use (\ref{eq:midpoint}) a function of $\boldsymbol{x}_i$ gets evaluated at an in-between point given by:
-\begin{equation}
+\begin{equation}\label{eq:midpoint_x}
   \boldsymbol{x}_i^{n+\lambda} = \lambda \boldsymbol{x}_i^{n+1} + (1-\lambda)\boldsymbol{x}_i^n.
@@
 The notion of a total elastic potential energy function $\Psi$ was introduced in chapter \ref{sec:cor_hyper}. The MPM approximation to this function can be defined by:
-\begin{equation}
+\begin{equation}\label{eq:energy_discr}
   e(\hat{\boldsymbol{x}}) = \sum_pV^0_p \Psi(\hat{\boldsymbol{F}}_{Ep}(\hat{\boldsymbol{x}})).
@@
 \end{equation}
-The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the Material Point Method differs from this by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
+A variable of $n+\frac{1}{2}$ then gets evaluated at the midpoint (\ref{eq:midpoint_x}):
+\begin{equation}
+  \boldsymbol{f}^{n+\frac{1}{2}}_i =  \boldsymbol{f}_i\left(\frac{\boldsymbol{x}_i^n + \boldsymbol{\hat{x}}_i^{n+1}}{2}\right).
+\end{equation}
+The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the Material Point Method differs by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
 \begin{equation}\label{eq:trapezoidal}
@@
 This modification allows for a more direct one-step scheme as shown in the following.
-Plugging in eq.
+Plugging $\boldsymbol{x}_i^{n+1}$ of eq.
 \ref{eq:vel_midpoint} into \ref{eq:force_midpoint}:
 \begin{equation}
-  m_i^n \frac{\boldsymbol{\hat{v}}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} = \boldsymbol{f}_i \left( \frac{\boldsymbol{x_i}^n+\boldsymbol{x_i}^n + \Delta t \left( \frac{\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n}}{2}
-\right)
-    }{2}\right)
-\end{equation}
-\cite{MPM:APIC}
+  h(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.
+\end{equation}
+And substituting
+One can recast this method back to an energy function by integrating:
+\begin{equation}
+  E(\boldsymbol{v}_i) = \sum_i \frac{m_i^n}{8} \|\boldsymbol{v}_i - \boldsymbol{v}_i^n\|^2_2 + e(\boldsymbol{x}_i^n + \frac{\Delta t}{4}(\boldsymbol{v}_i + \boldsymbol{v}_i^n)),
+\end{equation}
+where $e$ is just the discretized elastic potential energy out of \ref{eq:energy_discr}.
+In general this is a optimization objective that needs minimizing to solve for the updated velocities.
+Recasting allows solving in a general scheme by an optimization integrator.
+
+In general starting off with an energy description is a more physical approach of the problem.
+The first term can be identified clearly as the kinetic energy.
+Moreover it easily allows adding potential terms for gravity or collision \cite{MPM:OPTIMI_INTEGR}.
+Or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
+The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
+\begin{equation}
+  \argmin_{\forall \boldsymbol{v}_i}(E(\boldsymbol{v}_i)) \Leftrightarrow g(\boldsymbol{v}_i) = \frac{\partial E}{\partial \boldsymbol{v}_i} = 0.
+\end{equation}
+Since the objective from the start was minimizing $\boldsymbol{E}$ to find $\boldsymbol{v}_i$ the scale of this function can be chosen arbitrarily, i.e.
+a zero crossing does not scale.
+The scaling here can be identified as:
+\begin{equation}
+  g(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
+\end{equation}
+\cite{MPM:COURSE}\cite{MPM:APIC}
+
 \subsection{GPU memory}

### 2018-11-27 mscthesis.tex
hessian

@@
   $$
-\begin{equation}
+  \begin{equation}\label{eq:nodal_force}
   = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}.
@@
 \begin{equation}
-  h(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.
+  \boldsymbol{h}(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.
 \end{equation}
@@
 \begin{equation}
-  g(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
+  \boldsymbol{g}(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
 \end{equation}
 \cite{MPM:COURSE}\cite{MPM:APIC}
-
-\subsection{GPU memory}
-\subsubsection{Optimization opportunities}
-\cite{NVIDIA:BEST:PRACTICE}
-\cite{AMD:GPU_OPEN}
-\begin{itemize}
-  \item Minimize CPU-GPU transfers
-  \item Coalesced memory access
-  \item Bank conflicts
-\end{itemize}
-\subsubsection{SoA vs. AoS}
-\section{Implementation}
+\subsection{Newton's Method}
 \newcommand{\var}[1]{{\ttfamily#1}}% variable
@@
   \begin{algorithmic}[1]
-    \Procedure{CONJUGATE-GRADIENT}{$A,x,b$}
-    \State $\boldsymbol{r}\gets \boldsymbol{b} - \boldsymbol{Ax}$
+    \Procedure{CONJUGATE-GRADIENT}{$H,x,f$}
+    \State $\boldsymbol{x} \gets \Call{InitialGuess}$
+    \State $\boldsymbol{Hx} \gets \Call{ComputeHp}{x}$
+    \State $\boldsymbol{r}\gets \boldsymbol{f} - \boldsymbol{Hx}$
     \State $\boldsymbol{p}\gets \boldsymbol{r}$
@@
     \Repeat
-      \State $\boldsymbol { s } \gets \boldsymbol { A } \boldsymbol { p }$
+      \State $\boldsymbol{Hp} \gets \Call{ComputeHp}{p}$
+      \State $\boldsymbol { s } \gets \boldsymbol { H } \boldsymbol { p }$
       \State $\alpha \gets \frac { \gamma } { \langle \boldsymbol { p } , \boldsymbol { s } \rangle }$
@@
 \end{algorithm}
+Many minimization and finding zero crossing algorithms are available. One of such is Newton's Method, which allows for rapid quadratic convergence in a near local neighborhood.
+\begin{equation}
+  \boldsymbol{v}_i^{(i+1)} = \boldsymbol{v}^{(i)}_i + \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right]^{-1}\boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
+\end{equation}
+Computing the inverse is numerically irresponsible. Instead the following linear system is solved $\left(\Delta \boldsymbol{v} = \boldsymbol{v}_i^{(i+1)} - \boldsymbol{v}^{(i)}_i\right)$:
+\begin{equation}\label{eq:linear_system}
+  \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} =  \boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
+\end{equation}
+Using the Newton's Method however requires a computation of the Hessian of $E(\boldsymbol{v}_i)$:
+\begin{equation}
+  \frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j} = \frac{m_i^n}{4} - \frac{\Delta t^2}{16}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
+\end{equation}
+Computing $\frac{\partial f_i}{\partial x_j}$ for every combination $ij$ would be quite memory-intensive.
+Instead immediately the matrix-vector product on an increment $\delta \boldsymbol{u}_j$ is solved:
+\begin{equation}\label{eq:hessian-matrix-vector}
+  \sum_j\frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j}\delta \boldsymbol{u}_j = \frac{m_i^n}{4}\delta \boldsymbol{u}_j - \frac{\Delta t^2}{16}\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right) \delta \boldsymbol{u}_j.
+\end{equation}
+The linear system can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}) since $\frac{\partial g}{\partial v}$ is symmetric, positive definite.
+This is due to being the Hessian of the convex function $E$.
+There is to important things notice:
+\begin{enumerate}
+  \item Conservation of momentum on incomplete convergence is only fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $M^{-1}$.
+More in \cite{MPM:APIC}.
+Thus the equation becomes:
+    $$
+      \left(\frac{1}{4} - \frac{\Delta t^2}{16}\frac{1}{m_i^n}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right)\right) \Delta v
+      $$
+    \begin{equation}
+      = \frac{\boldsymbol{v}_i^{(i)}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4}\frac{1}{m_i^n}\boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right).
+    \end{equation}
+  \item Due to \ref{eq:hessian-matrix-vector} the product of the Hessian with an increment ($\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j$) needs to be computed anew every step, since a pure matrix store is too costly.
+\end{enumerate}
+The Hessian increment can be computed with a two-stage process:
+\begin{enumerate}
+  \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid to particle transfer:
+    \begin{equation}
+    \boldsymbol{A}_p = \frac { \partial ^ { 2 } \Psi } { \partial \boldsymbol { F } \partial \boldsymbol { F } } ( \hat { \boldsymbol { F } } _ { p } ( \hat { \boldsymbol { x } } )) : \left(\sum_j \delta \boldsymbol{u}_j (\nabla w_{jp}^n)^T \boldsymbol{F}^n_p\right).
+    \end{equation}
+  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle to grid transfer:
+\begin{equation}
+  - \delta \boldsymbol{f}_i = - \sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j = \sum_p V_p^0 \boldsymbol{A}_p(\boldsymbol{F}^n_p)^T \nabla w_{ip}^n.
+\end{equation}
+\end{enumerate}
+\subsection{GPU memory}
+\subsubsection{Optimization opportunities}
+\cite{NVIDIA:BEST:PRACTICE}
+\cite{AMD:GPU_OPEN}
+\begin{itemize}
+  \item Minimize CPU-GPU transfers
+  \item Coalesced memory access
+  \item Bank conflicts
+\end{itemize}
+\subsubsection{SoA vs. AoS}
+\section{Implementation}
 \subsection{Framework}

### 2018-11-28 mscthesis.tex
bug fixed

@@
 \documentclass[m,times]{cgMA}
-
 \usepackage{listings}
@@
 \usepackage{minted}
+\usepackage{verbatim}
 \addbibresource{mscthesis.bib}
@@
 \begin{document}
-
 \author{Fabian Meyer}
@@
 \section{Related Work}
-\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}
+\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}\\
+\cite{stantchev2008fast}\\
 benefits over radix sort /thrust sort-by-key -> counting sort
-\cite{NVIDIA:SHUFFLE}
-\cite{MPM:POLYPIC}
-\cite{MPM:GPU}
+\cite{NVIDIA:SHUFFLE}\\
+\cite{MPM:POLYPIC}\\
+\cite{MPM:GPU}\\
 %look again for MPM/PIC gpu impl
@@
 \begin{enumerate}
-
   \item Partition of unity:
@@
 As a reminder of chapter \ref{sec:notation}: $\boldsymbol{x}_p$ refers to a particle's position. $\boldsymbol{x}_i$ to a grid cell's position.
-\\
 Often dyadic products of one-dimensional interpolation functions suffice
@@
     \end{equation}
-
   \item Grid to particle transfer (in a $PIC$-manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
@@
     \end{equation}
-
     \begin{equation}
@@
     $$
-    \boldsymbol { B } _ { p } ^ { n + 1 } = \frac { 1 } { 2 } \sum _ { i } w _ { i p } ^ { n } (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } +  { \boldsymbol { x } } _ { i } ^ { n + 1 } - \boldsymbol { x } _ { p } ^ { n + 1 } ) ^ { T }
+    {^-}\Delta \boldsymbol{x} = \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } +   \boldsymbol { x }_ { i } ^ { n + 1 } - \boldsymbol { x } _ { p } ^ { n + 1 },
+    \quad {^+}\Delta \boldsymbol{x} = \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 }
     $$
 \begin{equation}\label{eq:apic_mom_g2p}
-+ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } ).
-    \end{equation}
+  \boldsymbol { B } _ { p } ^ { n + 1 } = \frac { 1 } { 2 } \sum _ { i } w _ { i p } ^ { n } (\boldsymbol { v }_ { i } ^ { n + 1 } ({^-}\Delta \boldsymbol{x}) ^ { T }) {+}  {^+}\Delta \boldsymbol{x} ( { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T }).
+\end{equation}
+
 \end{enumerate}
@@
 \end{equation}
-A simple proof (Appendices: \ref{app:dp_proof}) cancelling out the numerous polynomials can be done for instance using \textbf{SymPy} (\cite{Sympy}).
-\cite{MPM:APIC}\cite{MPM:OLD_APIC}
+A simple proof (Appendices: \ref{app:dp_proof}) cancelling out the numerous polynomials can be done for instance using SymPy (\cite{Sympy}).\cite{MPM:APIC}\cite{MPM:OLD_APIC}
 \subsubsection{CFL condition}
@@
 \subsection{GPU memory}
-\subsubsection{Optimization opportunities}
+\section{Implementation}
+\subsection{Framework}
+decorator pattern
+\subsection{Optimization opportunities}
 \cite{NVIDIA:BEST:PRACTICE}
@@
 \subsubsection{SoA vs. AoS}
-\section{Implementation}
-\subsection{Framework}
-decorator pattern
 \subsection{Parallel Reduction}

### 2018-11-29 mscthesis.tex
realted work begin

@@
 
-\DeclareMathOperator*{\argmin}{argmin}
+\DeclareMathOperator*{\argmin}{arg min}
 \makeatletter
@@
 \section{Related Work}
-\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}\\
-\cite{stantchev2008fast}\\
-benefits over radix sort /thrust sort-by-key -> counting sort
-\cite{NVIDIA:SHUFFLE}\\
-\cite{MPM:POLYPIC}\\
-\cite{MPM:GPU}\\
+\textbf{Material Point Method:}
+\cite{MPM:POLYPIC}
+\\
+\\
+There exist several implementations utilizing the GPU for Particle-in-Cell type techniques, which the Material point method is a part of:
+
+\textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods.
+After a particle position update becomes partially sorted.
+There exist methods like 'Message Passing Sort' and 'In Place Particle-QuickSort' to try to utilize the partial sorting and save time.
+\cite{PIC:GPU} also uses fixed bin-sizes which can be, depending on the extend of use, quite memory intensive.
+Fixed size binning certainly make the methods mentioned before easier to implement, which don't necessarily map well to the GPU due to its high dependence on coalescing.
+As a result they come to the conclusion to fall back to the CUDA library \texttt{thrust}.
+It provides the radix sort method \texttt{thrust::sort\_by\_key()}.
+Full or index sort can be done.
+A full sort reorders every variable of a particle.
+An index sort gives back an index with the offset in the particle list.
+\cite{NVIDIA:NNSEARCH} compare radix to counting sort utilizing a uniform grid and uses the latter for SPH-fluids. This master thesis implements counting sort.
+
+\textbf{Particle-Grid transfers:} The particle to grid transfer are generally done in two manners \cite{stantchev2008fast}:
+\begin{enumerate}
+  \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it. Since this results in a race condition \texttt{atomicAdd(float f)}'s are in use.
+  \item \textbf{Particle Pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
+\end{enumerate}
+The grid to particle transfer in generally are easier since a particle already knows, due to its position, the grid points in its local neighborhood.
+The grid points in a uniform grid are 'sorted' by nature.
+
+Generally these methods are augmented by splitting the grid in 2D/3D tiles making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}.
+Shared memory on GPUs typically ranges from 48-64KB.
+The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a tile.
+Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited tile-sizes are available \cite{MPM:SNOW}.
+An interesting prospect for filtering techniques is given by NVIDIA's new shuffle operations, which allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
+They are however not supported on other architecture yet.
 %look again for MPM/PIC gpu impl
@@
 Occasionally notation may be omitted where it is apparent from context.
+
 Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have subscript $p$ (e.g.
 $\boldsymbol{x}_p$).
-Grid cells will be assigned a subscript $i$ denoting the grid index (e.g.
+Grid cells will be assigned a subscript $i$ or additionally $j$ denoting the grid index (e.g.
 $\boldsymbol{x}_i$).
@@
 \section{Basics}
-\subsection{Reference and current configuration}
+\subsection{Reference and current configuration}\label{sec:ref_corr_config}
 Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
@@
 A particular body is composed of a set of particles and can adapt different configurations due to changes in shape.
-These changes are cause by external or internal effects (forces etc.) on it and deform the body over time.
+These changes are caused by external or internal effects (forces etc.) on it and deform the body over time.
 \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}
 
-When modelling solids changes of quantities from a initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
+When modelling solids changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
 In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
 This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
+
 Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.
@@
 The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
+
 These descriptions are different but they will yield the same measurements when related correctly.
 I.e.
-both configuration refer variables defined on them do the deformed state, but the position where we 'look up' that value is different.
+both configuration refer variables defined on them to the deformed state, but the position where the 'look up' of that value happens is different.
 In the reference configuration we look up at the initial position of the particle.
@@
 As can be seen the deformation gradient contains a rigid rotation.
-For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation.
+For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}).
 I.e.
@@
 Since singular values are positive, it is straightforward to see that the properties for $\boldsymbol{R}$ and $\boldsymbol{S}$ hold.
-The components of the singular value decomposition are important to gain an intuition: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors $\boldsymbol{u}_i,\boldsymbol{v}_i$, respectively \cite{MIT:SVD}.
+The components of the singular value decomposition are important to gain an intuition for its usefulness: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors $\boldsymbol{u}_i,\boldsymbol{v}_i$, respectively \cite{MIT:SVD}.
 For illustrating purposes imagine the manipulation of $\boldsymbol{v}_1$ due to $\boldsymbol{U\Sigma V}^T$:
@@
 For a positive definite matrix the singular value decomposition becomes even easier as $\boldsymbol{U}=\boldsymbol{V}$. Item \ref{it:tr_sing_vect} then effectively just becomes a transform 'back'.
-Following is a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3}. The proposed singular value decomposition is also called 'Polar SVD' and follows a specific convention.
-\begin{itemize}\label{ref:itemize_conv}
-  \item $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e. both $\text{det}(U), \text{det}(V) = 1$.
-  \item As a result if $\text{det}(\boldsymbol{F}) = -1$ the negative sign needs to move on to $\boldsymbol{\Sigma}$: The lowest singular value in magnitude will get a negative sign attached.
-\end{itemize}
+
+In the following a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3} is given.
+The proposed singular value decomposition is also called the 'Polar SVD' and follows a specific convention.
+\begin{enumerate}\label{ref:itemize_conv}
+  \item\label{item:refl_free} $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e. both $\text{det}(U), \text{det}(V) = 1$ hold.
+  \item If $\text{det}(\boldsymbol{F}) = -1$ the negative sign needs to move on to $\boldsymbol{\Sigma}$ as a result of item \ref{item:refl_free}.
+The lowest singular value in magnitude will get a negative sign attached.
+  \end{enumerate}
 This convention does not change the existence or uniqueness of the singular value or polar decomposition, although strictly speaking $\boldsymbol{S}$ in the Polar decomposition is not positive definite anymore.
+\\
+The algorithm proceeds as follows:
 \begin{enumerate}
@@
 The first aim will be to find the strain energy density $\Psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
-For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\Psi$ as:
-The most general linear stress-strain relationship is given by Hooke's Law in three dimensions
+The most general linear stress-strain relationship is given by Hooke's Law in three dimensions:
 \begin{equation}\label{eq:hook}
@@
 \end{equation}
-where $\boldsymbol{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order tensor with (3x3)x(3x3) = 81 elements.
+where $\boldsymbol{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order stiffness-tensor with (3x3)x(3x3) = 81 elements.
 Assuming following symmetries reduces the tensors to 6 and 21 unique elements respectively:
@@
       {\gamma_{12}}\\
-  \end{array} \right]
+  \end{array} \right].
 \end{equation}
@@
 \end{equation}
-The conservation of momentum (eq. \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} into \ref{eq:gen_ang_mom} leaving only:
+The conservation of momentum (eq. \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:
 \begin{equation}
@@
 Before deriving the weak form an explanation of what it achieves is in order.
+
 The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain (and is as such influenced by the whole domain in general).
@@
 The complete mathematical description of the weak formulation is beyond this thesis.
-Numeric books that discuess finite element methods will provide one (\cite{bathe2006finite}, \cite{dahmen2008numerik}).
-For simplicity: The weak formulation combined with Galerkin discretization restricts the globalility of the strong method due to so called 'test functions': $_0\boldsymbol{q}_h$ .
-A 'test function' generally only has limited support ($ _0\boldsymbol{q}_h\neq 0$ on a very small subset of the whole domain $\Omega^0$).
+Numeric books that discuss finite element methods will provide one (\cite{bathe2006finite}, \cite{dahmen2008numerik}).
+For simplicity: The weak formulation combined with Galerkin discretization restricts the globality of the strong method due to so called 'test functions': $_0\boldsymbol{q}_h$ .
+A 'test function' generally only has limited support, i.e.
+$ _0\boldsymbol{q}_h\neq 0$ on a very small subset of the whole domain $\Omega^0$.
 This is mostly used to gather information on a local neighborhood.
@@
 With the help of partial integration in multiple dimension the derivative moves over to the test-function.
-Reducing the order of derivatives by moving a derivative to the test function is the main motive of the weak form:
+Balancing out the order of derivatives, by moving a derivative to the test function, is the main motive of the weak form:
 $$
@@
 \end{equation}
-In the material point method stress computations are more naturally done in the current configuration.
+In the material point method stress computations are more naturally done in the current configuration, or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
 Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
@@
 Plugging $\boldsymbol{x}_i^{n+1}$ of eq.
-\ref{eq:vel_midpoint} into \ref{eq:force_midpoint}:
+\ref{eq:vel_midpoint} into \ref{eq:force_midpoint} leads to the discretized momentum equation:
 \begin{equation}
@@
 \end{equation}
-And substituting
 One can recast this method back to an energy function by integrating:
@@
 \end{equation}
-Since the objective from the start was minimizing $\boldsymbol{E}$ to find $\boldsymbol{v}_i$ the scale of this function can be chosen arbitrarily, i.e.
-a zero crossing does not scale.
-The scaling here can be identified as:
+Since the objective is minimizing the scale of $\boldsymbol{E}$ can be chosen arbitrarily, i.e. a zero crossing does not scale. The scaling here can be identified as:
 \begin{equation}
@@
 \end{equation}
-The linear system can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}) since $\frac{\partial g}{\partial v}$ is symmetric, positive definite.
-This is due to being the Hessian of the convex function $E$.
+The linear system of eq.
+\ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}) since $\frac{\partial g}{\partial v}$ is symmetric, positive definite.
+This is due it being the Hessian of a convex function $E$.
 There is to important things notice:

### 2018-11-30 mscthesis.tex
related work more

@@
 \section{Introduction} \label{intro}
-bib-test \cite{MPM:SNOW}
-$$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
 \section{Related Work}
-\textbf{Material Point Method:}
-\cite{MPM:POLYPIC}
-\\
-\\
-There exist several implementations utilizing the GPU for Particle-in-Cell type techniques, which the Material point method is a part of:
+\textbf{Particle-in-Cell (PIC):} PIC techniques were developed initially for hydrodynamics
+\cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations. Their key steps are identical to the MPM-method:
+\begin{enumerate}
+  \item Transfer particle data to a grid.
+  \item Solve the governing partial differential equations on the grid using numerical methods.
+  \item Transfer back grid data to particles.
+\end{enumerate}
+Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle (FLIP) was developed for fluid simulations and is still widely in use for large simulations.
+Their counterpart being the empirical Smoothed Particle Hydrodynamics (SPH) \cite{gingold1977smoothed}.
+
+For that matter APIC was developed as an improvement to PIC and FLIP solvers.
+Due to the definition of a local affine velocity field around each particle (instead of a single vector), APIC conserves angular momentum, reduces damping (of the underlying numerics) and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
+\cite{MPM:APIC} \cite{MPM:OLD_APIC}
+
+A step further goes PolyPIC giving the velocity field even more freedom by adding polynomial modes of increasing order to the description. \cite{MPM:POLYPIC}
+
+\textbf{Material Point Method:} The MPM-Method is like the PIC-Method very general in that it uses analysis and numerics to be derived without the need for empirical assumptions.
+Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids.
+\cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}.
+In rapid succession a variety materials and effects got modeled by the MPM-Method.
+Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature its fast rate of expenditure is not a surprise.
+To give a short summary of the developments in computer graphics:
+\cite{MPM:DRUCKER} %Sand Plasticity
+\cite{MPM:MULTI} %Sand+Water
+\cite{MPM:INVERT} %Neo-Hookean
+\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
+\cite{MPM:SHELLS} %Thin Shells
+\cite{MPM:PHASE_CHANGE} %Heat-Model
+
+There exist several implementations utilizing the GPU for Particle-in-Cell type techniques, which the Material Point Method is a part of:
 
@@
 \end{enumerate}
-The grid to particle transfer in generally are easier since a particle already knows, due to its position, the grid points in its local neighborhood.
+A grid to particle transfer is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
 The grid points in a uniform grid are 'sorted' by nature.
 
-Generally these methods are augmented by splitting the grid in 2D/3D tiles making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}.
+Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition.
 Shared memory on GPUs typically ranges from 48-64KB.
-The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a tile.
-Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited tile-sizes are available \cite{MPM:SNOW}.
-An interesting prospect for filtering techniques is given by NVIDIA's new shuffle operations, which allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
-They are however not supported on other architecture yet.
+The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
+Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited block sizes are available \cite{MPM:SNOW}.
+
+Alternatively one may also compute the partial sum of the transfer on the halo. Two different techniques exist to get the data between blocks coherent again:
+\begin{enumerate}
+\item Atomically add up the values of the halo.
+This results in two atomic adds on a side node, four on an edge node and eight on a corner-node; If the block extend is bigger than half the halo extend.
+\item Have every halo of every block correspond to actual global memory and do add steps along each of the three axes. This is infeasible for large supports. \cite{crassin2011interactive}
+\end{enumerate}
+Stencil computations and filtering techniques underlie a very similar process to MPM-transfers.
+Although these techniques employ a one-to-one grid relationship, since no particles are involved, it is worth taking a look at them.
+
+Widely used is a 2.5D blocking: A domain (of the domain decomposition) is split into 2D-planes along one axis.
+Data reuse of shared memory is maximized by a cyclic-queue strategy: Start a thread for each grid node in the first plane and load the halo along the axis into shared memory.
+Then each thread iterates along the axis over the 2D-planes synchronously, discarding the oldest plane in memory while loading the next plane (along the axis) into shared memory.
+Thus there are no z-dimension halos.
+\cite{brandvik2010sblock} \cite{williams2007scientific} \cite{krotkiewski2013efficient}
+
+An interesting prospect for filtering techniques on the GPU is given by NVI\-DIA's new shuffle operations.
+Shuffle allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
+They are however not supported on other architecture, yet.
+
+\textbf{Particle Activation:} If a particle is under a user-defined velocity threshold a particle becomes deactivated.
+It does not get recognized by any particle to grid transfer (nor any grid transfer not affected by the velocity transfer).
+They may of course be reactivated if the back transfer of the velocity exceeds that threshold again or they collide.
+Dividing the particles up can be realized with stream compaction algorithms: One implementation would be to compute a scan (based on the discussed condition) followed up by a modified reorder shader discussed in chapter \ref{sec:implementation}.
+The modified reorder shader would then be coalesced.
+\cite{MPM:GPU}
+
 %look again for MPM/PIC gpu impl
 %SPH, FEM methods
+\clearpage
 \section{Notation}\label{sec:notation}
@@
 \cite{MCGINTY:CONTINUUM}
+\clearpage
 \section{Basics}
@@
 \subsection{Constitutive Models}\label{seq:constitutive_models}
-\cite{MPM:DRUCKER} %Sand Plasticity
-\cite{MPM:MULTI} %Sand+Water
-\cite{MPM:INVERT} %Neo-Hookean
-\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
-\cite{MPM:SHELLS} %Thin Shells
-\cite{MPM:PHASE_CHANGE} %Heat-Model
-\cite{MPM:OPTIMI_INTEGR} %Opimization Methods, BDF-Semi-Impl-Midpoint rule
 $$\boldsymbol{F}_E \boldsymbol{F}_P$$ \cite{ochsner2014elasto}
@@
 \cite{MPM:APIC}\cite{MPM:COURSE}
-\subsubsection{APIC transfers}
+\subsubsection{APIC transfers}\label{sec:apic}
 The momentum transfer round trip could be defined as:
@@
 \end{enumerate}
+\clearpage
 \subsection{GPU memory}
-\section{Implementation}
+\section{Implementation}\label{sec:implementation}
 \subsection{Framework}

### 2018-12-02 mscthesis.tex
related work end, fixed cor. hyperel.,hessian

@@
   \item Transfer particle data to a grid.
-  \item Solve the governing partial differential equations on the grid using numerical methods.
-  \item Transfer back grid data to particles.
+\item \label{it:solve_gov_eq}Solve the governing partial differential equations on the grid to move forward in time.
+  \item Transfer grid data back to particles.
 \end{enumerate}
+In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}.
 Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle (FLIP) was developed for fluid simulations and is still widely in use for large simulations.
@@
 For that matter APIC was developed as an improvement to PIC and FLIP solvers.
-Due to the definition of a local affine velocity field around each particle (instead of a single vector), APIC conserves angular momentum, reduces damping (of the underlying numerics) and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
+Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
 \cite{MPM:APIC} \cite{MPM:OLD_APIC}
@@
 
-\textbf{Material Point Method:} The MPM-Method is like the PIC-Method very general in that it uses analysis and numerics to be derived without the need for empirical assumptions.
+\textbf{Material Point Method:} The MPM-Method is like the PIC-Method mathematical and physical in that it uses analysis and numerics to be derived without the need for empirical assumptions.
 Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids.
@@
 Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature its fast rate of expenditure is not a surprise.
-To give a short summary of the developments in computer graphics:
-\cite{MPM:DRUCKER} %Sand Plasticity
-\cite{MPM:MULTI} %Sand+Water
-\cite{MPM:INVERT} %Neo-Hookean
-\cite{MPM:OLROYDB} %Viscoelastic Foam,Sponge
-\cite{MPM:SHELLS} %Thin Shells
-\cite{MPM:PHASE_CHANGE} %Heat-Model
+
+The simulation of(hyper-)elastic material can be done with a number of different models utilizing fixed Linear strain \ref{sec:cor_hyper}, left Cauchy-Green strain \ref{sec:svd} or Hencky-strain(logarithmic) \cite{MPM:SHELLS} for the different requirements of small to large strains.
+Inserting these into the different energy density $\boldsymbol{\Psi}(\boldsymbol{\epsilon})$ models can already result in a constitutive model.
+Plasticity in general requires extra modeling in terms of a flow rule, yield-condition and a hardening rule which are not discussed here \cite{ochsner2014elasto}.
+To give a short impression of the developments in computer graphics:
+
+\cite{MPM:PHASE_CHANGE} model phase transition (melting) to a liquid with the fixed corotational hyperelasticity (\ref{sec:cor_hyper}) and discretizing the heat equation.
+\cite{MPM:OLROYDB} use an Olroyd-B model with Cauchy-Green strain to handle elasticity and plasticity for viscoelastic fluids, foams and sponges.
+\cite{MPM:DRUCKER} uses Hencky-strain in combination with a St.
+Venant-Kirchhoff model and Drucker-Prager elastoplascity to model sand.
+\cite{MPM:MULTI} combine particles of different species (sand and water) by one grid for each species and relating them with a heuristic momentum exchange.
+\cite{MPM:SHELLS} simulates thin shells (cloths,cups etc.) with a Kirchhoff-Love model using Hencky-strains.
+\\
 
@@
 \textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods.
-After a particle position update becomes partially sorted.
-There exist methods like 'Message Passing Sort' and 'In Place Particle-QuickSort' to try to utilize the partial sorting and save time.
-\cite{PIC:GPU} also uses fixed bin-sizes which can be, depending on the extend of use, quite memory intensive.
-Fixed size binning certainly make the methods mentioned before easier to implement, which don't necessarily map well to the GPU due to its high dependence on coalescing.
-As a result they come to the conclusion to fall back to the CUDA library \texttt{thrust}.
-It provides the radix sort method \texttt{thrust::sort\_by\_key()}.
-Full or index sort can be done.
-A full sort reorders every variable of a particle.
-An index sort gives back an index with the offset in the particle list.
-\cite{NVIDIA:NNSEARCH} compare radix to counting sort utilizing a uniform grid and uses the latter for SPH-fluids. This master thesis implements counting sort.
-
-\textbf{Particle-Grid transfers:} The particle to grid transfer are generally done in two manners \cite{stantchev2008fast}:
+Methods like 'Message Passing Sort' \cite{kong2011particle} \cite{decyk2011adaptable}, 'In Place Particle-QuickSort' \cite{stantchev2008fast} and 'Linked List Ordering' \cite{burau2010picongpu} try to utilize the partial sorting and only update particles that need to be moved within the list.
+These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic.
+Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL-condition, chapter \ref{sec:cfl}.
+These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management.
+Fixed size binning make the methods mentioned before easier to implement, which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
+As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes.
+\texttt{thrust} provides the radix sort method \texttt{thrust::sort\_by\_key()}, where one sorts by grid index as key.
+
+In general full or index sorts can be done. A full sort reorders every variable of a particle. An index sort gives back an index with an offset to access the particle variable.
+
+\cite{NVIDIA:NNSEARCH} compare radix to counting sort utilizing a uniform grid and use the latter for SPH-fluids.
+
+\textbf{Particle-Grid transfers:} The particle to grid transfer is generally done in two manners \cite{stantchev2008fast}:
 \begin{enumerate}
@@
 They may of course be reactivated if the back transfer of the velocity exceeds that threshold again or they collide.
+\cite{MPM:GPU}
+
 Dividing the particles up can be realized with stream compaction algorithms: One implementation would be to compute a scan (based on the discussed condition) followed up by a modified reorder shader discussed in chapter \ref{sec:implementation}.
 The modified reorder shader would then be coalesced.
-\cite{MPM:GPU}
-
+
+\textbf{Particle Resampling:} A method that could alleviate pressure off fixed bin sizes and could make the partial sorting methods as well as the shared-memory restrictions less of a factor is particle resampling.
+Their intention mostly is to fill numerical gaps in the material at large deformations.
+Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.
+
+A simple \textit{Split} method divides one particle into eight with a distance of half the diagonal length of a grid cell from the original particle.
+The eight particles span a cube that can be randomly rotated if desired.
+Mass and volume get equally distributed; Velocity and deformation gradient get copied.
+Refer to chapter \ref{sec:mat_point} for conservation of mass and momentum in MPM.
+\cite{gao2017adaptive}
+
+A \textit{Merge} method looks for the closest neighbors and creates the new merged particle at the geometric center.
+Mass and volume are accumulated.
+Velocity is computed from a mass-weighted average of the participating particles.
+The deformation gradient is accordingly to chapter \ref{sec:svd} decomposed into $\boldsymbol{F} = \boldsymbol{U\SigmaV}^T$.
+The quaternion average of $\boldsymbol{U}$ and $\boldsymbol{V}$ leads to an average rotation $\overline{\boldsymbol{U}}$ and $\overline{\boldsymbol{V}}$, respectively.
+The principal strains $\boldsymbol{\Sigma}$ are just averaged component-wise $\overline{\boldsymbol{\Sigma}}$.
+One can compose them in the same fashion $\overline{\boldsymbol{F}} = \overline{\boldsymbol{U}}\overline{\boldsymbol{\Sigma}}\overline{\boldsymbol{V}}^T$.\cite{gao2017adaptive}
+
+Unconditionally keeping particles on a fixed size is not desirable due to fracture requiring that gaps will be created.
+\cite{gao2017adaptive} produce a signed distance field to identify interior points.
+Since fracture typically starts at a surface point desired behavior is accomplished again.
+Furthermore the signed distance field can be used to adapt the grid size depending on the distance to the surface.
+This allows coarser levels with increasing distance from the surface.
 %look again for MPM/PIC gpu impl
@@
 Occasionally summation over grid index $i$ and particles $p$ may be implied.
-A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n} t^i$ will have a right superscript: $\boldsymbol{x}^n$.
+A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
 \subsection{Einstein-Notation}
@@
 \cite{MPM:COURSE}
-\subsection{Polar- and Singular Value Decomposition}
+\subsection{Polar- and Singular Value Decomposition}\label{sec:svd}
 The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$.
@@
 \begin{enumerate}\label{ref:itemize_conv}
-  \item\label{item:refl_free} $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e. both $\text{det}(U), \text{det}(V) = 1$ hold.
+  \item\label{item:refl_free} $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e.
+both $\text{det}(\boldsymbol{U}), \text{det}(\boldsymbol{V}) = 1$ hold.
   \item If $\text{det}(\boldsymbol{F}) = -1$ the negative sign needs to move on to $\boldsymbol{\Sigma}$ as a result of item \ref{item:refl_free}.
@@
 In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
+In general $\boldsymbol{R}$ is an upper triangular matrix.
+In item \ref{item:qr} it reduces to being diagonal.
 \cite{MPM:COURSE}
@@
 \subsection{Constitutive Models}\label{seq:constitutive_models}
-$$\boldsymbol{F}_E \boldsymbol{F}_P$$ \cite{ochsner2014elasto}
+It is common practice in elastoplascity theory to decompose the deformation gradient in its elastic and plastic parts $\boldsymbol{F} = \boldsymbol{F}_E \boldsymbol{F}_P$.
+The elastic part is recoverable.
+The plastic part is irreversibly lost.
+Plastic models are not covered here.
+They extend the energy densities discussed here with additional terms $\Psi(\boldsymbol{F}_E,\boldsymbol{F}_P)$.
+For more information refer to the literature \cite{ochsner2014elasto}.
 \subsubsection{Linear Elasticity}\label{sec:linear_elasticty}
@@
 \subsubsection{Fixed Corotated Hyperelasticity}
-Numerical Stepping algorithms\\
--> Newton Raphson scheme\\
--> Problem: Bad Primary contour\\
-Benefit: Easier description in terms of Invariants\\
+Numerical implicit stepping algorithms mostly rely on using the first and second derivative of $\Psi$, chapter \ref{sec:newton}.
+Furthermore one may describe an isotropic elastic model due to their principal stretches $\sigma_1,\sigma_2,\sigma_3$, chapter \ref{sec:svd}.
+The rest configuration is reached, when $\sigma_1,\sigma_2,\sigma_3 = 1$.
+I.e.
+no elastic forces will be exerted, because the material is completely relaxed (again).
+
+Under material relaxation these stepping algorithms are attracted to a material-dependent primary contour.
+For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume-preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$).
+Reaching the rest configuration is only the secondary goal.
+
+The primary contour can be formalized as $\boldsymbol{v}\cdot \boldsymbol{g} = 0$ of the gradient $g_i = \frac{\partial \Psi}{\partial \sigma_i}$ and the eigenvector with the largest-magnitude eigenvalue $\boldsymbol{v}$ of the Hessian $H_{ij} = \frac{\partial \Psi}{\partial \sigma_i \sigma_j}$.
+Problems arise in $\Psi_{CH}$ as the primary contour can easily cross into the inverted region ($\sigma_3 < 0$) leading to inverted configurations while relaxing.
+This is undesired behavior.
+Furthermore the primary contour can lead into energy kinks at extreme stretches leading to oscillations.
+
+Therefore they propose to use the Fixed Corotated Hyperelasticity energy density:
+\begin{equation}
+  \Psi_{FCH} = \mu \|\boldsymbol{\Sigma}_E - \boldsymbol{I}\|^2_F + \frac{\lambda}{2}(J_E-1)^2.
+\end{equation}
+This leads to the primary contour $J=1$, which does not cross the inverted region as only one singular value may be negative.
 \cite{MPM:INVERT}
@@
 Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$ and its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
-After \ref{eq:pull_back_vol} they are related as:
+After \ref{eq:volume_integral} they are related as:
 \begin{equation} \label{eq:density_pull_back}
@@
     \end{equation}
+  \item Non-negativity: $w \geq 0$. Negative weights can cause severe instability or non-physical behavior. \cite{gao2017adaptive}
+  \item Limited local support to reduce the number of discretizations, chapter \ref{sec:discretization}.
   \item $C^1$-continuity s.t. $\nabla w$ is continuous.
@@
 As a reminder of chapter \ref{sec:notation}: $\boldsymbol{x}_p$ refers to a particle's position. $\boldsymbol{x}_i$ to a grid cell's position.
-Often dyadic products of one-dimensional interpolation functions suffice
+For a more general discussion refer to \cite{gao2017adaptive}. Often dyadic products of one-dimensional interpolation functions suffice
 \begin{equation}
@@
 Since these function are used to weight (or filter) particles a shortening in notation may be employed as $ w_{ip}^n = w(\boldsymbol{x}_p^n-\boldsymbol{x}_i^n)$.
-\cite{MPM:APIC}\cite{MPM:COURSE}
+\cite{MPM:APIC}\cite{MPM:COURSE}\cite{steffen2008analysis}\cite{gao2017adaptive}
 \subsubsection{Mass transfer}
@@
 A simple proof (Appendices: \ref{app:dp_proof}) cancelling out the numerous polynomials can be done for instance using SymPy (\cite{Sympy}).\cite{MPM:APIC}\cite{MPM:OLD_APIC}
-\subsubsection{CFL condition}
+\subsubsection{CFL condition} \label{sec:cfl}
 The CFL condition is prominent for FEM-like methods: For a stable integration a particle should not travel father than the grid spacing $h$ in a discrete time-step $\Delta t$.
@@
 \cite{MPM:APIC}
-\subsection{Discretization}
+\subsection{Discretization}\label{sec:discretization}
 The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following for MPM preferable description:
@@
 The evolution of the deformation gradient \ref{eq:evol_def_grad_disc_x} for a general $\boldsymbol{\hat{x}}_i$  is necessary
-$$\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
+\begin{equation}\label{eq:partial_def_grad}
+\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
 = \delta_{\omega \alpha}
-\frac{\partial w_{ip}}{\partial x_\gamma}F_{\gamma\beta} = \delta_{\omega \alpha}F_{\gamma\beta} \frac{\partial{w_{ip}}}{\partial x_\gamma}$$
+\frac{\partial w_{ip}}{\partial x_\gamma}F_{\gamma\beta} = \delta_{\omega \alpha}F_{\gamma\beta} \frac{\partial{w_{ip}}}{\partial x_\gamma}
+\end{equation}
+
 for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of eq. \ref{eq:force_disc}:
@@
 \cite{MPM:COURSE}\cite{MPM:APIC}
-\subsection{Newton's Method}
+\subsection{Newton's Method}\label{sec:newton}
 \newcommand{\var}[1]{{\ttfamily#1}}% variable
@@
 \begin{enumerate}
-  \item Conservation of momentum on incomplete convergence is only fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $M^{-1}$.
-More in \cite{MPM:APIC}.
+  \item  One may want to use a preconditioner $\boldsymbol{P}^{-1}$ to reduce iteration times.
+However conservation of momentum on incomplete convergence is fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $\boldsymbol{M}^{-1}$ \cite{MPM:APIC}.
 Thus the equation becomes:
@@
 \end{enumerate}
+The Hessian is derived following \ref{eq:partial_def_grad} and \ref{eq:nodal_force}:
+\begin{equation}
+  \frac{\partial e}{\partial {x}_{i\alpha} \partial x_{j\tau}}
+= \sum_p V^0_p\frac { \partial ^ { 2 } \Psi } { \partial  { F}_{\alpha \beta} \partial { F } _{\tau \sigma}}  (F^n_p)_{\omega \sigma}\frac{\partial w_{jp}}{\partial x_\omega} (F^n_p)_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\frac{\partial {f}_{i\alpha}}{\partial x_{j\tau}}.
+\end{equation}
 The Hessian increment can be computed with a two-stage process:
@@
 \end{enumerate}
+%Optimally one would also try to derive a model-dependent preconditioner $\boldsymbol{P}$ for $\frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j}$ to reduce iterations of the conjugate gradient method:
+%\begin{equation}
+%\boldsymbol{P}^{-1}\boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} = \boldsymbol{P}^{-1} \boldsymbol{M}^{-1}\boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right).
+%\end{equation}
+%One of those is the Jacobi (diagonal) preconditioner: $\boldsymbol{P} = \text{diag}\left( \boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \right)$.
 \clearpage
-\subsection{GPU memory}
 \section{Implementation}\label{sec:implementation}
-\subsection{Framework}
-decorator pattern
 \subsection{Optimization opportunities}
@@
 \end{itemize}
-\subsubsection{SoA vs. AoS}
+\subsection{Framework}
+decorator pattern soa vs aos\\
+magic\_get reflection\\
+single + double\\
+GL\_EXT\_TIMER + NVIDIA Nsight\\
 \subsection{Parallel Reduction}
 \cite{NVIDIA:PARALLEL_REDUCTION}
-\subsection{Binning}
-\cite{NVIDIA:BINNING}
-\subsection{Scanning}
-\cite{NVIDIA:SCAN}
-\cite{NVIDIA:SCAN_MODERN}
-\subsection{Affine particle-in-cell}
-\cite{MPM:APIC}
-\subsection{Iterative solvers for linear systems}
-%GMRES, Conjugate Gradients
+\subsection{Counting sort}
+Map (Reset) \\
+\subsubsection{Bin}
+Binning\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
+\subsubsection{Scan}
+Scan\cite{NVIDIA:SCAN}\\
+Scan(Raking) \cite{NVIDIA:SCAN_MODERN}
+\subsubsection{Reordering}
+Coalesced, Uncoalesced
+\subsection{MPM-Transfers}
+Conjugate gradient method
+\subsubsection{Particle to grid transfers}
+\subsubsection{Grid to particle transfers}
 \subsection{Verifying Simulation}
@@
   \item Singular Value Decomposition
-  \item Parallel Reduction
+  \item Conjugate gradient
 \end{itemize}

### 2018-12-03 mscthesis.tex
implementation

@@
 \usepackage{verbatim}
+\usepackage{blindtext}
 \addbibresource{mscthesis.bib}
 
+\captionsetup{justification=raggedright,singlelinecheck=false}
 \DeclareMathOperator*{\argmin}{arg min}
@@
 
-The simulation of(hyper-)elastic material can be done with a number of different models utilizing fixed Linear strain \ref{sec:cor_hyper}, left Cauchy-Green strain \ref{sec:svd} or Hencky-strain(logarithmic) \cite{MPM:SHELLS} for the different requirements of small to large strains.
+The simulation of (hyper-)elastic material can be done with a number of different models utilizing fixed Linear strain \ref{sec:cor_hyper}, left Cauchy-Green strain \ref{sec:svd} or Hencky-strain(logarithmic) \cite{MPM:SHELLS} for the different requirements of small to large strains.
 Inserting these into the different energy density $\boldsymbol{\Psi}(\boldsymbol{\epsilon})$ models can already result in a constitutive model.
@@
 \begin{equation} \label{EINSTEIN:ADD}
-  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }
+  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }.
 \end{equation}
@@
 \begin{equation}
-  D^T _ { i j }  =D_{ji}
+  D^T _ { i j }  =D_{ji}.
 \end{equation}
@@
 \begin{equation}
-  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
+  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }.
 \end{equation}
@@
 \begin{equation} \label{EINSTEIN:FROBENIUS1}
-  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }
+  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }.
 \end{equation}
@@
 \begin{equation}
-  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
+  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }.
 \end{equation}
@@
 \begin{equation}
-  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }
+  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }.
 \end{equation}
@@
 \begin{equation}
-  a _ { i } \delta _ { i j } = a _ { j }
+  a _ { i } \delta _ { i j } = a _ { j }.
 \end{equation}
@@
 \begin{equation}
-  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}
+  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}.
 \end{equation}
@@
 \end{algorithm}
-Many minimization and finding zero crossing algorithms are available. One of such is Newton's Method, which allows for rapid quadratic convergence in a near local neighborhood.
+Many minimization and root finding algorithms are available. One of such is Newton's Method, which allows for rapid quadratic convergence in a near local neighborhood.
 \begin{equation}
@@
 \begin{equation}\label{eq:linear_system}
-  \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} =  \boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
+  \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} =  \boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right).
 \end{equation}
@@
 \end{equation}
-The linear system of eq.
-\ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}) since $\frac{\partial g}{\partial v}$ is symmetric, positive definite.
-This is due it being the Hessian of a convex function $E$.
-There is to important things notice:
+$\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}$ is symmetric, positive definite due it being the Hessian of a convex energy function $E$ with a hyper-elastic material.
+The linear system of eq. \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is to important things notice:
 \begin{enumerate}
@@
 \subsection{Optimization opportunities}
-\cite{NVIDIA:BEST:PRACTICE}
-\cite{AMD:GPU_OPEN}
-\begin{itemize}
-  \item Minimize CPU-GPU transfers
-  \item Coalesced memory access
-  \item Bank conflicts
-\end{itemize}
-\subsection{Framework}
+\begin{center}
+  \begin{tabular}{ | l | l | l | l | l |}
+    \hline
+    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
+    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
+    Local        &Off                    &L2       &1 thread             &thread     \\\hline
+    Shared       &On                     &n/a      &block                &block      \\\hline
+    Global       &Off                    &L2       &all threads+host     &host alloc \\
+    \hline
+  \end{tabular}
+\end{center}
+\begin{flushright}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
+The speed of these memory types is given in the following starting with the fastest:
+Register > Shared Memory > Local, Global.
+Therefore one should prefer register memory for intra-thread caching.
+Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block.
+Global memory should be mostly used at the start and end of a shader.
+
+\textbf{Register spilling:} It follows that one should avoid local memory.
+Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
+One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
+If the shader uses more registers the register spills.
+
+To our knowledge there is no way to find out the exact number of registers an OpenGL shader uses.
+The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
+It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version).
+One can identify TEMP variables as registers.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees this is not  further optimized.
+All other metrics discussed are readily availabe for OpenGL.
+
+\textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers.
+The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s.
+While the communication between device and GPU is 224GB/s (GTX 970).
+This includes doing calculations on the GPU that may run faster on the CPU.
+The API may also stall if the CPU is not issuing instructions to the GPU.
+This may be due to calculations on the CPU, or even worse calculations depending on GPU results.
+In this thesis the physics already runs on the GPU.
+Additionally we try to avoid transfers wherever possible.
+
+\textbf{Memory coalescing:} Between a block of threads and a single thread another important entity exists.
+NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront.
+Memory operations are issued per warp/wavefront.
+A memory access uses the L2 caching behavior.
+A L2-cache-line is 32 Bytes long.
+Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.
+
+If all threads of a warp access memory within this cache-line only one memory request of 32 bytes needs to be made.
+In contrast if a warp accesses N different cache-lines, N*32 bytes need to be loaded (assuming a cache miss).
+It does not matter in which order the cache-line is accessed by the warp.
+I.e.
+random or reversed access on a single cache-line are also coalesced.
+On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses.
+L2 is a write-back cache thus this assessment also qualifies for global memory writes.
+\cite{NVIDIA:MEMORY_BAND}
+
+\noindent\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}[caption={AoS Layout},language={GLSL},style={GL}]
+struct Particle {
+  vec4 position;
+  vec4 velocity;
+} Particles[n];
+\end{lstlisting}
+\end{minipage}
+\hfill\noindent\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}[caption={SoA Layout},label={llst:soa},language={GLSL},style={GL}]
+struct Container{
+  vec4 positions[n];
+  vec4 velocities[n];
+} Particles;
+\end{lstlisting}
+\end{minipage}
+
+Coalescing leads to the motivation of using structure of array layouts over arrays of struct layouts.
+The latter is the more commonly taught approach, while high performance CPU code may also prefer SoA or the more advanced entity component systems.
+This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.
+
+In short one might not access every variable of a particle in a shader.
+I.e.
+in physics one has a pipeline of different passes each requiring another subset of variables.
+Assume for instance one shader only requires the position then it is more favorable for consecutive threads to load consecutive position vectors.
+The cache-line is then filled with position vectors, which all will get used.
+In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the bandwidth.
+\noindent\begin{wrapfigure}{R}{0.5\textwidth}
+\hfill\begin{minipage}{0.45\textwidth}
+\begin{lstlisting}[caption={OpenGL Layout},label={llst:opengl},language={GLSL},style={GL}]
+//if SoA
+#define Particle PREC_VEC_TYPE
+#define Particle_position 0
+#define Particle_velocity 1
+#define Particle_size 2
+//if AoS
+struct Particle{
+  PREC_VEC_TYPE position;
+  PREC_VEC_TYPE velocity;
+};
+buffer Container{
+  Particle variables [ ];
+};
+\end{lstlisting}
+\end{minipage}
+\end{wrapfigure}
+
+Due to this we motivate the C++-library \textit{magic\_get} \cite{APOLUKHIN:MAGIC_GET} for basic reflection.
+Our code manipulates AoS layouts into SoA layouts and back using reflection.
+The MSVC-Compiler has a hard coded stack limit for these kind of operations.
+A struct may be not bigger than 256 Bytes before throwing a compiler error.
+With the optional use of double-precision this limit is quite restrictive towards that compiler.
+
+Furthermore we experienced that for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}.
+As an alternative we had to provide meta information in the style of reflection macros and collapse the container, Listing \ref{llst:opengl}.
+We can then access SoA by memory offset and AoS by variable or by memory offset for a more abstract view of the memory.
+We also share the AoS setup between C++ and OpenGL code.
+One could also build a shader using the abstract layout to convert from SoA to AoS.
+To name a few more cases that macros handle for us:
+\begin{enumerate}
+  \item Multiple buffering, e.g. double buffer for unsorted and sorted particles.
+  \item Single and double precision (PREC\_VEC\_TYPE in listing \ref{llst:opengl}).
+  \item A decorator pattern allows to add buffer information to existing buffer information: For instance index sort needs an extra buffer containing the indices.
+Full sort requires a double buffer.
+Thus a sorted buffer gets an attachment(decoration) to contain these information as well as signaling which sort was used in the first place for further operations down the pipeline.
+  \item Generalize methods to use variants of the same shader with different operations:
+
+    \texttt{\#define unary\_op(x) length(x)}
+
+    \texttt{\#define binary\_op(x,y) x+y}
+\end{enumerate}
+For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}.
+Except for the ommission of the glsl version or extension macros, we did not find any problems in this process.
+
+\textbf{Shared Memory Bank conflicts:}
+
+\textbf{Multiple elements per thread and block size:}
+
+\cite{NVIDIA:BEST:PRACTICE} \cite{AMD:GPU_OPEN}
 decorator pattern soa vs aos\\

### 2018-12-04 mscthesis.tex
metric,optim

@@
 Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition.
-Shared memory on GPUs typically ranges from 48-64KB.
-The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
+Shared memory is however limited per GPU (GTX 970, 48KB per block).The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
 Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited block sizes are available \cite{MPM:SNOW}.
@@
 \cite{MPM:COURSE}\cite{MPM:APIC}
-\subsection{Newton's Method}\label{sec:newton}
+\subsubsection{Newton's Method}\label{sec:newton}
 \newcommand{\var}[1]{{\ttfamily#1}}% variable
@@
 %One of those is the Jacobi (diagonal) preconditioner: $\boldsymbol{P} = \text{diag}\left( \boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \right)$.
-\clearpage
-\section{Implementation}\label{sec:implementation}
-\subsection{Optimization opportunities}
+\subsection{General-purpose computing on graphics processing units}
+\subsubsection{Optimization Opportunies}
 \begin{center}
@@
 Register > Shared Memory > Local, Global.
+
 Therefore one should prefer register memory for intra-thread caching.
@@
 Global memory should be mostly used at the start and end of a shader.
-
-\textbf{Register spilling:} It follows that one should avoid local memory.
-Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
-One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
-If the shader uses more registers the register spills.
-
-To our knowledge there is no way to find out the exact number of registers an OpenGL shader uses.
-The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
-It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version).
-One can identify TEMP variables as registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees this is not  further optimized.
-All other metrics discussed are readily availabe for OpenGL.
 
@@
 
-\textbf{Memory coalescing:} Between a block of threads and a single thread another important entity exists.
+\textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists.
 NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront.
-Memory operations are issued per warp/wavefront.
+(Memory) operations are issued per warp/wavefront.
 A memory access uses the L2 caching behavior.
@@
 
+\textbf{Shared Memory bank conflicts:} Shared memory is split into 32 banks of 4 bytes.
+The hardware groups together unique memory accesses on these banks.
+E.g.
+if two threads within a warp access bank 17 a bank conflict occurs and the access needs to be split into two groups.
+The one exception is a broadcast: all threads within a warp access the same bank.
+The number of banks is equal to the warp size, s.t.
+throughput is maximized if every thread does a unique bank access.
+\cite{NVIDIA:BEST:PRACTICE}
+
+\textbf{Register spilling:} One should avoid local memory.
+Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
+One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
+If the occupancy falls off to 0\% register spilling occurs.
+The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor.
+It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another.
+Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator.
+Non-optimal occupancy might not have any performance impact at all.
+\cite{NVIDIA:ACHIEVED_OCCUPANCY}
+
+To our knowledge there is no way to find out the exact number of registers an OpenGL shader uses.
+The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
+It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version).
+One can identify TEMP variables as registers.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees this is not further optimized.
+All other metrics discussed are readily availabe for OpenGL.
+
+\textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits.
+Though we in general had better experiences starting with a block size of 64 as can be seen in the Occupancy calculator.
+Increasing the block size is useful in combination with shared memory usage.
+Shared memory access is typically 100x faster than uncached global memory accesses.
+The more threads of a block can reuse shared memory data the faster the memory throughput will be.
+Therefore operators which involve filtering (also MPM transfers) can use this behavior.
+
+\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it.
+Otherwise merging sequential and parallel work (in shared memory) can improve performance heavily for instruction-limited shaders.
+Since the block size is limited by the GPU large amounts of data need to be split in tree structures to process in parallel.
+One can not avoid an indirection of writing-back and loading again from global memory.
+Therefore algorithms often can improve much further from more sequential work to avoid this indirection.
+\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
+\subsubsection{GPU Metrics}
+The GPU metric topping all others is the raw (averaged) time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
+\begin{enumerate}
+  \item For the GPU most interesting is the scalability with increasing input data sizes.
+  \item Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
+Therefore one might be interested how long the GPU needs to calculate one physical second:
+    \begin{equation}
+    \Delta t^c_{1s} = \frac{\Delta t_c}{\Delta t_p}.
+    \end{equation}
+    Otherwise one would calculate the speedup of procedure 1 over procedure 2 for a direct comparision:
+
+    \begin{equation}
+      \text{Speedup of Procedure 1} = \frac{\Delta t_{c1}}{\Delta t_{p1}} \frac{\Delta t_{p2}}{\Delta t_{c2}}.
+    \end{equation}
+\end{enumerate}
+  \begin{tabular}{ | l | l | l | l | l |l|}
+    \hline
+    $\Delta t_c$ & VRAM SOL\% & SM SOL\%  & L2 SOL\%& Tex SOL\% &Lifetime   \\
+    \hline
+  \end{tabular}
+\clearpage
+\section{Implementation}\label{sec:implementation}
 \noindent\begin{minipage}{0.45\textwidth}
@@
 For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}.
-Except for the ommission of the glsl version or extension macros, we did not find any problems in this process.
-
-\textbf{Shared Memory Bank conflicts:}
-
-\textbf{Multiple elements per thread and block size:}
-
-\cite{NVIDIA:BEST:PRACTICE} \cite{AMD:GPU_OPEN}
+Except for the omission of the glsl version or extension macros, we did not find any problems in this process.
+
 decorator pattern soa vs aos\\

### 2018-12-05 mscthesis.tex
gpu metrics

@@
 %One of those is the Jacobi (diagonal) preconditioner: $\boldsymbol{P} = \text{diag}\left( \boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \right)$.
-\subsection{General-purpose computing on graphics processing units}
-\subsubsection{Optimization Opportunies}
-\begin{center}
-  \begin{tabular}{ | l | l | l | l | l |}
-    \hline
-    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
-    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
-    Local        &Off                    &L2       &1 thread             &thread     \\\hline
-    Shared       &On                     &n/a      &block                &block      \\\hline
-    Global       &Off                    &L2       &all threads+host     &host alloc \\
-    \hline
-  \end{tabular}
-\end{center}
-\begin{flushright}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
+\subsection{General-purpose computing on graphics processing units(GPGPU)}
+\subsubsection {GPGPU Optimization opportunities}\label{sec:optim}
 The speed of these memory types is given in the following starting with the fastest:
@@
 
+\begin{table}
+  \begin{tabular}{ | l | l | l | l | l |}    \hline
+    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
+    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
+    Local        &Off                    &L2       &1 thread             &thread     \\\hline
+    Shared       &On                     &n/a      &block                &block      \\\hline
+    Global       &Off                    &L2       &all threads+host     &host alloc \\
+    \hline
+  \end{tabular}
+\caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}
+\end{table}
 \textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists.
@@
 A memory access uses the L2 caching behavior.
-A L2-cache-line is 32 Bytes long.
+A L2 cache-line is 32 Bytes long.
 Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.
-
 If all threads of a warp access memory within this cache-line only one memory request of 32 bytes needs to be made.
-In contrast if a warp accesses N different cache-lines, N*32 bytes need to be loaded (assuming a cache miss).
+In contrast, if a warp accesses N different cache-lines, N*32 bytes need to be loaded.
+
+The unified L1/texture cache acts as a coalescing buffer for threads of a warp (Maxwell architecture \cite{NVIDIA:MAXWELL}).
 It does not matter in which order the cache-line is accessed by the warp.
@@
 Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
-One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
+The available number of registers per thread is calculable with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
 If the occupancy falls off to 0\% register spilling occurs.
 The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor.
-It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another.
+It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another; this is called a warp stall, table \ref{table:metric}.
 Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator.
@@
 It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version).
-One can identify TEMP variables as registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees this is not further optimized.
+TEMP variables are registers.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees there are not further optimized.
 All other metrics discussed are readily availabe for OpenGL.
@@
 Shared memory access is typically 100x faster than uncached global memory accesses.
-The more threads of a block can reuse shared memory data the faster the memory throughput will be.
-Therefore operators which involve filtering (also MPM transfers) can use this behavior.
+The more threads of a block reuse shared memory data the faster the memory throughput will be.
+Therefore operators which involve filtering (also MPM-transfers) can use this behavior.
 
 \textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it.
-Otherwise merging sequential and parallel work (in shared memory) can improve performance heavily for instruction-limited shaders.
+Otherwise merging sequential and parallel work improves performance heavily.
 Since the block size is limited by the GPU large amounts of data need to be split in tree structures to process in parallel.
 One can not avoid an indirection of writing-back and loading again from global memory.
-Therefore algorithms often can improve much further from more sequential work to avoid this indirection.
+Therefore algorithms often improve much further from more sequential work to avoid this indirection.
 \begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
 \subsubsection{GPU Metrics}
-The GPU metric topping all others is the raw (averaged) time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
+The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
 \begin{enumerate}
   \item For the GPU most interesting is the scalability with increasing input data sizes.
-  \item Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
+  \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
 Therefore one might be interested how long the GPU needs to calculate one physical second:
@@
 \end{enumerate}
-  \begin{tabular}{ | l | l | l | l | l |l|}
-    \hline
-    $\Delta t_c$ & VRAM SOL\% & SM SOL\%  & L2 SOL\%& Tex SOL\% &Lifetime   \\
+\begin{table}
+  \begin{tabular}{ | l | l | l |}    \hline
+    \textbf{Metric}   & \textbf{Max}(GTX 970) & \textbf{Description} \\ \hline
+    Speedup           & -                     & item \ref{it:speedup}\\\hline
+    $\Delta t_c$      & -                     & computing time\\\hline
+    VRAM SOL\%        & 100                   & memory througput(, w.r.t. to hw.-limit)\\\hline
+    SM SOL\%          & 100                   & instruction throughput\\\hline
+    L2 SOL\%          & 100                   & L2-cache throughput\\\hline
+    Tex SOL\%         & 100                   & L1-cache throughput\\\hline
+    L2-Hitrate\%      & 100                   & L2 hit-miss ratio\\\hline
+    SM Active\%       & 100                   & at least one warp active avg. over SMs\\\hline
+    SM Occupancy      & 64.0                  & active warps avg. over SMs\\\hline
+    Read Active\%     & 100                   & \% of time spend on VRAM reads\\\hline
+    Write Active\%    & 100                   & \% of time spend on VRAM writes\\\hline
+    \multicolumn{3}{|l|}{\textbf{Warp Stall Reasons:}}\\\hline
+    Long Scoreboard\% & 100                   & memory latency\\\hline
+    Barrier\%         & 100                   & block synchronization \\
+    %No Instructions\% & 100                   & \\
     \hline
   \end{tabular}
+\caption{GPU Performance Metrics}\label{table:metric}
+\end{table}
+The other GPU metrics are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}. An active warp is also one that waits for memory.
+
+None of these metrics necessarily have a correlation to computing time.
+However they are good indicators on which operations/hardware units the GPU spends its time on.
+The top SOL\% metric measures fraction of the throughput to the bandwidth of that hardware unit.
+NVIDIA aims to get the top SOL\% metric between 60\% < top SOL\% < 80\%.
+Occupancy was discussed in \ref{sec:optim}.
+\cite{NVIDIA:PEAK_PERFORMANCE}
+
+Aiming to get shaders memory-bandwidth limited while minimizing the amount of needed accesses for any given workload is the top-priority for our shaders.
+Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% < top SOL\% < 80\%.
+The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.
 \clearpage
 \section{Implementation}\label{sec:implementation}
+GL\_EXT\_TIMER + NVIDIA Nsight\\
+\subsection{GPU memory layout}
 \noindent\begin{minipage}{0.45\textwidth}
@@
 
-Coalescing leads to the motivation of using structure of array layouts over arrays of struct layouts.
+Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts.
 The latter is the more commonly taught approach, while high performance CPU code may also prefer SoA or the more advanced entity component systems.
@@
   \item Multiple buffering, e.g. double buffer for unsorted and sorted particles.
-  \item Single and double precision (PREC\_VEC\_TYPE in listing \ref{llst:opengl}).
+  \item Single and double precision (PREC\_VEC\_TYPE in Listing \ref{llst:opengl}).
   \item A decorator pattern allows to add buffer information to existing buffer information: For instance index sort needs an extra buffer containing the indices.
@@
 
-decorator pattern soa vs aos\\
-magic\_get reflection\\
-single + double\\
-GL\_EXT\_TIMER + NVIDIA Nsight\\
 \subsection{Parallel Reduction}
@@
 \listoffigures
+\listoftables
 \lstlistoflistings

### 2018-12-06 mscthesis.tex
performance

@@
 \section{Implementation}\label{sec:implementation}
-GL\_EXT\_TIMER + NVIDIA Nsight\\
+The computing time $\Delta t_c$ of shaders is gained by OpenGL timer queries \cite{KHRONOS:TIMER_QUERY}.
+A small benchmarker takes care of queuing timer queries and resolving them a frame later.
+This is done to not stall the GPU pipeline waiting on those queries.
+At the end of testing data is gathered and statistical quantities are written into a log file.
 \subsection{GPU memory layout}

### 2018-12-08 mscthesis.tex
evaluation start

@@
 \usepackage{verbatim}
+\usepackage{graphicx}
 \usepackage{blindtext}
+\usepackage{svg}
 \addbibresource{mscthesis.bib}
@@
     Long Scoreboard\% & 100                   & memory latency\\\hline
-    Barrier\%         & 100                   & block synchronization \\
+    Barrier\%         & 100                   & block synchronization \\\hline
     %No Instructions\% & 100                   & \\
+    Drain\%           & 100                   & memory wait before warp exit \\
     \hline
@@
 The cache-line is then filled with position vectors, which all will get used.
-In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the bandwidth.
+In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the throughput.
+Chapter \ref{sec:eval}.
+
 \noindent\begin{wrapfigure}{R}{0.5\textwidth}
@@
 \end{itemize}
+\clearpage
 \section{Evaluation}
-\cite{NVIDIA:PEAK_PERFORMANCE}
-\cite{KHRONOS:TIMER_QUERY}
+\textbf{SoA vs. AoS}:
+
+\begin{table}
+  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
+    Description &  $\Delta t_c (\mu s)$ & Speedup &VRAM     & SM     & L2     & Tex \\\hline
+    AoS         & 278                   & -       &77.7\%   & 7.3\%  & 30.3\% & 14.7\% \\\hline
+    SoA         & 122                   & 2.26x   &75.4\%   & 14.3\% & 29.4\% & 17.9\% \\
+    \hline
+  \end{tabular}
+  \caption{Map operation with one instruction of form \texttt{float x = length2( (vec4) y)} on $1024 \times 1024$ \texttt{vec4}.}\label{table:map_1}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
+    Description &  $\Delta t_c (\mu s)$ & Speedup &VRAM   & SM     & L2     & Tex \\\hline
+    AoS         & 292                   & -       &61.3\% & 11.4\% & 41.8\% & 20.4\% \\\hline
+    SoA         & 252                   & 1.16x   &75.4\% & 14.3\% & 29.4\% & 17.9\% \\
+    \hline
+  \end{tabular}
+  \begin{tabular}{ | l | l | l | l | }    \hline
+    Description &  L2-Hit & Read Active \\\hline
+    AoS   & 53.8\%        & 48.9\%      \\\hline
+    SoA   & 20.0\%          & 62.3\%      \\
+    \hline
+  \end{tabular}
+  \caption{Map operation with two instructions of form \texttt{float x = length2( (vec4) y)} on $1024 \times 1024 \times 2$ \texttt{vec4}.}\label{table:map_2}
+\end{table}
+
+\begin{figure}[htbp]
+  \centering
+  \includesvg[svgpath =images/,width=\textwidth]{soa}
+  \caption{SoA-Layout in practive. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
+\end{figure}
+\begin{figure}[htbp]
+  \centering
+  \includesvg[svgpath =images/,width=\textwidth]{aos}
+  \caption{AoS-Layout in practice.
+The first instruction is shown in orange, the second instruction in green.
+Orange already loads both cache-lines.
+If green is left out throughput is halved.
+Also green is an L2-cache lookup requiring additional overhead over the direct SoA way.}
+\end{figure}
 \section{Conclusion}

### 2018-12-09 mscthesis.tex
red.+scan

@@
 \cite{MPM:APIC}\cite{MPM:COURSE}\cite{steffen2008analysis}\cite{gao2017adaptive}
-\subsubsection{Mass transfer}
+\subsubsection{Mass transfer}\label{sec:mass}
 Each material point will be assigned an initial volume $_0V_p$ as well as a initial mass $_0m_p$.
@@
 In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the throughput.
-Chapter \ref{sec:eval}.
+This process is show in Figure \ref{fig:soa} and \ref{fig:aos}.
 
@@
 One could also build a shader using the abstract layout to convert from SoA to AoS.
+
+\begin{figure}[htbp]
+  \centering
+  \includesvg[svgpath =images/,width=\textwidth]{soa}
+  \caption{SoA-Layout in practive. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
+\label{fig:soa}
+\end{figure}
+\begin{figure}[htbp]  \centering
+  \includesvg[svgpath =images/,width=\textwidth]{aos}
+  \caption{AoS-Layout in practice.
+The first instruction is shown in orange, the second instruction in green.
+Orange already loads both cache-lines.
+If green is left out throughput is halved.
+Also green is an L2-cache lookup requiring additional overhead over the direct SoA way.}
+\label{fig:aos}
+\end{figure}
+
 To name a few more cases that macros handle for us:
@@
 
-\subsection{Parallel Reduction}
-\cite{NVIDIA:PARALLEL_REDUCTION}
+\subsection{Parallel Reduction \& Scan}
+The schemes presented in the following are a result of the optimization opportunities discussed in \ref{sec:optim}.
+
+\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$, parallel reduction computes:
+\begin{equation}
+  r = a_0 \circ a_1 \circ ... \circ a_n.
+\end{equation}
+Our OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference.
+To avoid memory padding in shared memory the addressing scheme is given by the sequential addressing scheme in Figure \ref{fig:sequential}.
+\begin{figure}[htbp]
+    \centering
+  \includesvg[svgpath =images/,width=\textwidth]{sequential}
+  \caption{Sequential addressing for \texttt{vec4}s.
+Even threads are blue, odd are yellow.
+Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers.
+The second instruction is only green.
+The first and all following instructions can be handled fully concurrently.
+No memory padding required.}
+\label{fig:sequential}
+\end{figure}
+
+Parallel reduction schemes using shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}.
+They use registers instead of shared memory and thus do not need memory padding.
+\cite{NVIDIA:SHUFFLE}
+
+\cite{NVIDIA:PARALLEL_REDUCTION} also makes heavy use of combining parallel and sequential work.
+Every thread adds a multiple of two sequential adds into shared memory.
+After this the thread does its parallel work.
+Therefore the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.
+
+Parallel reduction solves assuming a \texttt{unary\_op(x)} directly after memory load:
+\begin{itemize}
+  \item CFL-condition of chapter \ref{sec:cfl} with operators
+  \begin{itemize}
+  \item \texttt{unary\_op(x) := length2(x)} and
+  \item \texttt{binary\_op(x,y) := max(x,y)}.
+  \end{itemize}
+\item Discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
+\begin{itemize}
+  \item \texttt{unary\_op(x) := length2(x)} and
+  \item \texttt{binary\_op(x,y) := x + y}.
+  \end{itemize}
+\end{itemize}
+
+\textbf{Scan:}  Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op} and an array of values $\left[a_0,a_1, ...
+,a_n\right]$, an exclusive scan computes the array:
+\begin{equation}
+  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
+\end{equation}
+The scan uses an interleaved addressing pattern, Figure \ref{fig:interleaved}. The interleaved addressing pattern has the benefit over the sequential one of already computing the
+\begin{wrapfigure}[15]{r}{0.20\textwidth}
+ \includesvg[svgpath =images/,width=0.20\textwidth]{scan}
+  \caption{Scan for addition.}
+  \label{fig:scan}
+\end{wrapfigure}
+partial scans for all odd indices (although misaligned).
+Note that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}.
+A full scan is then computed as \cite{NVIDIA:SCAN}:
+\begin{enumerate}
+  \item \label{it:up-sweep} Upsweep-Phase: Top-to-Bottom reduction to the right of the array.
+  \item Inserting the neutral element into the last element of the array.
+  \item Downsweep-Phase: Reverse execution order of item \ref{it:up-sweep} (inverted tree), and an additional right to left element swap.
+\end{enumerate}
+Again, sequential work can be combined with parallel work.
+\cite{NVIDIA:SCAN_MODERN} name this process Raking: Instead of loading one element per thread, one thread loads n elements and computes a sequential partial scan in register memory.
+The reduced last value of the array is written into shared memory and the parallel scan begins.
+Afterwards the parallel scan result is spread out with the sequential partial scan.\cite{NVIDIA:SCAN_MODERN}
+\begin{figure}[htbp]
+  \includesvg[svgpath =images/,width=\textwidth]{interleaved}
+  \caption{Interleaved addressing \texttt{vec4}s.
+Even threads are blue, odd are yellow.
+Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers.
+The second instruction is only green.
+Memory padding needs to be done for concurrent shared-memory access.
+The first and all following instructions can be handled fully concurrently.}
+\label{fig:interleaved}
+\end{figure}
+As mentioned, the interleaved addressing requires a memory padding for conflict-free addresses. For block sizes of 32 - 512 this becomes:
+\begin{equation}\label{eq:bank}
+  p = u + (u << \log_2(n)), \text{ or }   p = u + \frac{u}{n}
+\end{equation}
+where $p$ is the padded memory offset, $u$ is the unpadded memory offset and $n$ is the number of banks.
+For a block size of 1024 one would need additional padding for a full conflict-free access:
+\begin{equation}\label{eq:bank_1024}
+  p = u + (u << \log_2(n)) + (u << (2 * \log_2(n))), \text{ or }   p = u + \frac{u}{n} + \frac{u}{n*n}.
+\end{equation}
+However the added computation time does not outweigh the bank-conflicts.
+So one reverts to eq.
+\ref{eq:bank}.
+Note that this is a correction to \cite{NVIDIA:SCAN} where C++ Operator Precedence of $+$ over $<<$ is not recognized and additionally incorrectly $ u << n$ is used in the first part of eq.
+\ref{eq:bank_1024}.
 \subsection{Counting sort}
 Map (Reset) \\
-\subsubsection{Bin}
+\textbf{Bin}
 Binning\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
-\subsubsection{Scan}
-Scan\cite{NVIDIA:SCAN}\\
-Scan(Raking) \cite{NVIDIA:SCAN_MODERN}
-\subsubsection{Reordering}
-Coalesced, Uncoalesced
+\textbf{Reordering}
 \subsection{MPM-Transfers}
@@
 \clearpage
-\section{Evaluation}
+\section{Evaluation}\label{sec:eval}
 \textbf{SoA vs. AoS}:
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includesvg[svgpath =images/,width=\textwidth]{soa}
-  \caption{SoA-Layout in practive. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
-\end{figure}
-\begin{figure}[htbp]
-  \centering
-  \includesvg[svgpath =images/,width=\textwidth]{aos}
-  \caption{AoS-Layout in practice.
-The first instruction is shown in orange, the second instruction in green.
-Orange already loads both cache-lines.
-If green is left out throughput is halved.
-Also green is an L2-cache lookup requiring additional overhead over the direct SoA way.}
-\end{figure}
 \section{Conclusion}

### 2018-12-10 mscthesis.tex
tile_scan

@@
 Again, sequential work can be combined with parallel work.
-\cite{NVIDIA:SCAN_MODERN} name this process Raking: Instead of loading one element per thread, one thread loads n elements and computes a sequential partial scan in register memory.
+\cite{NVIDIA:SCAN_MODERN} name this process: Raking.
+Instead of loading one element per thread, one thread loads n elements and computes a sequential partial scan in register memory.
 The reduced last value of the array is written into shared memory and the parallel scan begins.
@@
 \ref{eq:bank_1024}.
+A scan is used to calculate the memory offsets of the by grid node sorted particles.
+
+\textbf{A Block Scan:} Assume we have already computed a global scan on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
+This again forms a uniform grid.
+For simplicity assume two dimensions:
+
+\noindent Given the initial two-dimensional pre-scan array:
+\begin{equation}
+  \boldsymbol{G}=
+  \left[
+  \begin{array} {c c c c }
+    a_{00} & a_{01} &  ... & a_{0n}\\
+
+    a_{10} & a_{11} &  ... & a_{1n}\\
+    ...    & ...    &  ... & ...   \\
+    a_{m0} & a_{m1} &  ... & a_{mn}\\
+\end{array}
+\right].
+\end{equation}
+We can identify a block with $i,j,k,l\geq 0$, $i+k<m$ and $j+l<n$ as:
+\begin{equation}
+  \boldsymbol{L}=
+  \left[
+  \begin{array} {c c c c }
+    a_{ij} & a_{(i)(j+1)} &  ... & a_{(i)(n-l)}\\
+
+    a_{(i+1)(j)} & a_{(i+1)(j+1)} &  ... & a_{(i+1)(n-l)}\\
+    ...    & ...    &  ... & ...   \\
+    a_{(m-k)(j)} & a_{(m-k)(j+1)} &  ... & a_{(m-k)(n-l)}\\
+\end{array}
+\right].
+\end{equation}
+We use the $\Sigma$-Notation for sums although any associative binary operator would hold. The global scan is computed as all previous rows + the current row to the current element:
+\begin{equation}
+  S^G_{st} = \sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(t-1)} a_{sr}
+\end{equation}
+with $0 \leq s,t \leq m,n$.
+The local scan is computed similiar:
+\begin{equation}
+  S^L_{bc} = \sum_{q,r=i,j}^{(b-1),(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br}.
+\end{equation}
+This however requires a full scan of $\boldsymbol{S}^L$, where $i,j \leq b,c \leq (m-k),(n-l)$.
+We can alternatively compute the local scan from the global scan with the following five steps:
+\begin{enumerate}
+\item Subtract the global scan's first element of every row-element and limit your view to indices $b,c$ then:
+  $$
+  T_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
+  $$
+\begin{equation}
+  \Rightarrow T_{bc} = \sum_{r=j}^{(c-1)} a_{br}
+\end{equation}
+\item Add the last value of $\boldsymbol{L}$ of the previous row (if it exists) to the current row.
+  \begin{equation}
+    \Rightarrow T^2_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
+  \end{equation}
+\item Extract the last column $c =(n-l)$:
+  \begin{equation}
+    t_b = a_{(b-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{br}.
+  \end{equation}
+\item Compute the in dimensionality-by-one reduced exclusive scan (of the last column):
+  \begin{equation}
+    s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right).
+\end{equation}
+\item Row-wise add back:
+  $$T_{bc} + s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right) + \sum_{r=j}^{(c-1)} a_{br}$$
+  \begin{equation}
+    = \sum_{q=i}^{(b-1)}\sum_{r=j}^{(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br} = S^L_{bc}.
+  \end{equation}
+\end{enumerate}
+\begin{wrapfigure}{r}{0.53\textwidth}
+  \includesvg[svgpath =images/,width=0.53\textwidth]{tile_scan}
+  \caption{.}
+  \label{fig:tile_scan}
+\end{wrapfigure}
+For a simpler understanding of this process Figure \ref{fig:tile_scan} shows the process with the same numbering.
 \subsection{Counting sort}

### 2019-01-07 mscthesis.tex
smol fix, momentum cons.
graphics done

@@
 \begin{enumerate}
-  \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it. Since this results in a race condition \texttt{atomicAdd(float f)}'s are in use.
+  \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it.
+Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary.
   \item \textbf{Particle Pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
@@
 
-Alternatively one may also compute the partial sum of the transfer on the halo. Two different techniques exist to get the data between blocks coherent again:
+In Particle Push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
 \begin{enumerate}
@@
 This results in two atomic adds on a side node, four on an edge node and eight on a corner-node; If the block extend is bigger than half the halo extend.
-\item Have every halo of every block correspond to actual global memory and do add steps along each of the three axes. This is infeasible for large supports. \cite{crassin2011interactive}
+\item Have every halo of every block correspond to actual global memory and do add steps along each of the three axes.
+This is infeasible for large supports due to the added memory requirements.
+\cite{crassin2011interactive}
 \end{enumerate}
@@
 TEMP variables are registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees there are not further optimized.
-All other metrics discussed are readily availabe for OpenGL.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees these are not further optimized.
+All other metrics discussed are readily available for OpenGL.
 
@@
 \begin{itemize}
-  \item \texttt{unary\_op(x) := length2(x)} and
+  \item \texttt{unary\_op(x) := x} and
   \item \texttt{binary\_op(x,y) := x + y}.
+  \end{itemize}
+\item Calculate the maximum count of particles of all grid nodes within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
+\begin{itemize}
+  \item \texttt{unary\_op(x) := x} and
+  \item \texttt{binary\_op(x,y) := max(x,y)}.
 \end{itemize}
@@
 \ref{eq:bank_1024}.
-A scan is used to calculate the memory offsets of the by grid node sorted particles.
-
-\textbf{A Block Scan:} Assume we have already computed a global scan on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
-This again forms a uniform grid.
-For simplicity assume two dimensions:
-
-\noindent Given the initial two-dimensional pre-scan array:
-\begin{equation}
-  \boldsymbol{G}=
-  \left[
-  \begin{array} {c c c c }
-    a_{00} & a_{01} &  ... & a_{0n}\\
-
-    a_{10} & a_{11} &  ... & a_{1n}\\
-    ...    & ...    &  ... & ...   \\
-    a_{m0} & a_{m1} &  ... & a_{mn}\\
-\end{array}
-\right].
-\end{equation}
-We can identify a block with $i,j,k,l\geq 0$, $i+k<m$ and $j+l<n$ as:
-\begin{equation}
-  \boldsymbol{L}=
-  \left[
-  \begin{array} {c c c c }
-    a_{ij} & a_{(i)(j+1)} &  ... & a_{(i)(n-l)}\\
-
-    a_{(i+1)(j)} & a_{(i+1)(j+1)} &  ... & a_{(i+1)(n-l)}\\
-    ...    & ...    &  ... & ...   \\
-    a_{(m-k)(j)} & a_{(m-k)(j+1)} &  ... & a_{(m-k)(n-l)}\\
-\end{array}
-\right].
-\end{equation}
-We use the $\Sigma$-Notation for sums although any associative binary operator would hold. The global scan is computed as all previous rows + the current row to the current element:
-\begin{equation}
-  S^G_{st} = \sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(t-1)} a_{sr}
-\end{equation}
-with $0 \leq s,t \leq m,n$.
-The local scan is computed similiar:
-\begin{equation}
-  S^L_{bc} = \sum_{q,r=i,j}^{(b-1),(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br}.
-\end{equation}
-This however requires a full scan of $\boldsymbol{S}^L$, where $i,j \leq b,c \leq (m-k),(n-l)$.
-We can alternatively compute the local scan from the global scan with the following five steps:
-\begin{enumerate}
-\item Subtract the global scan's first element of every row-element and limit your view to indices $b,c$ then:
-  $$
-  T_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
-  $$
-\begin{equation}
-  \Rightarrow T_{bc} = \sum_{r=j}^{(c-1)} a_{br}
-\end{equation}
-\item Add the last value of $\boldsymbol{L}$ of the previous row (if it exists) to the current row.
-  \begin{equation}
-    \Rightarrow T^2_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
-  \end{equation}
-\item Extract the last column $c =(n-l)$:
-  \begin{equation}
-    t_b = a_{(b-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{br}.
-  \end{equation}
-\item Compute the in dimensionality-by-one reduced exclusive scan (of the last column):
-  \begin{equation}
-    s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right).
-\end{equation}
-\item Row-wise add back:
-  $$T_{bc} + s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right) + \sum_{r=j}^{(c-1)} a_{br}$$
-  \begin{equation}
-    = \sum_{q=i}^{(b-1)}\sum_{r=j}^{(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br} = S^L_{bc}.
-  \end{equation}
-\end{enumerate}
-\begin{wrapfigure}{r}{0.53\textwidth}
-  \includesvg[svgpath =images/,width=0.53\textwidth]{tile_scan}
-  \caption{.}
-  \label{fig:tile_scan}
-\end{wrapfigure}
-For a simpler understanding of this process Figure \ref{fig:tile_scan} shows the process with the same numbering.
-\subsection{Counting sort}
-Map (Reset) \\
-\textbf{Bin}
-Binning\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
+
+A scan is used to calculate memory offsets for stream compaction or sorting algorithms.
+
+One can also calculate a per block scan (with halo) without doing a full scan for every block (Appendices \ref{sec:block_scan}).
+We did not end up using this more closely related approach to \cite{NVIDIA:NNSEARCH} since loading more particles does not improve performance for the implemented transfers (see chapter \ref{sec:eval}).
+\subsection{Counting Sort}
+\textbf{Binning}\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
 \textbf{Reordering}
@@
 \subsubsection{Particle to grid transfers}
+\begin{figure}[htbp]
+    \centering
+  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_par}
+  \caption{dwad}
+  \label{fig:p2g_transfer}
+\end{figure}
 \subsubsection{Grid to particle transfers}
+
+\begin{wrapfigure}{r}{0.55\textwidth}
+  \includesvg[svgpath =images/,width=0.55\textwidth]{block_transfer_inv_par}
+  \caption{dwad}
+  \label{fig:g2p_transfer}
+\end{figure}
+\subsubsection{Active Blocks /& Maximum Block Count:}\label{sec:active_blocks}
+\begin{figure}[htbp]
+  \includesvg[svgpath =images/,width=\textwidth]{indirect_dispatch}
+  \caption{dwad}
+  \label{fig:active_block}
+\end{figure}
+
 \subsection{Verifying Simulation}
@@
 \end{itemize}
+
 \clearpage
@@
   $$
+\section{Block Scan}\label{sec:block_scan}
+
+\textbf{A Block Scan:} Assume we have already computed a global scan on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
+This again forms a uniform grid.
+For simplicity assume two dimensions:
+
+\noindent Given the initial two-dimensional pre-scan array:
+\begin{equation}
+  \boldsymbol{G}=
+  \left[
+  \begin{array} {c c c c }
+    a_{00} & a_{01} &  ... & a_{0n}\\
+
+    a_{10} & a_{11} &  ... & a_{1n}\\
+    ...    & ...    &  ... & ...   \\
+    a_{m0} & a_{m1} &  ... & a_{mn}\\
+\end{array}
+\right].
+\end{equation}
+We can identify a block with $i,j,k,l\geq 0$, $i+k<m$ and $j+l<n$ as:
+\begin{equation}
+  \boldsymbol{L}=
+  \left[
+  \begin{array} {c c c c }
+    a_{ij} & a_{(i)(j+1)} &  ... & a_{(i)(n-l)}\\
+
+    a_{(i+1)(j)} & a_{(i+1)(j+1)} &  ... & a_{(i+1)(n-l)}\\
+    ...    & ...    &  ... & ...   \\
+    a_{(m-k)(j)} & a_{(m-k)(j+1)} &  ... & a_{(m-k)(n-l)}\\
+\end{array}
+\right].
+\end{equation}
+We use the $\Sigma$-Notation for sums although any associative binary operator would hold. The global scan is computed as all previous rows + the current row to the current element:
+\begin{equation}
+  S^G_{st} = \sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(t-1)} a_{sr}
+\end{equation}
+with $0 \leq s,t \leq m,n$.
+The local scan is computed similiar:
+\begin{equation}
+  S^L_{bc} = \sum_{q,r=i,j}^{(b-1),(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br}.
+\end{equation}
+This however requires a full scan of $\boldsymbol{S}^L$, where $i,j \leq b,c \leq (m-k),(n-l)$.
+We can alternatively compute the local scan from the global scan with the following five steps:
+\begin{enumerate}
+\item Subtract the global scan's first element of every row-element and limit your view to indices $b,c$ then:
+  $$
+  T_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
+  $$
+\begin{equation}
+  \Rightarrow T_{bc} = \sum_{r=j}^{(c-1)} a_{br}
+\end{equation}
+\item Add the last value of $\boldsymbol{L}$ of the previous row (if it exists) to the current row.
+  \begin{equation}
+    \Rightarrow T^2_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
+  \end{equation}
+\item Extract the last column $c =(n-l)$:
+  \begin{equation}
+    t_b = a_{(b-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{br}.
+  \end{equation}
+\item Compute the in dimensionality-by-one reduced exclusive scan (of the last column):
+  \begin{equation}
+    s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right).
+\end{equation}
+\item Row-wise add back:
+  $$T_{bc} + s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right) + \sum_{r=j}^{(c-1)} a_{br}$$
+  \begin{equation}
+    = \sum_{q=i}^{(b-1)}\sum_{r=j}^{(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br} = S^L_{bc}.
+  \end{equation}
+\end{enumerate}
+For a simpler understanding of this process Figure \ref{fig:tile_scan} shows the process with the same numbering:
+\begin{figure}[htbp]
+  \includesvg[svgpath =images/,width=0.53\textwidth]{tile_scan}
+  \caption{A per block scan following the above enumeration.}
+  \label{fig:tile_scan}
+\end{figure}
 \end{appendices}

### 2019-01-11 mscthesis.tex
counting sort

@@
 
-\subsection{Parallel Reduction \& Scan}
+\subsection{Parallel Reduction \& Scan} \label{sec:red_scan}
 The schemes presented in the following are a result of the optimization opportunities discussed in \ref{sec:optim}.
@@
 \end{equation}
-The scan uses an interleaved addressing pattern, Figure \ref{fig:interleaved}. The interleaved addressing pattern has the benefit over the sequential one of already computing the
-\begin{wrapfigure}[15]{r}{0.20\textwidth}
+\begin{wrapfigure}[13]{r}{0.20\textwidth}
+\vspace{-11}
  \includesvg[svgpath =images/,width=0.20\textwidth]{scan}
-  \caption{Scan for addition.}
+ \caption{Scan (add)}
   \label{fig:scan}
 \end{wrapfigure}
-partial scans for all odd indices (although misaligned).
+The scan uses an interleaved addressing pattern, Figure \ref{fig:interleaved}.
+The interleaved addressing pattern has the benefit over the sequential one of already computing the partial scans for all odd indices (although misaligned).
 Note that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}.
@@
 One can also calculate a per block scan (with halo) without doing a full scan for every block (Appendices \ref{sec:block_scan}).
-We did not end up using this more closely related approach to \cite{NVIDIA:NNSEARCH} since loading more particles does not improve performance for the implemented transfers (see chapter \ref{sec:eval}).
-\subsection{Counting Sort}
-\textbf{Binning}\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
-\textbf{Reordering}
-\subsection{MPM-Transfers}
+This implementation did not end up using this more closely related approach to \cite{NVIDIA:NNSEARCH} as it would commit shared memory bank-conflicts defying the architecture of the GPU.
+Instead preferred is batching a fixed amount of particles, more in chapter \ref{sec:transfers}.
+\clearpage
+\subsection{Counting Sort \& Stream Compaction}
+As the sorting algorithm of choice counting sort is used as in \cite{NVIDIA:NNSEARCH}. Counting sort is split into three parts:
+\begin{enumerate}
+  \item
+\parbox[t]{\dimexpr\textwidth-\leftmargin}{%
+      \vspace{-7.5}
+      \begin{wrapfigure}[10]{r}{0.425\textwidth}
+	\vspace{-33}
+ \includesvg[svgpath =images/,width=0.425\textwidth]{binning}
+  \caption{Binning: Particle adds one to the grid counter after storing it.}
+  \label{fig:binning}
+\end{wrapfigure}
+  \textbf{Binning}: Transform the particle position into the grid space, such that every integer in range corresponds to a grid node.
+Bin due to flooring the transformed particle: Store the particle count $c_i$ of this grid node on the particle as an offset $o_p$ into this grid node.$$o_p := c_i.$$Following that, up the counter on the grid by one:$$c_i := c_i +1.$$
+}
+  \item \textbf{Scan}: Compute a per grid node scan $s_i$ on the counter $c_i$ following \ref{sec:red_scan}.
+
+  \item \textbf{Reordering}: $i_p := s_i + o_p$ will now lead to a new ordering of the particle indices.
+Sorting is now similar to methods \texttt{sort\_by\_key(uint i)}, where ordering of particles within the same bin is undefined.
+Supported is either a full deep copy of all particle variables or alternatively an indexing to access an existing particle buffer in the specified sorted manner.
+\end{enumerate}
+The one-dimensional indexing of the grid nodes can be chosen as preferred. Possible are a per-block indexing or a global indexing which are both supported. \cite{NVIDIA:NNSEARCH}
+\subsection{MPM-Transfers}\label{sec:transfers}
 Conjugate gradient method

### 2019-01-12 mscthesis.tex
sort,stream compac, start mpm ops.

@@
 \section{Introduction} \label{intro}
-\section{Related Work}
+\section{Related Work}\label{sec:rel_work}
 \textbf{Particle-in-Cell (PIC):} PIC techniques were developed initially for hydrodynamics
@@
   \item Factoring out mass:
-    \begin{equation}
+    \begin{equation}\label{eq:mass_divide}
       \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
@@
   \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid to particle transfer:
-    \begin{equation}
+    \begin{equation}\label{eq:Ap}
     \boldsymbol{A}_p = \frac { \partial ^ { 2 } \Psi } { \partial \boldsymbol { F } \partial \boldsymbol { F } } ( \hat { \boldsymbol { F } } _ { p } ( \hat { \boldsymbol { x } } )) : \left(\sum_j \delta \boldsymbol{u}_j (\nabla w_{jp}^n)^T \boldsymbol{F}^n_p\right).
@@
   \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle to grid transfer:
-\begin{equation}
+    \begin{equation}\label{eq:delta_f}
   - \delta \boldsymbol{f}_i = - \sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j = \sum_p V_p^0 \boldsymbol{A}_p(\boldsymbol{F}^n_p)^T \nabla w_{ip}^n.
@@
 \subsection{Counting Sort \& Stream Compaction}
-As the sorting algorithm of choice counting sort is used as in \cite{NVIDIA:NNSEARCH}. Counting sort is split into three parts:
+Sorting on the GPU can dramatically increase workload performance of subsequent steps. If they can profit from following things:
+\begin{enumerate}
+  \item Accesses can now be handled in a coalesced fashion. \label{it:sort_1}
+  \item Data can now be reused due to L2-Cache and shared memory. \label{it:sort_2}
+\end{enumerate}
+This is the case for algorithms which require neighbor communication.
+Supported is either a full deep copy of all particle variables.
+A deep copy profits from item \ref{it:sort_1} and \ref{it:sort_2}, but takes the hit in performance of uncoalesced accesses for all variables.
+Or alternatively an indexing to access an existing particle buffer in the specified sorted manner.
+Indexing only profits from item \ref{it:sort_2} and subsequent shaders will be uncoalesced.
+
+\textbf{Stream Compaction} reduces applicable data for subsequent operations.
+The applicable data is checked by a condition.
+There are a lot of cases where stream compaction can reduce running times.
+In chapter \ref{sec:rel_work} particle activation is mentioned as one of such, chapter \ref{sec:active_blocks} will reduce to active blocks.
+Collision treatment could be implemented as an operation following a collision detection etc.
+Again indexing and deep copies can be executed for the same reasons mentioned in \ref{it:sort_1} and \ref{it:sort_2}.
+
+\textbf{Counting Sort}: As the sorting algorithm of choice counting sort is used.
+Counting sort is split into three parts:
 \begin{enumerate}
@@
  \includesvg[svgpath =images/,width=0.425\textwidth]{binning}
-  \caption{Binning: Particle adds one to the grid counter after storing it.}
+  \caption{Binning: Particle adds one to $c_i$ after storing it as $o_p$.}
   \label{fig:binning}
@@
 }
+This process is visualized in Figure \ref{fig:binning}.
   \item \textbf{Scan}: Compute a per grid node scan $s_i$ on the counter $c_i$ following \ref{sec:red_scan}.
@@
 Sorting is now similar to methods \texttt{sort\_by\_key(uint i)}, where ordering of particles within the same bin is undefined.
-Supported is either a full deep copy of all particle variables or alternatively an indexing to access an existing particle buffer in the specified sorted manner.
 \end{enumerate}
 The one-dimensional indexing of the grid nodes can be chosen as preferred. Possible are a per-block indexing or a global indexing which are both supported. \cite{NVIDIA:NNSEARCH}
-\subsection{MPM-Transfers}\label{sec:transfers}
-Conjugate gradient method
-\subsubsection{Particle to grid transfers}
+
+Particles could potentially access the same bin at the same time resulting in a race condition.
+A naive solution to the problem are atomic accesses.
+As long as the (atomic) writes can be handled coalesced this process can be done fast by the GPU, chapter \ref{sec:eval}.
+This is only the case for a full deep sort.
+Therefore binning in the case of a full sort acts upon the last sorted state.
+Double buffering particles allows to do so.
+However binning is done after a position update that changes the ordering.
+Full coalescing will therefore generally not occur.
+
+Alternatively one could work again with the CFL-condition. Since particles move at most one bin in the 27-neighborhood one could mimic a Particle-to-Grid transfer, chapter \ref{sec:p2g}:
+\begin{enumerate}
+  \item \textbf{Labeling}: Label the particles by the bin they want to change to.
+  \item \textbf{Neighborhood lookup}: A bin then assigns offsets to the particles in the neighborhood and increases its counter.
+\end{enumerate}
+Similarly to the Particle-to-Grid transfer one could choose from the different algorithms presented there to do so.
+However without additional endeavors the order within a bin is not preserved.
+One would need to rely on the L1-cache as a coalescing buffer as does the atomic global solution.
+
+Alternatively one could choose a closer approach to \cite{NVIDIA:NNSEARCH}'s neighborhood search, which greedily loads all particles.
+Ordering can easily be preserved then.
+Note that the support in their approach is smaller (27 instead of 64 nodes) and the data required per particle is only the label.
+
+Chapter \ref{sec:active_blocks} shows how to do this only for active blocks. Due to the difference in support the same pipeline cannot be used.
+
+In summary there is little gains to do so as binning is only executed once per step and is not the dominant performance cost.
+
+\subsection{MPM-Operations}\label{sec:transfers}
+Easy to parallelize are all operations that are executed entirely on the grid or entirely on the particles:
+$$\square_p = \square_p \circ \square_p \circ ... \circ \square_p,$$
+$$\square_i = \square_i \circ \square_i \circ ... \circ \square_i.$$
+One thread will just equal to one node or particle respectively.
+For one variable operations mostly used is a map-shader that applies a unary operation to any input element.
+This can be used for instance for the mass divide of momentum on the grid (equation \ref{eq:mass_divide}) or resetting buffers that are target of global atomic operations.
+
+In contrast much harder are transfers between the two structures, that are essential to the Material-Point-Method.
+
+Particle-to-Grid(P2G)-transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid; as well as the computation of the Hessian of equation \ref{eq:delta_f}.
+In general these can be summarized as:
+$$\square_i = \sum_p \square_p \circ \square_{ip}.$$
+The transfers back to the particles are called Grid-to-Particle(G2P)-transfers.
+Apart from the APIC-transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}) the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
+In short all transfers of the form:
+$$\square_p = \sum_i \square_i \circ \square_{ip}.$$
+
+\subsubsection{Particle-to-Grid transfers}\label{sec:p2g}
 \begin{figure}[htbp]
@@
 \end{figure}
-\subsubsection{Grid to particle transfers}
-
+\subsubsection{Grid to particle transfers}\label{sec:g2p}
 \begin{wrapfigure}{r}{0.55\textwidth}

### 2019-01-13 mscthesis.tex
g2p transfers

@@
 \clearpage
-\subsection{Counting Sort \& Stream Compaction}
+\subsection{Counting Sort \& Stream Compaction}\label{sec:sort}
 Sorting on the GPU can dramatically increase workload performance of subsequent steps. If they can profit from following things:
@@
 \end{enumerate}
-The one-dimensional indexing of the grid nodes can be chosen as preferred. Possible are a per-block indexing or a global indexing which are both supported. \cite{NVIDIA:NNSEARCH}
+\begin{flushright}\cite{NVIDIA:NNSEARCH}\cite{MPM:GPU}\end{flushright}
+
+The one-dimensional indexing of the grid nodes can be chosen as preferred. Possible are a per-block indexing or a global indexing which are both supported.
 
@@
 
-In contrast much harder are transfers between the two structures, that are essential to the Material-Point-Method.
+In contrast much harder are transfers between the two structures, that are essential to the Material-Point-Method:
 
@@
 $$\square_i = \sum_p \square_p \circ \square_{ip}.$$
+
 The transfers back to the particles are called Grid-to-Particle(G2P)-transfers.
@@
 
-\subsubsection{Particle-to-Grid transfers}\label{sec:p2g}
+The MPM-transfers are different from a typical stencil or filter operation in two main factors:
+\begin{enumerate}
+  \item \textbf{Dynamic weights}: The stencil is not static.
+In the MPM-Method the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
+This also means the position needs to be temporarily stored to apply the transfer.
+  \item \label{it:part_nodes} \textbf{Participating nodes}: A stencil operation reads from a fixed amount of nodes in its neighborhood given by the support of the stencil.
+This is a node to node relationship which all threads typically participate in from start to finish for the full range of the support.
+
+In a MPM-Transfer the number of participating nodes can vary.
+A node may have one, zero or any positive integer of particles associated with it.
+Thus in a transfer some nodes may not participate at all or have more load than any of the nodes in its neighborhood.
+\end{enumerate}
+Item \ref{it:part_nodes} does sound more problematic from a parallelization perspective than it is in practice.
+In chapter \ref{sec:rel_work} particle resampling is mentioned of a means to fill material gaps.
+This is an inherent mechanic to control the number of particles within a cell.
+Although it is designed to augment the numerics, the merge method can be easily used to control the upper boundary of particles within a cell.
+Nevertheless the number of particles within a cell is still variable.
+
+The sorting and binning employed in chapter \ref{sec:sort} allows to split the grid into a partitioning of blocks.
+Blocks map well to the thread group nature of GPUs.
+A single thread within a thread group thus corresponds to a grid node within that block.
+\subsubsection{Grid-to-Particle-transfers}\label{sec:g2p}
+The most straightforward solution to this problem is to handle all particles independently from each other.
+One particle adds up the contribution of each grid node variable in range of the interpolation function.
+In three dimensions this can be done for instance with a two times nested for-loop within a thread.
+Luckily grid nodes in a uniform grid are inherently sorted which takes care of coalescing.
+
+Furthermore sorting particles due to chapter \ref{sec:sort} increases caching behavior as variables of grid nodes are directly reused by particles in the same group.
+This results in direct L2-cache hits instead of going the long way of reloading from global memory.
+
+With the mentioned partitioning of the grid into blocks the shared memory architecture of the GPU can be exploited. This requires two-level binning of particles to nodes within blocks.
+
+\algblockdefx{FORALLP}{ENDFAP}[1]%
+  {\textbf{for all }#1 \textbf{do in parallel}}%
+  {\textbf{end for}}
+
+\begin{algorithm}[t]
+  \caption{G2P-Transfer}\label{alg:conj_grad}
+  \begin{algorithmic}[1]
+    \FORALLP{$blockNode \in grid$}
+      \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
+      \ForAll{$node\_id \in \text{assignedNodes}(blockNode, H\_FLAT)$}
+      \State $blockAndHalo[node\_id] \gets \text{LoadGlobalValue}(node\_id)$
+      \EndFor
+      \State \text{barrier()}
+      \ForAll{$particle \in \text{bin}(blockNode)$}
+      \State $sum \gets$ vec4(.0)
+      \ForAll{$neighbor \in \text{support}(blockNode)$}
+      \State $w_{ip} \gets weight(particle,neighbor)$
+      \State $sum \gets sum + blockAndHalo[neighbor]*w_{ip}$
+	\EndFor
+      \State $particle \gets sum$
+      \EndFor
+    \ENDFAP
+  \end{algorithmic}
+\end{algorithm}
+
+\begin{figure}[htbp]
+    \centering
+  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_inv_par}
+  \caption{dwad}
+  \label{fig:g2p_transfer}
+\end{figure}
+
+\subsubsection{Particle-to-Grid-transfers}\label{sec:p2g}
 \begin{figure}[htbp]
@@
 \end{figure}
-\subsubsection{Grid to particle transfers}\label{sec:g2p}
-\begin{wrapfigure}{r}{0.55\textwidth}
-  \includesvg[svgpath =images/,width=0.55\textwidth]{block_transfer_inv_par}
-  \caption{dwad}
-  \label{fig:g2p_transfer}
-\end{figure}
-\subsubsection{Active Blocks /& Maximum Block Count:}\label{sec:active_blocks}
+\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
 \begin{figure}[htbp]
-  \includesvg[svgpath =images/,width=\textwidth]{indirect_dispatch}
+  \incl[svgpath =images/,width=\textwidth]{indirect_dispatch}
   \caption{dwad}

### 2019-01-15 mscthesis.tex
p2g transfer algorithm
g2p transfer

@@
 \usepackage{svg}
+\usepackage{xcolor,colortbl}
 \addbibresource{mscthesis.bib}
@@
 
+The number of transfers per frame is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$).
+This motivates to make the transfers a very highly optimized operation.
+The Material-Point-Method requires for an elastic material as derived:
+\begin{enumerate}
+  \item One P2G-Transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
+  \item Per Conjugate Gradient Iteration:
+    \begin{enumerate}
+      \item One G2P-Transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
+      \item One P2G-Transfer of $\delta \boldsymbol{f}_i$ ($float4$)
+    \end{enumerate}
+  \item Per Newton Iteration: One G2P-Transfer for $\boldsymbol{F}^{n+1}_p$
+  \item One G2P-Transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
+\end{enumerate}
+In summary this is a lot of data to be transferred between grid and particles, which will only increase with more complex models.
+In contrast any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame.
+After that it helps each transfer running more performant.
+Therefore the transfers are the main focus of this implementation.
+
 The MPM-transfers are different from a typical stencil or filter operation in two main factors:
@@
 Nevertheless the number of particles within a cell is still variable.
+Nevertheless all methods presented in the following are not restricted to bins of a fixed size of particles.
 
@@
 \subsubsection{Grid-to-Particle-transfers}\label{sec:g2p}
-The most straightforward solution to this problem is to handle all particles independently from each other.
+The straightforward solution to this problem is to handle all particles independently from each other.
 One particle adds up the contribution of each grid node variable in range of the interpolation function.
@@
 
-With the mentioned partitioning of the grid into blocks the shared memory architecture of the GPU can be exploited. This requires two-level binning of particles to nodes within blocks.
+With the mentioned partitioning of the grid into blocks the shared memory architecture of the GPU can be exploited.
+This requires two-level binning of particles to nodes within blocks.
+Figure \ref{fig:g2p_transfer} show how the G2P-transfer using shared memory is realized.
+Algorithm \ref{alg:g2p} shows this process briefly.
+Due to binning a particle is associated with its lower left node.
+In Figure \ref{fig:g2p_transfer} four threads are active and form the thread group.
+All visible grid nodes need to be loaded by the threads.
+This includes the halo of the block, which is dependent upon the support of the interpolation function.
+For uneven polynomials of the interpolation function the floored position is very favorable.
+It allows the left support to be one smaller than the right support.
+As a result the weight of a most outer-edge particle (of the active particle region) just reaches zero for any neighborhood-node outside the halo.
+
+\begin{figure}[htbp]
+    \centering
+  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_inv_par}
+  \caption{G2P-Transfer: The transfer of a single particle is split into 64 sequential runs (using cubic weights). Shown are the first two runs.}
+  \label{fig:g2p_transfer}
+\end{figure}
+The bigger the block the less halo nodes need to be loaded in total.
+Given an interpolation function with left support $l$ and right support $r$, total grid size $g$ and block sizes $b_x,b_y,b_z$ the general function to calculate the number of elements loaded is:
+\begin{equation}
+  m(b_x,b_y,b_z) = \frac{g}{b_x*b_y*b_z}(l+b_x+r)(l+b_y+r)(l+b_z+r).
+\end{equation}
+ As an example, take a 128*128*128 grid and the cubic interpolation function, which has a left support of 1 and a right support of 2. Subdivide the grid into 4*4*4 or 8*4*4 elements respectively:
+ $$m(4,4,4): 11,239,424\text{ Elements, }m(8,4,4):  8,830,976\text{ Elements}$$
+Note however that in practice there is a good change that a major part of the additional elements will result in L2-cache hits.
+
+Next up is the transfer: Each thread corresponds to one particle within the blue nodes in Figure \ref{fig:g2p_transfer}.
+The threads will sequentially run over the neighbors within the support and collect the weighted contribution to this particle.
+This process needs to be repeated for each particle in the bin.
+Since the number of particles is variable, this loop is dynamic and will lead to branch divergence within a warp.
+One should rely on external mechanics like particle resampling to reduce branching.
 
@@
   {\textbf{end for}}
-
-\begin{algorithm}[t]
-  \caption{G2P-Transfer}\label{alg:conj_grad}
+\begin{algorithm}[htpb]
+  \caption{G2P-Transfer}\label{alg:g2p}
   \begin{algorithmic}[1]
@@
     \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
-      \ForAll{$node\_id \in \text{assignedNodes}(blockNode, H\_FLAT)$}
-      \State $blockAndHalo[node\_id] \gets \text{LoadGlobalValue}(node\_id)$
+    \ForAll{$(node,local\_id) \in \text{assignedNodes}(blockNode, H\_FLAT)$}
+      \State $blockAndHalo[local\_id] \gets node$\Comment{global load}
       \EndFor
@@
       \State $sum \gets$ vec4(.0)
-      \ForAll{$neighbor \in \text{support}(blockNode)$}
-      \State $w_{ip} \gets weight(particle,neighbor)$
-      \State $sum \gets sum + blockAndHalo[neighbor]*w_{ip}$
+      \ForAll{$neighbor\_id \in \text{support}(blockNode)$}\Comment{transfer}
+      \State $w_{ip} \gets weight(particle,neighbor\_id)$
+      \State $sum $ += $ blockAndHalo[neighbor\_id]*w_{ip}$
 	\EndFor
-      \State $particle \gets sum$
+	\State $particle \gets sum$ \Comment{global write}
       \EndFor
@@
 \end{algorithm}
-
-\begin{figure}[htbp]
-    \centering
-  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_inv_par}
-  \caption{dwad}
-  \label{fig:g2p_transfer}
-\end{figure}
-
+Alternatively one could come up with a per-warp load balancing scheme: One thread would then not directly correspond to one active node.
+They would need to act as another node in the same warp, when their load is done.
+This will inevitably lead to shared memory bank conflicts.
+For the P2G-Transfer this would introduce atomic operations on shared memory again.
+
+In Appendices \ref{sec:shared_transfer} the accesses of one warp on shared memory banks is shown.
+The interpolation function is assumed to be cubic and the first block dimension is $x=4$ or $x=8$.
+The shared memory can thus be handled as concurrently as possible.
+
+As an extension to this process batching is introduced.
+Instead of conservatively loading one particle at a time, each thread handles multiple particles together which it stores in register memory.
+Important to this process is to unroll the corresponding code that loads as well as processes these multiple particles.
+Instead of needing to rerun the transfer for each particle, the number of runs now done is the number of non-batched runs divided by the number of particles.
+This process adds to the amount of registers allocated and can thus also negatively impact performance.
 \subsubsection{Particle-to-Grid-transfers}\label{sec:p2g}
+The Particle-to-Grid transfer is
 \begin{figure}[htbp]
@@
 \end{figure}
+
+\begin{algorithm}[htpb]
+  \caption{P2G-Transfer (synchronized)}\label{alg:p2g}
+  \begin{algorithmic}[1]
+    \FORALLP{$blockNode \in grid$}
+    \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
+    \ForAll{$local_id\_id \in \text{assignedNodesID}(blockNode, H\_FLAT)$}
+      \State $blockAndHalo[node\_id] \gets $vec4(.0)
+      \EndFor
+      \State \text{barrier()}
+      \ForAll{$particle \in \text{bin}(blockNode)$} \Comment{global load}
+      \ForAll{$neighbor\_id \in \text{support}(blockNode)$} \Comment{transfer}
+	\State $w_{ip} \gets weight(particle,neighbor\_id)$
+	\State $blockAndHalo[neighbor\_id] $ += $ particle*w_{ip}$
+	\State \text{barrier()} \Comment{uniform flow control needed!}
+	\EndFor
+      \EndFor
+      \ForAll{$(node,local\_id) \in \text{assignedNodes}(blockNode, H\_FLAT)$}
+      \State atomicAdd($node$,$blockAndHalo[local\_id]$) \Comment{global write}
+      \EndFor
+    \ENDFAP
+  \end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
 \begin{figure}[htbp]
-  \incl[svgpath =images/,width=\textwidth]{indirect_dispatch}
+  \includesvg[svgpath =images/,width=\textwidth]{indirect_dispatch}
   \caption{dwad}
@@
   $$
-  T_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
-  $$
-\begin{equation}
-  \Rightarrow T_{bc} = \sum_{r=j}^{(c-1)} a_{br}
+  P_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
+  $$
+\begin{equation}
+  \Rightarrow P_{bc} = \sum_{r=j}^{(c-1)} a_{br}
 \end{equation}
@@
   \begin{equation}
-    \Rightarrow T^2_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
+    \Rightarrow T_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
   \end{equation}
@@
 \end{figure}
+
+\clearpage
+\section{Shared memory accesses on MPM-Transfers}\label{sec:shared_transfer}
+\newcommand{\mc}[2]{\multicolumn{#1}{c}{#2}}
+\definecolor{Gray}{gray}{0.85}
+
+\newcolumntype{a}{>{\columncolor{Gray}}c}
+\newcolumntype{b}{>{\columncolor{white}}c}
+\begin{table}[htpb]
+  \begin{tabular}{ | l | a | a | a | a | l | l | l |}\hline
+    \rowcolor{white}
+    BlockAndHalo(vec4) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
+    Banks (1.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
+    Banks (2.) & 7 & 0 & 1 & 2 & 3 & 4 & 5 \\\hline
+    Banks (3.) & 6 & 7 & 0 & 1 & 2 & 3 & 4 \\\hline
+    Banks (4.) & 5 & 6 & 7 & 0 & 1 & 2 & 3 \\\hline
+    Banks (5.) & 4 & 5 & 6 & 7 & 0 & 1 & 2 \\\hline
+    Banks (6.) & 3 & 4 & 5 & 6 & 7 & 0 & 1 \\\hline
+    Banks (7.) & 2 & 3 & 4 & 5 & 6 & 7 & 0 \\\hline
+    Banks (8.) & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\\hline
+    \rowcolor{white}
+    Banks (9.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
+  \end{tabular}
+  \caption{First run of transfer: Bank accesses by warp with block size $x=4$.
+Bank accesses are colored grey.
+The halo ranges due to the support from $0-7$.
+Every bank gets accessed four times.
+It is easy to see that a window shift would lead to the same amount of bank accesses.}
+\end{table}
+\begin{table}[htpb]
+  \begin{tabular}{ | l | a | a | a | a | a | a | a |a | l | l | l |}\hline
+    \rowcolor{white}
+    BlockAndHalo(vec4) & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\\hline
+    Banks (1.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2  \\\hline
+    Banks (2.) & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5  \\\hline
+    Banks (3.) & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0  \\\hline
+    Banks (4.) & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3  \\\hline
+    \rowcolor{white}
+    Banks (5.) & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\\hline
+  \end{tabular}
+  \caption{First run of transfer: Bank accesses by warp with block size $x=8$.
+Bank accesses are colored grey.
+The halo ranges due to the support from $0-10$.
+Every bank gets accessed four times.
+It is easy to see that a window shift would lead to the same amount of bank accesses.}
+\end{table}
 \end{appendices}

### 2019-01-15 mscthesis.tex
p2g transfer,  pyramid mention

@@
 Instead preferred is batching a fixed amount of particles, more in chapter \ref{sec:transfers}.
-\clearpage
+
+The input size for reduction and scan is limited.
+It's calculable by the $block\text{ }size$ $ \times sequential\text{ }loads$.
+To support bigger input sizes pyramid structures are in use to process the elements the blocks reduced.
+For our case we mostly reduce the input size s.t.
+the first level above in the pyramid can be computed within a single block.
+For the scan this result is then written back.
+This was enough for the input sizes we employed and can be similarly extended.
+
 \subsection{Counting Sort \& Stream Compaction}\label{sec:sort}
@@
 A single thread within a thread group thus corresponds to a grid node within that block.
-\subsubsection{Grid-to-Particle-transfers}\label{sec:g2p}
+\subsubsection{Grid-to-Particle transfers}\label{sec:g2p}
 The straightforward solution to this problem is to handle all particles independently from each other.
@@
 The threads will sequentially run over the neighbors within the support and collect the weighted contribution to this particle.
-This process needs to be repeated for each particle in the bin.
+In this process the particle's position need to be loaded to calculate the weight for its corresponding variables.
+The transfer needs to be repeated for each particle in the bin.
 Since the number of particles is variable, this loop is dynamic and will lead to branch divergence within a warp.
@@
 As an extension to this process batching is introduced.
-Instead of conservatively loading one particle at a time, each thread handles multiple particles together which it stores in register memory.
-Important to this process is to unroll the corresponding code that loads as well as processes these multiple particles.
+Instead of conservatively loading and writing one particle at a time, each thread handles multiple particles together which it stores in register memory.
+Important to this process is to unroll the corresponding loops that load, write as well as process these multiple particles.
 Instead of needing to rerun the transfer for each particle, the number of runs now done is the number of non-batched runs divided by the number of particles.
@@
 \subsubsection{Particle-to-Grid-transfers}\label{sec:p2g}
-The Particle-to-Grid transfer is
+Straightforward to implement are various variations on global memory.
+Consider first binning was note executed.
+A grid node would not know how to look up the particles in its neighborhood.
+The only solution to this is go from the perspective of a particle and write to the grid node in its support.
+Since any particle in the neighborhood of a grid node contribute to its variables all writes have to be atomic.
+The approach the could be simply parallelizing for every particle and atomically write to the neighbor nodes in a loop.
+In the unsorted case this stalls the warps immensely.
+In \cite{Meyer2015} the static loop over the support is also parallelized resulting in higher throughput.
+
+The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-transfers very significantly. The results of above get reversed and looping performs better. Stalling is much less of an issue.
+
+\textbf{P2G-Pull Transfer}: However sorting and binning again opens up the possibility of shared memory implementations.
+The equivalent of the G2P-Transfer for the P2G-Transfer is pulling the relevant data from the neighborhood.
+We will refer to this as the P2G-Pull transfer.
+The process is shown in Figure \ref{fig:p2g_transfer}.
+
 \begin{figure}[htbp]
@@
   \includesvg[svgpath =images/,width=\textwidth]{block_transfer_par}
-  \caption{dwad}
+  \caption{P2G-Transfer: The first run is shown.
+On the left the P2G-Sync transfer is shown that writes into the neighborhood.
+On the right the P2G-Pull approach is shown that instead reads out of the neighborhood.}
   \label{fig:p2g_transfer}
@@
 
+It is important to notice that the supports in the P2G-Pull approach are reversed: Left becomes right, right becomes left.
+Other than that the transfer works in the same way as the G2P-Transfer Instead of reading from grid nodes, reading from particles is done.
+Consequently one particle of each bin is loaded into shared memory.
+This typically puts a lot more strain onto shared memory.
+For a simple PIC-Transfer(equation \ref{eq:lin_p2g})
+$$
+(m\boldsymbol{v})_i^n = \sum_p w_{i}(\boldsymbol{x}_p)m_p\boldsymbol{v}^n_p.
+$$
+the position $\boldsymbol{x}_p$ and the velocity $\boldsymbol{v}_p$ need to be stored in shared memory for every node in the halo.
+Additionally the count $c_i$ and scan $s_i$ need to be available to load and access particles.
+For more complex transfers this will get even worse.
+
+By construction the Material-Point-Method tries to avoid matrices on the grid due to storage cost.
+Conversely the particles have matrices defined over them, e.g.
+$\boldsymbol{B}_p$,$\boldsymbol{F}_p$.
+In conclusion one should motivate approaches where grid nodes correspond to shared memory.
+
+There is one implementation detail left out which is solved in chapter \ref{sec:active_blocks}.
+That is grid nodes do not know if there are any particles in the neighborhood left processing.
+\ref{sec:active_blocks} solves this by finding the maximum particle count in the block (with halo).
+Thus all nodes within the block need to stay active during transfers.
+It is possible to delegate this decision down from the block to its subgroups (respecting their respective haloes).
+
+\textbf{P2G-Atomic Transfer}: One can invert the process of pulling into pushing.
+Instead of reading from particles this means now writing to grid nodes in shared memory.
+This immediately alarms one that accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
+
+As can be seen in Figure \ref{sec:p2g} push approaches have the big advantage of only needing to load particles within the block region.
+The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
+If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out of the transfer and \texttt{atomicAdds()} on shared memory are in use this process is straightforward to implement.
+One has to make sure to reset the shared memory as their contents are otherwise undefined.
+
+The write-back to memory however now also needs to feature atomics.
+This process will be the same in the following synchronized approach of line \ref{alg:atomicAdd} in algorithm \ref{alg:p2g}.
+For a node that will be part of any halo \texttt{atomicAdds()} are a strict requirement.
+Their results are only partially computed and need to be complemented by neighbor blocks.
+For such a node these are two writes for a side node, four for an edge node and eight for a corner node.
+
+\textbf{P2G-Sync Transfer}:
 \begin{algorithm}[htpb]
-  \caption{P2G-Transfer (synchronized)}\label{alg:p2g}
+  \caption{P2G-Sync Transfer}\label{alg:p2g}
   \begin{algorithmic}[1]
@@
     \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
-    \ForAll{$local_id\_id \in \text{assignedNodesID}(blockNode, H\_FLAT)$}
-      \State $blockAndHalo[node\_id] \gets $vec4(.0)
+    \ForAll{$local\_id \in \text{assignedNodesID}(blockNode, H\_FLAT)$}
+      \State $blockAndHalo[local\_id] \gets $vec4(.0)
       \EndFor
@@
 	\State $blockAndHalo[neighbor\_id] $ += $ particle*w_{ip}$
-	\State \text{barrier()} \Comment{uniform flow control needed!}
+	\State \text{barrier()} \Comment{uniform flow control needed!} \label{alg:p2g_uniform_barrier}
 	\EndFor
@@
       \ForAll{$(node,local\_id) \in \text{assignedNodes}(blockNode, H\_FLAT)$}
-      \State atomicAdd($node$,$blockAndHalo[local\_id]$) \Comment{global write}
+      \State atomicAdd($node$,$blockAndHalo[local\_id]$) \Comment{global write}\label{alg:atomicAdd}
       \EndFor

### 2019-01-16 mscthesis.tex
active blocks

@@
 
-\textbf{Particle-Grid transfers:} The particle to grid transfer is generally done in two manners \cite{stantchev2008fast}:
+\textbf{Particle-to-Grid=Transfers:} The particle to grid transfer is generally done in two manners \cite{stantchev2008fast}:
 \begin{enumerate}
   \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it.
-Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary.
+Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g.
+a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
   \item \textbf{Particle Pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
 \end{enumerate}
-A grid to particle transfer is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
+A Grid-to-Particle-Transfer is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
 The grid points in a uniform grid are 'sorted' by nature.
@@
 This process adds to the amount of registers allocated and can thus also negatively impact performance.
-\subsubsection{Particle-to-Grid-transfers}\label{sec:p2g}
+\subsubsection{Particle-to-Grid-Transfers}\label{sec:p2g}
 Straightforward to implement are various variations on global memory.
@@
 
-The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-transfers very significantly. The results of above get reversed and looping performs better. Stalling is much less of an issue.
-
-\textbf{P2G-Pull Transfer}: However sorting and binning again opens up the possibility of shared memory implementations.
-The equivalent of the G2P-Transfer for the P2G-Transfer is pulling the relevant data from the neighborhood.
-We will refer to this as the P2G-Pull transfer.
-The process is shown in Figure \ref{fig:p2g_transfer}.
-
-\begin{figure}[htbp]
+The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-Transfers very significantly. The results of above get reversed and looping performs better. Stalling is much less of an issue.
+
+\begin{figure}[t]
     \centering
@@
   \caption{P2G-Transfer: The first run is shown.
-On the left the P2G-Sync transfer is shown that writes into the neighborhood.
+On the left the P2G-Sync-Transfer is shown that writes into the neighborhood.
 On the right the P2G-Pull approach is shown that instead reads out of the neighborhood.}
@@
 \end{figure}
+
+\textbf{P2G-Pull-Transfer}: However sorting and binning again opens up the possibility of shared memory implementations.
+The equivalent of the G2P-Transfer for the P2G-Transfer is pulling the relevant data from the neighborhood.
+We will refer to this as the P2G-Pull-Transfer.
+The process is shown in Figure \ref{fig:p2g_transfer}.
 
@@
 It is possible to delegate this decision down from the block to its subgroups (respecting their respective haloes).
-
-\textbf{P2G-Atomic Transfer}: One can invert the process of pulling into pushing.
-Instead of reading from particles this means now writing to grid nodes in shared memory.
-This immediately alarms one that accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
-
-As can be seen in Figure \ref{sec:p2g} push approaches have the big advantage of only needing to load particles within the block region.
-The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
-If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out of the transfer and \texttt{atomicAdds()} on shared memory are in use this process is straightforward to implement.
-One has to make sure to reset the shared memory as their contents are otherwise undefined.
-
-The write-back to memory however now also needs to feature atomics.
-This process will be the same in the following synchronized approach of line \ref{alg:atomicAdd} in algorithm \ref{alg:p2g}.
-For a node that will be part of any halo \texttt{atomicAdds()} are a strict requirement.
-Their results are only partially computed and need to be complemented by neighbor blocks.
-For such a node these are two writes for a side node, four for an edge node and eight for a corner node.
-
-\textbf{P2G-Sync Transfer}:
-\begin{algorithm}[htpb]
-  \caption{P2G-Sync Transfer}\label{alg:p2g}
+Delegating this from the subgroup to the threads is possible as well.
+Each thread would calculate the summed up count in its neighborhood.
+But this will hardly matter as all other threads need to wait on the thread with the highest count in the subgroup anyway.
+
+\begin{algorithm}[t]
+  \caption{P2G-Sync-Transfer}\label{alg:p2g}
   \begin{algorithmic}[1]
@@
       \State \text{barrier()}
-      \ForAll{$particle \in \text{bin}(blockNode)$} \Comment{global load}
+      \ForAll{$particle \in \text{bin}(blockNode)$} \Comment{global load} \label{alg:p2g_dyn_loop}
       \ForAll{$neighbor\_id \in \text{support}(blockNode)$} \Comment{transfer}
@@
 
-\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
-\begin{figure}[htbp]
+\textbf{P2G-Atomic-Transfer}: One can invert the process of pulling into pushing.
+Instead of reading from particles this means now writing to grid nodes in shared memory.
+This immediately alarms one that accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
+
+As can be seen in Figure \ref{sec:p2g} push approaches have the big advantage of only needing to load particles within the block region.
+The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
+If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out of the transfer and \texttt{atomicAdds()} on shared memory are in use this process is straightforward to implement.
+One has to make sure to reset the shared memory as their contents are otherwise undefined.
+
+The write-back to memory however now also needs to feature atomics.
+This process will be the same in the following synchronized approach of line \ref{alg:atomicAdd} in algorithm \ref{alg:p2g}.
+For a node that will be part of any halo \texttt{atomicAdds()} are a strict requirement.
+Their results are only partially computed and need to be complemented by neighbor blocks.
+For such a node these are two writes for a side node, four for an edge node and eight for a corner node.
+
+\textbf{P2G-Sync-Transfer}: Warps/Subgroups are inherently synchronized.
+No race conditions are present between their threads as they all access different memory positions (as can be seen in Figure \ref{fig:p2g_transfer}).
+The problem arises between different warps of the same thread group.
+In a synchronized approach this is solved by a \texttt{barrier()}.
+As already mentioned the particle loop of line \ref{alg:p2g_dyn_loop} in algorithm \ref{alg:p2g} is dynamic.
+The OpenGL language makes the following requirements for barriers:
+
+"\texttt{barrier()} can be called from flow-control, but it can only be called from uniform flow control.
+(...) In short, (...) every execution must hit the exact same set of \texttt{barrier()} calls in the exact same order." \cite{KHRONOS:BARRIER}
+
+The dynamic loop has to become static within a group.
+The problem is solved in same fashion as in the P2G-Pull method with chapter \ref{sec:active_blocks}.
+Calculate the maximum particle count within the block without the halo.
+This is the minimum amount of times the loop needs to run across the threads to hit uniform flow control.
+
+Warps are still accelerated if all threads within that warp have lower count. They still however have to commit to wait at the same \texttt{barrier()}.
+
+In summary all transfers have the same optimal shared memory-access patterns as shown already by Appendices \ref{sec:shared_transfer}.
+Batching is again a valid option that is executed in the same manner by unrolling the corresponding loops.
+The block size as well as the level of batching can be chosen arbitrarily to optimize for the workload.
+
+\begin{figure}[!b]
   \includesvg[svgpath =images/,width=\textwidth]{indirect_dispatch}
-  \caption{dwad}
+  \caption{Pipeline to filter active blocks.
+Returned is the number of active blocks for the launch of a dispatch.
+A scan and stream compaction gives back the filtered indices to access the active blocks in that dispatch.}
   \label{fig:active_block}
@@
 
-\subsection{Verifying Simulation}
+\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
+The P2G-Pull- and P2G-Sync-Transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region.
+The active particle region is the region that needs to get reduced.
+Therefore it is sufficient for the P2G-Sync-Transfer to reduce the counter within the block, left of Figure \ref{fig:p2g_transfer}.
+The P2G-Pull-Transfer however also need to take the halo into account, right of Figure \ref{fig:p2g_transfer}.
+Both are achievable with a single level of parallel reduction (in contrast to a pyramid).
+We will call this the per-block count from now on, that uses as the functional input for the parallel reduction:
+\begin{enumerate}
+  \item \texttt{unary\_op(count) = count} .
+  \item \texttt{binary\_op(lc,rc) = max(lc,rc)}.
+\end{enumerate}
+
+For a P2G-Sync-Transfer one may just start with a parallelization level equal to the blocks employed in \ref{sec:transfers}.
+Then can narrow the shader down to do more sequential work to improve throughput.
+For the P2G-Pull-Transfer one may also do the first, but additionally load all rest elements (by the halo).
+From there on out more sequential work might be favorable.
+
+If a per block indexing mode is employed all blocks are directly behind each other in the buffer.
+Reduction then happens in the typical fashion.
+A global indexing needs to be permuted such that the indexing puts blocks first.
+
+The process of filtering active blocks can be easily appended.
+Figure \ref{fig:active_block} shows the process.
+An active block is one that has a non-zero per-block counter.
+The scan that will refer to the active block indices takes thus functional input:
+\begin{enumerate}
+  \item \texttt{unary\_op(count) = (count>0)?1:0}, which uses the ternary operator that increases count by one for any active block.
+  \item \texttt{binary\_op(lc,rc) = lc + rc}.
+\end{enumerate}
+
+Starting a dispatch for every block only the active blocks will now write their index at the scan's position.
+This process is not coalesced.
+Coalescing will still be guarenteed for all nodes within the block for the P2G-Transfers and G2P-Transfer.
+The only non-coalesced access within these shaders is the maximum per block count, which there are really few of in comparison.
+
+\section{Evaluation}\label{sec:eval}
+\subsection{Verifying Results}
 \begin{itemize}
@@
   \item Singular Value Decomposition
-  \item Conjugate gradient
 \end{itemize}
 
-\clearpage
-\section{Evaluation}\label{sec:eval}
 \textbf{SoA vs. AoS}:

### 2019-01-18 mscthesis.tex
performance tables

@@
   \begin{itemize}
-  \item \texttt{unary\_op(x) := length2(x)} and
+  \item \texttt{unary\_op(x) := length(x)} and
   \item \texttt{binary\_op(x,y) := max(x,y)}.
@@
 Coalescing will still be guarenteed for all nodes within the block for the P2G-Transfers and G2P-Transfer.
-The only non-coalesced access within these shaders is the maximum per block count, which there are really few of in comparison.
+The only non-coalesced access within these shaders is loading the maximum per block count from VRAM, which there are really few of in comparison.
 
@@
 \subsection{Verifying Results}
+\textbf{Parallel reduction}, \textbf{map} and \textbf{scan} are tested against \texttt{std::transform\_reduce} as it supplies the same functionality as a fully reduced parallel reduction.
+It is standard since C++17.
+\begin{table}[h]
+  \begin{tabular}{|l|r|r|r|}\hline
+    Shader       & \texttt{unary\_op} & Total abs. error & Total rel. error \\\hline
+    Map (float)   & x          & 0.0    & 0.0      \\\hline
+    Map (vec4,dvec4)   & length(x)  &0.0625 & 7.94e-08 \\\hline\hline
+    MapReduce (float)& x  &0.0183 & 3.51e-08 \\\hline
+    MapReduce (double)& x  &1.78e-08 & 3.42e-14 \\\hline
+    MapReduce (vec4)& length(x)  &212.813 & 2.71e-04 \\\hline
+    MapReduce (dvec4)& length(x)  &1.58e-05 & 2.02e-11 \\\hline
+  \end{tabular}
+  \caption{Error of various shaders on 1024*1024 random elements. Scalars are just between [0.0-1.0]. The vectors have length between [0.0-1.0].}
+\end{table}
+
 \begin{itemize}
@@
 \end{itemize}
-
+ \subsubsection{Performance}
 \textbf{SoA vs. AoS}:
-
+dvec4 same
 \begin{table}
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Description &  $\Delta t_c (\mu s)$ & Speedup &VRAM     & SM     & L2     & Tex \\\hline
+    Layout &  $\Delta t_c (\mu s)$ & Speedup &VRAM     & SM     & L2     & Tex \\\hline
     AoS         & 278                   & -       &77.7\%   & 7.3\%  & 30.3\% & 14.7\% \\\hline
@@
   \end{tabular}
-  \caption{Map operation with one instruction of form \texttt{float x = length2( (vec4) y)} on $1024 \times 1024$ \texttt{vec4}.}\label{table:map_1}
+\caption{Timings betweens different buffer layouts.
+Map operation with one instruction \texttt{float x = length(vel)} on $1024 \times 1024$ \texttt{vec4} buffer storing position and velocity (\texttt{vel})}.}\label{table:map_1}
 \end{table}
@@
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Description &  $\Delta t_c (\mu s)$ & Speedup &VRAM   & SM     & L2     & Tex \\\hline
-    AoS         & 292                   & -       &61.3\% & 11.4\% & 41.8\% & 20.4\% \\\hline
-    SoA         & 252                   & 1.16x   &75.4\% & 14.3\% & 29.4\% & 17.9\% \\
+    Layout &  $\Delta t_c (\mu s)$ & Speedup       &VRAM             & L2              & L2-Hit & Read Active \\\hline
+    AoS         & 292                   & -        &61.3\%           & \textbf{41.8}\% & 53.8\% & 48.9\% \\\hline
+    SoA         & 252                   & 1.16x    &\textbf{75.4}\%  & 29.4\%          & 20.0\% & 62.3\% \\
     \hline
   \end{tabular}
-  \begin{tabular}{ | l | l | l | l | }    \hline
-    Description &  L2-Hit & Read Active \\\hline
-    AoS   & 53.8\%        & 48.9\%      \\\hline
-    SoA   & 20.0\%          & 62.3\%      \\
+\caption{Timings betweens different buffer layouts.
+Map operation with instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel})}.}\label{table:map_1}
+\end{table}
+
+\begin{enumerate}
+Occupancy!
+\end{enumerate}
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
+    Method         &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM              & Top Warp-Stall Reas.\\\hline
+    Interleaved    & 305                   & -       &23.0\%            & \textbf{60.9}\% & Short Scoreb.(17.2\%)\\\hline
+    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%          & Long Scoreb.(17.3\%) \\\hline
+    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%          & Long Scoreb.(80.1\%)\\\hline
+    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%          & Long Scoreb.(84.4\%)\\\hline
+    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%          & Long Scoreb.(76.9\%)\\
     \hline
   \end{tabular}
-  \caption{Map operation with two instructions of form \texttt{float x = length2( (vec4) y)} on $1024 \times 1024 \times 2$ \texttt{vec4}.}\label{table:map_2}
+  \caption{Optimization of one parallel reduction dispatch on $1024\times1024$ instruction \texttt{float x= length(vec4 y)}.
+Note that the interleaved addressing intentionally does bank conflicts to show performance impacts.
+The number in parentheses states the number of sequential elements each thread loads.}\label{table:map_reduce}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
+    Method              &  $\Delta t_c$         & Speedup &VRAM              & SM      & L2     & Top Stall Reas.\\\hline
+    Conflicts           & 748                   & -       &17.5\%            & 51.7\%  & 8.2\%  & Barrier(49.5\%)\\\hline
+    No confl.           & 571                   & 1.31x   &25.7\%            & 50.1\%  & 5.4\%  & Barrier(31.9\%)\\\hline
+    part. confl.        & 546                   & 1.36x   &23.5\%            & 50.8\%  & 6.0\%  & Barrier(30.5\%) \\\hline\hline
+    Seq. (2x)           & 323                   & 2.31x   &41.7\%            & 43.0\%  & 17.2\% & S. Scoreb.(32.0\%)\\\hline
+    unrolled            & 258                   & 2.90x   &53.8\%            & 42.2\%  & 22.0\% & S. Scoreb.(32.0\%)\\\hline\hline
+    Seq. (4x)           & 311                   & 2.40x   &56.8\%            & 22.7\%  & 37.9\% & L. Scoreb.(43.3\%)\\\hline
+    unrolled            & 297                   & 2.52x   &61.1\%            & 21.2\%  & 40.6\% & L. Scoreb.(49.9\%)\\
+    \hline
+  \end{tabular}
+  \caption{Optimization of one scan dispatch on $4 \times 1024\times1024$ integers with \texttt{binary\_op(x,y) := x+y}. $\Delta t_c$ in $\mu s$. }\label{table:scan}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
+    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
+    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
+    Deep sort           &    218                & 6.95x   &75.3\%            & 24.4\%  & 35.0\% & 37.8\% \\
+    \hline
+ \end{tabular}
+\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
+    Best Case(*)          &  $\Delta t_c$         &VRAM  & Top Stall Reas. & Read & Write\\\hline
+    Index                 &  1,516                &25.6\%& L. Scoreb.(78.7\%) &&\\\hline
+    Full(8 vec4)     &  1,731                &75.3\%& L. Scoreb.(92.2\%) &&\\
+    \hline
+    \hline
+    Worst Case(*)          &  $\Delta t_c$         &VRAM              & Top Stall Reas.    &Read & Write\\\hline
+    Index                  &  1,516                &26.0\%            & L. Scoreb.(73.5\%) & &\\\hline
+    Deep Sort(8 vec4)      &  9,847                &24.0\%            & L. Scoreb.(82.2\%) & &\\
+    \hline
+ \end{tabular}
+\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
 \end{table}

### 2019-01-19 mscthesis.tex
performance

@@
 It is standard since C++17.
-\begin{table}[h]
+These shaders are exact on integers.
+Floating point precision can be seen in table \ref{tab:parallel_red}.
+\begin{table}[!b]
   \begin{tabular}{|l|r|r|r|}\hline
@@
   \caption{Error of various shaders on 1024*1024 random elements. Scalars are just between [0.0-1.0]. The vectors have length between [0.0-1.0].}
+  \label{tab:parallel_red}
 \end{table}
 
-\begin{itemize}
-  \item Conservation equations
-    \cite{MPM:APIC}
-  \item Singular Value Decomposition
-\end{itemize}
- \subsubsection{Performance}
+\textbf{Counting sort} is tested as a whole by giving back the bin keys of the particles. The bin keys are then tested on the CPU against \texttt{std::is\_sorted}, which is standard since C++11.
+
+The Eigen-library \cite{EIGEN:LIBRARY} is a well regarded linear algebra library.
+The \textbf{Singular value decomposition} is tested against \texttt{Eigen::JacobiSVD}.
+As a test one million random matrices are created by Eigen.
+Tests account for the 'Polar SVD'.
+There is not a single sign error.
+The average error between the singular values of Eigen and the implementation of chapter \ref{sec:svd} is 4.91782e-07.
+$det(\boldsymbol{U})$ and $det(\boldsymbol{V})$ should be as close as possible to 1.
+The errors are 3.55707e-07 and 3.26429e-07, respectively.
+
+The \textbf{MPM-Transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
+\begin{equation}
+  m_{p}^{tot}= \sum_p m_p =  \sum_i m_i = m_{g}^{tot}.
+\end{equation}
+Following this procedure, the same should be true for the total linear momentum $\boldsymbol{p}^{tot}$:
+\begin{equation}
+  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}.
+\end{equation}
+And more specifically for the case of APIC also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
+\begin{equation}
+  \boldsymbol{L}_{p}^{tot}= \sum_p \boldsymbol{x}_p \times m_p \boldsymbol{v}_p =  \sum_i \boldsymbol{x}_i \times m_i \boldsymbol{v}_i = \boldsymbol{L}_{g}^{tot}.
+\end{equation}
+An extreme test case spreads one million particles randomly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}.
+A roundtrip consists of a P2G-Transfer and a G2P-Transfer for relevant variables.
+In this process the (random) velocities will get filtered.
+In PIC-Transfers this regionally averages out velocities.
+Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.
+
+As a side note: Random particle positions as well as random velocities are a configuration that in practice will never occur. Even for a fundamental state of matter of gas this is unrealistic.
+
+The total errors vary little between the methods presented in chapter \ref{sec:transfers}.
+They are caused by floating-point inaccuracy.
+The order of execution on the GPU can vary.
+Thus the floating-point error will vary from execution to execution.
+\begin{table}[t]
+\centering
+  \begin{tabular}{|l|r|r|}\hline
+    Quantity& Abs. error  & Rel. error \\ \hline
+    $m^{tot}_i$ & 212.438     & 2.12438e-04     \\\hline
+    $\boldsymbol{p}^{tot}_i$ & (0.578,0.787,0.778)&(1.076,1.290,1.204)e-04 \\\hline
+   \end{tabular}
+   \caption{Error of 1,000 PIC-roundtrips of 1M randomly placed particles with random velocity in a $128\times128\times128$ grid.
+Used this instance is a P2G-Sync-Transfer and G2P-Transfer with batching = 4 and block size = (4,4,4).
+Particles are fixed.}
+  \label{tab:roundtrip}
+\end{table}
+
+\subsection{Performance}
 \textbf{SoA vs. AoS}:
 dvec4 same
-\begin{table}
+\begin{table}[]
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline

### 2019-01-20 mscthesis.tex
more tables, setting up eval. chapter

@@
 \newenvironment{code}{\captionsetup{type=algorithm}}{}
+ \caption{P2G-Transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+Block size is (8,4,4).
+Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_random}
 \begin{document}
@@
     SM Active\%       & 100                   & at least one warp active avg. over SMs\\\hline
+    SM Issue Util.\%  & 100                   & amount of cycles an instr. was issued\\\hline
     SM Occupancy      & 64.0                  & active warps avg. over SMs\\\hline
@@
     \multicolumn{3}{|l|}{\textbf{Warp Stall Reasons:}}\\\hline
-    Long Scoreboard\% & 100                   & memory latency\\\hline
+    Long Scoreboard\% & 100                   & VRAM latency (mostly reads)\\\hline
+    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\
+		      &                       & (mostly (atomic) writes)\\\hline
     Barrier\%         & 100                   & block synchronization \\\hline
-    %No Instructions\% & 100                   & \\
-    Drain\%           & 100                   & memory wait before warp exit \\
-    \hline
+    No Instruction\%  & 100                   & instruction fetching or\\
+		      &                       & waiting on instruction cache miss\\\hline
   \end{tabular}
@@
 
-Aiming to get shaders memory-bandwidth limited while minimizing the amount of needed accesses for any given workload is the top-priority for our shaders.
+Aiming to get shaders memory-bandwidth limited while minimizing the amount of needed accesses for any given workload is the top-priority for simple tasks.
 Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% < top SOL\% < 80\%.
 The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.
+
+The metric SM Issue Utilization (per active cycle) is important for compute-bound work as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data.
+This metric tells if the SM can issue instructions fast.
+NVIDIA also aims between 60\% to 80\% for this metric if the shader is compute-bound.
+Furthermore for compute-bound work often the top warp stall reason is waiting on instruction fetches.
+
 \clearpage
@@
   \includesvg[svgpath =images/,width=\textwidth]{soa}
-  \caption{SoA-Layout in practive. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
+  \caption{SoA-Layout in practice. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
 \label{fig:soa}
@@
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Layout &  $\Delta t_c (\mu s)$ & Speedup &VRAM     & SM     & L2     & Tex \\\hline
-    AoS         & 278                   & -       &77.7\%   & 7.3\%  & 30.3\% & 14.7\% \\\hline
-    SoA         & 122                   & 2.26x   &75.4\%   & 14.3\% & 29.4\% & 17.9\% \\
+    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util.\% \\\hline
+    AoS         & 278                   & -       &77.7\%   & 7.3\%  & 30.3\% & 14.0\% \\\hline
+    SoA         & 122                   & 2.26x   &75.4\%   & 14.3\% & 29.4\% & 6.8\% \\
     \hline
   \end{tabular}
-\caption{Timings betweens different buffer layouts.
+\caption{Timings between different buffer layouts.
 Map operation with one instruction \texttt{float x = length(vel)} on $1024 \times 1024$ \texttt{vec4} buffer storing position and velocity (\texttt{vel})}.}\label{table:map_1}
@@
   \end{tabular}
-\caption{Timings betweens different buffer layouts.
-Map operation with instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel})}.}\label{table:map_1}
+\caption{Timings between different buffer layouts.
+Map operation with two instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel})}.}\label{table:map_1}
 \end{table}
@@
 \begin{table}
-  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
-    Best Case(*)          &  $\Delta t_c$         &VRAM  & Top Stall Reas. & Read & Write\\\hline
-    Index                 &  1,516                &25.6\%& L. Scoreb.(78.7\%) &&\\\hline
-    Full(8 vec4)     &  1,731                &75.3\%& L. Scoreb.(92.2\%) &&\\
+  \begin{tabular}{ | l | r | r | r | r | r| }    \hline
+    Best Case    &  $\Delta t_c (\mu s)$&VRAM  & Top Stall Reas.    & Read   & Write  \\\hline
+    Index        &  1,516       &25.6\%& L. Scoreb.(78.7\%) & 16.30\%& 9.30\% \\\hline
+    Full(8 vec4) &  1,731       &75.3\%& L. Scoreb.(92.2\%) & 38.60\%& 34.60\%\\
     \hline
     \hline
-    Worst Case(*)          &  $\Delta t_c$         &VRAM              & Top Stall Reas.    &Read & Write\\\hline
-    Index                  &  1,516                &26.0\%            & L. Scoreb.(73.5\%) & &\\\hline
-    Deep Sort(8 vec4)      &  9,847                &24.0\%            & L. Scoreb.(82.2\%) & &\\
+    Worst Case             &  $\Delta t_c(\mu s)$         &VRAM              & Top Stall Reas.    & Read   & Write\\\hline
+    Index                  &  1,516                &26.0\%            & L. Scoreb.(73.5\%) & 16.10\%& 9.40\%\\\hline
+    Full(8 vec4)           &  9,847                &24.0\%            & L. Scoreb.(82.2\%) & 9.90\% & 14.0\%\\
     \hline
  \end{tabular}
-\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
+\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.
+Best case is already sorted, worst case is unsorted and still fully random.}\label{table:bin}
 \end{table}
 
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |r|r|r|}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2    &SM     \\\hline
+    G2P global             &   38,318             &-               &24.3\%&34.4\% & 9.6\% \\\hline
+    sorted                 &    2,991		  &12.81x          & 7.5\%&52.3\% &67.1\% \\\hline
+    G2P-Shared             &    9,665             & 3.96x          & 3.4\%& 2.5\% &65.0\% \\\hline
+ \end{tabular}
+ \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
+Block size is (8,4,4).}\label{table:p2g_uniform4}
+\end{table}
+
+VRAM-L.Scoreboard 79.9\%
+Wait No Instructions
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
+    G2P global             &   10,316             &-               &2.3\%&34.4\%&32.4\% \\\hline
+    sorted                 &    2,629		  &3.92x           &4.1\%&9.4\%&66.9\%  \\\hline
+    G2P-Shared             &    2,232             &4.62x           &7.3\%&4.1\% &63.6\% \\\hline
+ \end{tabular}
+
+ \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).
+Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
+\end{table}
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.\\\hline
+    P2G-Atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)\\\hline
+    indirect		   &    6,276             & 1.10x   &90.8\%&S.Scoreb.(31.7\%)\\\hline
+    batching = 4           &    4,929             & 1.41x   &65.1\%&S. Scoreb.(24.3\%)\\\hline
+    both                   &    3,148             & 2.20x   &65.0\%&No Instr.(23.2\%)\\\hline
+    \hline
+    P2G-Sync               &    3,161             & -       &72.4\%&No Instr.(17.7\%) \\\hline
+    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) \\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).}\label{table:p2g_optim}
+\end{table}
+
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
+    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
+    \cite{Meyer2015}       &   45,342             &0.98x           &25.1\%&42.4\%&11.7\% \\\hline
+    \cite{Meyer2015} sorted&   23,007		  &1.97x           &43.8\%&59.0\%&23.9\% \\\hline
+    global sorted          &   20,484		  &2.21x           &7.0 \%&44.0\%&16.1\% \\\hline
+    \hline
+    P2G-Pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
+    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &65.0\%\\\hline
+    P2G-Sync*              &    2,595             &17.47x          &5.9\%&7.6\%  &67.0\%\\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).
+Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
+\end{table}
+Barrier 26.3\% on P2G-Sync
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
+    global                 &  254,410             &-               &17.8\%&9.6\% &1.7\%  \\\hline
+    \cite{Meyer2015}       &  217,997             &1.17x           &21.7\%&3.7\% &13.2\% \\\hline
+    \cite{Meyer2015} sorted&   30,477		  &8.35x           &64.2\%&64.2\%&22.1\% \\\hline
+    global sorted          &   21,413		  &11.88x          &7.2\% &65.8\%&19.4\% \\\hline
+    \hline
+    P2G-Pull               &   26,822             &9.46x           &1.6\% &2.3\% &35.2\%\\\hline
+    P2G-Atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &65.4\%\\\hline
+    P2G-Sync*              &   12,801             &19.87x          &3.6\% &7.3\% &61.3\%\\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+Block size is (8,4,4).
+Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_random}
+\end{table}
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r |}    \hline
+    Method                 &  Area=47.6\% & Area=23.8\% &Area=11.9\%\\\hline
+    P2G-Sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
+    batching = 2           & -&2,942$\mu s$ & 2,745$\mu s$\\\hline
+    batching = 4           & -&- &2,622$\mu s$\\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered.
+Block size is (4,4,4).}\label{table:p2g_optim}
+\end{table}
+
 \section{Conclusion}
-\subsection{Review}
+\subsection{Summary}
 \subsection{Future Work}

### 2019-01-21 mscthesis.tex
eval. soa,aos;scan,reduce

@@
 Therefore algorithms often improve much further from more sequential work to avoid this indirection.
-\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
-\subsubsection{GPU Metrics}
+
+\textbf{Loop unrolling:} Foremost loop unrolling reduces instructions and can thus speed up compute-bound work.
+Moreover loop unrolling might help the compiler reorder instructions and thus execute code faster.
+Instructions benefiting greatly from this are global loads.
+\cite{NVIDIA:PEAK_PERFORMANCE} shows at the example of motion blur and ray-marching that this greatly can increase throughput.
+Whereas the motion blur example uses a fixed amount of samples and therefore is easily unrollable, ray-marching is a dynamic process.
+Dynamic loops can however be partially unrolled.
+In the case of ray-marching this would for instance fetch two samples at once, although it might process only one.
+In short, this merges two consecutive loop iterations.
+The latency of the second instruction might be partly hidden by the first one and thus speed up the work load.
+Mind however that loop unrolling can increase register pressure.
+More registers need to store and process the relevant code, where a dynamic loop reuses the same registers for any iteration.
+
+\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
+\subsubsection{GPU Metrics}\label{sec:metrics}
 The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
@@
 At the end of testing data is gathered and statistical quantities are written into a log file.
-\subsection{GPU memory layout}
+\subsection{GPU memory layout}\label{sec:soa_aos}
 \noindent\begin{minipage}{0.45\textwidth}
@@
 \subsection{Performance}
-\textbf{SoA vs. AoS}:
-dvec4 same
-\begin{table}[]
+The given metrics in \ref{sec:metrics} lead to valuable insight into how well the GPU can utilize a given shader or program.
+The GPU was a black box for a long time in this regard.
+The metrics provided by NVIDIA are a first step to open up the GPUs inner workings.
+Theory about optimization for the architecture such as \cite{NVIDIA:BEST:PRACTICE} can be more readily applied.
+However for abstraction purposes theory as well as descriptions about the architecture might be deliberately vague.
+These metrics give the programmer more power of identifying the right optimizations for a given code.
+All metrics in the following are gathered from a NVIDIA GTX970.
+
+\textbf{SoA vs.
+AoS}: In chapter \ref{sec:soa_aos} was already hinted that throughput could be halved in an AoS-Layout.
+These effects can be hideous without knowledge of the architecture.
+Take for instance a map operation (or equivalently a copy operation).
+Figure \ref{fig:soa} and \ref{fig:aos} in chapter \ref{sec:soa_aos} already gave an overview of the procedure for \texttt{vec4}.
+Table \ref{tab:map_1} compares differences between an AoS- and an SoA-Layout with a buffer of two attributes.
+\begin{table}[htbp]
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util.\% \\\hline
-    AoS         & 278                   & -       &77.7\%   & 7.3\%  & 30.3\% & 14.0\% \\\hline
-    SoA         & 122                   & 2.26x   &75.4\%   & 14.3\% & 29.4\% & 6.8\% \\
+    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
+    AoS         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
+    SoA         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\
     \hline
   \end{tabular}
-\caption{Timings between different buffer layouts.
-Map operation with one instruction \texttt{float x = length(vel)} on $1024 \times 1024$ \texttt{vec4} buffer storing position and velocity (\texttt{vel})}.}\label{table:map_1}
+\caption{Metrics between different buffer layouts.
+Map operation with one instruction \texttt{float x = length(vel)} on $1024 \times 1024$ \texttt{vec4} buffer storing position and velocity (\texttt{vel}).}
+\label{tab:map_1}
 \end{table}
 
-\begin{table}
+Immediately apparent becomes that the SoA-Layout is 2.26 times as fast.
+However both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout.
+Though the big difference is in the memory access, the wrong access may be conversely spotted in the non-memory subunits.
+Since an AoS-Layout loads twice as many \texttt{vec4}, every non-memory related metric shrinks by half.
+An AoS-Layout however issues the same instructions.
+Though the SM should do equal work and be active for the same amount of cycles which is not the case.
+\begin{table}[htpb]
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
     Layout &  $\Delta t_c (\mu s)$ & Speedup       &VRAM             & L2              & L2-Hit & Read Active \\\hline
-    AoS         & 292                   & -        &61.3\%           & \textbf{41.8}\% & 53.8\% & 48.9\% \\\hline
-    SoA         & 252                   & 1.16x    &\textbf{75.4}\%  & 29.4\%          & 20.0\% & 62.3\% \\
+    AoS         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
+    SoA         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
     \hline
   \end{tabular}
-\caption{Timings between different buffer layouts.
-Map operation with two instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel})}.}\label{table:map_1}
+\caption{Metrics between different buffer layouts.
+Map operation with two instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
+\label{tab:map_2}
 \end{table}
 
-\begin{enumerate}
-Occupancy!
-\end{enumerate}
-\begin{table}
+The case of using both attributes is also visualized in Figure \ref{fig:soa} and \ref{fig:aos}.
+One map instruction for each attribute is done within the same shader.
+Table \ref{tab:map_2} shows the results.
+Now AoS and SoA load the same amount of elements from VRAM.
+The SoA-Layout is still faster by 35 $\mu s$.
+
+In the SoA-Layout the second attribute lays consecutively in memory and is directly fetched.
+The SoA timing from table \ref{tab:map_1} is doubled.
+As expected the two operations are independent of each other and could thus be separated into two shaders with minimal performance impact; Assuming invocation cost is minimal.
+
+In the AoS-Layout the second attribute lays in the L2-Cache from fetching the first attribute.
+Thus using the second attribute will yield to cache hits.
+But going the extra way of checking for the element in the L2-cache requires an additional look up.
+Thus in table \ref{tab:map_2} L2-SOL and L2 Hit rate go up.
+
+The view given was restricted to \texttt{vec4} as its the most commonly occurring type.
+For float/uint data structures the impacts of AoS vs.
+SoA can be even more dramatic.
+Instead of two \texttt{vec4}s eight scalars fit into a cache-line.
+In contrast only one \texttt{dvec4} (double-precision) fits into a cache-line.
+Performance of AoS and SoA in this case is equal.
+
+As a closure the reader is again referred to the subject of data-oriented design.
+While for non-random access SoA is (as shown) the more efficient layout.
+Random access algorithms can improve performance by using an AoS or mixed setup.
+In a random access pattern the nearby elements of an SoA-Layout will be left unused in the cache-line.
+
+\textbf{Parallel Reduction \& Scan:} Chapter \ref{sec:red_scan} presents a shared memory implementation of parallel reduction.
+To further speed up the work load a sequential pattern is preferred to avoid bank conflicts.
+Table \ref{tab:map_reduce} compares this to an interleaved addressing without padding to illustrate the effects of bank conflicts on performance.
+Similarly the same can be done for a scan, where however the interleaved addressing is preferred anyway.
+Table \ref{tab:scan} shows the comparison for a scan.
+
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
-    Method         &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM              & Top Warp-Stall Reas.\\\hline
-    Interleaved    & 305                   & -       &23.0\%            & \textbf{60.9}\% & Short Scoreb.(17.2\%)\\\hline
-    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%          & Long Scoreb.(17.3\%) \\\hline
-    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%          & Long Scoreb.(80.1\%)\\\hline
-    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%          & Long Scoreb.(84.4\%)\\\hline
-    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%          & Long Scoreb.(76.9\%)\\
+    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
+    Interleaved    & 305                   & -       &23.0\%          & \textbf{60.9}\%  & S. Scoreb.(17.2\%)\\\hline
+    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0}\%) \\\hline
+    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%         & L. Scoreb.(80.1\%)\\\hline
+    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%         & L. Scoreb.(84.4\%)\\\hline
+    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%         & L. Scoreb.(76.9\%)\\
     \hline
   \end{tabular}
-  \caption{Optimization of one parallel reduction dispatch on $1024\times1024$ instruction \texttt{float x= length(vec4 y)}.
-Note that the interleaved addressing intentionally does bank conflicts to show performance impacts.
-The number in parentheses states the number of sequential elements each thread loads.}\label{table:map_reduce}
+  \caption{Optimization of one parallel reduction dispatch on $1024\times1024$ \texttt{vec4} with \texttt{unary\_op(vec4 x)=length(x)} with block\_size = 1024.
+$\Delta t_c$ is in $\mu s$ Note that the interleaved addressing in this example intentionally commits bank conflicts to show performance impacts.
+The number in parentheses states the number of additional added sequential elements.
+Reduced is by block\_size$\times$ sequential elements.
+Thus they do more work, while also performing better!}\label{tab:map_reduce}
 \end{table}
 
-\begin{table}
+The metrics between the parallel reduction and scan differ quite a lot in their comparison to the their bank-conflict reduced solutions.
+Where the scans issue utilization goes drastically up from 34.2\% to 51.3\%, the same metric goes down for parallel reduction from 62.2\% to 33.0\%.
+SM throughput decreases as VRAM throughput increases, although little for the scan.
+The top warp-stall reason for both is the barrier synchronization between thread groups.
+
+The warp-stall reason however showing the increased throughput due to respecting the shared memory architecture is the short scoreboard.
+The shaders in their original state have no reason to stall for the short scoreboard as any memory access is directly fetched.
+In the parallel reduction case the change is quite visible as this metric drops from 17.2\% to only 2.0\%.
+
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
-    Method              &  $\Delta t_c$         & Speedup &VRAM              & SM      & L2     & Top Stall Reas.\\\hline
-    Conflicts           & 748                   & -       &17.5\%            & 51.7\%  & 8.2\%  & Barrier(49.5\%)\\\hline
-    No confl.           & 571                   & 1.31x   &25.7\%            & 50.1\%  & 5.4\%  & Barrier(31.9\%)\\\hline
-    part. confl.        & 546                   & 1.36x   &23.5\%            & 50.8\%  & 6.0\%  & Barrier(30.5\%) \\\hline\hline
-    Seq. (2x)           & 323                   & 2.31x   &41.7\%            & 43.0\%  & 17.2\% & S. Scoreb.(32.0\%)\\\hline
-    unrolled            & 258                   & 2.90x   &53.8\%            & 42.2\%  & 22.0\% & S. Scoreb.(32.0\%)\\\hline\hline
+    Interleaved       &  $\Delta t_c$         & Speedup &VRAM              & SM      & L2     & Sel. Stall Reas.\\\hline
+    Conflicts           & 748                   & -       &17.5\%            & 51.7\%  & 8.2\%  & S. Scoreb.(6.3\%)\\\hline
+    No confl.           & 571                   & 1.31x   &25.7\%            & 50.1\%  & 5.4\%  & \textbf{S. Scoreb.(2.6\%)}\\\hline
+    part. confl.        & 546                   & 1.36x   &23.5\%            & 50.8\%  & 6.0\%  & \textbf{S. Scoreb.(3.2\%)} \\\hline\hline
+    Seq. (2x)           & 323                   & 2.31x   &41.7\%            & \textbf{43.0}\%  & 17.2\% & S. Scoreb.(32.0\%)\\\hline
+    unrolled            & 258                   & 2.90x   &\textbf{53.8}\%   & 42.2\%  & \textbf{22.0}\% & S. Scoreb.(32.0\%)\\\hline\hline
     Seq. (4x)           & 311                   & 2.40x   &56.8\%            & 22.7\%  & 37.9\% & L. Scoreb.(43.3\%)\\\hline
@@
   \end{tabular}
-  \caption{Optimization of one scan dispatch on $4 \times 1024\times1024$ integers with \texttt{binary\_op(x,y) := x+y}. $\Delta t_c$ in $\mu s$. }\label{table:scan}
+  \caption{Optimization of one scan dispatch on $4 \times 1024\times1024$ integers with \texttt{binary\_op(x,y) := x+y}.
+$\Delta t_c$ in $\mu s$, block size is 1024.
+The number in parentheses states the number of additionally loaded sequential elements.
+Segment of each scan is block\_size$\times$ sequential elements.
+Higher reduction factor corresponds to more work being done.}\label{tab:scan}
 \end{table}
-
+The second optimization is giving these algorithms more sequential work.
+This will lighten the SM throughput and increase VRAM throughput as can be seen in both cases.
+Note, that each thread in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done.
+However due to the additional sequential work parallelization a block level goes down by the same factor.
+As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this balances the relation of I/O latency to work better and thus improves performance.
+
+Doing sequential work in parallel reduction adds almost no strain.
+Thus very high sequential work can be done by each thread.
+The SM work will shrink, while the shader gets more and more memory bound.
+For a reduction factor of 256 first signs of performance decrease can be seen as only four groups are involved in the dispatch.
+The GTX970 has 13 SMs.
+Therefore work is not parallelized enough on a block level anymore.
+The metric SM active\% will go rapidly down at this point.
+For four groups its 27.8\%, which is roughly $\frac{4}{13}$.
+
+For the scan this is a little bit harder as elements need to be stored in registers.
+The effects can be seen in table \ref{tab:scan}.
+The segment each scan can handle is by far not as extensive as in parallel reduction.
+But the speedup achieved from a partially conflict free to an implementation with sequential work is in contrast much higher.
+Since each thread requires the nearest by elements, L2-cache hits increase rapidly as does the L2 utilization.Thus the scoreboard stall reasons rise again.
+
+\textbf{Counting Sort:}
 \begin{table}

### 2019-01-21 mscthesis.tex
sorting

@@
 \newenvironment{code}{\captionsetup{type=algorithm}}{}
- \caption{P2G-Transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-Block size is (8,4,4).
-Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_random}
 \begin{document}
@@
 This is the case for algorithms which require neighbor communication.
-Supported is either a full deep copy of all particle variables.
+Supported is a full deep copy of all particle variables.
 A deep copy profits from item \ref{it:sort_1} and \ref{it:sort_2}, but takes the hit in performance of uncoalesced accesses for all variables.
-Or alternatively an indexing to access an existing particle buffer in the specified sorted manner.
+Or, alternatively an indexing to access an existing particle buffer in the specified sorted manner.
 Indexing only profits from item \ref{it:sort_2} and subsequent shaders will be uncoalesced.
@@
     Interleaved    & 305                   & -       &23.0\%          & \textbf{60.9}\%  & S. Scoreb.(17.2\%)\\\hline
-    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0}\%) \\\hline
+    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0\%)} \\\hline
     Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%         & L. Scoreb.(80.1\%)\\\hline
@@
 This will lighten the SM throughput and increase VRAM throughput as can be seen in both cases.
-Note, that each thread in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done.
-However due to the additional sequential work parallelization a block level goes down by the same factor.
-As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this balances the relation of I/O latency to work better and thus improves performance.
+Note, that each thread group in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done by each thread.
+However due to the additional sequential work parallelization the amount of thread groups go down by the same factor.
+As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this saturates the threads and thus improves performance.
 
@@
 But the speedup achieved from a partially conflict free to an implementation with sequential work is in contrast much higher.
-Since each thread requires the nearest by elements, L2-cache hits increase rapidly as does the L2 utilization.Thus the scoreboard stall reasons rise again.
-
-\textbf{Counting Sort:}
-\begin{table}
-  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
+Since each thread requires the nearest by elements, L2-cache hits increase rapidly as does the L2 utilization.
+Thus the scoreboard stall reasons rise again.
+Unrolling the static loop which loads the sequential elements additionally hides memory latency.
+
+\textbf{Counting Sort:} The aim of sorting is coalesced as well as increased caching behavior for all following shaders.
+As a motivation the binning's similarity to the MPM-Operations might already be highlighted.
+A key difference is that binning does not query neighbor information (in the global atomic approach).
+Table \ref{table:bin} shows how a sorted particle position increases coalescing and L2-Cache hits.
+The VRAM-throughput increases to 75.3\% due to coalescing.
+Nearby particles hit the same bins and thus L2 rises up to 35.0\%.
+The 6.95x speed up shows how important the sorting as a preprocess is.
+\begin{table}[htbp]
+  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
     Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
     Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
-    Deep sort           &    218                & 6.95x   &75.3\%            & 24.4\%  & 35.0\% & 37.8\% \\
+    Deep sort           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
     \hline
@@
 
-\begin{table}
+Table \ref{table:sort} shows the performance of index and deep sorting as is.
+The index sorting will have constant timing across cases.
+Their writing access will stay uncoalesced.
+Moreover shaders down the pipeline will also not benefit from coalescing.
+The binning in table \ref{table:bin} thus for instance always runs on random ordering.
+Global atomic binning does not use neighbor data.
+Thus index sorting is not affecting it.
+
+Again differences similarly to the binning can be seen when already working with a sorted state.
+Coalescing highly increases VRAM throughput.
+So much that a deep sort on eight \texttt{vec4}s is almost on equal footing with a random index sort.
+The speedup from an uncoalesced to a fully coalesced state is 5.68x.
+The full sort suffers for each (particle) variable from an uncoalesced access.
+However if sorting is not done the MPM-Transfers will then suffer from those.
+
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r| }    \hline
@@
     Index        &  1,516       &25.6\%& L. Scoreb.(78.7\%) & 16.30\%& 9.30\% \\\hline
-    Full(8 vec4) &  1,731       &75.3\%& L. Scoreb.(92.2\%) & 38.60\%& 34.60\%\\
+    Full(8 vec4) &  1,731       &\textbf{75.3}\%& L. Scoreb.(92.2\%) & 38.60\%& 34.60\%\\
     \hline
@@
  \end{tabular}
-\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.
-Best case is already sorted, worst case is unsorted and still fully random.}\label{table:bin}
+\caption{Order dependency of reordering of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.
+Best case is already sorted, worst case is unsorted and still fully random.}\label{table:sort}
 \end{table}
 
-\begin{table}
+Binning as well as reordering will rarely run fully sorted as mentioned in \ref{sec:sort}.
+Sorting works on the last sorted state.
+The CFL-condition encourages limited reach.
+The Material-Point-Method encourages uniform movement.
+These two properties in general will guarantee higher coalescing as well as L2-cache hit rates.
+It is however up to the dynamics of the problem how much time is spend sorting.
+
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r |r|r|r|}    \hline
@@
 Wait No Instructions
-\begin{table}
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r |}    \hline
@@
 \end{table}
-\begin{table}
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r |}    \hline
@@
 
-\begin{table}
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r |}    \hline
@@
 Barrier 26.3\% on P2G-Sync
-\begin{table}
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r |}    \hline

### 2019-01-22 mscthesis.tex
g2p transfer

@@
     Long Scoreboard\% & 100                   & VRAM latency (mostly reads)\\\hline
-    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\
-		      &                       & (mostly (atomic) writes)\\\hline
+    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\\hline
     Barrier\%         & 100                   & block synchronization \\\hline
@@
 These metrics give the programmer more power of identifying the right optimizations for a given code.
+
 All metrics in the following are gathered from a NVIDIA GTX970.
+All timings are averaged over at least 1000 runs with OpenGL Timer Queries \cite{KHRONOS:TIMER_QUERY}.
+As more runs are done the GPU stays very consistent on bringing the average runtime closer to the minimum recorded time.
+This indicates that the first few runs can take comparably longer.
 
@@
 The CFL-condition encourages limited reach.
-The Material-Point-Method encourages uniform movement.
+Solids and fluids will mostly move uniformly.
 These two properties in general will guarantee higher coalescing as well as L2-cache hit rates.
 It is however up to the dynamics of the problem how much time is spend sorting.
+\\
+
+The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to build off with only one \texttt{vec4} transfer.
+The MPM-Operations are evaluated with two easily reproducible examples.
+
+For the \textbf{G2P-Transfers} the different datasets can be seen in table \ref{table:g2p_random} and \ref{table:g2p_uniform4}.
+The first one uses random particle positions and therefore is overly pessimistic for the Material Point Method as explained earlier.
+The second one is optimistic as warp branch divergence is minimized.
+
+The global G2P-Transfers does not reach desirable throughput in any of the GPU subunits.
+The top warp-stall reason is Long Scoreboard: Nearly all information needs to be reloaded uncoalesced from global memory.
+The data in \ref{table:g2p_uniform4} has an ordering that encourages data reuse.
+But is not aligned with the grid.
+However the SM already runs on three times its throughput compared to the random data.
 
 \begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |r|r|r|}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2    &SM     \\\hline
-    G2P global             &   38,318             &-               &24.3\%&34.4\% & 9.6\% \\\hline
-    sorted                 &    2,991		  &12.81x          & 7.5\%&52.3\% &67.1\% \\\hline
-    G2P-Shared             &    9,665             & 3.96x          & 3.4\%& 2.5\% &65.0\% \\\hline
+  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         & L2    &SM     &Top Warp-Stall Reas. \\\hline
+    G2P-global             &   38,318             &-               &34.4\% & 9.6\% &L. Scoreb.(95.1\%)  \\\hline
+    sorted                 &    \textbf{2,991}    &12.81x          &\textbf{52.3\%}&\textbf{67.1}\% &No Instr.(14.8\%)   \\\hline
+    G2P-shared             &    9,665             & 3.96x          & 2.5\% &65.0\% &No Instr.(14.2\%)\\\hline
  \end{tabular}
  \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
-Block size is (8,4,4).}\label{table:p2g_uniform4}
+Block size is (8,4,4).}\label{table:g2p_random}
 \end{table}
 
-VRAM-L.Scoreboard 79.9\%
-Wait No Instructions
+As elucidated in chapter \ref{sec:sort} deep sorting increases coalescing and caching behavior.
+The effects of sorting impact performance by a 12.81x speedup in the randomized case.
+The density of particles per bin is 0.5.
+Heavy use is made of the L2-Cache rising up to 52.3\% reading from many distinctive, successive memory positions.
+Meanwhile the density of four particles per bin in table \ref{table:g2p_uniform4} puts comparably low strain on the L2-Cache with 9.4\%.
+The runtime on random data is only 12.2\% slower than the uniform data.
+
+The G2P-shared method lies miles beyond that as its execution time is 9,665 $\mu s$.
+It can however get an edge on the more favored uniform data where 2,232 is the best recorded time.
+The G2P-shared method relies on shared memory instead of the L2-Cache to read data.
+Thus this method can be faster when little branch divergence between warps is the case.
+The random data however has highly divergent branching and is cause for the high execution times.
+
 \begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
-    G2P global             &   10,316             &-               &2.3\%&34.4\%&32.4\% \\\hline
-    sorted                 &    2,629		  &3.92x           &4.1\%&9.4\%&66.9\%  \\\hline
-    G2P-Shared             &    2,232             &4.62x           &7.3\%&4.1\% &63.6\% \\\hline
+  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         & L2   &SM             & Top Warp-Stall Reas.\\\hline
+    G2P-global             &   10,316             &-               &34.4\%&\textbf{32.4}\%         & L. Scoreb.(79.9\%) \\\hline
+    sorted                 &    2,629		  &3.92x           &\textbf{9.4}\%&66.9\%          & No Instr.(14.1\%) \\\hline
+    G2P-shared             &    \textbf{2,232}             &4.62x           &4.1\% &\textbf{63.6}\%& No Instr.(22.7\%) \\\hline
  \end{tabular}
-
  \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
 They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).
-Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
+Block size is (8,4,4).}\label{table:g2p_uniform4}
 \end{table}
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.\\\hline
-    P2G-Atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)\\\hline
-    indirect		   &    6,276             & 1.10x   &90.8\%&S.Scoreb.(31.7\%)\\\hline
-    batching = 4           &    4,929             & 1.41x   &65.1\%&S. Scoreb.(24.3\%)\\\hline
-    both                   &    3,148             & 2.20x   &65.0\%&No Instr.(23.2\%)\\\hline
-    \hline
-    P2G-Sync               &    3,161             & -       &72.4\%&No Instr.(17.7\%) \\\hline
-    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) \\\hline
- \end{tabular}
- \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).}\label{table:p2g_optim}
-\end{table}
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
-    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
-    \cite{Meyer2015}       &   45,342             &0.98x           &25.1\%&42.4\%&11.7\% \\\hline
-    \cite{Meyer2015} sorted&   23,007		  &1.97x           &43.8\%&59.0\%&23.9\% \\\hline
-    global sorted          &   20,484		  &2.21x           &7.0 \%&44.0\%&16.1\% \\\hline
-    \hline
-    P2G-Pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
-    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &65.0\%\\\hline
-    P2G-Sync*              &    2,595             &17.47x          &5.9\%&7.6\%  &67.0\%\\\hline
- \end{tabular}
- \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).
-Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
-\end{table}
-Barrier 26.3\% on P2G-Sync
+
+In NVIDIA-Nsight run times recorded for the G2P-Transfers were comparably lower while making use of batching.
+However OpenGL timer queries recorded little difference in trade-off for the added register pressure.
+Thus G2P-Transfers should stay away of batching.
+
+For the sorted G2P-global as well as the G2P-shared method following statements are also of interest.
+The top-warp stall reason is waiting on the instruction fetches which is typical for compute-bound shaders with high SM usage.
+The SM Issue Utilization per Active Cycles is consistent around 65\%.
+This indicates that neither increasing occupancy nor reducing warp-stalls would increase performance.
+
+As a summary, relying on the comparably fast L2-Cache memory transactions over shared memory is consistent across data for the sorted G2P-global method.
+The main benefit of the global solution is that warps do not suffer from branch divergence in that method.
+However the G2P-shared method may still give an edge in performance.
+
+The \textbf{P2G-Transfer}
 \begin{table}[htpb]
@@
 \end{table}
+
+\begin{table}[htpb]
+  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM     \\\hline
+    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
+    \cite{Meyer2015}       &   45,342             &0.98x           &25.1\%&42.4\%&11.7\% \\\hline
+    \cite{Meyer2015} sorted&   23,007		  &1.97x           &43.8\%&59.0\%&23.9\% \\\hline
+    global sorted          &   20,484		  &2.21x           &7.0 \%&44.0\%&16.1\% \\\hline
+    \hline
+    P2G-Pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
+    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &65.0\%\\\hline
+    P2G-Sync*              &    2,595             &17.47x          &5.9\%&7.6\%  &67.0\%\\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).
+Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
+\end{table}
+
+Barrier 26.3\% on P2G-Sync
+\begin{table}[htpb]
+  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occupancy\\\hline
+    P2G-Atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
+    indirect		   &    6,276             & 1.10x   &90.8\%&S. Scoreb.(31.7\%) & 43.4\\\hline
+    batching = 4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& 34.9\\\hline
+    both                   &    3,148             & 2.20x   &65.0\%&No Instr.(23.2\%) & 35.4\\\hline
+    \hline
+    P2G-Sync               &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
+    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & 35.4\\\hline
+ \end{tabular}
+ \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).}\label{table:p2g_optim}
+\end{table}
+
 \begin{table}

### 2019-01-22 mscthesis.tex
p2g transfer start

@@
 
-As a summary, relying on the comparably fast L2-Cache memory transactions over shared memory is consistent across data for the sorted G2P-global method.
+As a summary, relying on the comparably fast L2-Cache memory transactions over shared memory is consistent across the reduced examples for the sorted G2P-global method.
 The main benefit of the global solution is that warps do not suffer from branch divergence in that method.
@@
 
-The \textbf{P2G-Transfer}
+The \textbf{P2G-Transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4} respectively.
+The global atomic approaches suffer from low throughput across all units.
+Cause are uncoalesced accesses and low cache utilization.
+
+The top stall reason for all global operations is the short scoreboard.
+This is caused by the high number of atomic writes which are accumulated in L2-Cache.
+The sorting brings forth this problem more acutely as L2-Caches throughput rises to the top SOL with 44.0\% in the uniform case of table \ref{table:p2g_uniform4}.
+The random data example in table \ref{table:p2g_random} relies even more heavily on the L2-Cache with 65.8 SOL\% similarly to what has been seen in the G2P-Transfer.
+
+Nevertheless this already speeds up the random case by a factor of 11.88x as it followed no particular ordering. The sorting makes sure that the incoming data is preprocessed for optimal access.
+
 \begin{table}[htpb]
@@
     \cite{Meyer2015} sorted&   30,477		  &8.35x           &64.2\%&64.2\%&22.1\% \\\hline
-    global sorted          &   21,413		  &11.88x          &7.2\% &65.8\%&19.4\% \\\hline
+    global sorted          &   21,413		  &11.88x          &7.2\% &\textbf{65.8}\%&19.4\% \\\hline
     \hline
     P2G-Pull               &   26,822             &9.46x           &1.6\% &2.3\% &35.2\%\\\hline
-    P2G-Atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &65.4\%\\\hline
-    P2G-Sync*              &   12,801             &19.87x          &3.6\% &7.3\% &61.3\%\\\hline
+    P2G-Atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &\textbf{65.4}\%\\\hline
+    P2G-Sync*              &   \textbf{12,801}    &19.87x          &3.6\% &7.3\% &\textbf{61.3}\%\\\hline
  \end{tabular}
@@
 \end{table}
+
+Using the P2G-Pull method on the random case performs worse than the global method.
+It can significantly save time however in the uniform case.
+The P2G-Pull-Transfer is 4.31 times faster than the global P2G-Transfer on the sorted uniform data.
+
+Top SOL\% is now the SM, but it lacks throughput with 39.4\%.
+Its SM Issue Utilization per active cycle is 40\% which is below the desired threshold of 60\%.
+The top warp stall reason is instruction based.
+So as long as the algorithm does not change no increase can be achieved here.
+As explained in chapter \ref{sec:p2g} the pull method requires high amount of shared memory.
+The occupancy calculator as well as NVIDIA-Nsight report an occupancy of 16.0.
+This is $\frac{1}{4}$-th of the maximum occupancy of the GTX970.
+This is however optimized for thread group size.
 
@@
     \cite{Meyer2015} sorted&   23,007		  &1.97x           &43.8\%&59.0\%&23.9\% \\\hline
-    global sorted          &   20,484		  &2.21x           &7.0 \%&44.0\%&16.1\% \\\hline
+    global sorted          &   20,484		  &2.21x           &7.0 \%&\textbf{44.0}\%&16.1\% \\\hline
     \hline
     P2G-Pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
-    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &65.0\%\\\hline
-    P2G-Sync*              &    2,595             &17.47x          &5.9\%&7.6\%  &67.0\%\\\hline
+    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &\textbf{65.0}\%\\\hline
+    P2G-Sync*              &    \textbf{2,595}             &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
  \end{tabular}
@@
 
+A closer look is taken of the P2G-Atomic- and P2G-Sync-Transfers to illustrate the performance improvements of batching and filtering active blocks.
 Barrier 26.3\% on P2G-Sync

### 2019-01-23 mscthesis.tex
evaluation finished

@@
 
-Using the P2G-Pull method on the random case performs worse than the global method.
+Using the \textbf{P2G-Pull method} on the random case performs worse than the global method.
 It can significantly save time however in the uniform case.
@@
 This is $\frac{1}{4}$-th of the maximum occupancy of the GTX970.
-This is however optimized for thread group size.
+This is however already optimized for thread group size.
 
@@
 
-A closer look is taken of the P2G-Atomic- and P2G-Sync-Transfers to illustrate the performance improvements of batching and filtering active blocks.
-Barrier 26.3\% on P2G-Sync
+A closer look is taken of the P2G-Atomic- and P2G-Sync-Transfers to illustrate the performance improvements of batching and filtering for active blocks in table \ref{table:p2g_optim}.
+
+The \textbf{P2G-Atomic-Transfer} in its initial state uses one atomic transfer per particle on shared memory.
+Its effects are quite visible on the SM SOL\% with 82.9\% to 90.9\% in the call for active blocks.
+The top stall reason is short scoreboard which is indicative of the high number of stalls due to atomic writes on shared memory.
+
+Batching reduces the atomic writes.
+Particles get handled together.
+In succession one atomic write is done.
+This changes the top stall reason to the more favorable instruction fetches as the main stall reason with 20.8\%.
+Short scoreboard goes down to 11.6\%.
+As previously the SM was strained too hard throughput goes down to 65.0\%.
+The greatest time improvement can be seen when batching and indirect call lead to a combined speedup of 2.2x.
+This is much higher than the multiplicative speedup of 1.51x one could expect.
+
+The \textbf{P2G-Sync-Transfer} also may use batching and save time.
+The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}.
+Batching in this case reduces the amount of thread group barriers needed for synchronization.
+The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process.
+The external synchronization is however much less invasive than atomic writes.
+The P2G-Sync method without batching has roughly the same timing as the P2G-Atomic-Transfer with batching of four particles.
 \begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occupancy\\\hline
+    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occup.\\\hline
     P2G-Atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
-    indirect		   &    6,276             & 1.10x   &90.8\%&S. Scoreb.(31.7\%) & 43.4\\\hline
-    batching = 4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& 34.9\\\hline
-    both                   &    3,148             & 2.20x   &65.0\%&No Instr.(23.2\%) & 35.4\\\hline
+    indirect		   &    6,276             & 1.10x   &90.8\%&\textbf{S. Scoreb.(31.7\%)} & 43.4\\\hline
+    batching=4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& \textbf{34.9}\\\hline
+    both                   &    3,148             & \textbf{2.20x}   &\textbf{65.0}\%&No Instr.(23.2\%) & 35.4\\\hline
     \hline
     P2G-Sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
-    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & 35.4\\\hline
+    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & \textbf{35.4}\\\hline
  \end{tabular}
@@
 They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).}\label{table:p2g_optim}
+Block size is (8,4,4).
+Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:p2g_optim}
 \end{table}
 
-\begin{table}
+Batching is a process that adds register pressure as can be seen by the reduced occupancy of 34.9-35.4.
+Using batching every new particle variable involved in the transfer takes the batching amount of that variables size more registers.
+
+The P2G-Atomic-Transfer is very reliant on batching where the P2G-Sync-Transfer is not.
+This is further highlighted in table \ref{table:p2g_area}.
+With one particle per cell(area 47.6\%) batching does not help at all.
+The P2G-Sync-Transfer however can still work comparably fast to the high density of four particles per cell losing about 7.2\% speed.
+This also means that the P2G-Sync-Transfer only scales slightly with the area covered, which is due to the overhead of loading and writing more blocks.
+
+\begin{table}[htpb]
   \begin{tabular}{ | l | r | r | r | r |}    \hline
@@
  \end{tabular}
- \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered.
-Block size is (4,4,4).}\label{table:p2g_optim}
+ \caption{P2G-Sync-Transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered with 47.6\%/23.8\%/11.9\%.
+Block size is (4,4,4).}
+ \label{table:p2g_area}
 \end{table}
-
+The P2G-Atomic- and the even more favorable P2G-Sync-Transfer give great speedups over the global atomic methods.
+Even in a random case (table \ref{table:p2g_random}) they can outdo the global sorted method despite the high branch divergence.
+They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
 \section{Conclusion}

### 2019-01-24 mscthesis.tex
introduction+conclusion

@@
 \section{Introduction} \label{intro}
+General Purpose Computation on Graphics Processing Unit(GPGPU) has elevated Graphics Processing Units(GPUs) to any computational field utilizing massively parallelizable algorithms.
+As such is the nature of discretizations in physics.
+GPUs performance still follows Moore's Law.
+In unison with efficient algorithms they still have much potential to offer.
+
+The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it.
+Conversely algorithms which are not parallelizable perform much worse and should be avoided.
+Dividing up these tasks between CPU and GPU is not recommendable.
+Transferring data over the PCI-Bus between the two is slow for big amounts of data and requires synchronization.
+Tasks in such a pipeline should thus stay on the GPU although they might perform worse.
+
+Programming for the GPU is another hurdle.
+GPU code is very reliable on the generation of the GPU to use relevant extensions.
+Backwards compatibility therefore is mostly ignored and written towards a certain GPU (generation).
+This is amplified due to the GPU code often being reliant on architecture limits for maximum performance.
+The major languages to target the GPU are CUDA, OpenCL, Direct 3D, OpenGL and Vulkan.
+They build on top of a C/C++-language subset.
+Whereas CUDA offers the most libraries and optimized code it is only available on NVIDIA GPUs.
+OpenCL's development on NVIDIA GPUs is dragging far behind the standard.
+It's merge into the Vulkan roadmap therefore could be an incentive to support a more in-depth language universally \cite{OPENCL:ROADMAP}.
+Direct3D is only available on Windows.
+Vulkan is the successor of OpenGL and removes work from the driver to the developer.
+Main benefit over OpenGL is less CPU usage, which helps little for the almost full GPU implementation here.
+This implementation opts for OpenGL as Vulkan is still relatively new.
+Everything presented could be implemented in one of the other languages as well.
+
+The simulation of the dynamics of materials is still extremely challenging in different applications and fields.
+Multi-physics interactions between materials are not realized on a big scale and often are not possible or feasible.
+The Material-Point-Method has already reliable tackled many different models, materials and even interaction between materials.
+Chapter \ref{sec:rel_work} will give a short summary of such.
+
+Pure Eulerian methods like the Finite-Element-Method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability.
+Pure Langrangian methods form the end other of the spectrum.
+Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and of empirical nature.
+Particle-in-Cell(PIC) techniques like the Material-Point-Method are hybrid Lagrangian/Eulerian methods.
+They try to benefit from the nature of both.
+The Material-Point-Method utilizes Langrangian particles as a consistent storage of the dynamics.
+The governing equations of the physical problem can be discretized to and solved on a uniform Eulerian grid.
+
+The necessary transfers between the Lagrangian and Eulerian representation are the focus of the Particle-in-Cell techniques.
+These transfers, realized due to interpolation, create (estimable) numerical error and lead to dissipation.
+Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.
+
+This $p$ particles to one node relationship of a Particle-to-Grid-Transfer(P2G) or inversely one node to $p$ particles of a Grid-to-Particle-Transfer(G2P) is the main problem when realizing these operations (on the GPU).
+When faced with the static memory architecture of the GPU this will inevitably lead to branch divergence as some nodes are assigned more particles than others.
+
+Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as wel as Material-Point-Method extensions.
+In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
+The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method.
+Furthermore GPU intrinsics, metrics and optimization opportunities are discussed.
+Chapter \ref{sec:implementation} tends to the implementation.
+The evaluation of the implementation follows up in chapter \ref{sec:eval}.
+As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.
+
+\clearpage
 \section{Related Work}\label{sec:rel_work}
-\textbf{Particle-in-Cell (PIC):} PIC techniques were developed initially for hydrodynamics
+\textbf{Particle-in-Cell:} PIC techniques were developed initially for hydrodynamics
 \cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations. Their key steps are identical to the MPM-method:
@@
 Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle (FLIP) was developed for fluid simulations and is still widely in use for large simulations.
-Their counterpart being the empirical Smoothed Particle Hydrodynamics (SPH) \cite{gingold1977smoothed}.
+Their counterpart being the empirical Smoothed Particle Hydrodynamics \cite{gingold1977smoothed}.
 
@@
 \clearpage
-\section{Basics}
+\section{Basics}\label{sec:basics}
 \subsection{Reference and current configuration}\label{sec:ref_corr_config}
@@
 They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
-\section{Conclusion}
-\subsection{Summary}
-\subsection{Future Work}
-Particle Activation: \cite{MPM:GPU}
-
-Sparse: \cite{OPENVDB} \cite{NVIDIA:GVDB_VOXELS}
+\section{Conclusion \& Future Work}\label{sec:conclusion}
+The derivation of the Material-Point-Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
+Although the MPM-Method consists of many mathematical tasks four operations on the GPU can be identified.
+Exclusive grid and particle operations are parallelized with a 1:1 correspondence to threads.
+The optional use of stream compaction reduces applicable particles/nodes where necessary.
+
+The other two operations are particle and grid transfers which are the main focus of this thesis.
+Different particle and grid transfers utilizing the hierarchical memory structure of the GPU are compared.
+Presented is a pipeline to efficiently execute these transfers on the GPU.
+All transfers do not rely on fixed bin sizes in any stages of the pipeline.
+Since these operations are executed multiple times per frame these operations are augmented with two preprocessing steps.
+The preprocessing steps need to be done once and are outweighed by the transfers.
+
+Deep sorting particles with counting sort in a double-buffered manner allows coalesced accesses and high cache hit rates on transfers.
+Although possible this process does not rely on the CFL-condition.
+While the shared memory architecture can give an edge in performance it produces branch divergence and may perform worse over a simpler algorithm relying on the L2-cache.
+The P2G-Sync-Transfer is shown to be an easily extendable and even in edge cases well performing operation.
+Particle-Push-Transfers could be shown to profit from batching as multiple particles are handled in unison.
+The sorted global G2P-Transfer's consistency of utilizing the L2-Cache is a surprise further to be monitored.
+The second preprocessing step is filtering the active blocks with a stream compaction pipeline where no particles are present.
+It also supplies the P2G-Pull- and P2G-Sync-Transfer with a maximum per block bin counter.
+
+To show the weaknesses and strengths of the transfers tests on an overly pessimistic and an optimistic case are performed and examined.
+While this strongly furthers understanding it leaves real simulation performance comparisons to be desired.
+However before this process can be concluded merge and split techniques need to be taken into consideration.
+They are designed to augment the numerics of the Material-Point-Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
+As such they will stray from the optimistic example.
+How much they will stray from it is however dependent on the dynamics of the problem.
+The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.
 \clearpage
 \begin{appendices}
-  \section{"Inertia Tensor" for Cubic Splines}\label{app:dp_proof}
+\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
   \begin{code}

### 2019-01-25 mscthesis.tex
rereading almost halfway through

@@
 As such is the nature of discretizations in physics.
-GPUs performance still follows Moore's Law.
+GPU's performance still follows Moore's Law.
 In unison with efficient algorithms they still have much potential to offer.
@@
 Programming for the GPU is another hurdle.
-GPU code is very reliable on the generation of the GPU to use relevant extensions.
-Backwards compatibility therefore is mostly ignored and written towards a certain GPU (generation).
+GPU code is reliant on the generation and hardware vendor of the GPU to use relevant extensions.
+Backwards compatibility therefore is mostly ignored and written towards a certain GPU generation.
 This is amplified due to the GPU code often being reliant on architecture limits for maximum performance.
-The major languages to target the GPU are CUDA, OpenCL, Direct 3D, OpenGL and Vulkan.
+The major languages to target the GPU are CUDA, OpenCL, Direct 3D, OpenGL, and Vulkan.
 They build on top of a C/C++-language subset.
-Whereas CUDA offers the most libraries and optimized code it is only available on NVIDIA GPUs.
+
+Whereas CUDA offers the most libraries and optimized code, it is only available on NVIDIA GPUs.
 OpenCL's development on NVIDIA GPUs is dragging far behind the standard.
-It's merge into the Vulkan roadmap therefore could be an incentive to support a more in-depth language universally \cite{OPENCL:ROADMAP}.
+Its merge into the Vulkan roadmap could be an incentive to support a more in-depth language universally \cite{OPENCL:ROADMAP}.
 Direct3D is only available on Windows.
 Vulkan is the successor of OpenGL and removes work from the driver to the developer.
-Main benefit over OpenGL is less CPU usage, which helps little for the almost full GPU implementation here.
+Main benefit over OpenGL is less CPU usage which helps little for the almost full GPU implementation here.
 This implementation opts for OpenGL as Vulkan is still relatively new.
@@
 The simulation of the dynamics of materials is still extremely challenging in different applications and fields.
-Multi-physics interactions between materials are not realized on a big scale and often are not possible or feasible.
-The Material-Point-Method has already reliable tackled many different models, materials and even interaction between materials.
-Chapter \ref{sec:rel_work} will give a short summary of such.
+Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible.
+The Material-Point-Method has already reliable tackled many different models, materials, and even interaction between materials.
 
@@
 Pure Langrangian methods form the end other of the spectrum.
-Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and of empirical nature.
+Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature.
 Particle-in-Cell(PIC) techniques like the Material-Point-Method are hybrid Lagrangian/Eulerian methods.
 They try to benefit from the nature of both.
-The Material-Point-Method utilizes Langrangian particles as a consistent storage of the dynamics.
-The governing equations of the physical problem can be discretized to and solved on a uniform Eulerian grid.
-
-The necessary transfers between the Lagrangian and Eulerian representation are the focus of the Particle-in-Cell techniques.
-These transfers, realized due to interpolation, create (estimable) numerical error and lead to dissipation.
+Advection on particles is trivial.
+Thus the Material-Point-Method utilizes Lagrangian particles as a consistent storage of dynamic properties.
+The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid.
+The grid is unmoving and does not store any information between frames.
+
+Transfers between the Lagrangian and Eulerian are of need to bring the two together.
+These transfers are the focus of the Particle-in-Cell techniques.
+These transfers are realized due to interpolation.
+Thus they create (estimable) numerical error and lead to dissipation.
 Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.
 
-This $p$ particles to one node relationship of a Particle-to-Grid-Transfer(P2G) or inversely one node to $p$ particles of a Grid-to-Particle-Transfer(G2P) is the main problem when realizing these operations (on the GPU).
-When faced with the static memory architecture of the GPU this will inevitably lead to branch divergence as some nodes are assigned more particles than others.
-
-Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as wel as Material-Point-Method extensions.
+The $p$ particles to one node relationship of a Particle-to-Grid(P2G)-Transfer or inversely one node to $p$ particles of a Grid-to-Particle(G2P)-Transfer is the main problem when realizing these operations on the GPU.
+When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.
+
+Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material-Point-Method extensions.
 In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
-The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method.
+The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method to get an understanding of the prominent operations.
 Furthermore GPU intrinsics, metrics and optimization opportunities are discussed.
@@
 \section{Related Work}\label{sec:rel_work}
-\textbf{Particle-in-Cell:} PIC techniques were developed initially for hydrodynamics
-\cite{evans1957particle}. They are also widely used in plasma simulations to solve Maxwell's equations. Their key steps are identical to the MPM-method:
+\textbf{Particle-in-Cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}.
+They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}.
+Their key steps are identical to the MPM-method:
 \begin{enumerate}
@@
 In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}.
-Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle (FLIP) was developed for fluid simulations and is still widely in use for large simulations.
+I.e.
+the grid could be deleted each frame.
+Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations.
 Their counterpart being the empirical Smoothed Particle Hydrodynamics \cite{gingold1977smoothed}.
 
-For that matter APIC was developed as an improvement to PIC and FLIP solvers.
+As a further development the Affine-Particle-in-Cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers.
 Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
@@
 
-A step further goes PolyPIC giving the velocity field even more freedom by adding polynomial modes of increasing order to the description. \cite{MPM:POLYPIC}
-
-\textbf{Material Point Method:} The MPM-Method is like the PIC-Method mathematical and physical in that it uses analysis and numerics to be derived without the need for empirical assumptions.
+A step further goes PolyPIC giving the velocity field even more freedom by adding polynomial velocity modes of increasing order to the description. \cite{MPM:POLYPIC}
+
+The \textbf{Material Point Method} is a further development to the PIC methods.
+It uses analysis and numerics to be derived without the need for empirical assumptions.
 Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids.
@@
 In rapid succession a variety materials and effects got modeled by the MPM-Method.
-Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature its fast rate of expenditure is not a surprise.
-
-The simulation of (hyper-)elastic material can be done with a number of different models utilizing fixed Linear strain \ref{sec:cor_hyper}, left Cauchy-Green strain \ref{sec:svd} or Hencky-strain(logarithmic) \cite{MPM:SHELLS} for the different requirements of small to large strains.
+Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.
+
+The simulation of (hyper-)elastic materials can be done with a number of different strain measures utilizing fixed Linear strain (chapter \ref{sec:cor_hyper}), Green-Lagrangian strain tensor (chapter \ref{sec:svd}), or logarithmic Hencky-strain (\cite{MPM:SHELLS}) for the different requirements of small to large strains.
 Inserting these into the different energy density $\boldsymbol{\Psi}(\boldsymbol{\epsilon})$ models can already result in a constitutive model.
-Plasticity in general requires extra modeling in terms of a flow rule, yield-condition and a hardening rule which are not discussed here \cite{ochsner2014elasto}.
+Plasticity in general requires extra modeling in terms of a flow rule, yield-condition, and a hardening rule which are not discussed here \cite{ochsner2014elasto}.
 To give a short impression of the developments in computer graphics:
 
-\cite{MPM:PHASE_CHANGE} model phase transition (melting) to a liquid with the fixed corotational hyperelasticity (\ref{sec:cor_hyper}) and discretizing the heat equation.
-\cite{MPM:OLROYDB} use an Olroyd-B model with Cauchy-Green strain to handle elasticity and plasticity for viscoelastic fluids, foams and sponges.
+\cite{MPM:PHASE_CHANGE} model phase transition (melting) to a liquid with using the fixed corotational hyperelasticity model (chapter \ref{sec:cor_hyper}) in combination with a discretization of the heat equation.
+\cite{MPM:OLROYDB} use an Olroyd-B model with Cauchy-Green strain to handle elasticity and plasticity for viscoelastic fluids, foams, and sponges.
 \cite{MPM:DRUCKER} uses Hencky-strain in combination with a St.
@@
 \cite{MPM:SHELLS} simulates thin shells (cloths,cups etc.) with a Kirchhoff-Love model using Hencky-strains.
-\\
-
-There exist several implementations utilizing the GPU for Particle-in-Cell type techniques, which the Material Point Method is a part of:
+
+There exist several implementations utilizing the \textbf{GPU for Particle-in-Cell} techniques, which the Material Point Method is a part of:
 
@@
 These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management.
-Fixed size binning make the methods mentioned before easier to implement, which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
+Fixed size binning make the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
 As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes.
@@
 
-In general full or index sorts can be done. A full sort reorders every variable of a particle. An index sort gives back an index with an offset to access the particle variable.
+In general deep or index sorts can be done. A deep sort reorders every variable of a particle. An index sort gives back an index with an offset to access the particle variable.
 
@@
 
-\textbf{Particle-to-Grid=Transfers:} The particle to grid transfer is generally done in two manners \cite{stantchev2008fast}:
+This implementation compares index and deep counting sort and uses the latter.
+Although possible, this counting sort stays away from the limited movement implied by the CFL-condition.
+Fixed size bins is an assumption this implementation although does not want to impose on the method.
+
+\textbf{Particle-to-Grid-Transfers} are generally done in two manners \cite{stantchev2008fast}:
 \begin{enumerate}
@@
 The grid points in a uniform grid are 'sorted' by nature.
+This implementation tests one Particle Pull and two Particle Push methods.
 
 Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition.
-Shared memory is however limited per GPU (GTX 970, 48KB per block).The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
-Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited block sizes are available \cite{MPM:SNOW}.
-
-In Particle Push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
-\begin{enumerate}
-\item Atomically add up the values of the halo.
-This results in two atomic adds on a side node, four on an edge node and eight on a corner-node; If the block extend is bigger than half the halo extend.
-\item Have every halo of every block correspond to actual global memory and do add steps along each of the three axes.
-This is infeasible for large supports due to the added memory requirements.
-\cite{crassin2011interactive}
-\end{enumerate}
+Shared memory is however limited per GPU (GTX 970, 48KB per block).
+The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
+Since the number of particles per grid point in the MPM-method is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.
+
 Stencil computations and filtering techniques underlie a very similar process to MPM-transfers.
 Although these techniques employ a one-to-one grid relationship, since no particles are involved, it is worth taking a look at them.
+An interesting prospect for filtering techniques on the GPU is given by NVI\-DIA's new shuffle operations.
+Shuffle allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
+They are however not supported on other architecture, yet.
 
@@
 Thus there are no z-dimension halos.
+This however requires all particles in the halo to be present in the limited shared memory.
+As mentioned, this implementation wants to stay away from assumptions limiting particles similar to fixed bin sizes.
 \cite{brandvik2010sblock} \cite{williams2007scientific} \cite{krotkiewski2013efficient}
 
-An interesting prospect for filtering techniques on the GPU is given by NVI\-DIA's new shuffle operations.
-Shuffle allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
-They are however not supported on other architecture, yet.
-
-\textbf{Particle Activation:} If a particle is under a user-defined velocity threshold a particle becomes deactivated.
-It does not get recognized by any particle to grid transfer (nor any grid transfer not affected by the velocity transfer).
-They may of course be reactivated if the back transfer of the velocity exceeds that threshold again or they collide.
+In Particle Push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
+\begin{enumerate}
+\item Atomically add up the values of the halo.
+This results in two atomic adds on a side node, four on an edge node, and eight on a corner node; If the block extend is bigger than half the halo extend, otherwise more.
+\item Every halo of every block corresponds to actual global memory.
+After a transfer, do add steps along each of the three axes.
+This is infeasible for large supports due to the added memory requirements.
+\cite{crassin2011interactive}
+\end{enumerate}
+
+\textbf{Particle Activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated.
+It does not get recognized by any Particle-to-Grid-Transfer, nor the grid solve(, nor any Grid-to-Particle-Transfer not affected by the velocity transfer).
+They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide.
 \cite{MPM:GPU}
 
-Dividing the particles up can be realized with stream compaction algorithms: One implementation would be to compute a scan (based on the discussed condition) followed up by a modified reorder shader discussed in chapter \ref{sec:implementation}.
-The modified reorder shader would then be coalesced.
-
-\textbf{Particle Resampling:} A method that could alleviate pressure off fixed bin sizes and could make the partial sorting methods as well as the shared-memory restrictions less of a factor is particle resampling.
-Their intention mostly is to fill numerical gaps in the material at large deformations.
+Dividing the particles up can be realized with stream compaction algorithms discussed in chapter \ref{sec:implementation}.
+
+\textbf{Particle Resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling.
+Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet.
 Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.
@@
 Mass and volume get equally distributed; Velocity and deformation gradient get copied.
-Refer to chapter \ref{sec:mat_point} for conservation of mass and momentum in MPM.
+Refer to chapter \ref{sec:mat_point} or for conservation of mass and momentum in MPM.
 \cite{gao2017adaptive}
@@
 Velocity is computed from a mass-weighted average of the participating particles.
-The deformation gradient is accordingly to chapter \ref{sec:svd} decomposed into $\boldsymbol{F} = \boldsymbol{U\SigmaV}^T$.
-The quaternion average of $\boldsymbol{U}$ and $\boldsymbol{V}$ leads to an average rotation $\overline{\boldsymbol{U}}$ and $\overline{\boldsymbol{V}}$, respectively.
-The principal strains $\boldsymbol{\Sigma}$ are just averaged component-wise $\overline{\boldsymbol{\Sigma}}$.
-One can compose them in the same fashion $\overline{\boldsymbol{F}} = \overline{\boldsymbol{U}}\overline{\boldsymbol{\Sigma}}\overline{\boldsymbol{V}}^T$.\cite{gao2017adaptive}
+The deformation gradient is accordingly to chapter \ref{sec:svd} decomposed into $\boldsymbol{F} = \boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T$.
+The quaternion average of $\boldsymbol{U}$s and $\boldsymbol{V}$s of the particles lead to an average rotation $\overline{\boldsymbol{U}}$ and $\overline{\boldsymbol{V}}$, respectively.
+The principal strains within $\boldsymbol{\Sigma}$ are just averaged component-wise to $\overline{\boldsymbol{\Sigma}}$.
+One can compose them in the same fashion leading to an average deformation gradient with $\overline{\boldsymbol{F}} = \overline{\boldsymbol{U}}\overline{\boldsymbol{\Sigma}}\overline{\boldsymbol{V}}^T$.\cite{gao2017adaptive}
 
@@
 \cite{gao2017adaptive} produce a signed distance field to identify interior points.
-Since fracture typically starts at a surface point desired behavior is accomplished again.
-Furthermore the signed distance field can be used to adapt the grid size depending on the distance to the surface.
-This allows coarser levels with increasing distance from the surface.
+Surface points also should not merge or split due to visual artifacts.
+Furthermore, the signed distance field can be used to adapt the grid size depending on the distance to the surface.
+This allows for coarser levels with increasing distance from the surface.
 %look again for MPM/PIC gpu impl
@@
 Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$.
-Vectors furthermore use lower-case letters $\boldsymbol{v}$, while Matrices use upper-case letters $\boldsymbol{M}$.
-A variable in the Lagrangian form has a subscript 0 $ _0{\boldsymbol{x}}$, the Eulerian description will be written with an 'overline' $_t\boldsymbol{x}$ if distinction is needed.
+Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$.
+A variable in the Lagrangian form has a left subscript 0: $ _0{\boldsymbol{x}}$.
+The Eulerian description has a left subscript t: $_t\boldsymbol{x}$.
 Occasionally notation may be omitted where it is apparent from context.
 
-Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have subscript $p$ (e.g.
+Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have right subscript $p$ (e.g.
 $\boldsymbol{x}_p$).
-Grid cells will be assigned a subscript $i$ or additionally $j$ denoting the grid index (e.g.
+Grid cells will be assigned a right subscript $i$ or additionally $j$ denoting the grid index (e.g.
 $\boldsymbol{x}_i$).
@@
 Occasionally summation over grid index $i$ and particles $p$ may be implied.
-A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
+A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
 \subsection{Einstein-Notation}
 Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors.
+
+Repeated indices, that are also defined on the variable, imply component-wise operations.
 Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors.
-Repeated indices that are also defined on the variable imply component-wise operations.
-$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
+$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$.
+Vector and tensor/matrix addition thus become:
 \begin{equation} \label{EINSTEIN:ADD}
@@
 \end{equation}
-Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
+Repeated indices, that are not otherwise defined, however imply summation on that index. The vector dot product becomes:
 \begin{equation}
@@
       { 0 } & {\text{if}} &  { i \neq j }
-  \end{array} \right.
+  \end{array}. \right
 \end{equation}
@@
 \end{equation}
-I.e.
-this component is only evaluated if $i = j$.
-In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$.
-The Kronecker Delta comes up in differentiating a variable by itself with different indexing:
+I.e. this component is only evaluated if $i = j$. The Kronecker Delta comes up in differentiating a variable by itself with same or different index:
 \begin{equation}
@@
 \section{Basics}\label{sec:basics}
+fill with nonsense
 \subsection{Reference and current configuration}\label{sec:ref_corr_config}
 Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
-Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t.
+Rather the continuum assumption holds: each particle represents a continuous piece of material s.t.
 a microscopic view does not need to be adopted.
@@
 
-When modelling solids changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
-In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
+When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
+In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
 This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
@@
 Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the reference configuration to the current configuration.
-If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
+If for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
 \begin{equation}
@@
 \end{equation}
-If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
+If, for instance, the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
 \begin{equation}\label{eq:rigid}
@@
 \end{equation}
-The velocity of a material point in $_0\boldsymbol{x}$ can be defined using this mapping
+The velocity of a material point in the reference configuration can be defined using this mapping
 \begin{equation}\label{eq:velocity}
-  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial {^t_0}\phi}{\partial t}(_0\boldsymbol{x},t)
+  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial ({^t_0}\phi)}{\partial t}(_0\boldsymbol{x},t),
 \end{equation}
@@
 \begin{equation}
-  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 {^t_0}\phi}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
+  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 ({^t_0}\phi)}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
 \end{equation}
@@
 It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed.
-The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
+The Eulerian description is static: Variables of particles moving by are measured while staying in a fixed position.
 
@@
 I.e.
-both configuration refer variables defined on them to the deformed state, but the position where the 'look up' of that value happens is different.
-In the reference configuration we look up at the initial position of the particle.
-In the current configuration we look up at the particle's 'world position'.
+both configurations refer variables defined on them to the deformed state but the position, where the 'look up' of that value happens, is different.
+In the reference confguration look up is done at the initial position of the particle.
+In the current configuration look up happens at the particle's world position.
 These relations for some particle quantity $f$ are called the (Lagrangian) pull back
@@
   \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
-  \frac { \partial  {_tf_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tf_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t )
-\end{equation}
-Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative:
+  \frac { \partial  {_tf_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tf_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t ).
+\end{equation}
+Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative in the current configuration:
 \begin{equation} \label{eq:material_derivative}
-  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } )
-\end{equation}
-The Jacobian of the deformation map $\phi$ is the deformation gradient $\boldsymbol{F}$ and is one of the key components to measure strain:
-\begin{equation}
-  {^t_0 F_{ij}} ( _0\boldsymbol {x} ,  t  ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , t ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , t )
-\end{equation}
-Intuitively it measures the local deformation of all particles in a small neighborhood of $_0B_\epsilon$ to $_tB_\epsilon$.
-Topology specifies the neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$.
+  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } ).
+\end{equation}
+The Jacobian of the deformation map $\phi$ is the deformation gradient $\boldsymbol{F}$ and is one of the key components to measure strain for material models:
+\begin{equation}
+  {^t_0 F_{ij}} ( _0\boldsymbol {x} ,  t  ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , t ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , t ).
+\end{equation}
+Topology specifies a neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$.
 $_0B_\epsilon$ becomes the pre-image of $_tB_\epsilon$ under $_0^t\phi$.
+Intuitively the deformation gradient measures the local deformation of all particles in a small neighborhood $_0B_\epsilon$ to $_tB_\epsilon$.
 This allows to describe infinitesimal changes in position from the reference to the current configuration
@@
 With this quantity in place volume and area changes are calculable.
-In a typical analytical fashion a coordinate system change $_0x \rightarrow {_tx}$ is done using the Jacobian (determinant) matrix.
+In a typical analytical fashion a coordinate system change $_0x \rightarrow {_tx}$ for a quantity $_0g$ is done using the determinant of the Jacobian matrix.
 The determinant is given a separate name $^t_0J =\text{det}(^t_0\boldsymbol{F})$.
-The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ is $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
+The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ thus becomes $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
 \begin{equation} \label{eq:volume_integral}
@@
 \end{equation}
-The area change is given by Nanson's Formula. Where $d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area:
-$$ d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}, \quad d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}} $$
+The area change is given by Nanson's Formula.
+$d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area and $d_t\boldsymbol{l} ={}^t_0\boldsymbol{F}d_0\boldsymbol{l}$ an arbitrary line element:
+$$ d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}},\quad d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}$$
 $$ \xRightarrow{\ref{eq:def_grad_pos},\ref{eq:j}} {^t_0J} d_0\boldsymbol{A} \cdot d_0\boldsymbol{l} = d{_t\boldsymbol{A}} \cdot ({^t_0\boldsymbol{F}}d{_0\boldsymbol{l}})$$
@@
 where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$.
-$d_0\boldsymbol{A}$,$d_t\boldsymbol{A}$ point in the direction of the surface normal of $\partial {_0B}(_0\boldsymbol{x})$, $\partial {_tB}(_t\boldsymbol{x})$, respectively.
+$d_0\boldsymbol{A}$, $d_t\boldsymbol{A}$ point in the direction of the surface normal of $\partial {_0B}(_0\boldsymbol{x})$, $\partial {_tB}(_t\boldsymbol{x})$, respectively.
 \cite{MIT:CONTINUUM_MECHANICS}
@@
 In equation \ref{eq:rigid} a rigid body movement was introduced.
+Let $b_i(t) = v_it$ be more generally some translation.
 A problem arises when calculating the deformation gradient of this equation.
-Let $b_i(t) = v_it$ be more generally a translation:
-\begin{equation}
-  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t)
+\begin{equation}
+  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t).
 \end{equation}
@@
 \begin{enumerate}
-  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-strain with quadratic components:
+  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-Lagrangian strain tensor with quadratic components:
     \begin{equation}\label{eq:green}
@@
     \end{equation}
-  \item \label{it:polar} Polar decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric positive definite) distortion $\boldsymbol{S}$ part $\boldsymbol{F}=\boldsymbol{R}\boldsymbol{S}$.
+  \item \label{it:polar} Polar decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric positive definite) distortional $\boldsymbol{S}$ parts $\boldsymbol{F}=\boldsymbol{R}\boldsymbol{S}$.
 \end{enumerate}
@@
 \end{equation}
-is already computed, where $\boldsymbol{U}$,$\boldsymbol{V}$ are orthogonal matrices and $\boldsymbol{\Sigma}$ is a diagonal matrix containing the singular values $\sigma_1 \geq \sigma_2 \geq ...
+is already computed where $\boldsymbol{U}$,$\boldsymbol{V}$ are orthogonal matrices and $\boldsymbol{\Sigma}$ is a diagonal matrix containing the singular values $\sigma_1 \geq \sigma_2 \geq ...
 \geq \sigma_r$ of $\boldsymbol{F}$.
@@
 \begin{enumerate}
-  \item Transform from the right singular vector space to standard basis space: $\boldsymbol{V}^T\boldsymbol{v}_1 = \boldsymbol{e}_1$
-  \item Scale by singular values to transform to principal stretch space: $\boldsymbol{\Sigma} \boldsymbol{e}_1 = \sigma_1 \boldsymbol{e}_1$
-  \item \label{it:tr_sing_vect} Transform to left singular vector space: $\boldsymbol{U}\sigma_1\boldsymbol{e}_1 = \sigma_1 \boldsymbol{u}_{1}$
+  \item Transform from the right singular vector space to standard basis space: $\boldsymbol{V}^T\boldsymbol{v}_1$ $ = \boldsymbol{e}_1$.
+  \item Scale by singular values to transform to principal stretch space: $\boldsymbol{\Sigma} \boldsymbol{e}_1 = \sigma_1 \boldsymbol{e}_1$.
+  \item \label{it:tr_sing_vect} Transform to left singular vector space: $\boldsymbol{U}\sigma_1\boldsymbol{e}_1 = \sigma_1 \boldsymbol{u}_{1}$.
 \end{enumerate}
@@
   \end{enumerate}
-This convention does not change the existence or uniqueness of the singular value or polar decomposition, although strictly speaking $\boldsymbol{S}$ in the Polar decomposition is not positive definite anymore.
+This convention does not change the existence or uniqueness of the singular value or polar decomposition although strictly speaking $\boldsymbol{S}$ is not positive definite anymore.
 \\
@@
     \begin{enumerate}[label*=\arabic*.]
-      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$, where $\boldsymbol{Q}$ is a Givens rotation aiming to eliminate off $S_{12}$.
+      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$ where $\boldsymbol{Q}$ is a Givens-Rotation aiming to eliminate off $S_{12}$.
 Store $\boldsymbol{V}^{(k+1)} = \boldsymbol{V}^{(k)}  \boldsymbol{Q}^{(k)}$.
@@
 Enforce the convention mentioned above.
-  \item \label{item:qr} \textbf{QR-Factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-Factorization with Givens Rotations, where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
+  \item \label{item:qr} \textbf{QR-Factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-Factorization with Givens-Rotations where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
 The $\boldsymbol{QR}$-Factorization is done once in the same fashion as in item \ref{item:sk} - \ref{item:sii}.
@@
 In item \ref{item:qr} it reduces to being diagonal.
-\cite{MPM:COURSE}
+\cite{MPM:COURSE}\cite{SVD:3x3}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{enumerate}
-  \item Conservation of angular momentum: $\sigma_{ij} = \sigma_{ji} \Rightarrow C_{ijkl} = C_{jikl} $
-  \item No-net-rotation: $\epsilon_{kl} = \epsilon_{lk} \Rightarrow C_{ijkl} = C_{ijlk}$
+  \item Conservation of angular momentum: $\sigma_{ij} = \sigma_{ji} \Rightarrow C_{ijkl} = C_{jikl}$.
+  \item No-net-rotation: $\epsilon_{kl} = \epsilon_{lk} \Rightarrow C_{ijkl} = C_{ijlk}$.
   \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\Psi$:
@@
 \end{enumerate}
-where item \ref{it:mixed_partials} holds for the strain energy density functional of an (hyper-)elastic material. The stress may then also be calculated by
+where item \ref{it:mixed_partials} holds for the strain energy density functional of an (hyper-)elastic material.
+This potential $\Psi$ is thus similar to potentials for example involved in gravitation, electrodynamics or fluids.
+The stress may then also be calculated by
 \begin{equation}\label{eq:partial_energy}
@@
 \end{equation}
-Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$ and $C_{22} = \mu$.
-$\gamma$ is also referred to as the engineering strain.
+Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$, and $C_{22} = \mu$.
+$\boldsymbol{\gamma}$ is also referred to as the engineering strain.
 Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\Psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:
@@
 \end{equation}
-\cite{MIT:LINEAR_ELASTICITY}
+\begin{flushright}\cite{MIT:LINEAR_ELASTICITY}\end{flushright}
 \subsubsection{Corotated Hyperelasticity}\label{sec:cor_hyper}
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
-  \boldsymbol{\epsilon} = \frac { 1 } { 2 } \left( \boldsymbol { F }_E + \boldsymbol { F }_E ^ { T } \right) - \boldsymbol { I }
-\end{equation}
-The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t.
-to $\boldsymbol{F}_E$: $\Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$.
+  \boldsymbol{\epsilon}_l = \frac { 1 } { 2 } \left( \boldsymbol { F }_E + \boldsymbol { F }_E ^ { T } \right) - \boldsymbol { I }.
+\end{equation}
+The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant w.r.t.
+to $\boldsymbol{F}_E$: $\Psi_{LE}(\boldsymbol{\epsilon}_l(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \Psi_{LE}(\boldsymbol{\epsilon}_l(\boldsymbol{F}_E))$.
 Rigid body motions however don't result in strain and consequently don't need to be recovered from.
 So energy should not change.
-Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:
-\begin{equation}
-  { \boldsymbol { \epsilon } } ( \boldsymbol { F }_E ) = \boldsymbol { \epsilon } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) = \frac { 1 } { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E + \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) ^ { T } \right) - \boldsymbol { I } = \boldsymbol { S }_E - \boldsymbol { I }
+Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$ an alternate strain measure may be defined as:
+\begin{equation}
+ \boldsymbol { \epsilon } \left( \boldsymbol { F }_E \right) = \boldsymbol { \epsilon }_l \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) = \frac { 1 } { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E + \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) ^ { T } \right) - \boldsymbol { I } = \boldsymbol { S }_E - \boldsymbol { I }.
 \end{equation}
@@
 \begin{equation}\label{eq:corot_S}
-  \Psi_{CH} = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )
+  \Psi_{CH} = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } ).
 \end{equation}
@@
 Equation \ref{eq:corot_sing} uses the diagonal matrix $\boldsymbol{\Sigma}_E$ containing the principal stretches typically acquired by the singular value decomposition $\boldsymbol{F}_E \boldsymbol{= U \Sigma}_E\boldsymbol{ V ^T }.$ Since the energy density in \ref{eq:corot_sing} is a function of only three (singular) values, which describe stretch/compression of the material, isotropy of the material is underlined.
-\cite{ADAMS:ELASTICITY}
+\begin{flushright}\cite{ADAMS:ELASTICITY}\end{flushright}
 \subsubsection{Fixed Corotated Hyperelasticity}
@@
 I.e.
-no elastic forces will be exerted, because the material is completely relaxed (again).
+no elastic forces will be exerted because the material is completely relaxed (again).
 
@@
 This is undesired behavior.
-Furthermore the primary contour can lead into energy kinks at extreme stretches leading to oscillations.
+Furthermore, the primary contour can lead into energy kinks at extreme stretches leading to oscillations.
 
@@
 \end{equation}
-This leads to the primary contour $J=1$, which does not cross the inverted region as only one singular value may be negative.
-\cite{MPM:INVERT}
+This leads to the primary contour $J=1$ which does not cross the inverted region as only one singular value may be negative.
+\begin{flushright}\cite{MPM:INVERT}\end{flushright}
 \subsection{Governing equations}
+fill with nonsense
 deriving the weak form
@@
 \begin{equation} \label{eq:density_pull_back}
-  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}
+  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}.
 \end{equation}
@@
 \end{equation}
-Pushing forward \ref{eq:euler_density_evol} with the result of \ref{eq:evol_jacobian} using the material derivative formulation \ref{eq:material_derivative}:
-\begin{equation}
-  \frac{D}{Dt}\rho(_t\boldsymbol{x},t) + \rho(_t\boldsymbol{x},t) \vec{\nabla} \cdot \boldsymbol{v}(_t\boldsymbol{x},t)=0
+Pushing forward \ref{eq:euler_density_evol} with the result of \ref{eq:evol_jacobian} using the material derivative formulation \ref{eq:material_derivative} leads to the Eulerian conservation of mass:
+\begin{equation}
+  \frac{D}{Dt}\rho(_t\boldsymbol{x},t) + \rho(_t\boldsymbol{x},t) \vec{\nabla} \cdot \boldsymbol{v}(_t\boldsymbol{x},t)=0.
 \end{equation}
@@
 \end{equation}
-Where the assumption due to angular momentum conservation $\boldsymbol{\sigma}^T  = \boldsymbol{\sigma} $ out of the second part of this section has already been made for completion.
-Beginning with a mix out of Lagrangian and Eulerian view
+Already assumed is angular momentum conservation $\boldsymbol{\sigma}^T  = \boldsymbol{\sigma} $ out of the second part of this section. Beginning with a mix out of Lagrangian and Eulerian view
 $$\left(\frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} \stackrel{\text{\ref{eq:eulerian_general}}}{=} \right)
@@
   \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{a} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}
-  = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x},
+  = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}
 \end{equation}
@@
 \end{equation}
-Summarized the Lagrangian view of the momentum equation with an initial momentum is ($^0_0J=1$):
+Summarized, the Lagrangian view of the momentum equation with an initial momentum is ($^0_0J=1$):
 \begin{equation}\label{eq:lagr_mom}
@@
 This can be shown as follows.
-The description of angular momentum follows that of linear momentum where product rule generally applies $\left(\frac{D_t\boldsymbol{x}}{Dt} \times {_t\boldsymbol{v}} = {_t\boldsymbol{v}} \times {_t\boldsymbol{v}} = 0\right)$ :
+The description of angular momentum follows that of linear momentum ($ _t\boldsymbol{x}\times \ref{eq:eul_transport}$) where product rule generally applies $\left(\frac{D_t\boldsymbol{x}}{Dt} \times {_t\boldsymbol{v}} = {_t\boldsymbol{v}} \times {_t\boldsymbol{v}} = 0\right)$ :
 \begin{equation}
@@
 as assumed before. Note that $\boldsymbol{P}$ however is not constrained to be symmetric.
+\begin{flushright}\cite{MPM:COURSE}\cite{MIT:CONTINUUM_MECHANICS}\end{flushright}
 \subsubsection{Weak formulation}\label{sec:weak}
 Before deriving the weak form an explanation of what it achieves is in order.
-
-The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain (and is as such influenced by the whole domain in general).
+The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain and is as such influenced by the whole domain in general.
 Such a solution can be found for simplified models analytically.
@@
 $ _0\boldsymbol{q}_h\neq 0$ on a very small subset of the whole domain $\Omega^0$.
-This is mostly used to gather information on a local neighborhood.
-
-The weak formulation requires an observation of the dot product. Consider the conservation of momentum in the strong form as formulated before (\ref{eq:lagr_mom}) as
+This is mostly used to gather information on a local neighborhood and not on the whole domain.
+
+The weak formulation requires an observation of the dot product. Consider the conservation of momentum in the strong form as formulated before \ref{eq:lagr_mom} as:
 \begin{equation}\label{eq:strong}
@@
 \end{equation}
-With the help of partial integration in multiple dimension the derivative moves over to the test-function.
+With the help of partial integration in multiple dimension, the derivative moves over to the test-function.
 Balancing out the order of derivatives, by moving a derivative to the test function, is the main motive of the weak form:
@@
 - \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
-d_0\boldsymbol{x}
+d_0\boldsymbol{x}.
 $$
@@
 \end{equation}
-In the material point method stress computations are more naturally done in the current configuration, or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
+In the Material-Point-Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
 Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
@@
 \subsection{Material Point Method}\label{sec:mat_point}
-The key idea of the material point method is to use (Lagrangian) particles as a consistent storage of material properties.
+The key idea of the Material-Point-Method is to use (Lagrangian) particles as a consistent storage of material properties.
 All stress based forces are computed on a Eulerian grid however.
@@
 After stresses are computed there also needs to be a way to get back the relevant information from the grid to the particles.
-After the information is transferred back the particles are moved.
-This advection is typically hard to do in Eulerian/FEM-like methods and cause of a lot of problems down the development pipeline.
+Advection is typically hard to do in Eulerian/FEM-like methods and cause of a lot of problems down the development pipeline.
 In a Lagrangian view particle advection is trivial.
@@
 Its is very important that the two transfers as well as the grid solver are in compliance with all governing equations.
-While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that also conserve the properties defined in them.
+While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that conserve the properties defined in them.
 \cite{MPM:COURSE}
@@
     \end{equation}
-  \item Non-negativity: $w \geq 0$. Negative weights can cause severe instability or non-physical behavior. \cite{gao2017adaptive}
+  \item Non-negativity: $w \geq 0$. Negative weights can cause severe instability or non-physical behavior, unlike in FEM where they are used. \cite{gao2017adaptive}
   \item Limited local support to reduce the number of discretizations, chapter \ref{sec:discretization}.
@@
     \end{array}
-  \right),
+  \right)
 \end{equation}
@@
 \subsubsection{Mass transfer}\label{sec:mass}
-Each material point will be assigned an initial volume $_0V_p$ as well as a initial mass $_0m_p$.
+Each material point will be assigned an initial volume $_0V_p$ as well as an initial mass $_0m_p$.
 The volume the material point occupies may change in time due to eq.
 \ref{eq:j}: ${^t_0}J _0V_p = {_tV_p}$.
-But due to conservation of mass \ref{eq:lagr_mass} it will have a constant, initial mass associated with it.
+But, due to conservation of mass \ref{eq:lagr_mass} it will have a constant, initial mass associated with it.
 I.e.
@@
 In doing so no information can be lost on the particles.
-Proving that the mass transfer to the grid is conserving is enough.
-The stress based-solver may manipulate this information further (mass-lumping strategies).
+Proving, that the mass transfer to the grid is conserving, is enough.
+The stress based-solver may manipulate this information further.
+Mass-lumping strategies typically will use this information directly, equation \ref{eq:mass_discr}.
 \begin{equation}
@@
 \end{enumerate}
-While $PIC$-Transfers are very stable they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
+While $PIC$-Transfers are very stable, they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
 This causes a heavy loss in angular momentum and velocity modes.
@@
 $$
-Bearing in mind that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion, where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
+Bearing in mind, that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
 In classical mechanics the angular velocity $\boldsymbol{\omega_p}$ can be then described using the inertia tensor $\boldsymbol{I}_p$ with the help of the angular momentum $\boldsymbol{L}_p$:
@@
 \end{enumerate}
-For a full proof that these transfers preserve linear and angular momentum consult \cite{MPM:APIC}.
+For a full proof, that these transfers preserve linear and angular momentum, consult \cite{MPM:APIC}.
 For the choice of dyadic products of cubic b-splines (\ref{eq:cubic_weights}) $\boldsymbol{D}_p$ takes on the simple form:
@@
 \end{equation}
-Assuming $\|\boldsymbol{x}_i^n -\boldsymbol{x}_p^n\| \leq \kappa h$, where $\kappa$ is determined by the interpolation stencil support (cubic 3D: $\kappa = 2\sqrt{3}$) and additionally assuming $\boldsymbol{D}_p^n = k\boldsymbol{I} \Rightarrow (\boldsymbol{D}_p^n)^{-1} = \frac{1}{k}\boldsymbol{I}$ (cubic 3D: $k=\frac{1}{3}h^2$).
-$\|\boldsymbol{v}^n_i\|_2$ can be estimated on the particles. Typically the number of particles is lower. Given Eq. \ref{eq:apic_mom_p2g}:
+Assuming $\|\boldsymbol{x}_i^n -\boldsymbol{x}_p^n\| \leq \kappa h$, where $\kappa$ is determined by the interpolation stencil support (cubic 3D: $\kappa = 2\sqrt{3}$) and additionally assuming $\boldsymbol{D}_p^n = k\boldsymbol{I} \Rightarrow (\boldsymbol{D}_p^n)^{-1} = \frac{1}{k}\boldsymbol{I}$ (cubic 3D: $k=\frac{1}{3}h^2$),
+$\|\boldsymbol{v}^n_i\|_2$ can be estimated on the particles. Typically the number of particles is lower. Given Eq. \ref{eq:apic_mom_p2g} this leads to the following estimate:
 $$
@@
 \subsection{Discretization}\label{sec:discretization}
-The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following for MPM preferable description:
+The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following, for MPM preferable, description:
 \begin{equation} \label{eq:weak_mpm}
@@
 \end{equation}
-The boundary integral is mostly due to collisions and will be ignored for now.
+The boundary integral is mostly due to collisions and will be ignored.
 \cite{MPM:OPTIMI_INTEGR} discusses level set collisions due to constraint collisions, object penalty collision and penalty self-collisions.
-An handling of collision would need to take them into account when solving the equation.
-A simple though less accurate method is to process collisions separately in a typical computer graphics manner, which is assumed for now.
+Collision treatment would need to be involved in the solving of the equation.
+A simple though less accurate method is to process particle collisions separately in a typical computer graphics manner which is assumed for now.
 \subsubsection{Discretize time}
@@
 \subsubsection{Discretize space}
-A Galerkin discretization brings all spatial terms of equation \ref{eq:time_disc} to a finite-dimensional space: $\boldsymbol{q} \rightarrow \boldsymbol{q}_h$.
+A Galerkin discretization brings all spatial terms of equation \ref{eq:time_disc} to a finite-dimen\-sion\-al space: $\boldsymbol{q} \rightarrow \boldsymbol{q}_h$.
 To not further clutter up the notation, the $h$ will be omitted.
@@
 \end{equation}
-Further chapter \ref{sec:weak} mentions that $\boldsymbol{q}$ can be chosen arbitrarily.
+Further, chapter \ref{sec:weak} mentions $\boldsymbol{q}$ can be chosen arbitrarily.
 The Galerkin discretization of a $d$-dimensional space with $m$ grid nodes therefore uses the standard basis functions $e_1,e_2, ...
@@
 Numerically this matrix is mostly not used as is due to possibility of it being singular.
-This is solved commonly due to a mass lumping strategy.
-Replace $m^n_{ii}$ with the $i$-th row sum to get:
+This is solved commonly due to a mass-lumping strategy.
+Replacing $m^n_{ii}$ with the $i$-th row sum and clearing all other elements leads to a diagonalization
 \begin{equation}\label{eq:mass_discr}
@@
 where partition of unity $\sum_j w_{jp} = 1$ is used. This is exactly the mass transfer as in eq. \ref{eq:mpm:mass}, s.t. no further assembling of a mass matrix is needed.
-Discretizing the right side of \ref{eq:galerkin} with an estimated per particle stress $\boldsymbol{\sigma}_p^{n+\lambda}$:
+The discretization of the right side of \ref{eq:galerkin} happens with an estimated per particle stress $\boldsymbol{\sigma}_p^{n+\lambda}$:
 \begin{equation}\label{eq:stress_discr}
@@
 \end{equation}
-The momentum change of the left side is by construction equal to a (grid node) force.\\
-Given that the material point method keeps track of the deformation by a deformation gradient.
-Each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
+The momentum change of the left side is by construction equal to a (grid node) force.
+
+Given, that the material point method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
 Based on this one may also gain a volume change measure around the particle as $J^n_p = \text{det}(\boldsymbol{F}^n_p)$.
@@
 \end{equation}
-In eq. \ref{eq:piola} an alternate measure for the stress by the first Piola Kirchhoff stress. The results of \ref{eq:space_discr} may therefore equally expressed by it:
+In eq. \ref{eq:piola} an alternate measure for the stress by the first Piola-Kirchhoff stress is given. The results of \ref{eq:space_discr} may therefore equally expressed by it:
 $$
@@
 \end{equation}
-\cite{MPM:COURSE} \cite{MPM:APIC} \cite{bathe2006finite}
+\begin{flushright}\cite{MPM:COURSE} \cite{MPM:APIC} \cite{bathe2006finite}\end{flushright}
 \subsubsection{Deformation gradient evolution}
@@
 The grid never actually gets deformed (unlike in FEM-methods).
-Therefore the discretized evolution of the deformation gradient  \ref{eq:evol_def_grad_disc} is directly a function of $\boldsymbol{\hat{x}}$.
+The discretized evolution of the deformation gradient  \ref{eq:evol_def_grad_disc} is directly a function of $\boldsymbol{\hat{x}}$.
 For the point $\boldsymbol{\hat{x}}_i^{n+1}$ this becomes:
@@
 \end{equation}
-This
-\cite{MPM:COURSE}
+\begin{flushright}\cite{MPM:COURSE}\end{flushright}
 \subsubsection{Grid nodal forces}
@@
 \end{equation}
-
 for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of eq. \ref{eq:force_disc}:
@@
 \end{equation}
-\cite{MPM:COURSE}\cite{MPM:APIC}
+\begin{flushright}\cite{MPM:COURSE}\cite{MPM:APIC}\end{flushright}
 \subsubsection{Symplectic midpoint scheme}
@@
 \end{equation}
-One can recast this method back to an energy function by integrating:
-\begin{equation}
-  E(\boldsymbol{v}_i) = \sum_i \frac{m_i^n}{8} \|\boldsymbol{v}_i - \boldsymbol{v}_i^n\|^2_2 + e(\boldsymbol{x}_i^n + \frac{\Delta t}{4}(\boldsymbol{v}_i + \boldsymbol{v}_i^n)),
+One can recast this method back to an energy function by integrating
+\begin{equation}
+  E(\boldsymbol{v}_i) = \sum_i \frac{m_i^n}{8} \|\boldsymbol{v}_i - \boldsymbol{v}_i^n\|^2_2 + e(\boldsymbol{x}_i^n + \frac{\Delta t}{4}(\boldsymbol{v}_i + \boldsymbol{v}_i^n))
 \end{equation}
 where $e$ is just the discretized elastic potential energy out of \ref{eq:energy_discr}.
-In general this is a optimization objective that needs minimizing to solve for the updated velocities.
-Recasting allows solving in a general scheme by an optimization integrator.
+In general this is an optimization objective that needs minimizing to solve for the updated velocities.
+Recasting allows to solve the energy objective in a general manner by an optimization integrator.
 
@@
 The first term can be identified clearly as the kinetic energy.
-Moreover it easily allows adding potential terms for gravity or collision \cite{MPM:OPTIMI_INTEGR}.
-Or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
-The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
+Moreover, it easily allows adding potential terms for gravity or collisions \cite{MPM:OPTIMI_INTEGR}, or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
+The analysis of minimizing the objective is equivalent to findino the zero crossing of the derivative:
 \begin{equation}
@@
 \end{equation}
-Since the objective is minimizing the scale of $\boldsymbol{E}$ can be chosen arbitrarily, i.e. a zero crossing does not scale. The scaling here can be identified as:
+Since the objective is minimizing, the scale of $\boldsymbol{E}$ can be chosen arbitrarily. I.e. a zero crossing does not scale. The scaling here can be identified as:
 \begin{equation}
@@
 \end{equation}
-\cite{MPM:COURSE}\cite{MPM:APIC}
+\begin{flushright}\cite{MPM:COURSE}\cite{MPM:APIC}\end{flushright}
 \subsubsection{Newton's Method}\label{sec:newton}
@@
 \end{equation}
-$\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}$ is symmetric, positive definite due it being the Hessian of a convex energy function $E$ with a hyper-elastic material.
-The linear system of eq. \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is to important things notice:
+$\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}$ is symmetric, positive definite due it being the Hessian of a convex energy function $E$ of a hyper-elastic material.
+The linear system of eq. \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is two important things to notice:
 \begin{enumerate}
   \item  One may want to use a preconditioner $\boldsymbol{P}^{-1}$ to reduce iteration times.
-However conservation of momentum on incomplete convergence is fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $\boldsymbol{M}^{-1}$ \cite{MPM:APIC}.
+However, conservation of momentum on incomplete convergence is fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $\boldsymbol{M}^{-1}$ \cite{MPM:APIC}.
 Thus the equation becomes:
@@
     \end{equation}
-  \item Due to \ref{eq:hessian-matrix-vector} the product of the Hessian with an increment ($\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j$) needs to be computed anew every step, since a pure matrix store is too costly.
+  \item Due to \ref{eq:hessian-matrix-vector} the product of the Hessian with an increment ($\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j$) needs to be computed anew every step since a pure matrix store is too costly.
 \end{enumerate}
@@
 \begin{enumerate}
-  \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid to particle transfer:
+  \item Compute a particle quantity $\boldsymbol{A}_p$ with a Grid-to-Particle-Transfer:
     \begin{equation}\label{eq:Ap}
@@
     \end{equation}
-  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle to grid transfer:
+  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a Particle-to-Grid-Transfer:
     \begin{equation}\label{eq:delta_f}
@@
 \end{enumerate}
-%Optimally one would also try to derive a model-dependent preconditioner $\boldsymbol{P}$ for $\frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j}$ to reduce iterations of the conjugate gradient method:
-%\begin{equation}
-%\boldsymbol{P}^{-1}\boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} = \boldsymbol{P}^{-1} \boldsymbol{M}^{-1}\boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right).
-%\end{equation}
-%One of those is the Jacobi (diagonal) preconditioner: $\boldsymbol{P} = \text{diag}\left( \boldsymbol{M}^{-1}\left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \right)$.
 \subsection{General-purpose computing on graphics processing units(GPGPU)}
+fill with nonsense
 \subsubsection {GPGPU Optimization opportunities}\label{sec:optim}
@@
 
-Therefore one should prefer register memory for intra-thread caching.
+Therefore, one should prefer register memory for intra-thread caching.
 Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block.
@@
 This includes doing calculations on the GPU that may run faster on the CPU.
-The API may also stall if the CPU is not issuing instructions to the GPU.
-This may be due to calculations on the CPU, or even worse calculations depending on GPU results.
-In this thesis the physics already runs on the GPU.
-Additionally we try to avoid transfers wherever possible.
+The API may also stall if the CPU is not issuing instructions fast enough to the GPU.
+This may be due to calculations on the CPU or, even worse, calculations depending on GPU results.
+This thesis tries to avoid PCI-transfers wherever possible.
 
@@
 NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront.
-(Memory) operations are issued per warp/wavefront.
-A memory access uses the L2 caching behavior.
-A L2 cache-line is 32 Bytes long.
+Generally they are called subgroups.
+All operations are issued per subgroup.
+A memory access uses the L2-caching behavior.
+A L2-cache-line is 32 bytes long.
 Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.
-If all threads of a warp access memory within this cache-line only one memory request of 32 bytes needs to be made.
+If all threads of a warp access memory within this cache-line, only one memory request of 32 bytes needs to be made.
 In contrast, if a warp accesses N different cache-lines, N*32 bytes need to be loaded.
 
-The unified L1/texture cache acts as a coalescing buffer for threads of a warp (Maxwell architecture \cite{NVIDIA:MAXWELL}).
+The unified L1/texture-cache acts as a coalescing buffer for the thread of a subgroup (Maxwell architecture \cite{NVIDIA:MAXWELL}).
 It does not matter in which order the cache-line is accessed by the warp.
@@
 On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses.
-L2 is a write-back cache thus this assessment also qualifies for global memory writes.
+L2 is a write-back cache.
+Thus this assessment also qualifies for global memory writes.
 \cite{NVIDIA:MEMORY_BAND}
@@
 The hardware groups together unique memory accesses on these banks.
-E.g.
-if two threads within a warp access bank 17 a bank conflict occurs and the access needs to be split into two groups.
-The one exception is a broadcast: all threads within a warp access the same bank.
-The number of banks is equal to the warp size, s.t.
+E.g., if two threads within a subgroup access bank 17, a bank conflict occurs, and the access needs to be split into two groups.
+The one exception is a broadcast: all threads within a subgroup access the same bank.
+The number of banks is equal to the warp size on the GTX970 s.t.
 throughput is maximized if every thread does a unique bank access.
@@
 The available number of registers per thread is calculable with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
-If the occupancy falls off to 0\% register spilling occurs.
+If the occupancy falls off to 0\%, register spilling occurs.
 The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor.
-It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another; this is called a warp stall, table \ref{table:metric}.
+It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another.
+This is called a warp stall, table \ref{table:metric}.
 Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator.
-Non-optimal occupancy might not have any performance impact at all.
+Non optimal occupancy might not have any performance impact at all.
 \cite{NVIDIA:ACHIEVED_OCCUPANCY}
@@
 The more threads of a block reuse shared memory data the faster the memory throughput will be.
-Therefore operators which involve filtering (also MPM-transfers) can use this behavior.
+Therefore, operators which involve filtering (also MPM-transfers) can use this behavior.
 

### 2019-01-25 mscthesis.tex
optimization chapter finished

@@
 
-To our knowledge there is no way to find out the exact number of registers an OpenGL shader uses.
+There is no way to find out the exact number of registers an OpenGL shader uses.
 The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
@@
 TEMP variables are registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages there is however no guarentees these are not further optimized.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentees these are not further optimized.
+Otherwise close monitoring of the occupancy is recommended.
 All other metrics discussed are readily available for OpenGL.
@@
 \textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits.
-Though we in general had better experiences starting with a block size of 64 as can be seen in the Occupancy calculator.
+One should start however with a block size of at least 64 as can be seen in the Occupancy calculator.
 Increasing the block size is useful in combination with shared memory usage.
-Shared memory access is typically 100x faster than uncached global memory accesses.
+Shared memory accesses are typically faster than global memory accesses.
 The more threads of a block reuse shared memory data the faster the memory throughput will be.
@@
 
-\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it.
-Otherwise merging sequential and parallel work improves performance heavily.
-Since the block size is limited by the GPU large amounts of data need to be split in tree structures to process in parallel.
-One can not avoid an indirection of writing-back and loading again from global memory.
+\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure, if the algorithm allows it.
+Otherwise merging sequential and parallel work improves I/O latency hiding.
+Since the block size is limited by the GPU, some algorithms need to be split in tree structures to be processed.
+One can not avoid an indirection of writing back and loading again from global memory.
 Therefore algorithms often improve much further from more sequential work to avoid this indirection.
 
-\textbf{Loop unrolling:} Foremost loop unrolling reduces instructions and can thus speed up compute-bound work.
-Moreover loop unrolling might help the compiler reorder instructions and thus execute code faster.
+\textbf{Loop unrolling:} Foremost, loop unrolling reduces instructions and can thus speed up compute-bound work.
+Moreover, loop unrolling might help the compiler reorder instructions and thus execute code faster.
 Instructions benefiting greatly from this are global loads.
@@
 Dynamic loops can however be partially unrolled.
-In the case of ray-marching this would for instance fetch two samples at once, although it might process only one.
+In the case of ray-marching this would for instance fetch two samples at once although it might process only one.
 In short, this merges two consecutive loop iterations.
@@
 Mind however that loop unrolling can increase register pressure.
-More registers need to store and process the relevant code, where a dynamic loop reuses the same registers for any iteration.
+More registers need to stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.
 

### 2019-01-26 mscthesis.tex
rest checked + abstract

@@
 \newenvironment{code}{\captionsetup{type=algorithm}}{}
+
+\setcounter{biburllcpenalty}{7000}
+\setcounter{biburlucpenalty}{8000}
+
 \begin{document}
@@
 \selectlanguage{english}
+
+\vspace*{\fill}~
+\begin{center}
 \subsubsection*{Abstract}
-
-\noindent Physic
-
-%\vfill~
-%ggf. noch'n \vfill~
-
+\end{center}
+The Material-Point-Method is a technique allowing for physically accurate simulations.
+It has found its way into computer graphics and since then rapidly expanded.
+The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.
+It has found its way into computer graphics and since then rapidly expanded.
+
+\noindent Core of this process are the grid and particles transfers that interpolate between the two structures.
+These transfers are executed multiple times per frame so preprocessing steps might be taken if they are outweighed.
+
+\noindent The Method however suffers from high execution times and make the method only viable for hero shots.
+The Material-Point-Method including the transfers are highly parallelizable.
+As such this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.
+
+\noindent Deep sorting with counting sort increases coalescing and L2 hit rates.
+Binning allows for shared memory filtering techniques using blocking.
+As another preprocessing step only grid blocks are executed which have particles in them.
+All transfers do not rely on fixed size bins.
+\vfill
 \clearpage
@@
 \subsubsection{Conservation of mass}
-Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$ and its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
+Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$. Similarly, let its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
 After \ref{eq:volume_integral} they are related as:
@@
 \end{enumerate}
-\subsection{General-purpose computing on graphics processing units(GPGPU)}
+\subsection{General-purpose computing on graphics processing units}
 fill with nonsense
@@
 Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block.
-Global memory should be mostly used at the start and end of a shader.
+Global memory is the persistent storage on the GPU.
 
 \textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers.
-The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s.
-While the communication between device and GPU is 224GB/s (GTX 970).
+The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s while the communication between device and GPU is 224GB/s (GTX 970).
 This includes doing calculations on the GPU that may run faster on the CPU.
-The API may also stall if the CPU is not issuing instructions fast enough to the GPU.
+The API may also stall the GPU if the CPU is not issuing instructions fast enough to the GPU.
 This may be due to calculations on the CPU or, even worse, calculations depending on GPU results.
@@
 NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront.
-Generally they are called subgroups.
+In general they are called subgroups.
 All operations are issued per subgroup.
@@
 On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses.
-L2 is a write-back cache.
+L2 is a write back cache.
 Thus this assessment also qualifies for global memory writes.
@@
 E.g., if two threads within a subgroup access bank 17, a bank conflict occurs, and the access needs to be split into two groups.
-The one exception is a broadcast: all threads within a subgroup access the same bank.
+The one exception is a broadcast: All threads within a subgroup access the same bank.
 The number of banks is equal to the warp size on the GTX970 s.t.
-throughput is maximized if every thread does a unique bank access.
+throughput for 32 floats is maximized if every thread does a unique bank access.
 \cite{NVIDIA:BEST:PRACTICE}
 
-\textbf{Register spilling:} One should avoid local memory.
+\textbf{Register spilling:} Local memory should be avoided.
 Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
-The available number of registers per thread is calculable with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}.
+The available number of registers per thread is calculable with the CUDA Occupancy Calculator \cite{NVIDIA:OCCUPANCY} on NVIDIA hardware.
 If the occupancy falls off to 0\%, register spilling occurs.
-The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor.
-It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another.
+The occupancy is either given absolute due the amount of active warp or relative divided by the possible maximum amount of warps on a Streaming Multiprocessor.
+Occupancy allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.), the warp scheduler will switch to another.
 This is called a warp stall, table \ref{table:metric}.
-Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator.
-Non optimal occupancy might not have any performance impact at all.
+Occupancy is mostly affected due to resource limits on Streaming Multiprocessors (shared memory, register), as can be seen in the Occupancy Calculator.
+Non optimal occupancy might not have any performance impact at all when the GPU has enough warps to switch to at all times.
 \cite{NVIDIA:ACHIEVED_OCCUPANCY}
 
-There is no way to find out the exact number of registers an OpenGL shader uses.
+There is no exact way to find out the exact number of registers for any OpenGL shader.
 The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
-It will output an intermediate language of one the families of \textit{NV\_gpu\_programX} (where $X$ stands for the version).
+It will output an intermediate language of one of the families of \textit{NV\_gpu\_\-programX} (where $X$ stands for the version).
 TEMP variables are registers.
 Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentees these are not further optimized.
-Otherwise close monitoring of the occupancy is recommended.
+Otherwise, close monitoring of the occupancy is recommended.
 All other metrics discussed are readily available for OpenGL.
@@
 \textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits.
-One should start however with a block size of at least 64 as can be seen in the Occupancy calculator.
+One should start however with a block size of at least 64, as can be seen in the Occupancy Calculator.
 Increasing the block size is useful in combination with shared memory usage.
 Shared memory accesses are typically faster than global memory accesses.
-The more threads of a block reuse shared memory data the faster the memory throughput will be.
+The more threads of a block reuse shared memory data, the faster the memory throughput will be.
 Therefore, operators which involve filtering (also MPM-transfers) can use this behavior.
 
-\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure, if the algorithm allows it.
-Otherwise merging sequential and parallel work improves I/O latency hiding.
+\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it.
+Otherwise, merging sequential and parallel work improves I/O latency hiding.
 Since the block size is limited by the GPU, some algorithms need to be split in tree structures to be processed.
-One can not avoid an indirection of writing back and loading again from global memory.
-Therefore algorithms often improve much further from more sequential work to avoid this indirection.
+An indirection cannot be avoided for writing back and loading again from global memory.
+Therefore, algorithms often improve much further from more sequential work to avoid this indirection.
 
@@
 The latency of the second instruction might be partly hidden by the first one and thus speed up the work load.
-Mind however that loop unrolling can increase register pressure.
-More registers need to stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.
+Mind however, that loop unrolling can increase register pressure.
+More registers need to be stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.
 
@@
 \begin{enumerate}
-  \item For the GPU most interesting is the scalability with increasing input data sizes.
+  \item For the GPU most interesting is the scalability with increasing input data sizes. All algorithms here however scale linear with the input size.
   \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
@@
     \end{equation}
-    Otherwise one would calculate the speedup of procedure 1 over procedure 2 for a direct comparision:
+    Otherwise, one would calculate the speedup of procedure 1 over procedure 2 for a direct comparison:
 
@@
     $\Delta t_c$      & -                     & computing time\\\hline
-    VRAM SOL\%        & 100                   & memory througput(, w.r.t. to hw.-limit)\\\hline
+    VRAM SOL\%        & 100                   & memory througput w.r.t. to hw.-limit\\\hline
     SM SOL\%          & 100                   & instruction throughput\\\hline
@@
     L2-Hitrate\%      & 100                   & L2 hit-miss ratio\\\hline
-    SM Active\%       & 100                   & at least one warp active avg. over SMs\\\hline
+    SM Active\%       & 100                   & one warp active avg. over SMs\\\hline
     SM Issue Util.\%  & 100                   & amount of cycles an instr. was issued\\\hline
@@
 \end{table}
-The other GPU metrics are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}. An active warp is also one that waits for memory.
+The other GPU metrics in table \ref{table:metric} are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}.
 
 None of these metrics necessarily have a correlation to computing time.
-However they are good indicators on which operations/hardware units the GPU spends its time on.
+They are however good indicators on which operations/hardware units the GPU spends its time on.
 The top SOL\% metric measures fraction of the throughput to the bandwidth of that hardware unit.
-NVIDIA aims to get the top SOL\% metric between 60\% < top SOL\% < 80\%.
+NVIDIA aims to get the top SOL\% metric between 60\% and 80\%.
 Occupancy was discussed in \ref{sec:optim}.
@@
 
-Aiming to get shaders memory-bandwidth limited while minimizing the amount of needed accesses for any given workload is the top-priority for simple tasks.
-Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% < top SOL\% < 80\%.
+Aiming to get shaders memory-bandwidth limited, and minimizing the amount of needed accesses for any given workload, is the top priority for simple tasks.
+Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% and 80\%.
 The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.
 
-The metric SM Issue Utilization (per active cycle) is important for compute-bound work as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data.
+The metric SM Issue Utilization (per active cycle) is important for compute-bound work, as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data.
 This metric tells if the SM can issue instructions fast.
 NVIDIA also aims between 60\% to 80\% for this metric if the shader is compute-bound.
-Furthermore for compute-bound work often the top warp stall reason is waiting on instruction fetches.
+Furthermore, for compute-bound work often the top warp stall reason is waiting on instruction fetches.
 
@@
 Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts.
-The latter is the more commonly taught approach, while high performance CPU code may also prefer SoA or the more advanced entity component systems.
+The latter is the more commonly taught approach while high performance CPU code may also prefer SoA or the more advanced entity component systems.
 This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.
 
-In short one might not access every variable of a particle in a shader.
+In short, one might not access every variable of a particle in a shader.
 I.e.
-in physics one has a pipeline of different passes each requiring another subset of variables.
-Assume for instance one shader only requires the position then it is more favorable for consecutive threads to load consecutive position vectors.
-The cache-line is then filled with position vectors, which all will get used.
-In contrast in an AoS layout the cache-line will have a position and a velocity loaded, but the shader never uses the velocity, halfing the throughput.
-This process is show in Figure \ref{fig:soa} and \ref{fig:aos}.
+in most compute work one has a pipeline of different passes each requiring another subset of the variables.
+Assume for instance, one shader only requires the position.
+Then it is more favorable for consecutive threads to load consecutive position vectors.
+The cache-line is then filled with position vectors which all will get used.
+In contrast, in an AoS layout the cache-line will have a position and a velocity loaded but the shader never uses the velocity.
+This halfes the throughput.
+This process is shown in Figure \ref{fig:soa} and \ref{fig:aos}.
 
@@
 
-Furthermore we experienced that for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}.
-As an alternative we had to provide meta information in the style of reflection macros and collapse the container, Listing \ref{llst:opengl}.
-We can then access SoA by memory offset and AoS by variable or by memory offset for a more abstract view of the memory.
-We also share the AoS setup between C++ and OpenGL code.
+Furthermore, for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}.
+As an alternative meta information is provided in the style of reflection macros.
+Therefore, the container is collapsed as in Listing \ref{llst:opengl}.
+Memory is then accessible by SoA by memory offset and by AoS by variable or by memory offset for a more abstract view of the memory.
+The AoS setup is shared between C++ and OpenGL code.
 One could also build a shader using the abstract layout to convert from SoA to AoS.
+For more coherency converting the whole process to reflection macros is sensible.
 
@@
 \end{enumerate}
-For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}.
-Except for the omission of the glsl version or extension macros, we did not find any problems in this process.
+For debugging the shader is output with all processed includes.
+The gcc-pre\-pro\-ces\-sor can preprocess the shader with options: \texttt{gcc -E -x c -P}.
+Except for the omission of the glsl version or extension macros no problems were encountered in this process.
 
@@
 
-\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$, parallel reduction computes:
+\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$ reduction computes:
 \begin{equation}
@@
 \end{equation}
-Our OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference.
+This OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference.
 To avoid memory padding in shared memory the addressing scheme is given by the sequential addressing scheme in Figure \ref{fig:sequential}.
@@
   \caption{Sequential addressing for \texttt{vec4}s.
-Even threads are blue, odd are yellow.
+Even threads are blue.
+Odd threads are yellow.
 Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers.
@@
 The first and all following instructions can be handled fully concurrently.
-No memory padding required.}
+No memory padding is required.}
 \label{fig:sequential}
@@
 
-Parallel reduction schemes using shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}.
-They use registers instead of shared memory and thus do not need memory padding.
-\cite{NVIDIA:SHUFFLE}
-
-\cite{NVIDIA:PARALLEL_REDUCTION} also makes heavy use of combining parallel and sequential work.
+Parallel reduction schemes using warp shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}
+
+\cite{NVIDIA:PARALLEL_REDUCTION} also make use of combining parallel and sequential work.
 Every thread adds a multiple of two sequential adds into shared memory.
@@
 
-Parallel reduction solves assuming a \texttt{unary\_op(x)} directly after memory load:
+Parallel reduction solves, assuming a \texttt{unary\_op(x)} directly after memory load, the following problems in MPM:
 \begin{itemize}
-  \item CFL-condition of chapter \ref{sec:cfl} with operators
+  \item CFL-condition of chapter \ref{sec:cfl} with operators,
   \begin{itemize}
@@
   \end{itemize}
-\item Discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
+\item discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
 \begin{itemize}
@@
   \end{itemize}
-\item Calculate the maximum count of particles of all grid nodes within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
+\item calculate the maximum count of particles of all bins within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
 \begin{itemize}
@@
 
-\textbf{Scan:}  Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op} and an array of values $\left[a_0,a_1, ...
-,a_n\right]$, an exclusive scan computes the array:
+\textbf{Scan:}  Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ...
+,a_n\right]$ an exclusive scan computes the array:
 \begin{equation}
@@
 The interleaved addressing pattern has the benefit over the sequential one of already computing the partial scans for all odd indices (although misaligned).
-Note that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}.
+Note, that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}.
 A full scan is then computed as \cite{NVIDIA:SCAN}:
@@
   \item Inserting the neutral element into the last element of the array.
-  \item Downsweep-Phase: Reverse execution order of item \ref{it:up-sweep} (inverted tree), and an additional right to left element swap.
+  \item Downsweep-Phase: Reverse execution order of item \ref{it:up-sweep} (inverted tree), and in addition overwrite the left element with the right one's.
 \end{enumerate}
 Again, sequential work can be combined with parallel work.
-\cite{NVIDIA:SCAN_MODERN} name this process: Raking.
-Instead of loading one element per thread, one thread loads n elements and computes a sequential partial scan in register memory.
+\cite{NVIDIA:SCAN_MODERN} name the process: Raking.
+The general technique is also called register blocking.
+Instead of loading one element per thread, one thread loads N elements and computes a sequential partial scan in register memory.
 The reduced last value of the array is written into shared memory and the parallel scan begins.
-Afterwards the parallel scan result is spread out with the sequential partial scan.\cite{NVIDIA:SCAN_MODERN}
+Afterwards the parallel scan result is spread out(raking) with the sequential partial scan.\cite{NVIDIA:SCAN_MODERN}
 \begin{figure}[htbp]
@@
 where $p$ is the padded memory offset, $u$ is the unpadded memory offset and $n$ is the number of banks.
-For a block size of 1024 one would need additional padding for a full conflict-free access:
+For a block size of 1024 additional padding is needed for a full conflict-free access:
 \begin{equation}\label{eq:bank_1024}
@@
 \end{equation}
-However the added computation time does not outweigh the bank-conflicts.
-So one reverts to eq.
+However, the added computation time does not outweigh the bank-conflicts.
+So the process is reverted to eq.
 \ref{eq:bank}.
-Note that this is a correction to \cite{NVIDIA:SCAN} where C++ Operator Precedence of $+$ over $<<$ is not recognized and additionally incorrectly $ u << n$ is used in the first part of eq.
+Note, that this is a correction to \cite{NVIDIA:SCAN} where C++-Operator-Precedence of $+$ over $<<$ is not recognized, and additionally incorrectly $ u << n$ is used in the first part of eq.
 \ref{eq:bank_1024}.
@@
 This implementation did not end up using this more closely related approach to \cite{NVIDIA:NNSEARCH} as it would commit shared memory bank-conflicts defying the architecture of the GPU.
+It also puts restriction on the maximum allowed particles per block.
 Instead preferred is batching a fixed amount of particles, more in chapter \ref{sec:transfers}.
@@
 To support bigger input sizes pyramid structures are in use to process the elements the blocks reduced.
-For our case we mostly reduce the input size s.t.
+In this implementation the input size is mostly reduced s.t.
 the first level above in the pyramid can be computed within a single block.
 For the scan this result is then written back.
-This was enough for the input sizes we employed and can be similarly extended.
+This was enough for the input sizes we employed and can be similarly extended by pyramid structures.
 
 \subsection{Counting Sort \& Stream Compaction}\label{sec:sort}
-Sorting on the GPU can dramatically increase workload performance of subsequent steps. If they can profit from following things:
+Sorting on the GPU can dramatically increase workload performance of subsequent steps if they can profit from following things:
 \begin{enumerate}
   \item Accesses can now be handled in a coalesced fashion. \label{it:sort_1}
-  \item Data can now be reused due to L2-Cache and shared memory. \label{it:sort_2}
+  \item Data can now be reused due to L2-Cache and/or shared memory. \label{it:sort_2}
 \end{enumerate}
@@
 Supported is a full deep copy of all particle variables.
-A deep copy profits from item \ref{it:sort_1} and \ref{it:sort_2}, but takes the hit in performance of uncoalesced accesses for all variables.
-Or, alternatively an indexing to access an existing particle buffer in the specified sorted manner.
+A deep copy profits from item \ref{it:sort_1} and \ref{it:sort_2} but takes a hit in performance of uncoalesced accesses for all variables.
+Alternatively, an indexing to access an existing particle buffer in the specified sorted manner is although supported.
 Indexing only profits from item \ref{it:sort_2} and subsequent shaders will be uncoalesced.
@@
 The applicable data is checked by a condition.
-There are a lot of cases where stream compaction can reduce running times.
-In chapter \ref{sec:rel_work} particle activation is mentioned as one of such, chapter \ref{sec:active_blocks} will reduce to active blocks.
+There are a lot of cases where stream compaction can reduce running times: In chapter \ref{sec:rel_work} particle activation is mentioned as one of such.
+Chapter \ref{sec:active_blocks} will reduce active blocks with an indexed stream compaction.
 Collision treatment could be implemented as an operation following a collision detection etc.
@@
 \end{wrapfigure}
-\textbf{Binning}: Transform the particle position into the grid space, such that every integer in range corresponds to a grid node.
+\textbf{Binning}: Transform the particle position into the grid space s.t.
+every integer in range corresponds to a grid node.
 Bin due to flooring the transformed particle: Store the particle count $c_i$ of this grid node on the particle as an offset $o_p$ into this grid node.$$o_p := c_i.$$Following that, up the counter on the grid by one:$$c_i := c_i +1.$$
@@
   \item \textbf{Reordering}: $i_p := s_i + o_p$ will now lead to a new ordering of the particle indices.
-Sorting is now similar to methods \texttt{sort\_by\_key(uint i)}, where ordering of particles within the same bin is undefined.
+Sorting is now similar to methods \texttt{sort\_by\_key(uint i)} where ordering of particles within the same bin is undefined.
 \end{enumerate}
@@
 A naive solution to the problem are atomic accesses.
-As long as the (atomic) writes can be handled coalesced this process can be done fast by the GPU, chapter \ref{sec:eval}.
-This is only the case for a full deep sort.
-Therefore binning in the case of a full sort acts upon the last sorted state.
+As long as the (atomic) writes can be handled coalesced, this process can be done fast by the GPU, chapter \ref{sec:eval}.
+This is however only the case for a full deep sort.
+Therefore, binning in the case of a full sort acts upon the last sorted state.
 Double buffering particles allows to do so.
@@
 
-Alternatively one could work again with the CFL-condition. Since particles move at most one bin in the 27-neighborhood one could mimic a Particle-to-Grid transfer, chapter \ref{sec:p2g}:
+Alternatively, one could work again with the CFL-condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a Particle-to-Grid-Transfer, chapter \ref{sec:p2g}:
 \begin{enumerate}
@@
 
-Alternatively one could choose a closer approach to \cite{NVIDIA:NNSEARCH}'s neighborhood search, which greedily loads all particles.
-Ordering can easily be preserved then.
+Alternatively, one could choose a closer approach to \cite{NVIDIA:NNSEARCH}'s neighborhood search which greedily loads all particles. Ordering can easily be preserved then.
+
 Note that the support in their approach is smaller (27 instead of 64 nodes) and the data required per particle is only the label.
-
-Chapter \ref{sec:active_blocks} shows how to do this only for active blocks. Due to the difference in support the same pipeline cannot be used.
-
-In summary there is little gains to do so as binning is only executed once per step and is not the dominant performance cost.
+Chapter \ref{sec:active_blocks} shows how to do this only for active blocks.
+Due to the difference in support the same pipeline cannot be used.
+
+In summary, there is little gains to do so yet as binning is only executed once per step and is not the dominant performance cost.
 
@@
 $$\square_i = \square_i \circ \square_i \circ ... \circ \square_i.$$
-One thread will just equal to one node or particle respectively.
-For one variable operations mostly used is a map-shader that applies a unary operation to any input element.
+One thread will just equal one node or particle, respectively.
+For one variable assignments a map-shader is used that applies a unary operation to any input element.
 This can be used for instance for the mass divide of momentum on the grid (equation \ref{eq:mass_divide}) or resetting buffers that are target of global atomic operations.
 
-In contrast much harder are transfers between the two structures, that are essential to the Material-Point-Method:
-
-Particle-to-Grid(P2G)-transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid; as well as the computation of the Hessian of equation \ref{eq:delta_f}.
+In contrast much harder are transfers between the two structures that are essential to the Material-Point-Method:
+
+Particle-to-Grid(P2G)-Transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid, as well as the computation of the Hessian of equation \ref{eq:delta_f}.
 In general these can be summarized as:
@@
 
-The transfers back to the particles are called Grid-to-Particle(G2P)-transfers.
-Apart from the APIC-transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}) the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
-In short all transfers of the form:
+The transfers back to the particles are called Grid-to-Particle(G2P)-Transfers.
+Apart from the APIC-Transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}), the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
+In short, all transfers of the form:
 $$\square_p = \sum_i \square_i \circ \square_{ip}.$$
+This concludes all necessary operations for the MPM-Method as derived.
 
 The number of transfers per frame is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$).
-This motivates to make the transfers a very highly optimized operation.
+This motivates making the transfers a very highly optimized operation.
 The Material-Point-Method requires for an elastic material as derived:
@@
 \end{enumerate}
-In summary this is a lot of data to be transferred between grid and particles, which will only increase with more complex models.
-In contrast any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame.
-After that it helps each transfer running more performant.
-Therefore the transfers are the main focus of this implementation.
+In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models.
+In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame.
+Each transfer can profit from it.
+Therefore, the transfers are the main focus of this implementation.
 
@@
 Nevertheless the number of particles within a cell is still variable.
-Nevertheless all methods presented in the following are not restricted to bins of a fixed size of particles.
+All methods presented in the following are not restricted to bins of a fixed size of particles.
 
@@
 Blocks map well to the thread group nature of GPUs.
-A single thread within a thread group thus corresponds to a grid node within that block.
+A single thread within a thread group thus corresponds to a grid node within that block, respectively.
+
 \subsubsection{Grid-to-Particle transfers}\label{sec:g2p}
 The straightforward solution to this problem is to handle all particles independently from each other.
-One particle adds up the contribution of each grid node variable in range of the interpolation function.
+One particle adds up the contributions from each of the grid nodes in range of the interpolation function.
 In three dimensions this can be done for instance with a two times nested for-loop within a thread.
-Luckily grid nodes in a uniform grid are inherently sorted which takes care of coalescing.
-
-Furthermore sorting particles due to chapter \ref{sec:sort} increases caching behavior as variables of grid nodes are directly reused by particles in the same group.
+Luckily, grid nodes in a uniform grid are inherently sorted which takes care of coalescing.
+
+Furthermore, sorting particles due to chapter \ref{sec:sort} increases caching behavior as variables of grid nodes are directly reused by particles in the same group.
 This results in direct L2-cache hits instead of going the long way of reloading from global memory.
 
-With the mentioned partitioning of the grid into blocks the shared memory architecture of the GPU can be exploited.
-This requires two-level binning of particles to nodes within blocks.
+With the mentioned partitioning of the grid into blocks, the shared memory architecture of the GPU can be exploited.
+This requires two level binning of particles to nodes within blocks \cite{MPM:GPU}.
 Figure \ref{fig:g2p_transfer} show how the G2P-transfer using shared memory is realized.
@@
 All visible grid nodes need to be loaded by the threads.
-This includes the halo of the block, which is dependent upon the support of the interpolation function.
+This includes the halo of the block which is dependent upon the support of the interpolation function.
 For uneven polynomials of the interpolation function the floored position is very favorable.
 It allows the left support to be one smaller than the right support.
-As a result the weight of a most outer-edge particle (of the active particle region) just reaches zero for any neighborhood-node outside the halo.
+As a result, the weight of a most outer-edge particle (of the active particle region) just reaches zero for any neighborhood-node outside the halo.
 
@@
 The bigger the block the less halo nodes need to be loaded in total.
-Given an interpolation function with left support $l$ and right support $r$, total grid size $g$ and block sizes $b_x,b_y,b_z$ the general function to calculate the number of elements loaded is:
+Given an interpolation function with left support $l$ and right support $r$, total grid size $g$, and block sizes $b_x,b_y,b_z$ the general function to calculate the number of elements loaded is:
 \begin{equation}
@@
 \end{equation}
- As an example, take a 128*128*128 grid and the cubic interpolation function, which has a left support of 1 and a right support of 2. Subdivide the grid into 4*4*4 or 8*4*4 elements respectively:
+ As an example, take a 128*128*128 grid and the cubic interpolation function which has a left support of 1, and a right support of 2. Subdivide the grid into 4*4*4 or 8*4*4 elements respectively:
  $$m(4,4,4): 11,239,424\text{ Elements, }m(8,4,4):  8,830,976\text{ Elements}$$
-Note however that in practice there is a good change that a major part of the additional elements will result in L2-cache hits.
+Note however, that in practice there is a good chance, that a majority of the additional elements results in L2-cache hits.
 
 Next up is the transfer: Each thread corresponds to one particle within the blue nodes in Figure \ref{fig:g2p_transfer}.
-The threads will sequentially run over the neighbors within the support and collect the weighted contribution to this particle.
-In this process the particle's position need to be loaded to calculate the weight for its corresponding variables.
+The threads will in unison run over their relative neighbors within the support and collect the weighted contribution to this particle.
+In this process the particle's position need to be loaded to the weighted corresponding contribution.
 The transfer needs to be repeated for each particle in the bin.
@@
 \end{algorithm}
-Alternatively one could come up with a per-warp load balancing scheme: One thread would then not directly correspond to one active node.
-They would need to act as another node in the same warp, when their load is done.
-This will inevitably lead to shared memory bank conflicts.
-For the P2G-Transfer this would introduce atomic operations on shared memory again.
+Alternatively, one could come up with a per warp load balancing scheme: One thread would then not directly correspond to one active node.
+They would need to act as another node in the same warp when their load is done.
+This will inevitably lead to shared memory bank conflicts and other difficulties.
+For the P2G-Push-Transfers this would necessarily introduce atomic operations on shared memory again.
 
@@
 Instead of conservatively loading and writing one particle at a time, each thread handles multiple particles together which it stores in register memory.
-Important to this process is to unroll the corresponding loops that load, write as well as process these multiple particles.
+Important to this process is to unroll the corresponding loops that load, write, as well as process these particles.
 Instead of needing to rerun the transfer for each particle, the number of runs now done is the number of non-batched runs divided by the number of particles.
@@
 Straightforward to implement are various variations on global memory.
-Consider first binning was note executed.
+Consider first binning was not introduced.
 A grid node would not know how to look up the particles in its neighborhood.
-The only solution to this is go from the perspective of a particle and write to the grid node in its support.
-Since any particle in the neighborhood of a grid node contribute to its variables all writes have to be atomic.
-The approach the could be simply parallelizing for every particle and atomically write to the neighbor nodes in a loop.
+This would require the particle to find and write to the grid node in its support.
+Since any particle in the neighborhood of a grid node contributes to it, all writes have to be atomic.
+A typical approach is thus simply parallelizing for each particle and atomically write to the neighbor nodes in a loop.
 In the unsorted case this stalls the warps immensely.
 In \cite{Meyer2015} the static loop over the support is also parallelized resulting in higher throughput.
-
-The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-Transfers very significantly. The results of above get reversed and looping performs better. Stalling is much less of an issue.
+The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-Transfers very significantly.
+The results of above get reversed and looping performs better.
+Stalling is much less of an issue.
 
@@
 
-\textbf{P2G-Pull-Transfer}: However sorting and binning again opens up the possibility of shared memory implementations.
+\textbf{P2G-Pull-Transfer}: Sorting and binning however open up the possibility of shared memory implementations.
 The equivalent of the G2P-Transfer for the P2G-Transfer is pulling the relevant data from the neighborhood.
@@
 It is important to notice that the supports in the P2G-Pull approach are reversed: Left becomes right, right becomes left.
-Other than that the transfer works in the same way as the G2P-Transfer Instead of reading from grid nodes, reading from particles is done.
-Consequently one particle of each bin is loaded into shared memory.
+Other than that the transfer works in the same way as the G2P-Transfer.
+Instead of reading from grid nodes, reading from particles is done.
+Consequently, one particle of each bin is loaded into shared memory.
 This typically puts a lot more strain onto shared memory.
@@
 By construction the Material-Point-Method tries to avoid matrices on the grid due to storage cost.
-Conversely the particles have matrices defined over them, e.g.
+Conversely, the particles have matrices defined over them, e.g.
 $\boldsymbol{B}_p$,$\boldsymbol{F}_p$.
-In conclusion one should motivate approaches where grid nodes correspond to shared memory.
-
-There is one implementation detail left out which is solved in chapter \ref{sec:active_blocks}.
+In conclusion one should motivate approaches where grid nodes correspond to shared memory due to the shared memory size limitation.
+
+There is one implementation detail left out for the P2G-Pull-Transfer which is solved in chapter \ref{sec:active_blocks}.
 That is grid nodes do not know if there are any particles in the neighborhood left processing.
@@
 Thus all nodes within the block need to stay active during transfers.
+
 It is possible to delegate this decision down from the block to its subgroups (respecting their respective haloes).
@@
 Each thread would calculate the summed up count in its neighborhood.
-But this will hardly matter as all other threads need to wait on the thread with the highest count in the subgroup anyway.
+But this hardly matters for compute-bound shaders as all other threads need to wait on the thread with the highest count in the subgroup anyway.
 
@@
 
+In the following the two Particle-Push-Transfers are examined as they put less strain on shared memory:
+
 \textbf{P2G-Atomic-Transfer}: One can invert the process of pulling into pushing.
-Instead of reading from particles this means now writing to grid nodes in shared memory.
+Instead of reading from shared memory corresponding to particles, this means now writing to shared memory corresponding grid nodes.
 This immediately alarms one that accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
@@
 The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
-If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out of the transfer and \texttt{atomicAdds()} on shared memory are in use this process is straightforward to implement.
+If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out, and \texttt{atomicAdd()}s on shared memory are in use, this process is straightforward to implement.
 One has to make sure to reset the shared memory as their contents are otherwise undefined.
 
-The write-back to memory however now also needs to feature atomics.
-This process will be the same in the following synchronized approach of line \ref{alg:atomicAdd} in algorithm \ref{alg:p2g}.
-For a node that will be part of any halo \texttt{atomicAdds()} are a strict requirement.
+The write back to memory however now also needs to feature atomics.
+This process will be the same in the following synchronized approach.
+For a node that is part of any halo, \texttt{atomicAdds()} are a strict requirement.
 Their results are only partially computed and need to be complemented by neighbor blocks.
-For such a node these are two writes for a side node, four for an edge node and eight for a corner node.
+For such a node these are two writes for a side node, four for an edge node, and eight for a corner node.
 
@@
 In a synchronized approach this is solved by a \texttt{barrier()}.
-As already mentioned the particle loop of line \ref{alg:p2g_dyn_loop} in algorithm \ref{alg:p2g} is dynamic.
-The OpenGL language makes the following requirements for barriers:
-
-"\texttt{barrier()} can be called from flow-control, but it can only be called from uniform flow control.
-(...) In short, (...) every execution must hit the exact same set of \texttt{barrier()} calls in the exact same order." \cite{KHRONOS:BARRIER}
-
-The dynamic loop has to become static within a group.
+Care needs to be taken as the particle loop of line \ref{alg:p2g_dyn_loop} in algorithm \ref{alg:p2g} is dynamic.
+
+The OpenGL language states the following requirements for external synchronization:
+\textit{"\texttt{barrier()} can be called from flow-control, but it can only be called from uniform flow control.
+(...) In short, (...) every execution must hit the exact same set of \texttt{barrier()} calls in the exact same order."} \cite{KHRONOS:BARRIER}
+
+The dynamic loop has to become static within a thread group.
 The problem is solved in same fashion as in the P2G-Pull method with chapter \ref{sec:active_blocks}.
-Calculate the maximum particle count within the block without the halo.
-This is the minimum amount of times the loop needs to run across the threads to hit uniform flow control.
+The maximum particle of each bin within the block without the halo is the minimum amount of times the loop needs to run across the threads to hit uniform flow control.
 
@@
 
-In summary all transfers have the same optimal shared memory-access patterns as shown already by Appendices \ref{sec:shared_transfer}.
+In summary all transfers have the same optimal shared-memory access patterns as shown already by Appendices \ref{sec:shared_transfer}.
 Batching is again a valid option that is executed in the same manner by unrolling the corresponding loops.
@@
 Returned is the number of active blocks for the launch of a dispatch.
-A scan and stream compaction gives back the filtered indices to access the active blocks in that dispatch.}
+A scan and an indexed stream compaction give back the filtered indices to access the active blocks in that dispatch.}
   \label{fig:active_block}
@@
 The active particle region is the region that needs to get reduced.
-Therefore it is sufficient for the P2G-Sync-Transfer to reduce the counter within the block, left of Figure \ref{fig:p2g_transfer}.
-The P2G-Pull-Transfer however also need to take the halo into account, right of Figure \ref{fig:p2g_transfer}.
+Therefore it is sufficient for the P2G-Sync-Transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}.
+The P2G-Pull-Transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}.
 Both are achievable with a single level of parallel reduction (in contrast to a pyramid).
-We will call this the per-block count from now on, that uses as the functional input for the parallel reduction:
+This is the \textit{per block bin count} from now on that uses as the functional input for the parallel reduction:
 \begin{enumerate}
@@
 Then can narrow the shader down to do more sequential work to improve throughput.
-For the P2G-Pull-Transfer one may also do the first, but additionally load all rest elements (by the halo).
+For the P2G-Pull-Transfer one may also do the first but additionally load all rest elements (by the halo).
 From there on out more sequential work might be favorable.
 
-If a per block indexing mode is employed all blocks are directly behind each other in the buffer.
+If a per block indexing mode is employed, all blocks are directly behind each other in the buffer.
 Reduction then happens in the typical fashion.
@@
 Starting a dispatch for every block only the active blocks will now write their index at the scan's position.
+It results in an indexed stream compaction.
 This process is not coalesced.
 Coalescing will still be guarenteed for all nodes within the block for the P2G-Transfers and G2P-Transfer.
-The only non-coalesced access within these shaders is loading the maximum per block count from VRAM, which there are really few of in comparison.
-
+The only uncoalesced access within these shaders is loading the maximum per block count from VRAM which there are few of in comparison.
+
+\clearpage
 \section{Evaluation}\label{sec:eval}
+fill with nonsense
 \subsection{Verifying Results}
-\textbf{Parallel reduction}, \textbf{map} and \textbf{scan} are tested against \texttt{std::transform\_reduce} as it supplies the same functionality as a fully reduced parallel reduction.
+\textbf{Parallel reduction}, \textbf{map}, and \textbf{scan} verify against \texttt{std::transform\_reduce} as it supplies the same functionality as a fully reduced parallel reduction.
 It is standard since C++17.
@@
 All timings are averaged over at least 1000 runs with OpenGL Timer Queries \cite{KHRONOS:TIMER_QUERY}.
-As more runs are done the GPU stays very consistent on bringing the average runtime closer to the minimum recorded time.
+As more runs are done, the GPU stays very consistent on bringing the average runtime closer to the minimum recorded time.
 This indicates that the first few runs can take comparably longer.
@@
 \textbf{SoA vs.
-AoS}: In chapter \ref{sec:soa_aos} was already hinted that throughput could be halved in an AoS-Layout.
+AoS}: In chapter \ref{sec:soa_aos} is already hinted that throughput could be halved in an AoS-Layout.
 These effects can be hideous without knowledge of the architecture.
@@
 An AoS-Layout however issues the same instructions.
-Though the SM should do equal work and be active for the same amount of cycles which is not the case.
+The SM should do equal work and be active for the same amount of cycles which is not the case.
 \begin{table}[htpb]
@@
 \caption{Metrics between different buffer layouts.
-Map operation with two instructions \texttt{float x = length(vel), float y = length(pos)} on $1024 \times 1024$ \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
+Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length}\texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
 \label{tab:map_2}
@@
 As a closure the reader is again referred to the subject of data-oriented design.
-While for non-random access SoA is (as shown) the more efficient layout.
-Random access algorithms can improve performance by using an AoS or mixed setup.
+While for non-random access SoA is (as shown) the more efficient layout, random access algorithms can improve performance by using an AoS or mixed setup.
 In a random access pattern the nearby elements of an SoA-Layout will be left unused in the cache-line.
@@
 Table \ref{tab:map_reduce} compares this to an interleaved addressing without padding to illustrate the effects of bank conflicts on performance.
-Similarly the same can be done for a scan, where however the interleaved addressing is preferred anyway.
+Similarly the same can be done for a scan where however the interleaved addressing is preferred anyway.
 Table \ref{tab:scan} shows the comparison for a scan.
@@
 
-The metrics between the parallel reduction and scan differ quite a lot in their comparison to the their bank-conflict reduced solutions.
+The metrics between the parallel reduction and scan differ quite a lot when it comes to the resolve of bank-conflicts.
 Where the scans issue utilization goes drastically up from 34.2\% to 51.3\%, the same metric goes down for parallel reduction from 62.2\% to 33.0\%.
-SM throughput decreases as VRAM throughput increases, although little for the scan.
+SM throughput decreases as VRAM throughput increases although little for the scan.
 The top warp-stall reason for both is the barrier synchronization between thread groups.
 
-The warp-stall reason however showing the increased throughput due to respecting the shared memory architecture is the short scoreboard.
-The shaders in their original state have no reason to stall for the short scoreboard as any memory access is directly fetched.
+The warp-stall reason however showing the increased throughput due to respecting the shared memory architecture is the Short Scoreboard.
+The shaders in their original state have no reason to stall for the Short Scoreboard as any memory access is directly fetched.
 In the parallel reduction case the change is quite visible as this metric drops from 17.2\% to only 2.0\%.
@@
 Note, that each thread group in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done by each thread.
-However due to the additional sequential work parallelization the amount of thread groups go down by the same factor.
-As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this saturates the threads and thus improves performance.
+However due to the additional sequential work the amount of thread groups go down by the same factor.
+As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this saturates I/O-Latency and thus improves performance.
 
 Doing sequential work in parallel reduction adds almost no strain.
-Thus very high sequential work can be done by each thread.
-The SM work will shrink, while the shader gets more and more memory bound.
+Thus high amounts sequential work can be done by each thread.
+The SM work will shrink while the shader get more and more memory-bound.
 For a reduction factor of 256 first signs of performance decrease can be seen as only four groups are involved in the dispatch.
@@
 The metric SM active\% will go rapidly down at this point.
-For four groups its 27.8\%, which is roughly $\frac{4}{13}$.
+For four groups its 27.8\% which is roughly $\frac{4}{13}$.
 
@@
 The segment each scan can handle is by far not as extensive as in parallel reduction.
-But the speedup achieved from a partially conflict free to an implementation with sequential work is in contrast much higher.
+But the speedup achieved from a partially conflict free to an implementation with sequential work, is in contrast much higher.
 Since each thread requires the nearest by elements, L2-cache hits increase rapidly as does the L2 utilization.
@@
 
-\textbf{Counting Sort:} The aim of sorting is coalesced as well as increased caching behavior for all following shaders.
+\textbf{Counting Sort:} The aim of sorting is coalescing as well as increased caching behavior for all following shaders.
 As a motivation the binning's similarity to the MPM-Operations might already be highlighted.
-A key difference is that binning does not query neighbor information (in the global atomic approach).
 Table \ref{table:bin} shows how a sorted particle position increases coalescing and L2-Cache hits.
@@
 
-Table \ref{table:sort} shows the performance of index and deep sorting as is.
+Table \ref{table:sort} shows the performance of reordering with index and deep sorting as is.
 The index sorting will have constant timing across cases.
@@
 The binning in table \ref{table:bin} thus for instance always runs on random ordering.
-Global atomic binning does not use neighbor data.
-Thus index sorting is not affecting it.
+Index sorting is not affecting it.
 
@@
 The CFL-condition encourages limited reach.
-Solids and fluids will mostly move uniformly.
+Most physical processes will move uniformly.
 These two properties in general will guarantee higher coalescing as well as L2-cache hit rates.
@@
 
-The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to build off with only one \texttt{vec4} transfer.
+The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to build off of with only one \texttt{vec4} transfer.
 The MPM-Operations are evaluated with two easily reproducible examples.
@@
 
-The \textbf{P2G-Transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4} respectively.
+The \textbf{P2G-Transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4}, respectively.
 The global atomic approaches suffer from low throughput across all units.
@@
 
-Top SOL\% is now the SM, but it lacks throughput with 39.4\%.
+Top SOL\% is now the SM but it lacks throughput with 39.4\%.
 Its SM Issue Utilization per active cycle is 40\% which is below the desired threshold of 60\%.
@@
 Its effects are quite visible on the SM SOL\% with 82.9\% to 90.9\% in the call for active blocks.
-The top stall reason is short scoreboard which is indicative of the high number of stalls due to atomic writes on shared memory.
+The top stall reason is Short Scoreboard which is indicative of the high number of stalls due to atomic writes on shared memory.
 
@@
 Short scoreboard goes down to 11.6\%.
-As previously the SM was strained too hard throughput goes down to 65.0\%.
+As previously the SM was strained, throughput goes down to 65.0\%.
 The greatest time improvement can be seen when batching and indirect call lead to a combined speedup of 2.2x.
@@
 With one particle per cell(area 47.6\%) batching does not help at all.
-The P2G-Sync-Transfer however can still work comparably fast to the high density of four particles per cell losing about 7.2\% speed.
-This also means that the P2G-Sync-Transfer only scales slightly with the area covered, which is due to the overhead of loading and writing more blocks.
+The P2G-Sync-Transfer however can still work comparably fast to either the high density of four particles per cell losing about 7.2\% speed or atomic operations because batching is not available.
+This also means that the P2G-Sync-Transfer only scales slightly with the area covered which is due to the overhead of loading and writing more blocks.
 
@@
 They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
+\clearpage
 \section{Conclusion \& Future Work}\label{sec:conclusion}
@@
 Although possible this process does not rely on the CFL-condition.
+The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.
+
+The second preprocessing step is filtering the active blocks of grid and particle transfers with a stream compaction pipeline where no particles are present.
+It also supplies the P2G-Pull- and P2G-Sync-Transfer with a maximum per block bin counter.
+
 While the shared memory architecture can give an edge in performance it produces branch divergence and may perform worse over a simpler algorithm relying on the L2-cache.
@@
 The sorted global G2P-Transfer's consistency of utilizing the L2-Cache is a surprise further to be monitored.
-The second preprocessing step is filtering the active blocks with a stream compaction pipeline where no particles are present.
-It also supplies the P2G-Pull- and P2G-Sync-Transfer with a maximum per block bin counter.
 
@@
 While this strongly furthers understanding it leaves real simulation performance comparisons to be desired.
-However before this process can be concluded merge and split techniques need to be taken into consideration.
+
+However, before this process can be concluded merge and split techniques need to be taken into consideration.
 They are designed to augment the numerics of the Material-Point-Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
-As such they will stray from the optimistic example.
+As such they will stray less from the optimistic example.
 How much they will stray from it is however dependent on the dynamics of the problem.
-The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.
 \clearpage
@@
       xi_xp = np.array([ai,bj,ck])
-      # each outer product weighted by interpolation functions
+      # each outer product weighted by interpolation
+      # functions
       this_outer = wip(i,ai)*wip(j,bj)*wip(k,ck)*
@@
 
-D = np.array([[0.0,0.0,0.0],[0.0,0.0,0.0],[0.0,0.0,0.0]])
+D = np.array([[0.0,0.0,0.0],
+	      [0.0,0.0,0.0],
+	      [0.0,0.0,0.0]])
 

### 2019-01-27 mscthesis.tex
final fill things

@@
 \section{Basics}\label{sec:basics}
-fill with nonsense
+The first chapters will derive the Material-Point-Method and necessary extensions for an elastic model.
+An elastic model is an interesting case displaying much of the needed computations for a typical model.
+The last chapter \ref{sec:gpgpu} tends to the GPGPU side.
 \subsection{Reference and current configuration}\label{sec:ref_corr_config}
@@
 \subsection{Governing equations}
-fill with nonsense
-deriving the weak form
+Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and material point method.
 \subsubsection{Conservation of mass}
@@
 \end{equation}
-\cite{MPM:COURSE}
-\cite{strang2007computational}
-\cite{bathe2006finite}
+\begin{flushright}\cite{MPM:COURSE}\cite{strang2007computational}\cite{bathe2006finite}\end{flushright}
 \subsection{Material Point Method}\label{sec:mat_point}
@@
 \end{equation}
-\cite{MPM:APIC}\cite{MPM:COURSE}
+\begin{flushright}\cite{MPM:APIC}\cite{MPM:COURSE}\end{flushright}
 \subsubsection{APIC transfers}\label{sec:apic}
@@
 \end{enumerate}
-\subsection{General-purpose computing on graphics processing units}
-fill with nonsense
+\subsection{General-purpose computing on graphics processing units}\label{sec:gpgpu}
+This chapter is split into two parts.
+\ref{sec:optim} tends to the memory hierarchy of the GPU and typical optimizations done in GPGPU work.
+Furthermore, chapter \ref{sec:metrics} will elaborate metrics to quantify optimizations.
 \subsubsection {GPGPU Optimization opportunities}\label{sec:optim}
@@
 \section{Evaluation}\label{sec:eval}
-fill with nonsense
-\subsection{Verifying Results}
+An important part of any operation is to verify them.
+Chapter \ref{sec:verify} does that.
+Following up, chapter \ref{sec:perf} will use the metrics out of chapter \ref{sec:metrics} to evaluate the performance of the implementation.
+\subsection{Verifying Results}\label{sec:verify}
 \textbf{Parallel reduction}, \textbf{map}, and \textbf{scan} verify against \texttt{std::transform\_reduce} as it supplies the same functionality as a fully reduced parallel reduction.
@@
 
-\subsection{Performance}
+\subsection{Performance}\ref{sec:perf}
 The given metrics in \ref{sec:metrics} lead to valuable insight into how well the GPU can utilize a given shader or program.

### 2019-01-27 mscthesis.tex
rework abstract, zusammenfassung

@@
 
-\vspace*{\fill}~
+\selectlanguage{english}
+\vfill
 \begin{center}
@@
 \end{center}
-The Material-Point-Method is a technique allowing for physically accurate simulations.
+The Material-Point-Method is allowing for physically based simulations.
 It has found its way into computer graphics and since then rapidly expanded.
 The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.
-It has found its way into computer graphics and since then rapidly expanded.
-
-\noindent Core of this process are the grid and particles transfers that interpolate between the two structures.
-These transfers are executed multiple times per frame so preprocessing steps might be taken if they are outweighed.
-
-\noindent The Method however suffers from high execution times and make the method only viable for hero shots.
-The Material-Point-Method including the transfers are highly parallelizable.
-As such this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.
+
+\noindent The Material-Point-Method suffers from high execution times and is thus only viable for hero shots.
+The method is however highly parallelizable.
+Thus this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.
+
+\noindent Core of the Material-Point-Method are grid and particles transfers that interpolate between the two structures.
+These transfers are executed multiple times per physical time step.
+Preprocessing steps might be taken if their computing time is outweighed.
 
 \noindent Deep sorting with counting sort increases coalescing and L2 hit rates.
-Binning allows for shared memory filtering techniques using blocking.
+Binning allows to divide the grid into blocks for shared memory filtering techniques.
+All operations do not rely on fixed bin size.
 As another preprocessing step only grid blocks are executed which have particles in them.
-All transfers do not rely on fixed size bins.
+\vfill
+\selectlanguage{german}
+\begin{center}
+\subsubsection*{Zusammenfassung}
+\end{center}
+Die Material-Point-Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert.
+Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter.
+Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.
+
+\noindent Die Material-Point-Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht.
+Sie ist jedoch hoch-parallelisierbar.
+Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.
+
+\noindent Der Datenaustausch von Partikel und Gitter wird über interpolierende Transfers erreicht.
+Pro physikalischen Zeitschritt werden diese mehrmals ausgeführt.
+Vorverarbeitungsschritte können unternommen werden, um diese Transfers performanter zu machen.
+
+\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Hitraten.
+Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en.
+Die Größen der Bins sind für keine Operation beschränkt.
+Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.
+
 \vfill

### 2019-01-27 mscthesis.tex
small fixed

@@
 
-\selectlanguage{english}
+\vfill
+\begin{center}
+\subsubsection*{Zusammenfassung}
+\end{center}
+Die Material Point Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert.
+Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter.
+Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.
+
+\noindent Die Material Point Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht.
+Sie ist jedoch hoch-parallelisierbar.
+Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.
+
+\noindent Der Datenaustausch von Partikel und Gitter wird über interpolierende Transfers erreicht.
+Pro physikalischen Zeitschritt werden diese mehrmals ausgeführt.
+Vorverarbeitungsschritte können unternommen werden, um diese Transfers performanter zu machen.
+
+\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Hitraten.
+Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en.
+Die Größen der Bins sind für keine Operation beschränkt.
+Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.
+
 \vfill
@@
 \end{center}
-The Material-Point-Method is allowing for physically based simulations.
+The Material Point Method is allowing for physically based simulations.
 It has found its way into computer graphics and since then rapidly expanded.
@@
 
-\noindent The Material-Point-Method suffers from high execution times and is thus only viable for hero shots.
+\noindent The Material Point Method suffers from high execution times and is thus only viable for hero shots.
 The method is however highly parallelizable.
-Thus this thesis proposes how to accelerate the Material-Point-Method using GPGPU techniques.
-
-\noindent Core of the Material-Point-Method are grid and particles transfers that interpolate between the two structures.
+Thus this thesis proposes how to accelerate the Material Point Method using GPGPU techniques.
+
+\noindent Core of the Material Point Method are grid and particles transfers that interpolate between the two structures.
 These transfers are executed multiple times per physical time step.
@@
 As another preprocessing step only grid blocks are executed which have particles in them.
-\vfill
-\selectlanguage{german}
-\begin{center}
-\subsubsection*{Zusammenfassung}
-\end{center}
-Die Material-Point-Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert.
-Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter.
-Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.
-
-\noindent Die Material-Point-Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht.
-Sie ist jedoch hoch-parallelisierbar.
-Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.
-
-\noindent Der Datenaustausch von Partikel und Gitter wird über interpolierende Transfers erreicht.
-Pro physikalischen Zeitschritt werden diese mehrmals ausgeführt.
-Vorverarbeitungsschritte können unternommen werden, um diese Transfers performanter zu machen.
-
-\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Hitraten.
-Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en.
-Die Größen der Bins sind für keine Operation beschränkt.
-Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.
-
 \vfill
@@
 Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible.
-The Material-Point-Method has already reliable tackled many different models, materials, and even interaction between materials.
+The Material Point Method has already reliable tackled many different models, materials, and even interaction between materials.
 
@@
 Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature.
-Particle-in-Cell(PIC) techniques like the Material-Point-Method are hybrid Lagrangian/Eulerian methods.
+Particle-in-Cell(PIC) techniques like the Material Point Method are hybrid Lagrangian/Eulerian methods.
 They try to benefit from the nature of both.
 Advection on particles is trivial.
-Thus the Material-Point-Method utilizes Lagrangian particles as a consistent storage of dynamic properties.
+Thus the Material Point Method utilizes Lagrangian particles as a consistent storage of dynamic properties.
 The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid.
@@
 
-Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material-Point-Method extensions.
+Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material Point Method extensions.
 In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
@@
 \section{Basics}\label{sec:basics}
-The first chapters will derive the Material-Point-Method and necessary extensions for an elastic model.
+The first chapters will derive the Material Point Method and necessary extensions for an elastic model.
 An elastic model is an interesting case displaying much of the needed computations for a typical model.
@@
 \end{equation}
-In the Material-Point-Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
+In the Material Point Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
 Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
@@
 \subsection{Material Point Method}\label{sec:mat_point}
-The key idea of the Material-Point-Method is to use (Lagrangian) particles as a consistent storage of material properties.
+The key idea of the Material Point Method is to use (Lagrangian) particles as a consistent storage of material properties.
 All stress based forces are computed on a Eulerian grid however.
@@
 
-In contrast much harder are transfers between the two structures that are essential to the Material-Point-Method:
+In contrast much harder are transfers between the two structures that are essential to the Material Point Method:
 
@@
 This motivates making the transfers a very highly optimized operation.
-The Material-Point-Method requires for an elastic material as derived:
+The Material Point Method requires for an elastic material as derived:
 \begin{enumerate}
@@
 
-By construction the Material-Point-Method tries to avoid matrices on the grid due to storage cost.
+By construction the Material Point Method tries to avoid matrices on the grid due to storage cost.
 Conversely, the particles have matrices defined over them, e.g.
@@
 \section{Conclusion \& Future Work}\label{sec:conclusion}
-The derivation of the Material-Point-Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
+The derivation of the Material Point Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
 Although the MPM-Method consists of many mathematical tasks four operations on the GPU can be identified.
@@
 However, before this process can be concluded merge and split techniques need to be taken into consideration.
-They are designed to augment the numerics of the Material-Point-Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
+They are designed to augment the numerics of the Material Point Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
 As such they will stray less from the optimistic example.

### 2019-01-27 mscthesis.tex
l2 hitrates

@@
 
-\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Hitraten.
+\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten.
 Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en.
@@
 
-\noindent Deep sorting with counting sort increases coalescing and L2 hit rates.
+\noindent Deep sorting with counting sort increases coalescing and L2 Cache hit rates.
 Binning allows to divide the grid into blocks for shared memory filtering techniques.

### 2019-01-27 mscthesis.tex
changed all occurences of Material Point Method to be written the same

@@
 In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
-The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM-Method to get an understanding of the prominent operations.
+The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations.
 Furthermore GPU intrinsics, metrics and optimization opportunities are discussed.
@@
 They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}.
-Their key steps are identical to the MPM-method:
+Their key steps are identical to the MPM:
 \begin{enumerate}
@@
 \cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}.
-In rapid succession a variety materials and effects got modeled by the MPM-Method.
+In rapid succession a variety materials and effects got modeled by the MPM.
 Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.
@@
 The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
-Since the number of particles per grid point in the MPM-method is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.
+Since the number of particles per grid point in the MPM is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.
 
@@
 When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
-In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
+In the Material Point Method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
 This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
@@
 \subsection{Governing equations}
-Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and material point method.
+Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and Material Point Method.
 \subsubsection{Conservation of mass}
@@
 Such a solution can be found for simplified models analytically.
-These act as a ground truth to numerical methods like the material point method.
+These act as a ground truth to numerical methods like the Material Point Method.
 Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.
@@
 
-Given, that the material point method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
+Given, that the Material Point Method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
 Based on this one may also gain a volume change measure around the particle as $J^n_p = \text{det}(\boldsymbol{F}^n_p)$.
@@
 $$\square_p = \sum_i \square_i \circ \square_{ip}.$$
-This concludes all necessary operations for the MPM-Method as derived.
+This concludes all necessary operations for the MPM as derived.
 
@@
   \item \textbf{Dynamic weights}: The stencil is not static.
-In the MPM-Method the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
+In the MPM the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
 This also means the position needs to be temporarily stored to apply the transfer.
@@
 The derivation of the Material Point Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
-Although the MPM-Method consists of many mathematical tasks four operations on the GPU can be identified.
+Although the MPM consists of many mathematical tasks four operations on the GPU can be identified.
 Exclusive grid and particle operations are parallelized with a 1:1 correspondence to threads.

### 2019-01-27 mscthesis.tex
added apic transfers

@@
   keywordstyle=[3]\color[rgb]{0.127,0.427,0.514},
-  keywordstyle=[4]\color[rgb]{0.4,0.4,0.4},
-  commentstyle=\color[rgb]{0.133,0.545,0.133},
-  stringstyle=\color[rgb]{0.639,0.082,0.082},
-  numbers=left,
-}
-
-\newenvironment{code}{\captionsetup{type=algorithm}}{}
+  keywordstyle=[4]\color[rgb]{0.4,0.4,0.4}}
 
@@
 The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid.
-The grid is unmoving and does not store any information between frames.
+The grid is unmoving and does not store any information between physical time steps.
 
@@
 I.e.
-the grid could be deleted each frame.
+the grid could be deleted each physical times step.
 Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations.
@@
 These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic.
-Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL-condition, chapter \ref{sec:cfl}.
+Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL condition, chapter \ref{sec:cfl}.
 These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management.
@@
 This implementation compares index and deep counting sort and uses the latter.
-Although possible, this counting sort stays away from the limited movement implied by the CFL-condition.
+Although possible, this counting sort stays away from the limited movement implied by the CFL condition.
 Fixed size bins is an assumption this implementation although does not want to impose on the method.
@@
 \end{equation}
-\cite{MPM:APIC}
+\begin{flushright}\cite{MPM:APIC}\end{flushright}
 \subsection{Discretization}\label{sec:discretization}
@@
 \end{enumerate}
+\clearpage
 \subsection{General-purpose computing on graphics processing units}\label{sec:gpgpu}
@@
 \begin{itemize}
-  \item CFL-condition of chapter \ref{sec:cfl} with operators,
+  \item CFL condition of chapter \ref{sec:cfl} with operators,
   \begin{itemize}
@@
 
-Alternatively, one could work again with the CFL-condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a Particle-to-Grid-Transfer, chapter \ref{sec:p2g}:
+Alternatively, one could work again with the CFL condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a Particle-to-Grid-Transfer, chapter \ref{sec:p2g}:
 \begin{enumerate}
@@
 
-The number of transfers per frame is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$).
+The number of transfers per physical time step is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$).
 This motivates making the transfers a very highly optimized operation.
@@
 In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models.
-In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every frame.
+In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every physical times step.
 Each transfer can profit from it.
@@
 \end{equation}
-An extreme test case spreads one million particles randomly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}.
+The test case spreads one million particles uniformly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}.
 A roundtrip consists of a P2G-Transfer and a G2P-Transfer for relevant variables.
 In this process the (random) velocities will get filtered.
-In PIC-Transfers this regionally averages out velocities.
+In PIC-Transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$.
+APIC-Transfers can conserve the angular momentum better as they define a local velocity field around the particle.
 Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.
-
-As a side note: Random particle positions as well as random velocities are a configuration that in practice will never occur. Even for a fundamental state of matter of gas this is unrealistic.
 
@@
     Quantity& Abs. error  & Rel. error \\ \hline
-    $m^{tot}_i$ & 212.438     & 2.12438e-04     \\\hline
-    $\boldsymbol{p}^{tot}_i$ & (0.578,0.787,0.778)&(1.076,1.290,1.204)e-04 \\\hline
+    \multicolumn{3}{|c|}{PIC-Transfer}\\\hline
+    $m^{tot}_i$ & 6.438     & 6.437e-06     \\\hline
+    $\boldsymbol{p}^{tot}_i$  & (0.882,0.977,0.018)&(1.417,1.722,0.003)e-04 \\\hline
+    $\boldsymbol{L}^{tot}_i$  & \textbf{(27.81,43.55,24.35)}&(0.024,0.036,0.273)\\\hline
+    \multicolumn{3}{|c|}{APIC-Transfer}\\\hline
+    $m^{tot}_i$ & 7.188     & 7.188e-06     \\\hline
+    $\boldsymbol{p}^{tot}_i$  & (1.230,0.990,0.855)&(1.975,1.745,1.371)e-04 \\\hline
+    $\boldsymbol{L}^{tot}_i$  & \textbf{(0.081,0.184,5.6e-04)}&(0.7043,1.510,0.063)e-04\\\hline
    \end{tabular}
-   \caption{Error of 1,000 PIC-roundtrips of 1M randomly placed particles with random velocity in a $128\times128\times128$ grid.
+   \caption{Error of 1,000 PIC/APIC-roundtrips of 1M uniformly placed particles with random velocity in a $128\times128\times128$ grid.
 Used this instance is a P2G-Sync-Transfer and G2P-Transfer with batching = 4 and block size = (4,4,4).
@@
 \end{table}
-
-\subsection{Performance}\ref{sec:perf}
+\clearpage
+\subsection{Performance}\label{sec:perf}
 The given metrics in \ref{sec:metrics} lead to valuable insight into how well the GPU can utilize a given shader or program.
@@
 Sorting works on the last sorted state.
-The CFL-condition encourages limited reach.
+The CFL condition encourages limited reach.
 Most physical processes will move uniformly.
@@
 For the \textbf{G2P-Transfers} the different datasets can be seen in table \ref{table:g2p_random} and \ref{table:g2p_uniform4}.
-The first one uses random particle positions and therefore is overly pessimistic for the Material Point Method as explained earlier.
+The first one uses random particle positions and therefore is overly pessimistic for the Material Point Method.
+Random particle positions as well as random velocities are a configuration that in practice will never occur.
+Even for a fundamental state of matter of gas this is unrealistic.
 The second one is optimistic as warp branch divergence is minimized.
@@
 \end{table}
+To underline the extensibility of all shaders table \ref{table:apic} implements the APIC-Transfers of chapter \ref{sec:apic}.
+The G2P-shared algorithm continues to lose ground on the global sorted G2P method as its execution time now only measures $300\mu s$ instead of the $400\mu s$ in table \ref{table:g2p_uniform4}.
+Its SM throughput also overtakes the global method with 81.8\%.
+\begin{table}[htpb]
+  \begin{tabular}{ | l | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &SM    & Top Stall Reas.  & Occup.\\\hline
+    G2P(sorted)		   &    3,647             &74.9\%&No Instr.(14.2\%)& 57.7\\\hline
+    G2P-shared             &    3,384             &\textbf{81.8}\%&No Instr.(18.7\%) & 43.3\\\hline
+    \hline
+    P2G-Sync               &    3,843             &68.9\%&No Instr.(24.9\%) & 43.3\\\hline
+    batching=2             &    3,634             &67.3\%&No Instr.(26.5\%) & 31.6\\\hline
+ \end{tabular}
+ \caption{APIC-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+They form a rotated (unsorted) cube with four particles per cell.
+Block size is (8,4,4).
+Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:apic}
+\end{table}
+
 The P2G-Atomic- and the even more favorable P2G-Sync-Transfer give great speedups over the global atomic methods.
@@
 All transfers do not rely on fixed bin sizes in any stages of the pipeline.
-Since these operations are executed multiple times per frame these operations are augmented with two preprocessing steps.
+Since these operations are executed multiple times per physical time step these operations are augmented with two preprocessing steps.
 The preprocessing steps need to be done once and are outweighed by the transfers.
@@
 Deep sorting particles with counting sort in a double-buffered manner allows coalesced accesses and high cache hit rates on transfers.
-Although possible this process does not rely on the CFL-condition.
+Although possible this process does not rely on the CFL condition.
 The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.

### 2019-01-28 mscthesis.tex
moore out
first review done

@@
 \noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten.
-Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen ermög\-lich\-en.
+Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen er\-mög\-lich\-en.
 Die Größen der Bins sind für keine Operation beschränkt.
@@
 \end{center}
-The Material Point Method is allowing for physically based simulations.
+The material point method is allowing for physically based simulations.
 It has found its way into computer graphics and since then rapidly expanded.
-The Material Point Method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enable solving of various partial differential equations with ease.
-
-\noindent The Material Point Method suffers from high execution times and is thus only viable for hero shots.
+The material point method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enables solving of various partial differential equations with ease.
+
+\noindent The material point method suffers from high execution times and is thus only viable for hero shots.
 The method is however highly parallelizable.
-Thus this thesis proposes how to accelerate the Material Point Method using GPGPU techniques.
-
-\noindent Core of the Material Point Method are grid and particles transfers that interpolate between the two structures.
+Thus, this thesis proposes how to accelerate the material point method using GPGPU techniques.
+
+\noindent Core of the material point method are grid and particles transfers that interpolate between the two structures.
 These transfers are executed multiple times per physical time step.
@@
 
-\noindent Deep sorting with counting sort increases coalescing and L2 Cache hit rates.
+\noindent Deep sorting with counting sort increases coalescing and L2 cache hit rates.
 Binning allows to divide the grid into blocks for shared memory filtering techniques.
 All operations do not rely on fixed bin size.
-As another preprocessing step only grid blocks are executed which have particles in them.
+As another preprocessing step, only grid blocks are executed which have particles in them.
 \vfill
@@
 \section{Introduction} \label{intro}
-General Purpose Computation on Graphics Processing Unit(GPGPU) has elevated Graphics Processing Units(GPUs) to any computational field utilizing massively parallelizable algorithms.
+General purpose computation on graphics processing unit(GPGPU) has elevated graphics processing units(GPUs) to any computational field utilizing massively parallelizable algorithms.
 As such is the nature of discretizations in physics.
-GPU's performance still follows Moore's Law.
+
+The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it.
 In unison with efficient algorithms they still have much potential to offer.
-
-The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it.
-Conversely algorithms which are not parallelizable perform much worse and should be avoided.
+Conversely, algorithms which are not parallelizable perform much worse and should be avoided.
 Dividing up these tasks between CPU and GPU is not recommendable.
@@
 Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible.
-The Material Point Method has already reliable tackled many different models, materials, and even interaction between materials.
-
-Pure Eulerian methods like the Finite-Element-Method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability.
+The material point method(MPM) has already reliable tackled many different models, materials, and even interaction between materials.
+
+Pure Eulerian methods like the finite element method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability.
 Pure Langrangian methods form the end other of the spectrum.
-Prominent are Smoothed Particle Hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature.
-Particle-in-Cell(PIC) techniques like the Material Point Method are hybrid Lagrangian/Eulerian methods.
+Prominent are smoothed-particle hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature.
+Particle in cell(PIC) techniques like the material point method are hybrid Lagrangian/Eulerian methods.
 They try to benefit from the nature of both.
 Advection on particles is trivial.
-Thus the Material Point Method utilizes Lagrangian particles as a consistent storage of dynamic properties.
+Thus the material point method utilizes Lagrangian particles as a consistent storage of dynamic properties.
 The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid.
@@
 Transfers between the Lagrangian and Eulerian are of need to bring the two together.
-These transfers are the focus of the Particle-in-Cell techniques.
+These transfers are the focus of particle in cell techniques.
 These transfers are realized due to interpolation.
@@
 
-The $p$ particles to one node relationship of a Particle-to-Grid(P2G)-Transfer or inversely one node to $p$ particles of a Grid-to-Particle(G2P)-Transfer is the main problem when realizing these operations on the GPU.
+The $p$ particles to one node relationship of a particle-to-grid(P2G)-transfer or inversely one node to $p$ particles of a grid-to-particle(G2P)-transfer is the main problem when realizing these operations on the GPU.
 When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.
 
-Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as Material Point Method extensions.
+Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as material point method extensions.
 In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
-The Basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations.
+The basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations.
 Furthermore GPU intrinsics, metrics and optimization opportunities are discussed.
@@
 \section{Related Work}\label{sec:rel_work}
-\textbf{Particle-in-Cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}.
+\textbf{Particle in cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}.
 They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}.
@@
 Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations.
-Their counterpart being the empirical Smoothed Particle Hydrodynamics \cite{gingold1977smoothed}.
-
-As a further development the Affine-Particle-in-Cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers.
+Their counterpart being the empirical smoothed-particle hydrodynamics \cite{gingold1977smoothed}.
+
+As a further development the Affine-particle in cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers.
 Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
@@
 
-The \textbf{Material Point Method} is a further development to the PIC methods.
-It uses analysis and numerics to be derived without the need for empirical assumptions.
+The \textbf{material point method} is a further development to the PIC methods.
+Both use analysis and numerics to be derived without the need for empirical assumptions.
 Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids.
@@
 
-There exist several implementations utilizing the \textbf{GPU for Particle-in-Cell} techniques, which the Material Point Method is a part of:
+There exist several implementations utilizing the \textbf{GPU for particle in cell} techniques, which the material point method is a part of:
 
@@
 These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management.
-Fixed size binning make the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
+Fixed size binning makes the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
 As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes.
@@
 This implementation compares index and deep counting sort and uses the latter.
-Although possible, this counting sort stays away from the limited movement implied by the CFL condition.
-Fixed size bins is an assumption this implementation although does not want to impose on the method.
-
-\textbf{Particle-to-Grid-Transfers} are generally done in two manners \cite{stantchev2008fast}:
+Although possible, this counting sort stays away from the limited movement implied by the Courant–Friedrichs–Lewy(CFL) condition.
+Fixed size bins are an assumption this implementation also does not want to impose on the method.
+
+\textbf{Particle-to-grid-transfers} are generally done in two manners \cite{stantchev2008fast}:
 \begin{enumerate}
-  \item \textbf{Particle Push:} X particles in a local neighborhood of a grid point push their attributes onto it.
+  \item \textbf{Particle push:} X particles in a local neighborhood of a grid point push their attributes onto it.
 Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g.
 a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
-  \item \textbf{Particle Pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
+  \item \textbf{Particle pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
 \end{enumerate}
-A Grid-to-Particle-Transfer is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
+A \textbf{grid-to-particle-transfer} is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
 The grid points in a uniform grid are 'sorted' by nature.
-This implementation tests one Particle Pull and two Particle Push methods.
+This implementation tests one particle pull and two particle push methods.
 
 Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition.
-Shared memory is however limited per GPU (GTX 970, 48KB per block).
+Shared memory is however limited per GPU (GTX 970, 48KB per block \cite{NVIDIA:GTX970}).
 The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
@@
 
-In Particle Push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
+In particle push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
 \begin{enumerate}
@@
 
-\textbf{Particle Activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated.
-It does not get recognized by any Particle-to-Grid-Transfer, nor the grid solve(, nor any Grid-to-Particle-Transfer not affected by the velocity transfer).
+\textbf{Particle activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated.
+It does not get recognized by any particle-to-grid-transfer, nor the grid solve(, nor any grid-to-particle-transfer not affected by the velocity transfer).
 They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide.
@@
 
-\textbf{Particle Resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling.
+\textbf{Particle resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling.
 Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet.
@@
 Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$.
-A variable in the Lagrangian form has a left subscript 0: $ _0{\boldsymbol{x}}$.
-The Eulerian description has a left subscript t: $_t\boldsymbol{x}$.
-Occasionally notation may be omitted where it is apparent from context.
+A variable in the Lagrangian form has a left subscript $0$: $ _0{\boldsymbol{x}}$.
+The Eulerian description has a left subscript $t$: $_t\boldsymbol{x}$.
+Occasionally, notation may be omitted where it is apparent from context.
 
@@
 $\boldsymbol{x}_i$).
-This is not to be confused with the following Einstein-Notation.
-Therefore beginning with chapter \ref{sec:mat_point} Einstein-Notation for components will use greek letters ($\alpha,\beta$).
+This is not to be confused with the following Einstein-notation.
+Therefore, beginning with chapter \ref{sec:mat_point} Einstein-notation for components will use greek letters ($\alpha,\beta$).
 Occasionally summation over grid index $i$ and particles $p$ may be implied.
 A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
-\subsection{Einstein-Notation}
-Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors.
+
+Used throughout this thesis is at instances the \textbf{Einstein-notation} when using tensors and vectors.
 
@@
 \end{equation}
-\cite{MCGINTY:CONTINUUM}
+\begin{flushright}\cite{MCGINTY:CONTINUUM}\end{flushright}
 \clearpage
 \section{Basics}\label{sec:basics}
-The first chapters will derive the Material Point Method and necessary extensions for an elastic model.
+The first chapters will derive the material point method and necessary extensions for an elastic model.
 An elastic model is an interesting case displaying much of the needed computations for a typical model.
@@
 Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
-Rather the continuum assumption holds: each particle represents a continuous piece of material s.t.
+Rather, the continuum assumption holds: Each particle represents a continuous piece of material s.t.
 a microscopic view does not need to be adopted.
@@
 When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
-In the Material Point Method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
+In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
 This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
@@
 Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.
-Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the reference configuration to the current configuration.
-If for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
+Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$ which relates the reference configuration to the current configuration.
+Let for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations.
+Then this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
 \begin{equation}
@@
 \end{equation}
-If, for instance, the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
+If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
 \begin{equation}\label{eq:rigid}
@@
 I.e.
-both configurations refer variables defined on them to the deformed state but the position, where the 'look up' of that value happens, is different.
-In the reference confguration look up is done at the initial position of the particle.
-In the current configuration look up happens at the particle's world position.
+both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different.
+In the reference confguration lookup is done at the initial position of the particle.
+In the current configuration lookup happens at the particle's world position.
 These relations for some particle quantity $f$ are called the (Lagrangian) pull back
@@
 \cite{MPM:COURSE}
-\subsection{Polar- and Singular Value Decomposition}\label{sec:svd}
+\subsection{Polar and singular value decomposition}\label{sec:svd}
 The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$.
@@
 \end{equation}
-As can be seen the deformation gradient contains a rigid rotation.
+As can be seen, the deformation gradient contains a rigid rotation.
 For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}).
@@
 the deformation gradient has two components a constant rotation and the actual distortion or strain.
-There is two ways to deal with this:
+There are two ways to deal with this:
 \begin{enumerate}
@@
 Enforce the convention mentioned above.
-  \item \label{item:qr} \textbf{QR-Factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-Factorization with Givens-Rotations where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
-The $\boldsymbol{QR}$-Factorization is done once in the same fashion as in item \ref{item:sk} - \ref{item:sii}.
+  \item \label{item:qr} \textbf{QR-factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-factorization with Givens-Rotations where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
+The $\boldsymbol{QR}$-factorization is done once in the same fashion as in item \ref{item:sk} - \ref{item:sii}.
 \end{enumerate}
 Due to inherent normalization, fast multiplication and storage efficiency quaternions are preferred over actual rotation matrices.
-In item \ref{item:qr} a $\boldsymbol{QR}$-Factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
+In item \ref{item:qr} a $\boldsymbol{QR}$-factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
 In general $\boldsymbol{R}$ is an upper triangular matrix.
@@
 
-\subsection{Constitutive Models}\label{seq:constitutive_models}
+\subsection{Constitutive models}\label{seq:constitutive_models}
 It is common practice in elastoplascity theory to decompose the deformation gradient in its elastic and plastic parts $\boldsymbol{F} = \boldsymbol{F}_E \boldsymbol{F}_P$.
@@
 For more information refer to the literature \cite{ochsner2014elasto}.
-\subsubsection{Linear Elasticity}\label{sec:linear_elasticty}
+\subsubsection{Linear elasticity}\label{sec:linear_elasticty}
 The first aim will be to find the strain energy density $\Psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
@@
 \begin{flushright}\cite{MIT:LINEAR_ELASTICITY}\end{flushright}
-\subsubsection{Corotated Hyperelasticity}\label{sec:cor_hyper}
+\subsubsection{Corotated hyperelasticity}\label{sec:cor_hyper}
 The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
@@
 \begin{flushright}\cite{ADAMS:ELASTICITY}\end{flushright}
-\subsubsection{Fixed Corotated Hyperelasticity}
+\subsubsection{Fixed corotated hyperelasticity}
 Numerical implicit stepping algorithms mostly rely on using the first and second derivative of $\Psi$, chapter \ref{sec:newton}.
@@
 Under material relaxation these stepping algorithms are attracted to a material-dependent primary contour.
-For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume-preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$).
+For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$).
 Reaching the rest configuration is only the secondary goal.
@@
 
-Therefore they propose to use the Fixed Corotated Hyperelasticity energy density:
+Therefore, they propose to use the fixed corotated hyperelasticity energy density:
 \begin{equation}
@@
 \subsection{Governing equations}
-Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and Material Point Method.
+Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and material point method.
 \subsubsection{Conservation of mass}
@@
 
-Conservation of mass dictates dictates that mass does not depend on motion or time. Only the space occupied by this mass may be more or less.
+Conservation of mass dictates that mass does not depend on motion or time. Only the space occupied by this mass may be more or less.
 \begin{equation}
@@
 \end{equation}
-The conservation of momentum (eq. \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:
+The conservation of momentum (equation \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:
 \begin{equation}
@@
 Such a solution can be found for simplified models analytically.
-These act as a ground truth to numerical methods like the Material Point Method.
+These act as a ground truth to numerical methods like the material point method.
 Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.
@@
 \end{equation}
-In the Material Point Method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
+In the material point method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
 Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
@@
 \begin{flushright}\cite{MPM:COURSE}\cite{strang2007computational}\cite{bathe2006finite}\end{flushright}
-\subsection{Material Point Method}\label{sec:mat_point}
-The key idea of the Material Point Method is to use (Lagrangian) particles as a consistent storage of material properties.
+\subsection{Material point method}\label{sec:mat_point}
+The key idea of the material point method is to use (Lagrangian) particles as a consistent storage of material properties.
 All stress based forces are computed on a Eulerian grid however.
@@
 After stresses are computed there also needs to be a way to get back the relevant information from the grid to the particles.
-Advection is typically hard to do in Eulerian/FEM-like methods and cause of a lot of problems down the development pipeline.
+Advection is typically hard to do in Eulerian/FE-like methods and cause of a lot of problems down the development pipeline.
 In a Lagrangian view particle advection is trivial.
 
-Its is very important that the two transfers as well as the grid solver are in compliance with all governing equations.
+It is very important that the two transfers as well as the grid solver are in compliance with all governing equations.
 While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that conserve the properties defined in them.
@@
 Each material point will be assigned an initial volume $_0V_p$ as well as an initial mass $_0m_p$.
-The volume the material point occupies may change in time due to eq.
-\ref{eq:j}: ${^t_0}J _0V_p = {_tV_p}$.
+The volume the material point occupies may change in time due to equation \ref{eq:j}: ${^t_0}J _0V_p = {_tV_p}$.
 But, due to conservation of mass \ref{eq:lagr_mass} it will have a constant, initial mass associated with it.
@@
 In doing so no information can be lost on the particles.
-Proving, that the mass transfer to the grid is conserving, is enough.
+Proving that the mass transfer to the grid is conserving is enough.
 The stress based-solver may manipulate this information further.
@@
 \begin{flushright}\cite{MPM:APIC}\cite{MPM:COURSE}\end{flushright}
-\subsubsection{APIC transfers}\label{sec:apic}
+\subsubsection{APIC-transfers}\label{sec:apic}
 The momentum transfer round trip could be defined as:
@@
 \end{enumerate}
-While $PIC$-Transfers are very stable, they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
+While $PIC$-transfers are very stable, they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
 This causes a heavy loss in angular momentum and velocity modes.
 $FLIP$-Transfers avoid dissipation and loss of angular momentum by only updating the velocity with a difference \ref{eq:FLIP}.
-However some velocity modes are also not recognized on the grid and may cause unpredictable and unstable behavior in following steps.
-Therefore often a combination of both is taken.
+However, some velocity modes are also not recognized on the grid and may cause unpredictable and unstable behavior in following steps.
+Therefore, often a combination of both is taken.
 
@@
 \begin{equation}
-\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T },
+\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T }
 \end{equation}
@@
 \end{equation}
-This motivates the velocity derivative to be equally defined by a similar relationship, where $\boldsymbol{B}^n_p$ holds momentum information.
+This motivates the velocity derivative to be equally defined by a similar relationship where $\boldsymbol{B}^n_p$ holds momentum information.
 \begin{equation}
@@
 \end{equation}
-The transfers of the $APIC$-scheme are then summarized:
+The transfers of the $APIC$ scheme are then summarized:
 \begin{enumerate}
@@
     \end{equation}
-  \item Grid to particle transfer (in a $PIC$-manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
+  \item Grid to particle transfer (in a $PIC$ manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
     \begin{equation}
@@
 \subsubsection{CFL condition} \label{sec:cfl}
-The CFL condition is prominent for FEM-like methods: For a stable integration a particle should not travel father than the grid spacing $h$ in a discrete time-step $\Delta t$.
-$\Delta t$ is thus limited by:
+The CFL condition is prominent for FEM: For a stable integration a particle should not travel father than the grid spacing $h$ in a discrete time-step $\Delta t$. $\Delta t$ is thus limited by:
 \begin{equation}
@@
 The boundary integral is mostly due to collisions and will be ignored.
-\cite{MPM:OPTIMI_INTEGR} discusses level set collisions due to constraint collisions, object penalty collision and penalty self-collisions.
+\cite{MPM:OPTIMI_INTEGR} discuss level set collisions due to constraint collisions, object penalty collision and penalty self-collisions.
 Collision treatment would need to be involved in the solving of the equation.
@@
 A prominent member of this class is the implicit midpoint rule ($\lambda =\frac{1}{2}$).
-Replacing the Lagrangian acceleration $_0a_\alpha$ in eq.
-\ref{eq:weak_mpm} with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
+Replacing the Lagrangian acceleration $_0a_\alpha$ in equation \ref{eq:weak_mpm} with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
 \begin{equation}\label{eq:time_disc}
@@
 \end{equation}
-where partition of unity $\sum_j w_{jp} = 1$ is used. This is exactly the mass transfer as in eq. \ref{eq:mpm:mass}, s.t. no further assembling of a mass matrix is needed.
+where partition of unity $\sum_j w_{jp} = 1$ is used. This is exactly the mass transfer as in equation \ref{eq:mpm:mass}, s.t. no further assembling of a mass matrix is needed.
 The discretization of the right side of \ref{eq:galerkin} happens with an estimated per particle stress $\boldsymbol{\sigma}_p^{n+\lambda}$:
@@
 
-Given, that the Material Point Method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
+Given, that the material point method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
 Based on this one may also gain a volume change measure around the particle as $J^n_p = \text{det}(\boldsymbol{F}^n_p)$.
@@
 \end{equation}
-In eq. \ref{eq:piola} an alternate measure for the stress by the first Piola-Kirchhoff stress is given. The results of \ref{eq:space_discr} may therefore equally expressed by it:
+In equation \ref{eq:piola} an alternate measure for the stress by the first Piola-Kirchhoff stress is given. The results of \ref{eq:space_discr} may therefore equally expressed by it:
 $$
@@
 \subsubsection{Deformation gradient evolution}
-In eq. \ref{eq:evol_def_grad} the evolution of the deformation gradient is shown to be:
+In equation \ref{eq:evol_def_grad} the evolution of the deformation gradient is shown to be:
 \begin{equation}
@@
 \end{equation}
-Discretizing the Lagrangian deformation gradient in time with eq. \ref{eq:midpoint} results in:
+Discretizing the Lagrangian deformation gradient in time with equation \ref{eq:midpoint} results in:
 \begin{equation}
@@
 \end{equation}
-The discretization of the position will also be of need to advance the particles and weight them back:
+The discretization of the position will also be necessary to advance the particles and weight them back:
 \begin{equation}  \frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1}
@@
   \begin{equation} \label{eq:def_grad_lambda}
-  = (1-\lambda)\boldsymbol{F}^n_p + \lambda \boldsymbol{F}^{n+1}_p
+  = (1-\lambda)\boldsymbol{F}^n_p + \lambda \boldsymbol{F}^{n+1}_p.
 \end{equation}
@@
 \end{equation}
-for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of eq. \ref{eq:force_disc}:
+for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of equation \ref{eq:force_disc}:
 $$
@@
  \end{equation}
-Due to eq. \ref{eq:def_grad_lambda} the stress computation is summarized for the class of functions in use as:
+Due to equation \ref{eq:def_grad_lambda} the stress computation is summarized for the class of functions in use as:
 \begin{equation}
@@
 \subsubsection{Symplectic midpoint scheme}
-Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
+Starting from the last update on the grid which is typically the position with equation \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
 \begin{equation}\label{eq:vel_midpoint}
@@
 \end{equation}
-The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the Material Point Method differs by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
+The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the material point method differs by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
 \begin{equation}\label{eq:trapezoidal}
@@
 This modification allows for a more direct one-step scheme as shown in the following.
-Plugging $\boldsymbol{x}_i^{n+1}$ of eq.
-\ref{eq:vel_midpoint} into \ref{eq:force_midpoint} leads to the discretized momentum equation:
+Plugging $\boldsymbol{x}_i^{n+1}$ of equation \ref{eq:vel_midpoint} into \ref{eq:force_midpoint} leads to the discretized momentum equation:
 \begin{equation}
@@
 Moreover, it easily allows adding potential terms for gravity or collisions \cite{MPM:OPTIMI_INTEGR}, or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
-The analysis of minimizing the objective is equivalent to findino the zero crossing of the derivative:
+The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
 \begin{equation}
@@
 $\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}$ is symmetric, positive definite due it being the Hessian of a convex energy function $E$ of a hyper-elastic material.
-The linear system of eq. \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is two important things to notice:
+The linear system of equations \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is two important things to notice:
 \begin{enumerate}
@@
 \begin{enumerate}
-  \item Compute a particle quantity $\boldsymbol{A}_p$ with a Grid-to-Particle-Transfer:
+  \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid-to-particle-transfer:
     \begin{equation}\label{eq:Ap}
@@
     \end{equation}
-  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a Particle-to-Grid-Transfer:
+  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle-to-grid-transfer:
     \begin{equation}\label{eq:delta_f}
@@
 The speed of these memory types is given in the following starting with the fastest:
-Register > Shared Memory > Local, Global.
+Register > Shared Memory > Local, Global. Table \ref{table:gpu_mem} gives an overview over their capabilities which is referred to in the following.
 
@@
 \textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers.
-The attainable bandwidth for PCIe x16 3.0 between host and device is roughly 12GB/s while the communication between device and GPU is 224GB/s (GTX 970).
+The attainable bandwidth for PCIe x16 3.0 between host and device is 16GB/s while the communication between device and GPU is 224GB/s (GTX 970) \cite{NVIDIA:GTX970}.
 This includes doing calculations on the GPU that may run faster on the CPU.
@@
   \end{tabular}
-\caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}
+  \caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}\label{table:gpu_mem}
 \end{table}
@@
 If the occupancy falls off to 0\%, register spilling occurs.
-The occupancy is either given absolute due the amount of active warp or relative divided by the possible maximum amount of warps on a Streaming Multiprocessor.
-Occupancy allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.), the warp scheduler will switch to another.
+The occupancy is either given absolute due the amount of active warp or relative divided by the possible maximum amount of warps on a Streaming Multiprocessor(SM).
+Occupancy allows the SM to schedule between warps: If the current warp cannot continue (memory latency, barrier etc.), the warp scheduler will switch to another.
 This is called a warp stall, table \ref{table:metric}.
-Occupancy is mostly affected due to resource limits on Streaming Multiprocessors (shared memory, register), as can be seen in the Occupancy Calculator.
+Occupancy is mostly affected due to resource limits on the SM (shared memory, register), as can be seen in the Occupancy Calculator.
 Non optimal occupancy might not have any performance impact at all when the GPU has enough warps to switch to at all times.
@@
 TEMP variables are registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentees these are not further optimized.
+Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentee these are not further optimized.
 Otherwise, close monitoring of the occupancy is recommended.
@@
 Increasing the block size is useful in combination with shared memory usage.
-Shared memory accesses are typically faster than global memory accesses.
+Shared memory accesses arinstructionspically faster than global memory accesses.
 The more threads of a block reuse shared memory data, the faster the memory throughput will be.
@@
 \textbf{Loop unrolling:} Foremost, loop unrolling reduces instructions and can thus speed up compute-bound work.
-Moreover, loop unrolling might help the compiler reorder instructions and thus execute code faster.
+Moreover, loop unrolling might help the compiler reorder instructions and thus execute instructions faster.
 Instructions benefiting greatly from this are global loads.
@@
   \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
-Therefore one might be interested how long the GPU needs to calculate one physical second:
+Therefore, one might be interested how long the GPU needs to calculate one physical second:
     \begin{equation}
@@
     \end{equation}
-    Otherwise, one would calculate the speedup of procedure 1 over procedure 2 for a direct comparison:
+    Otherwise, one would calculate the speedup of one procedure over a another procedure for a direct comparison:
 
@@
 They are however good indicators on which operations/hardware units the GPU spends its time on.
-The top SOL\% metric measures fraction of the throughput to the bandwidth of that hardware unit.
+The top speed of light(SOL)\% metric measures fraction of the throughput to the bandwidth of that hardware unit.
 NVIDIA aims to get the top SOL\% metric between 60\% and 80\%.
@@
 \noindent\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[caption={AoS Layout},language={GLSL},style={GL}]
+\begin{lstlisting}[caption={AoS-Layout},language={GLSL},style={GL}]
 struct Particle {
@@
 \hfill\noindent\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[caption={SoA Layout},label={llst:soa},language={GLSL},style={GL}]
+\begin{lstlisting}[caption={SoA-Layout},label={llst:soa},language={GLSL},style={GL}]
 struct Container{
@@
 Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts.
-The latter is the more commonly taught approach while high performance CPU code may also prefer SoA or the more advanced entity component systems.
+The latter is the more commonly taught approach while high performance CPU code may also prefer SoA.
 This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.
@@
 In contrast, in an AoS layout the cache-line will have a position and a velocity loaded but the shader never uses the velocity.
-This halfes the throughput.
+This halves the throughput.
 This process is shown in Figure \ref{fig:soa} and \ref{fig:aos}.
@@
 
-Furthermore, for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA Layout as in Listing \ref{llst:soa}.
+Furthermore, for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA-Layout as in Listing \ref{llst:soa}.
 As an alternative meta information is provided in the style of reflection macros.
@@
 \end{enumerate}
-For debugging the shader is output with all processed includes.
-The gcc-pre\-pro\-ces\-sor can preprocess the shader with options: \texttt{gcc -E -x c -P}.
+For debugging purposes the shader is written to the hard drive with all processed includes.
+The gcc-pre\-pro\-ces\-sor(\cite{GCC:PREP}) can preprocess the shader with options: \texttt{gcc -E -x c -P}.
 Except for the omission of the glsl version or extension macros no problems were encountered in this process.
@@
 After this the thread does its parallel work.
-Therefore the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.
+Therefore, the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.
 
@@
 However, the added computation time does not outweigh the bank-conflicts.
-So the process is reverted to eq.
-\ref{eq:bank}.
-Note, that this is a correction to \cite{NVIDIA:SCAN} where C++-Operator-Precedence of $+$ over $<<$ is not recognized, and additionally incorrectly $ u << n$ is used in the first part of eq.
-\ref{eq:bank_1024}.
+So the process is reverted to equation \ref{eq:bank}.
+Note, that this is a correction to \cite{NVIDIA:SCAN} where C++-Operator-Precedence of $+$ over $<<$ is not recognized, and additionally incorrectly $ u << n$ is used in the first part of equation \ref{eq:bank_1024}.
 
@@
 For the scan this result is then written back.
-This was enough for the input sizes we employed and can be similarly extended by pyramid structures.
-
-\subsection{Counting Sort \& Stream Compaction}\label{sec:sort}
+This was enough for the input sizes employed and can be similarly extended by pyramid structures.
+
+\subsection{Counting sort \& stream compaction}\label{sec:sort}
 Sorting on the GPU can dramatically increase workload performance of subsequent steps if they can profit from following things:
@@
 
-\textbf{Stream Compaction} reduces applicable data for subsequent operations.
+\textbf{Stream compaction} reduces applicable data for subsequent operations.
 The applicable data is checked by a condition.
@@
 
-\textbf{Counting Sort}: As the sorting algorithm of choice counting sort is used.
+\textbf{Counting sort}: As the sorting algorithm of choice counting sort is used.
 Counting sort is split into three parts:
@@
 Double buffering particles allows to do so.
-However binning is done after a position update that changes the ordering.
+However, binning is done after a position update that changes the ordering.
 Full coalescing will therefore generally not occur.
 
-Alternatively, one could work again with the CFL condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a Particle-to-Grid-Transfer, chapter \ref{sec:p2g}:
+Alternatively, one could work again with the CFL condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a particle-to-grid-transfer, chapter \ref{sec:p2g}:
 \begin{enumerate}
@@
 \end{enumerate}
-Similarly to the Particle-to-Grid transfer one could choose from the different algorithms presented there to do so.
+Similarly to the particle-to-grid-transfer one could choose from the different algorithms presented there to do so.
 However without additional endeavors the order within a bin is not preserved.
@@
 
-\subsection{MPM-Operations}\label{sec:transfers}
+\subsection{MPM operations}\label{sec:transfers}
 Easy to parallelize are all operations that are executed entirely on the grid or entirely on the particles:
@@
 One thread will just equal one node or particle, respectively.
-For one variable assignments a map-shader is used that applies a unary operation to any input element.
+For one variable assignments a map shader is used that applies a unary operation to any input element.
 This can be used for instance for the mass divide of momentum on the grid (equation \ref{eq:mass_divide}) or resetting buffers that are target of global atomic operations.
 
-In contrast much harder are transfers between the two structures that are essential to the Material Point Method:
-
-Particle-to-Grid(P2G)-Transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid, as well as the computation of the Hessian of equation \ref{eq:delta_f}.
+In contrast much harder are transfers between the two structures that are essential to the material point method:
+
+Particle-to-grid(P2G)-transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid, as well as the computation of the Hessian of equation \ref{eq:delta_f}.
 In general these can be summarized as:
@@
 
-The transfers back to the particles are called Grid-to-Particle(G2P)-Transfers.
-Apart from the APIC-Transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}), the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
+The transfers back to the particles are called grid-to-particle(G2P)-Transfers.
+Apart from the APIC-transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}), the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
 In short, all transfers of the form:
@@
 This motivates making the transfers a very highly optimized operation.
-The Material Point Method requires for an elastic material as derived:
+The material point method requires for an elastic material as derived:
 \begin{enumerate}
-  \item One P2G-Transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
-  \item Per Conjugate Gradient Iteration:
+  \item One P2G-transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
+  \item Per conjugate gradient iteration:
     \begin{enumerate}
-      \item One G2P-Transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
-      \item One P2G-Transfer of $\delta \boldsymbol{f}_i$ ($float4$)
+      \item One G2P-transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
+      \item One P2G-transfer of $\delta \boldsymbol{f}_i$ ($float4$)
     \end{enumerate}
-  \item Per Newton Iteration: One G2P-Transfer for $\boldsymbol{F}^{n+1}_p$
-  \item One G2P-Transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
+  \item Per Newton iteration: One G2P-transfer for $\boldsymbol{F}^{n+1}_p$
+  \item One G2P-transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
 \end{enumerate}
 In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models.
-In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks(chapter \ref{sec:active_blocks}) need to be executed only once every physical times step.
+In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks (chapter \ref{sec:active_blocks}) need to be executed only once every physical times step.
 Each transfer can profit from it.
@@
   \item \textbf{Dynamic weights}: The stencil is not static.
-In the MPM the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
+In MPM the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
 This also means the position needs to be temporarily stored to apply the transfer.
@@
 
-In a MPM-Transfer the number of participating nodes can vary.
+In a MPM-transfer the number of participating nodes can vary.
 A node may have one, zero or any positive integer of particles associated with it.
@@
 
-\subsubsection{Grid-to-Particle transfers}\label{sec:g2p}
+\subsubsection{Grid-to-particle transfers}\label{sec:g2p}
 The straightforward solution to this problem is to handle all particles independently from each other.
@@
 This requires two level binning of particles to nodes within blocks \cite{MPM:GPU}.
-Figure \ref{fig:g2p_transfer} show how the G2P-transfer using shared memory is realized.
+Figure \ref{fig:g2p_transfer} shows how the G2P-transfer using shared memory is realized.
 Algorithm \ref{alg:g2p} shows this process briefly.
@@
   \includesvg[svgpath =images/,width=\textwidth]{block_transfer_inv_par}
-  \caption{G2P-Transfer: The transfer of a single particle is split into 64 sequential runs (using cubic weights). Shown are the first two runs.}
+  \caption{G2P-transfer: The transfer of a single particle is split into 64 sequential runs (using cubic weights). Shown are the first two runs.}
   \label{fig:g2p_transfer}
@@
 The threads will in unison run over their relative neighbors within the support and collect the weighted contribution to this particle.
-In this process the particle's position need to be loaded to the weighted corresponding contribution.
+In this process the particle's position needs to be loaded to the weighted corresponding contribution.
 The transfer needs to be repeated for each particle in the bin.
@@
 \begin{algorithm}[htpb]
-  \caption{G2P-Transfer}\label{alg:g2p}
+  \caption{G2P-transfer}\label{alg:g2p}
   \begin{algorithmic}[1]
@@
 This will inevitably lead to shared memory bank conflicts and other difficulties.
-For the P2G-Push-Transfers this would necessarily introduce atomic operations on shared memory again.
+For the particle push transfers this would necessarily introduce atomic operations on shared memory again.
 
@@
 Important to this process is to unroll the corresponding loops that load, write, as well as process these particles.
-Instead of needing to rerun the transfer for each particle, the number of runs now done is the number of non-batched runs divided by the number of particles.
+Instead of needing to rerun the transfer for each particle, the number of runs now is the number of non-batched runs divided by the number of particles.
 This process adds to the amount of registers allocated and can thus also negatively impact performance.
-\subsubsection{Particle-to-Grid-Transfers}\label{sec:p2g}
+\subsubsection{Particle-to-grid-transfers}\label{sec:p2g}
 Straightforward to implement are various variations on global memory.
 Consider first binning was not introduced.
-A grid node would not know how to look up the particles in its neighborhood.
+A grid node would not know how to lookup the particles in its neighborhood.
 This would require the particle to find and write to the grid node in its support.
@@
 In \cite{Meyer2015} the static loop over the support is also parallelized resulting in higher throughput.
-The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-Transfers very significantly.
+The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-transfers very significantly.
 The results of above get reversed and looping performs better.
@@
   \includesvg[svgpath =images/,width=\textwidth]{block_transfer_par}
-  \caption{P2G-Transfer: The first run is shown.
-On the left the P2G-Sync-Transfer is shown that writes into the neighborhood.
-On the right the P2G-Pull approach is shown that instead reads out of the neighborhood.}
+  \caption{P2G-transfer: The first run is shown.
+On the left the P2G-sync-transfer is shown that writes into the neighborhood.
+On the right the P2G-pull approach is shown that instead reads out of the neighborhood.}
   \label{fig:p2g_transfer}
@@
 
-\textbf{P2G-Pull-Transfer}: Sorting and binning however open up the possibility of shared memory implementations.
-The equivalent of the G2P-Transfer for the P2G-Transfer is pulling the relevant data from the neighborhood.
-We will refer to this as the P2G-Pull-Transfer.
+\textbf{P2G-pull-transfer}: Sorting and binning however open up the possibility of shared memory implementations.
+The equivalent of the G2P-transfer for the P2G-transfer is pulling the relevant data from the neighborhood.
 The process is shown in Figure \ref{fig:p2g_transfer}.
 
-It is important to notice that the supports in the P2G-Pull approach are reversed: Left becomes right, right becomes left.
-Other than that the transfer works in the same way as the G2P-Transfer.
+It is important to notice that the supports in the P2G-pull approach are reversed: Left becomes right, right becomes left.
+Other than that the transfer works in the same way as the G2P-transfer.
 Instead of reading from grid nodes, reading from particles is done.
@@
 This typically puts a lot more strain onto shared memory.
-For a simple PIC-Transfer(equation \ref{eq:lin_p2g})
-$$
-(m\boldsymbol{v})_i^n = \sum_p w_{i}(\boldsymbol{x}_p)m_p\boldsymbol{v}^n_p.
+For a simple PIC-transfer (equation \ref{eq:lin_p2g})
+$$
+(m\boldsymbol{v})_i^n = \sum_p w_{i}(\boldsymbol{x}_p)m_p\boldsymbol{v}^n_p
 $$
@@
 
-By construction the Material Point Method tries to avoid matrices on the grid due to storage cost.
+By construction the material point method tries to avoid matrices on the grid due to storage cost.
 Conversely, the particles have matrices defined over them, e.g.
@@
 
-There is one implementation detail left out for the P2G-Pull-Transfer which is solved in chapter \ref{sec:active_blocks}.
+There is one implementation detail left out for the P2G-pull-transfer which is solved in chapter \ref{sec:active_blocks}.
 That is grid nodes do not know if there are any particles in the neighborhood left processing.
@@
 \begin{algorithm}[t]
-  \caption{P2G-Sync-Transfer}\label{alg:p2g}
+  \caption{P2G-sync-transfer}\label{alg:p2g}
   \begin{algorithmic}[1]
@@
 
-\textbf{P2G-Atomic-Transfer}: One can invert the process of pulling into pushing.
+\textbf{P2G-atomic-transfer}: One can invert the process of pulling into pushing.
 Instead of reading from shared memory corresponding to particles, this means now writing to shared memory corresponding grid nodes.
-This immediately alarms one that accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
-
-As can be seen in Figure \ref{sec:p2g} push approaches have the big advantage of only needing to load particles within the block region.
+Accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
+
+As can be seen in Figure \ref{sec:p2g}, push approaches have the big advantage of only needing to load particles within the block region.
 The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
@@
 This process will be the same in the following synchronized approach.
-For a node that is part of any halo, \texttt{atomicAdds()} are a strict requirement.
+For a node that is part of any halo, \texttt{atomicAdd()}s are a strict requirement.
 Their results are only partially computed and need to be complemented by neighbor blocks.
@@
 
-\textbf{P2G-Sync-Transfer}: Warps/Subgroups are inherently synchronized.
+\textbf{P2G-sync-transfer}: Warps/Subgroups are inherently synchronized.
 No race conditions are present between their threads as they all access different memory positions (as can be seen in Figure \ref{fig:p2g_transfer}).
@@
 The dynamic loop has to become static within a thread group.
-The problem is solved in same fashion as in the P2G-Pull method with chapter \ref{sec:active_blocks}.
+The problem is solved in same fashion as in the P2G-pull method with chapter \ref{sec:active_blocks}.
 The maximum particle of each bin within the block without the halo is the minimum amount of times the loop needs to run across the threads to hit uniform flow control.
@@
 \subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
-The P2G-Pull- and P2G-Sync-Transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region.
+The P2G-pull- and P2G-sync-transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region.
 The active particle region is the region that needs to get reduced.
-Therefore it is sufficient for the P2G-Sync-Transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}.
-The P2G-Pull-Transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}.
+Therefore, it is sufficient for the P2G-sync-transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}.
+The P2G-pull-transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}.
 Both are achievable with a single level of parallel reduction (in contrast to a pyramid).
@@
 
-For a P2G-Sync-Transfer one may just start with a parallelization level equal to the blocks employed in \ref{sec:transfers}.
-Then can narrow the shader down to do more sequential work to improve throughput.
-For the P2G-Pull-Transfer one may also do the first but additionally load all rest elements (by the halo).
+For a P2G-sync-transfer one may just start with a parallelization level equal to the blocks employed in \ref{sec:transfers}.
+Then the shader can be narrowed down to do more sequential work to improve throughput.
+For the P2G-pull-transfer one may also do the first but additionally load all rest elements (by the halo).
 From there on out more sequential work might be favorable.
@@
 This process is not coalesced.
-Coalescing will still be guarenteed for all nodes within the block for the P2G-Transfers and G2P-Transfer.
+Coalescing will still be guarenteed for all nodes within the block for the P2G-transfers and G2P-transfer.
 The only uncoalesced access within these shaders is loading the maximum per block count from VRAM which there are few of in comparison.
@@
   \end{tabular}
-  \caption{Error of various shaders on 1024*1024 random elements. Scalars are just between [0.0-1.0]. The vectors have length between [0.0-1.0].}
+  \caption{Error of various shaders on 1024*1024 random elements. Scalars are just between [0.0-1.0]. The vectors have a length between [0.0-1.0].}
   \label{tab:parallel_red}
@@
 \begin{equation}
-  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}.
-\end{equation}
-And more specifically for the case of APIC also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
+  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}
+\end{equation}
+and also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
 \begin{equation}
@@
 The test case spreads one million particles uniformly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}.
-A roundtrip consists of a P2G-Transfer and a G2P-Transfer for relevant variables.
+A roundtrip consists of a P2G-transfer and a G2P-transfer for relevant variables.
 In this process the (random) velocities will get filtered.
@@
    \caption{Error of 1,000 PIC/APIC-roundtrips of 1M uniformly placed particles with random velocity in a $128\times128\times128$ grid.
-Used this instance is a P2G-Sync-Transfer and G2P-Transfer with batching = 4 and block size = (4,4,4).
+Used this instance is a P2G-sync-transfer and G2P-transfer with batching = 4 and block size = (4,4,4).
 Particles are fixed.}
@@
 \subsection{Performance}\label{sec:perf}
-The given metrics in \ref{sec:metrics} lead to valuable insight into how well the GPU can utilize a given shader or program.
+The given metrics in \ref{sec:metrics} give valuable insight into how well the GPU can utilize a given shader or program.
 The GPU was a black box for a long time in this regard.
@@
 Figure \ref{fig:soa} and \ref{fig:aos} in chapter \ref{sec:soa_aos} already gave an overview of the procedure for \texttt{vec4}.
-Table \ref{tab:map_1} compares differences between an AoS- and an SoA-Layout with a buffer of two attributes.
+Table \ref{tab:map_1} compares differences between an AoS- and a SoA-Layout with a buffer of two attributes.
 \begin{table}[htbp]
@@
 Immediately apparent becomes that the SoA-Layout is 2.26 times as fast.
-However both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout.
+However, both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout.
 Though the big difference is in the memory access, the wrong access may be conversely spotted in the non-memory subunits.
@@
 Thus using the second attribute will yield to cache hits.
-But going the extra way of checking for the element in the L2-cache requires an additional look up.
+But going the extra way of checking for the element in the L2-cache requires an additional lookup.
 Thus in table \ref{tab:map_2} L2-SOL and L2 Hit rate go up.
@@
   \caption{Optimization of one parallel reduction dispatch on $1024\times1024$ \texttt{vec4} with \texttt{unary\_op(vec4 x)=length(x)} with block\_size = 1024.
-$\Delta t_c$ is in $\mu s$ Note that the interleaved addressing in this example intentionally commits bank conflicts to show performance impacts.
+$\Delta t_c$ is in $\mu s$.
+Note that, the interleaved addressing in this example intentionally commits bank conflicts to show performance impacts.
 The number in parentheses states the number of additional added sequential elements.
@@
 Note, that each thread group in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done by each thread.
-However due to the additional sequential work the amount of thread groups go down by the same factor.
+However, due to the additional sequential work the amount of thread groups go down by the same factor.
 As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this saturates I/O-Latency and thus improves performance.
@@
 The GTX970 has 13 SMs.
-Therefore work is not parallelized enough on a block level anymore.
+Therefore, work is not parallelized enough on a block level anymore.
 The metric SM active\% will go rapidly down at this point.
@@
 
-The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to build off of with only one \texttt{vec4} transfer.
+The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to based on only one \texttt{vec4} transfer.
 The MPM-Operations are evaluated with two easily reproducible examples.
 
-For the \textbf{G2P-Transfers} the different datasets can be seen in table \ref{table:g2p_random} and \ref{table:g2p_uniform4}.
-The first one uses random particle positions and therefore is overly pessimistic for the Material Point Method.
+For the \textbf{G2P-transfers} the different datasets can be seen in table \ref{table:g2p_random} and \ref{table:g2p_uniform4}.
+The first one uses random particle positions and therefore is overly pessimistic for the material point method.
 Random particle positions as well as random velocities are a configuration that in practice will never occur.
@@
 
-The global G2P-Transfers does not reach desirable throughput in any of the GPU subunits.
+The global G2P-transfers does not reach desirable throughput in any of the GPU subunits.
 The top warp-stall reason is Long Scoreboard: Nearly all information needs to be reloaded uncoalesced from global memory.
@@
 But is not aligned with the grid.
-However the SM already runs on three times its throughput compared to the random data.
+However, the SM already runs on three times its throughput compared to the random data.
 
@@
  \end{tabular}
- \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
+ \caption{G2P-transfers of a $128\times 128\times128$ grid back to one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
 Block size is (8,4,4).}\label{table:g2p_random}
@@
  \end{tabular}
- \caption{G2P-Transfers of a $128\times 128\times128$ grid back to one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
+ \caption{G2P-transfers of a $128\times 128\times128$ grid back to one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
 They form a rotated (unsorted) cube with four particles per cell.
@@
 
-In NVIDIA-Nsight run times recorded for the G2P-Transfers were comparably lower while making use of batching.
-However OpenGL timer queries recorded little difference in trade-off for the added register pressure.
-Thus G2P-Transfers should stay away of batching.
+In NVIDIA-Nsight run times recorded for the G2P-transfers were comparably lower while making use of batching.
+However, OpenGL timer queries recorded little difference in trade-off for the added register pressure.
+Thus G2P-transfers should stay away of batching.
 
@@
 The main benefit of the global solution is that warps do not suffer from branch divergence in that method.
-However the G2P-shared method may still give an edge in performance.
-
-The \textbf{P2G-Transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4}, respectively.
+However, the G2P-shared method may still give an edge in performance.
+
+The \textbf{P2G-transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4}, respectively.
 The global atomic approaches suffer from low throughput across all units.
@@
 The sorting brings forth this problem more acutely as L2-Caches throughput rises to the top SOL with 44.0\% in the uniform case of table \ref{table:p2g_uniform4}.
-The random data example in table \ref{table:p2g_random} relies even more heavily on the L2-Cache with 65.8 SOL\% similarly to what has been seen in the G2P-Transfer.
-
-Nevertheless this already speeds up the random case by a factor of 11.88x as it followed no particular ordering. The sorting makes sure that the incoming data is preprocessed for optimal access.
+The random data example in table \ref{table:p2g_random} relies even more heavily on the L2-Cache with 65.8 SOL\% similarly to what has been seen in the G2P-transfer.
+
+Nevertheless, this already speeds up the random case by a factor of 11.88x as it followed no particular ordering. The sorting makes sure that the incoming data is preprocessed for optimal access.
 
@@
     \hline
-    P2G-Pull               &   26,822             &9.46x           &1.6\% &2.3\% &35.2\%\\\hline
-    P2G-Atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &\textbf{65.4}\%\\\hline
-    P2G-Sync*              &   \textbf{12,801}    &19.87x          &3.6\% &7.3\% &\textbf{61.3}\%\\\hline
+    P2G-pull               &   26,822             &9.46x           &1.6\% &2.3\% &35.2\%\\\hline
+    P2G-atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &\textbf{65.4}\%\\\hline
+    P2G-sync*              &   \textbf{12,801}    &19.87x          &3.6\% &7.3\% &\textbf{61.3}\%\\\hline
  \end{tabular}
- \caption{P2G-Transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+ \caption{P2G-transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
 Block size is (8,4,4).
@@
 
-Using the \textbf{P2G-Pull method} on the random case performs worse than the global method.
+Using the \textbf{P2G-pull method} on the random case performs worse than the global method.
 It can significantly save time however in the uniform case.
-The P2G-Pull-Transfer is 4.31 times faster than the global P2G-Transfer on the sorted uniform data.
+The P2G-pull-transfer is 4.31 times faster than the global P2G-transfer on the sorted uniform data.
 
@@
     \hline
-    P2G-Pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
-    P2G-Atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &\textbf{65.0}\%\\\hline
-    P2G-Sync*              &    \textbf{2,595}    &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
+    P2G-pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
+    P2G-atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &\textbf{65.0}\%\\\hline
+    P2G-sync*              &    \textbf{2,595}    &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
  \end{tabular}
- \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+ \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
 They form a rotated (unsorted) cube with four particles per cell.
@@
 
-A closer look is taken of the P2G-Atomic- and P2G-Sync-Transfers to illustrate the performance improvements of batching and filtering for active blocks in table \ref{table:p2g_optim}.
-
-The \textbf{P2G-Atomic-Transfer} in its initial state uses one atomic transfer per particle on shared memory.
+A closer look is taken of the P2G-atomic- and P2G-sync-transfers to illustrate the performance improvements of batching and filtering for active blocks in table \ref{table:p2g_optim}.
+
+The \textbf{P2G-atomic-Transfer} in its initial state uses one atomic transfer per particle on shared memory.
 Its effects are quite visible on the SM SOL\% with 82.9\% to 90.9\% in the call for active blocks.
@@
 
-The \textbf{P2G-Sync-Transfer} also may use batching and save time.
+The \textbf{P2G-sync-transfer} also may use batching and save time.
 The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}.
@@
 The external synchronization is however much less invasive than atomic writes.
-The P2G-Sync method without batching has roughly the same timing as the P2G-Atomic-Transfer with batching of four particles.
+The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.
 \begin{table}[htpb]
@@
     Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occup.\\\hline
-    P2G-Atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
+    P2G-atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
     indirect		   &    6,276             & 1.10x   &90.8\%&\textbf{S. Scoreb.(31.7\%)} & 43.4\\\hline
@@
     \hline
-    P2G-Sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
+    P2G-sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
     batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & \textbf{35.4}\\\hline
  \end{tabular}
- \caption{P2G-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+ \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
 They form a rotated (unsorted) cube with four particles per cell.
@@
 
-The P2G-Atomic-Transfer is very reliant on batching where the P2G-Sync-Transfer is not.
+The P2G-atomic-Transfer is very reliant on batching where the P2G-sync-transfer is not.
 This is further highlighted in table \ref{table:p2g_area}.
 With one particle per cell(area 47.6\%) batching does not help at all.
-The P2G-Sync-Transfer however can still work comparably fast to either the high density of four particles per cell losing about 7.2\% speed or atomic operations because batching is not available.
-This also means that the P2G-Sync-Transfer only scales slightly with the area covered which is due to the overhead of loading and writing more blocks.
+The P2G-sync-transfer however can still work comparably fast to either the high density of four particles per cell losing about 7.2\% speed or atomic operations because batching is not available.
+This also means that the P2G-sync-transfer only scales slightly with the area covered which is due to the overhead of loading and writing more blocks.
 
@@
     Method                 &  Area=47.6\% & Area=23.8\% &Area=11.9\%\\\hline
-    P2G-Sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
+    P2G-sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
     batching = 2           & -&2,942$\mu s$ & 2,745$\mu s$\\\hline
@@
  \end{tabular}
- \caption{P2G-Sync-Transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
+ \caption{P2G-sync-transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
 They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered with 47.6\%/23.8\%/11.9\%.
@@
     \hline
-    P2G-Sync               &    3,843             &68.9\%&No Instr.(24.9\%) & 43.3\\\hline
+    P2G-sync               &    3,843             &68.9\%&No Instr.(24.9\%) & 43.3\\\hline
     batching=2             &    3,634             &67.3\%&No Instr.(26.5\%) & 31.6\\\hline
@@
 
-The P2G-Atomic- and the even more favorable P2G-Sync-Transfer give great speedups over the global atomic methods.
+The P2G-atomic- and the even more favorable P2G-sync-transfer give great speedups over the global atomic methods.
 Even in a random case (table \ref{table:p2g_random}) they can outdo the global sorted method despite the high branch divergence.
@@
 \section{Conclusion \& Future Work}\label{sec:conclusion}
-The derivation of the Material Point Method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
+The derivation of the material point method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
 Although the MPM consists of many mathematical tasks four operations on the GPU can be identified.
@@
 Deep sorting particles with counting sort in a double-buffered manner allows coalesced accesses and high cache hit rates on transfers.
-Although possible this process does not rely on the CFL condition.
+Although possible, this process does not rely on the CFL condition.
 The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.
@@
 The second preprocessing step is filtering the active blocks of grid and particle transfers with a stream compaction pipeline where no particles are present.
-It also supplies the P2G-Pull- and P2G-Sync-Transfer with a maximum per block bin counter.
+It also supplies the P2G-pull- and P2G-sync-transfer with a maximum per block bin counter.
 
 While the shared memory architecture can give an edge in performance it produces branch divergence and may perform worse over a simpler algorithm relying on the L2-cache.
-The P2G-Sync-Transfer is shown to be an easily extendable and even in edge cases well performing operation.
-Particle-Push-Transfers could be shown to profit from batching as multiple particles are handled in unison.
-The sorted global G2P-Transfer's consistency of utilizing the L2-Cache is a surprise further to be monitored.
+The P2G-sync-transfer is shown to be an easily extendable and even in edge cases well performing operation.
+Particle push transfers could be shown to profit from batching as multiple particles are handled in unison.
+The sorted global G2P-transfer's consistency of utilizing the L2-Cache is a surprise further to be monitored.
 
@@
 However, before this process can be concluded merge and split techniques need to be taken into consideration.
-They are designed to augment the numerics of the Material Point Method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
+They are designed to augment the numerics of the material point method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
 As such they will stray less from the optimistic example.
-How much they will stray from it is however dependent on the dynamics of the problem.
+The extent to which they will stray from it dependends however on the dynamics of the problem.
 \clearpage
@@
 
-\textbf{A Block Scan:} Assume we have already computed a global scan on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
+\textbf{A Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
 This again forms a uniform grid.
@@
 with $0 \leq s,t \leq m,n$.
-The local scan is computed similiar:
+The local scan is computed similar:
 \begin{equation}

### 2019-01-28 mscthesis.tex
smol fixes

@@
 \selectlanguage{english}
-
 \vfill
@@
 Next up is the transfer: Each thread corresponds to one particle within the blue nodes in Figure \ref{fig:g2p_transfer}.
-The threads will in unison run over their relative neighbors within the support and collect the weighted contribution to this particle.
+The threads run in unison over their relative neighbors within the support and collect the weighted contribution to this particle.
 In this process the particle's position needs to be loaded to the weighted corresponding contribution.
@@
 
-Immediately apparent becomes that the SoA-Layout is 2.26 times as fast.
+Immediately apparent becomes that the SoA-Layout is 2.26 times faster.
 However, both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout.

### 2019-01-29 mscthesis.tex
gao et al 2018

@@
 both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different.
-In the reference confguration lookup is done at the initial position of the particle.
+In the reference configuration lookup is done at the initial position of the particle.
 In the current configuration lookup happens at the particle's world position.
@@
 
-Parallel reduction schemes using warp shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}
+Parallel reduction schemes using warp shuffle-operations may use sequential or interleaved addressing.
+Their stride is limited by the memory restrictions imposed due to a warp, Figure \ref{fig:interleaved}.
+\cite{NVIDIA:SHUFFLE}
 
@@
 
-\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
+\subsubsection{Active blocks \& maximum block count:}\label{sec:active_blocks}
 The P2G-pull- and P2G-sync-transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region.
@@
 
-The \textbf{MPM-Transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
+The \textbf{MPM-transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
 \begin{equation}
@@
 In this process the (random) velocities will get filtered.
-In PIC-Transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$.
-APIC-Transfers can conserve the angular momentum better as they define a local velocity field around the particle.
+In PIC-transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$.
+APIC-transfers can conserve the angular momentum better as they define a local velocity field around the particle.
 Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.
@@
 \end{table}
-\clearpage
 \subsection{Performance}\label{sec:perf}
@@
 \caption{Metrics between different buffer layouts.
-Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length}\texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
-\label{tab:map_2}
+Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length} \texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}\label{tab:map_2}
 \end{table}
@@
 
-The \textbf{P2G-sync-transfer} also may use batching and save time.
-The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}.
-Batching in this case reduces the amount of thread group barriers needed for synchronization.
-The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process.
-The external synchronization is however much less invasive than atomic writes.
-The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.
 \begin{table}[htpb]
@@
 \end{table}
+
+The \textbf{P2G-sync-transfer} also may use batching and save time.
+The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}.
+Batching in this case reduces the amount of thread group barriers needed for synchronization.
+The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process.
+The external synchronization is however much less invasive than atomic writes.
+The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.
 
@@
 As such they will stray less from the optimistic example.
-The extent to which they will stray from it dependends however on the dynamics of the problem.
+The extent to which they will stray from it depends however on the dynamics of the problem.
+
+At the end of this thesis awareness was raised to the following publication \cite{Gao:2018} which was simultaneously being worked on. Their decision making is very much alike:
+
+Counting/Histogram sort is also their sorting algorithm of choice in the fashion of \cite{NVIDIA:NNSEARCH}.
+Instead of reordering any variable, they choose per variable if an index or deep sort is required.
+Their binning also floors in the same fashion.
+
+The domain decomposition consists of $4\times4\times4$ blocks, also one of the best options employed here.
+Instead of filtering the active blocks in the pipeline, they motivate this process earlier by employing a sparse grid structure with fast neighbor queries altogether.
+
+Implemented is a quadratic interpolation instead of a cubic one.
+The domain shift thus becomes $\frac{h}{2}$ instead of $h$.
+Accounting for the larger cubic weighting support, uniform distribution works as fast as the one shown here.
+
+Their main difference in computation is the use of the aforementioned NVIDIA exclusive shuffle operations in the fashion of \cite{NVIDIA:SHUFFLE}.
+These operations largely solve two of the main problems the transfers have.
+
+Instead of needing to rely on shared memory for inter-thread communication, the shuffle operation may be used.
+Each particle is assigned to a single thread.
+If for instance the particle size within the $4\times4\times4$ block exceeds the thread group size, another virtual block may be assigned to the same geometric block.
+
+Each particle in a particle-to-grid-transfer then scatters their contribution to the neighboring grid nodes in a particle push fashion.
+The transfers are then solved due to segmented parallel reduction where the segments are the variable particle bin sizes.
+In shared memory this would cause numerous bank conflicts in both the scattering as well as the segmented parallel reduction as elucidated with the block scan of chapter \ref{sec:red_scan}.
+The scatter write would also need to be atomic.
+
+This implementation tried to avoid these steps and schedule over grid nodes.
+This however is the cause of thread divergence which slows down computation for varying bin sizes.
+Therefore, \cite{Gao:2018} will achieve more throughput and better performance in these cases in exchange for a (at the moment) hardware vendor exclusive operation.
+
 \clearpage
@@
 
-\textbf{A Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
+\textbf{Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid.
+A local scan of an n-dimensional tile/block in that grid can be computed based on this global scan.
 This again forms a uniform grid.

### 2019-02-07 mscthesis.tex
pres begin

@@
 Furthermore, chapter \ref{sec:metrics} will elaborate metrics to quantify optimizations.
-\subsubsection {GPGPU Optimization opportunities}\label{sec:optim}
+\subsubsection{GPGPU Optimization opportunities}\label{sec:optim}
 The speed of these memory types is given in the following starting with the fastest:
@@
 \begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
-\subsubsection{GPU Metrics}\label{sec:metrics}
+\subsubsection{GPU metrics}\label{sec:metrics}
 The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
@@
     Barrier\%         & 100                   & block synchronization \\\hline
-    No Instruction\%  & 100                   & instruction fetching or\\
+    No Instruction\%  & 100                   & instruction fetching or \\
 		      &                       & waiting on instruction cache miss\\\hline
   \end{tabular}
-\caption{GPU Performance Metrics}\label{table:metric}
+\caption{GPU performance metrics}\label{table:metric}
 \end{table}
@@
 Occupancy was discussed in \ref{sec:optim}.
-\cite{NVIDIA:PEAK_PERFORMANCE}
 
@@
 Furthermore, for compute-bound work often the top warp stall reason is waiting on instruction fetches.
-
+\begin{flushright}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
 \clearpage
@@
 
-\subsection{Parallel Reduction \& Scan} \label{sec:red_scan}
+\subsection{Parallel reduction \& scan} \label{sec:red_scan}
 The schemes presented in the following are a result of the optimization opportunities discussed in \ref{sec:optim}.
 
-\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$ reduction computes:
+\textbf{Parallel reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$ reduction computes:
 \begin{equation}
@@
 
-\textbf{Counting Sort:} The aim of sorting is coalescing as well as increased caching behavior for all following shaders.
+\textbf{Counting sort:} The aim of sorting is coalescing as well as increased caching behavior for all following shaders.
 As a motivation the binning's similarity to the MPM-Operations might already be highlighted.
@@
 
-The \textbf{P2G-atomic-Transfer} in its initial state uses one atomic transfer per particle on shared memory.
+The \textbf{P2G-atomic-transfer} in its initial state uses one atomic transfer per particle on shared memory.
 Its effects are quite visible on the SM SOL\% with 82.9\% to 90.9\% in the call for active blocks.
@@
 
-At the end of this thesis awareness was raised to the following publication \cite{Gao:2018} which was simultaneously being worked on. Their decision making is very much alike:
+At the end of this thesis awareness was raised to \cite{Gao:2018} which was simultaneously being worked on. Their decision making is very much alike:
 
@@
 Their main difference in computation is the use of the aforementioned NVIDIA exclusive shuffle operations in the fashion of \cite{NVIDIA:SHUFFLE}.
-These operations largely solve two of the main problems the transfers have.
-
-Instead of needing to rely on shared memory for inter-thread communication, the shuffle operation may be used.
+These operations solve two of the main problems the transfers have.
+
+Instead of needing to rely on shared memory for inter-thread communication, the shuffle (and ballot) operation may be used.
 Each particle is assigned to a single thread.
-If for instance the particle size within the $4\times4\times4$ block exceeds the thread group size, another virtual block may be assigned to the same geometric block.
+If the particle size within the $4\times4\times4$ block exceeds the thread group size, another virtual block may be assigned to the same geometric block.
 
@@
 In shared memory this would cause numerous bank conflicts in both the scattering as well as the segmented parallel reduction as elucidated with the block scan of chapter \ref{sec:red_scan}.
-The scatter write would also need to be atomic.
+The scattered write would also need to be atomic.
 
@@
   \begin{code}
-\caption{$\boldsymbol{D}_p$ proof}
   \begin{minted}{python}
@@
 \end{minted}
+\caption{$\boldsymbol{D}_p$ proof}
 \end{code}
@@
   $$
-\section{Block Scan}\label{sec:block_scan}
-
-\textbf{Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid.
+\section{Block scan}\label{sec:block_scan}
+
+\textbf{Block scan:} Assume a global scan is already computed on a n-dimensional uniform grid.
 A local scan of an n-dimensional tile/block in that grid can be computed based on this global scan.

### 2019-02-07 pres/pres.tex
pres begin

--- 2019-01-28 mscthesis.tex
@@
-\documentclass[m,times]{cgMA}
+\documentclass{beamer}
 \usepackage{listings}
@@
 \usepackage[backend=biber,style=alphabetic]{biblatex}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithms
-\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
 \usepackage{mathtools}
 \usepackage{subdepth}
-\usepackage{enumitem}
-\usepackage[toc,page]{appendix}
-\usepackage{minted}
-\usepackage{verbatim}
 \usepackage{graphicx}
@@
 \usepackage{xcolor,colortbl}
-\addbibresource{mscthesis.bib}
 
+\addbibresource{../mscthesis.bib}
 \captionsetup{justification=raggedright,singlelinecheck=false}
@@
 \begin{document}
-\author{Fabian Meyer}
-\title{GPU-Beschleunigung der Material Point Method}
-\zweitgutachter{Bastian Krayer, M.Sc.}
-\zweitgutachterInfo{(Institut f{\"u}r Computervisualistik, AG Computergraphik)}
-% Umschalten der Sprache (fuer englische Rubrikbezeichnungen etc.)
-
-\pagenumbering{roman}
-\maketitle
-
-\clearpage
-\selectlanguage{english}
-\vfill
-\begin{center}
-\subsubsection*{Zusammenfassung}
-\end{center}
-Die Material Point Method hat sich in der Computergrafik für physikalische basier\-te Simulationen etabliert.
-Sie benutzt ein hybrides Modell aus Lagrange-Partikeln und Euler-Gitter.
-Während Partikel als konsistenter Speicher fungieren, erlaubt das Gitter anfallende partielle Differentialgleichungen effizient zu lösen.
-
-\noindent Die Material Point Method unterliegt hohen Berechnungszeiten, die die Methode nur für Hero Shots rentabel macht.
-Sie ist jedoch hoch-parallelisierbar.
-Diese Arbeit zeigt, wie die Methode mit GPGPU-Techniken beschleunigt werden kann.
-
-\noindent Der Datenaustausch von Partikel und Gitter wird über interpolierende Transfers erreicht.
-Pro physikalischen Zeitschritt werden diese mehrmals ausgeführt.
-Vorverarbeitungsschritte können unternommen werden, um diese Transfers performanter zu machen.
-
-\noindent Countingsort für jede Partikel-Variable erhöht Coalescing und L2-Cache Hitraten.
-Binning teilt das Gitter in Blöcke auf, die Shared-Memory Implementationen er\-mög\-lich\-en.
-Die Größen der Bins sind für keine Operation beschränkt.
-Weiterhin werden nur Blöcke ausgeführt, die Informationen erhalten.
-
-\vfill
-\begin{center}
-\subsubsection*{Abstract}
-\end{center}
-The material point method is allowing for physically based simulations.
-It has found its way into computer graphics and since then rapidly expanded.
-The material point method's hybrid use of Lagrangian particles as a persistent storage and a background uniform Eulerian grid enables solving of various partial differential equations with ease.
-
-\noindent The material point method suffers from high execution times and is thus only viable for hero shots.
-The method is however highly parallelizable.
-Thus, this thesis proposes how to accelerate the material point method using GPGPU techniques.
-
-\noindent Core of the material point method are grid and particles transfers that interpolate between the two structures.
-These transfers are executed multiple times per physical time step.
-Preprocessing steps might be taken if their computing time is outweighed.
-
-\noindent Deep sorting with counting sort increases coalescing and L2 cache hit rates.
-Binning allows to divide the grid into blocks for shared memory filtering techniques.
-All operations do not rely on fixed bin size.
-As another preprocessing step, only grid blocks are executed which have particles in them.
-\vfill
-\clearpage
-\tableofcontents
-
-%\setcounter{page}{3}
-\clearpage         % oder \cleardoublepage bei zweiseitigem Druck
-% \listoffigures   % fuer ein eventuelles Abbildungsverzeichnis
-% \clearpage
-\pagenumbering{arabic}
-
-\section{Introduction} \label{intro}
-General purpose computation on graphics processing unit(GPGPU) has elevated graphics processing units(GPUs) to any computational field utilizing massively parallelizable algorithms.
-As such is the nature of discretizations in physics.
-
-The higher parallelization acceleration of GPUs is a blessing for those applications that can profit from it.
-In unison with efficient algorithms they still have much potential to offer.
-Conversely, algorithms which are not parallelizable perform much worse and should be avoided.
-Dividing up these tasks between CPU and GPU is not recommendable.
-Transferring data over the PCI-Bus between the two is slow for big amounts of data and requires synchronization.
-Tasks in such a pipeline should thus stay on the GPU although they might perform worse.
-
-Programming for the GPU is another hurdle.
-GPU code is reliant on the generation and hardware vendor of the GPU to use relevant extensions.
-Backwards compatibility therefore is mostly ignored and written towards a certain GPU generation.
-This is amplified due to the GPU code often being reliant on architecture limits for maximum performance.
-The major languages to target the GPU are CUDA, OpenCL, Direct 3D, OpenGL, and Vulkan.
-They build on top of a C/C++-language subset.
-
-Whereas CUDA offers the most libraries and optimized code, it is only available on NVIDIA GPUs.
-OpenCL's development on NVIDIA GPUs is dragging far behind the standard.
-Its merge into the Vulkan roadmap could be an incentive to support a more in-depth language universally \cite{OPENCL:ROADMAP}.
-Direct3D is only available on Windows.
-Vulkan is the successor of OpenGL and removes work from the driver to the developer.
-Main benefit over OpenGL is less CPU usage which helps little for the almost full GPU implementation here.
-This implementation opts for OpenGL as Vulkan is still relatively new.
-Everything presented could be implemented in one of the other languages as well.
-
-The simulation of the dynamics of materials is still extremely challenging in different applications and fields.
-Multiphysics interactions between materials are not realized on a big scale and often are not possible or feasible.
-The material point method(MPM) has already reliable tackled many different models, materials, and even interaction between materials.
-
-Pure Eulerian methods like the finite element method(FEM) have been in use for long in engineering tasks but face problems handling topology changes with numerical stability.
-Pure Langrangian methods form the end other of the spectrum.
-Prominent are smoothed-particle hydrodynamics(SPH) that are restricted to (viscous) fluids and are of empirical nature.
-Particle in cell(PIC) techniques like the material point method are hybrid Lagrangian/Eulerian methods.
-They try to benefit from the nature of both.
-Advection on particles is trivial.
-Thus the material point method utilizes Lagrangian particles as a consistent storage of dynamic properties.
-The governing equations of the physical problem can be discretized easily to and solved on a uniform Eulerian grid.
-The grid is unmoving and does not store any information between physical time steps.
-
-Transfers between the Lagrangian and Eulerian are of need to bring the two together.
-These transfers are the focus of particle in cell techniques.
-These transfers are realized due to interpolation.
-Thus they create (estimable) numerical error and lead to dissipation.
-Moreover grid nodes are mostly outnumbered by particles leading to more numerical error as modes of the single particles are not realized by the grid.
-
-The $p$ particles to one node relationship of a particle-to-grid(P2G)-transfer or inversely one node to $p$ particles of a grid-to-particle(G2P)-transfer is the main problem when realizing these operations on the GPU.
-When faced with the static architecture of the GPU, this will inevitably lead to branch divergence of threads within a subgroup as some nodes are assigned more particles than others.
-
-Chapter \ref{sec:rel_work} looks at the relevant literature of PIC-Methods specifically on the GPU as well as material point method extensions.
-In chapter \ref{sec:notation} notations which will be used throughout the thesis are laid out.
-The basics of chapter \ref{sec:basics} cover the derivation of an elastic material in the MPM to get an understanding of the prominent operations.
-Furthermore GPU intrinsics, metrics and optimization opportunities are discussed.
-Chapter \ref{sec:implementation} tends to the implementation.
-The evaluation of the implementation follows up in chapter \ref{sec:eval}.
-As a conclusion chapter \ref{sec:conclusion} gives a summarizing overview and tends to future work.
-
-\clearpage
-\section{Related Work}\label{sec:rel_work}
-\textbf{Particle in cell} techniques were developed initially for hydrodynamics \cite{evans1957particle}.
-They are also widely used in plasma simulations to solve Maxwell's equations \cite{PIC:GPU}.
-Their key steps are identical to the MPM:
+\begin{frame}
   \begin{enumerate}
-  \item Transfer particle data to a grid.
-\item \label{it:solve_gov_eq}Solve the governing partial differential equations on the grid to move forward in time.
-  \item Transfer grid data back to particles.
+    \item \cite{MPM:COURSE}
   \end{enumerate}
-In general the grid is regarded as a scratch-pad since its only a temporary storage for the means of computing item \ref{it:solve_gov_eq}.
-I.e.
-the grid could be deleted each physical times step.
-Based on PIC to reduce dampening and momentum conservation Fluid-Implicit-Particle(FLIP) was developed for fluid simulations and is still widely in use for large simulations.
-Their counterpart being the empirical smoothed-particle hydrodynamics \cite{gingold1977smoothed}.
-
-As a further development the Affine-particle in cell(APIC) technique was introduced as an improvement to PIC and FLIP solvers.
-Due to the definition of a local affine velocity field around each particle (instead of a single vector) APIC conserves angular momentum, reduces damping and improves stability over PIC and FLIP, more in chapter \ref{sec:apic}.
-\cite{MPM:APIC} \cite{MPM:OLD_APIC}
-
-A step further goes PolyPIC giving the velocity field even more freedom by adding polynomial velocity modes of increasing order to the description. \cite{MPM:POLYPIC}
-
-The \textbf{material point method} is a further development to the PIC methods.
-Both use analysis and numerics to be derived without the need for empirical assumptions.
-Originally it was developed by \cite{sulsky1995application} with a focus on the dynamics of solids.
-\cite{MPM:SNOW} brought the method into computer graphics for their animation film \textit{Frozen}.
-In rapid succession a variety materials and effects got modeled by the MPM.
-Since MPM is based on discretizing governing equations and using constitutive models covered by the different literature, its fast rate of expenditure is not a surprise.
-
-The simulation of (hyper-)elastic materials can be done with a number of different strain measures utilizing fixed Linear strain (chapter \ref{sec:cor_hyper}), Green-Lagrangian strain tensor (chapter \ref{sec:svd}), or logarithmic Hencky-strain (\cite{MPM:SHELLS}) for the different requirements of small to large strains.
-Inserting these into the different energy density $\boldsymbol{\Psi}(\boldsymbol{\epsilon})$ models can already result in a constitutive model.
-Plasticity in general requires extra modeling in terms of a flow rule, yield-condition, and a hardening rule which are not discussed here \cite{ochsner2014elasto}.
-To give a short impression of the developments in computer graphics:
-
-\cite{MPM:PHASE_CHANGE} model phase transition (melting) to a liquid with using the fixed corotational hyperelasticity model (chapter \ref{sec:cor_hyper}) in combination with a discretization of the heat equation.
-\cite{MPM:OLROYDB} use an Olroyd-B model with Cauchy-Green strain to handle elasticity and plasticity for viscoelastic fluids, foams, and sponges.
-\cite{MPM:DRUCKER} uses Hencky-strain in combination with a St.
-Venant-Kirchhoff model and Drucker-Prager elastoplascity to model sand.
-\cite{MPM:MULTI} combine particles of different species (sand and water) by one grid for each species and relating them with a heuristic momentum exchange.
-\cite{MPM:SHELLS} simulates thin shells (cloths,cups etc.) with a Kirchhoff-Love model using Hencky-strains.
-
-There exist several implementations utilizing the \textbf{GPU for particle in cell} techniques, which the material point method is a part of:
-
-\textbf{Particle sorting:} \cite{PIC:GPU} studies different particle list sorting methods.
-Methods like 'Message Passing Sort' \cite{kong2011particle} \cite{decyk2011adaptable}, 'In Place Particle-QuickSort' \cite{stantchev2008fast} and 'Linked List Ordering' \cite{burau2010picongpu} try to utilize the partial sorting and only update particles that need to be moved within the list.
-These methods often rely on the assumption that only a small number of particles change their grid node, where in contrast MPM can be quite dynamic.
-Those that do move, move at most to one neighboring grid-node or stay in the same node due to the CFL condition, chapter \ref{sec:cfl}.
-These methods often use fixed bin-sizes which can be quite memory intensive and add additional memory management.
-Fixed size binning makes the methods mentioned before easier to implement which don't necessarily map well to the GPU due to their high complexity and uncoalesced accesses.
-As a result \cite{PIC:GPU} come to the conclusion to fall back to the CUDA library \texttt{thrust}, although they do note that the aforementioned methods can perform faster for particles that move only a limited amount of grid nodes.
-\texttt{thrust} provides the radix sort method \texttt{thrust::sort\_by\_key()}, where one sorts by grid index as key.
-
-In general deep or index sorts can be done. A deep sort reorders every variable of a particle. An index sort gives back an index with an offset to access the particle variable.
-
-\cite{NVIDIA:NNSEARCH} compare radix to counting sort utilizing a uniform grid and use the latter for SPH-fluids.
-
-This implementation compares index and deep counting sort and uses the latter.
-Although possible, this counting sort stays away from the limited movement implied by the Courant–Friedrichs–Lewy(CFL) condition.
-Fixed size bins are an assumption this implementation also does not want to impose on the method.
-
-\textbf{Particle-to-grid-transfers} are generally done in two manners \cite{stantchev2008fast}:
-\begin{enumerate}
-  \item \textbf{Particle push:} X particles in a local neighborhood of a grid point push their attributes onto it.
-Since this relationship results in a race condition \texttt{atomicAdd(float f)}s or explicit synchronization is necessary, e.g.
-a\-tomic: \cite{PIC:GPU}, synchronized: \cite{FRANCESCO:ROSSI}.
-  \item \textbf{Particle pull:} A grid point pulls the particles in its local neighborhood from a sorted particle list. This has the benefit of no race conditions being present.
-\end{enumerate}
-A \textbf{grid-to-particle-transfer} is in general easier since a particle already knows, due to its position, the grid points in its local neighborhood.
-The grid points in a uniform grid are 'sorted' by nature.
-This implementation tests one particle pull and two particle push methods.
-
-Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition.
-Shared memory is however limited per GPU (GTX 970, 48KB per block \cite{NVIDIA:GTX970}).
-The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block.
-Since the number of particles per grid point in the MPM is generally between 4-10, only very limited block sizes are available for pull methods that greedily load all particles \cite{MPM:SNOW}.
-
-Stencil computations and filtering techniques underlie a very similar process to MPM-transfers.
-Although these techniques employ a one-to-one grid relationship, since no particles are involved, it is worth taking a look at them.
-An interesting prospect for filtering techniques on the GPU is given by NVI\-DIA's new shuffle operations.
-Shuffle allows to 'share' register memory between threads of a warp \cite{NVIDIA:SHUFFLE}.
-They are however not supported on other architecture, yet.
-
-Widely used is a 2.5D blocking: A domain (of the domain decomposition) is split into 2D-planes along one axis.
-Data reuse of shared memory is maximized by a cyclic-queue strategy: Start a thread for each grid node in the first plane and load the halo along the axis into shared memory.
-Then each thread iterates along the axis over the 2D-planes synchronously, discarding the oldest plane in memory while loading the next plane (along the axis) into shared memory.
-Thus there are no z-dimension halos.
-This however requires all particles in the halo to be present in the limited shared memory.
-As mentioned, this implementation wants to stay away from assumptions limiting particles similar to fixed bin sizes.
-\cite{brandvik2010sblock} \cite{williams2007scientific} \cite{krotkiewski2013efficient}
-
-In particle push methods partial sums are also computed on the halo. Two different techniques exist to get the data between blocks coherent again:
-\begin{enumerate}
-\item Atomically add up the values of the halo.
-This results in two atomic adds on a side node, four on an edge node, and eight on a corner node; If the block extend is bigger than half the halo extend, otherwise more.
-\item Every halo of every block corresponds to actual global memory.
-After a transfer, do add steps along each of the three axes.
-This is infeasible for large supports due to the added memory requirements.
-\cite{crassin2011interactive}
-\end{enumerate}
-
-\textbf{Particle activation:} If the particle's velocity is under a user-defined threshold, it becomes deactivated.
-It does not get recognized by any particle-to-grid-transfer, nor the grid solve(, nor any grid-to-particle-transfer not affected by the velocity transfer).
-They may of course be reactivated if the velocity in the neighborhood exceeds that threshold again or they collide.
-\cite{MPM:GPU}
-
-Dividing the particles up can be realized with stream compaction algorithms discussed in chapter \ref{sec:implementation}.
-
-\textbf{Particle resampling:} A method that could alleviate pressure off fixed bin sizes and make shared-memory restrictions less of a factor is particle resampling.
-Their intention mostly is to fill numerical gaps in the material at large deformations where the material is not supposed to break yet.
-Split and merge methods can directly control how many particles in a cell are allowed and accordingly increase or reduce them.
-
-A simple \textit{Split} method divides one particle into eight with a distance of half the diagonal length of a grid cell from the original particle.
-The eight particles span a cube that can be randomly rotated if desired.
-Mass and volume get equally distributed; Velocity and deformation gradient get copied.
-Refer to chapter \ref{sec:mat_point} or for conservation of mass and momentum in MPM.
-\cite{gao2017adaptive}
-
-A \textit{Merge} method looks for the closest neighbors and creates the new merged particle at the geometric center.
-Mass and volume are accumulated.
-Velocity is computed from a mass-weighted average of the participating particles.
-The deformation gradient is accordingly to chapter \ref{sec:svd} decomposed into $\boldsymbol{F} = \boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T$.
-The quaternion average of $\boldsymbol{U}$s and $\boldsymbol{V}$s of the particles lead to an average rotation $\overline{\boldsymbol{U}}$ and $\overline{\boldsymbol{V}}$, respectively.
-The principal strains within $\boldsymbol{\Sigma}$ are just averaged component-wise to $\overline{\boldsymbol{\Sigma}}$.
-One can compose them in the same fashion leading to an average deformation gradient with $\overline{\boldsymbol{F}} = \overline{\boldsymbol{U}}\overline{\boldsymbol{\Sigma}}\overline{\boldsymbol{V}}^T$.\cite{gao2017adaptive}
-
-Unconditionally keeping particles on a fixed size is not desirable due to fracture requiring that gaps will be created.
-\cite{gao2017adaptive} produce a signed distance field to identify interior points.
-Surface points also should not merge or split due to visual artifacts.
-Furthermore, the signed distance field can be used to adapt the grid size depending on the distance to the surface.
-This allows for coarser levels with increasing distance from the surface.
-%look again for MPM/PIC gpu impl
-%SPH, FEM methods
-\clearpage
-\section{Notation}\label{sec:notation}
-Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$.
-Vectors furthermore use lower-case letters $\boldsymbol{v}$ with the exception of angular momentum $\boldsymbol{L}$; Matrices use upper-case letters $\boldsymbol{M}$.
-A variable in the Lagrangian form has a left subscript $0$: $ _0{\boldsymbol{x}}$.
-The Eulerian description has a left subscript $t$: $_t\boldsymbol{x}$.
-Occasionally, notation may be omitted where it is apparent from context.
-
-Beginning from chapter \ref{sec:mat_point} variables defined on a particle will have right subscript $p$ (e.g.
-$\boldsymbol{x}_p$).
-Grid cells will be assigned a right subscript $i$ or additionally $j$ denoting the grid index (e.g.
-$\boldsymbol{x}_i$).
-This is not to be confused with the following Einstein-notation.
-Therefore, beginning with chapter \ref{sec:mat_point} Einstein-notation for components will use greek letters ($\alpha,\beta$).
-Occasionally summation over grid index $i$ and particles $p$ may be implied.
-A variable of the $n$-th time step with associated time $t^n = \sum_{i=1}^{n-1} \Delta t^i$ will have a right superscript: $\boldsymbol{x}^n$.
-
-Used throughout this thesis is at instances the \textbf{Einstein-notation} when using tensors and vectors.
-
-Repeated indices, that are also defined on the variable, imply component-wise operations.
-Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors.
-$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$.
-Vector and tensor/matrix addition thus become:
-\begin{equation} \label{EINSTEIN:ADD}
-  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }.
-\end{equation}
-A transpose operator swaps the indices:
-\begin{equation}
-  D^T _ { i j }  =D_{ji}.
-\end{equation}
-Repeated indices, that are not otherwise defined, however imply summation on that index. The vector dot product becomes:
-\begin{equation}
-  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }.
-\end{equation}
-Following this notation the Frobenius inner product between two second-order tensors is (also called Frobenius scalar product):
-\begin{equation} \label{EINSTEIN:FROBENIUS1}
-  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }.
-\end{equation}
-The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
-\begin{equation}
-  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }.
-\end{equation}
-Matrix-Vector and Matrix-Matrix multiplication can be expressed this way as:
-\begin{equation}
-  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }.
-\end{equation}
-The definition of the Kronecker Delta is as follows:
-\begin{equation}
-  \delta _ { i j } = \left\{
-  \begin{array} { l l l }
-      { 1 } & {\text{if}} &  {i = j}  \\
-      { 0 } & {\text{if}} &  { i \neq j }
-  \end{array}. \right
-\end{equation}
-The Kronecker Delta as an operator is most efficiently described as a substitution:
-\begin{equation}
-  a _ { i } \delta _ { i j } = a _ { j }.
-\end{equation}
-I.e. this component is only evaluated if $i = j$. The Kronecker Delta comes up in differentiating a variable by itself with same or different index:
-\begin{equation}
-  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}.
-\end{equation}
-The alternating tensor or Levi-Civita symbol
-\begin{equation}
-  \varepsilon _ { i j k } = \left\{ \begin{array} { r l } { + 1 } & { \text { if } ( i , j , k ) \text { is } ( 1,2,3 ) , ( 2,3,1 ) , \text { or } ( 3,1,2 ) } \\ { - 1 } & { \text { if } ( i , j , k ) \text { is } ( 3,2,1 ) , ( 1,3,2 ) , \text { or } ( 2,1,3 ) } \\ { 0 } & { \text { if } i = j , \text { or } j = k , \text { or } k = i } \end{array} \right.
-\end{equation}
-is used to express cross-products $\boldsymbol{c}=\boldsymbol{a} \times \boldsymbol{b}$:
-\begin{equation}
-  c_i = \varepsilon_{ijk} a_j b_k \left(= \sum _ { j = 1 } ^ { n } \sum _ { k = 1 } ^ { n } \varepsilon_{ijk} a_j b_k\right).
-\end{equation}
-A useful relation to the Kronecker-Delta is:
-\begin{equation} \label{eq:compact_levi}
-  \varepsilon _ { i j k } \varepsilon _ { i m n } = \delta _ { j m } \delta _ { k n } - \delta _ { j n } \delta _ { k m }.
-\end{equation}
-\begin{flushright}\cite{MCGINTY:CONTINUUM}\end{flushright}
-\clearpage
-\section{Basics}\label{sec:basics}
-The first chapters will derive the material point method and necessary extensions for an elastic model.
-An elastic model is an interesting case displaying much of the needed computations for a typical model.
-The last chapter \ref{sec:gpgpu} tends to the GPGPU side.
-\subsection{Reference and current configuration}\label{sec:ref_corr_config}
-Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle.
-Rather, the continuum assumption holds: Each particle represents a continuous piece of material s.t.
-a microscopic view does not need to be adopted.
-A particular body is composed of a set of particles and can adapt different configurations due to changes in shape.
-These changes are caused by external or internal effects (forces etc.) on it and deform the body over time.
-\cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}
-
-When modelling solids, changes of quantities from an initial reference configuration $_0\boldsymbol{x}$ to another current configuration $_t\boldsymbol{x}$ need to be measured.
-In the material point method the reference configuration $_0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$).
-This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
-
-Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the reference and current configuration respectively.
-Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$ which relates the reference configuration to the current configuration.
-Let for simplicity $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations.
-Then this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
-\begin{equation}
-  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
-\end{equation}
-If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , this mapping is defined as:
-\begin{equation}\label{eq:rigid}
-  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
-\end{equation}
-The velocity of a material point in the reference configuration can be defined using this mapping
-\begin{equation}\label{eq:velocity}
-  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial ({^t_0}\phi)}{\partial t}(_0\boldsymbol{x},t),
-\end{equation}
-and similarly the acceleration is defined
-\begin{equation}
-  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 ({^t_0}\phi)}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
-\end{equation}
-It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space.
-Physically this has the impact of moving with the particle in world space.
-Commonly known as the Lagrangian form.
-It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed.
-The Eulerian description is static: Variables of particles moving by are measured while staying in a fixed position.
-
-These descriptions are different but they will yield the same measurements when related correctly.
-I.e.
-both configurations refer variables defined on them to the deformed state but the position, where the 'lookup' of that value happens, is different.
-In the reference confguration lookup is done at the initial position of the particle.
-In the current configuration lookup happens at the particle's world position.
-These relations for some particle quantity $f$ are called the (Lagrangian) pull back
-\begin{equation}
-  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
-\end{equation}
-and the (Eulerian) push forward
-\begin{equation}\label{eq:push_forward}
-  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
-\end{equation}
-with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.
-To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t.
-an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.
-
-The difficulty in the eulerian formulation becomes apparent when differentiating (due to the chain rule):
-\begin{equation} \label{eq:eulerian_general}
-  \frac {\partial}{\partial t} {_0f_i } ( {_0\boldsymbol { x }} ,  t  ) =
-  \frac { \partial  {_tf_i} } { \partial t } ( {^t_0}\phi ( {_0\boldsymbol {x }} , t ) , t ) + \frac { \partial  {_tf_i }} { \partial {_tx_j} } ( {^t_0}\phi ( {_0\boldsymbol { x }} , t ) , t ) \frac { \partial {^t_0}\phi_j } { \partial t } ( {_0\boldsymbol { x }} , t ).
-\end{equation}
-Combining this with equation \ref{eq:velocity} and applying the push forward \ref{eq:push_forward} to cancel out mappings leads to the definition referred to as the material derivative in the current configuration:
-\begin{equation} \label{eq:material_derivative}
-  \frac {  { D } } {   { D }   { t } }   {  {f} } ( _t\boldsymbol { x } ,   { t } ) = \frac { \partial   {  {f} } } { \partial   { t } } ( _t\boldsymbol {x } ,   { t } ) + \frac { \partial   {  {f} } } { \partial x _ {   { j } } } ( _t{\boldsymbol { x }} ,   { t } )  {v} _ {   { j } } ( _t\boldsymbol { x } ,   { t } ).
-\end{equation}
-The Jacobian of the deformation map $\phi$ is the deformation gradient $\boldsymbol{F}$ and is one of the key components to measure strain for material models:
-\begin{equation}
-  {^t_0 F_{ij}} ( _0\boldsymbol {x} ,  t  ) = \frac { \partial {^t_0\phi_i} } { \partial _0 x _j } ( _0\boldsymbol { x } , t ) = \frac { \partial _t x _i } { \partial _0 x_j } ( _0\boldsymbol { x } , t ).
-\end{equation}
-Topology specifies a neighborhood using the open ball concept $_{0,t}B_\epsilon(\boldsymbol{x}) = \{\boldsymbol{y} \in {\Omega^{0,t} | d(\boldsymbol{x},\boldsymbol{y}) < \epsilon}\}$ given a distance measure $d$.
-$_0B_\epsilon$ becomes the pre-image of $_tB_\epsilon$ under $_0^t\phi$.
-Intuitively the deformation gradient measures the local deformation of all particles in a small neighborhood $_0B_\epsilon$ to $_tB_\epsilon$.
-This allows to describe infinitesimal changes in position from the reference to the current configuration
-\begin{equation}\label{eq:def_grad_pos}
-  d {_tx_i} = {^t_0}F{_{ij}} d{_0x_j}.
-\end{equation}
-With this quantity in place volume and area changes are calculable.
-In a typical analytical fashion a coordinate system change $_0x \rightarrow {_tx}$ for a quantity $_0g$ is done using the determinant of the Jacobian matrix.
-The determinant is given a separate name $^t_0J =\text{det}(^t_0\boldsymbol{F})$.
-The push forward of $_0g:\Omega^0\rightarrow \mathbb{R}^d$ thus becomes $_tg:\Omega^t\rightarrow \mathbb{R}^d$:
-\begin{equation} \label{eq:volume_integral}
-  \int _ { {_t\boldsymbol { B } }}  {_tg} ( {_t\boldsymbol { x }} ,t)  d {_t\boldsymbol { x }} = \int _ { { _0\boldsymbol { B }}} {_0 g } ( {_0\boldsymbol{x}},t) {^t_0 J}d {_0\boldsymbol{x}}.
-\end{equation}
-This can also be achieved by the cross product.
-A cube spanned by vectors $_0\boldsymbol{x}_i$ ($i=1,2,3$) becomes a parallelepiped in the deformed configuration $d_tV =|{^t_0\boldsymbol{F}}d_0\boldsymbol{x}_0 \cdot({^t_0\boldsymbol{F}}d_0\boldsymbol{x}_1 \times {^t_0\boldsymbol{F}}d_0\boldsymbol{x}_2)| {d_0V}$:
-\begin{equation}\label{eq:j}
-  d_tV = {^t_0J} d_0V.
-\end{equation}
-The area change is given by Nanson's Formula.
-$d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area and $d_t\boldsymbol{l} ={}^t_0\boldsymbol{F}d_0\boldsymbol{l}$ an arbitrary line element:
-$$ d{_0V} = d{_0\boldsymbol{A}}\cdot d{_0\boldsymbol{l}},\quad d{_tV} = d{_t\boldsymbol{A}} \cdot d{_t\boldsymbol{l}}$$
-$$ \xRightarrow{\ref{eq:def_grad_pos},\ref{eq:j}} {^t_0J} d_0\boldsymbol{A} \cdot d_0\boldsymbol{l} = d{_t\boldsymbol{A}} \cdot ({^t_0\boldsymbol{F}}d{_0\boldsymbol{l}})$$
-\begin{equation}
-  \Rightarrow d_t\boldsymbol{A} = {^t_0\boldsymbol{F}}^{-T} {^t_0}J d_0\boldsymbol{A} = {^0_t\boldsymbol{F}}^{T} {^t_0}J d_0\boldsymbol{A}.
-\end{equation}
-A surface integral may then be transformed to reference configuration by
-\begin{equation}\label{eq:surface_integral}
-  \int _ { \partial {_0B}  } \boldsymbol{h} ( {_t\boldsymbol{x}} , t ) \cdot d \boldsymbol{A} ( {_t\boldsymbol{x}} ) =  \int _ { \partial {_tB} } \boldsymbol{h} ( {_0\boldsymbol{x}} , t ) \cdot \boldsymbol{F}^{-T}J d \boldsymbol{A} ( {_0\boldsymbol{x}} )
-\end{equation}
-where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$.
-$d_0\boldsymbol{A}$, $d_t\boldsymbol{A}$ point in the direction of the surface normal of $\partial {_0B}(_0\boldsymbol{x})$, $\partial {_tB}(_t\boldsymbol{x})$, respectively.
-\cite{MIT:CONTINUUM_MECHANICS}
-\cite{MPM:COURSE}
-\subsection{Polar and singular value decomposition}\label{sec:svd}
-The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$.
-In equation \ref{eq:rigid} a rigid body movement was introduced.
-Let $b_i(t) = v_it$ be more generally some translation.
-A problem arises when calculating the deformation gradient of this equation.
-\begin{equation}
-  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t).
-\end{equation}
-As can be seen, the deformation gradient contains a rigid rotation.
-For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation (more in chapter \ref{sec:linear_elasticty}).
-I.e.
-the deformation gradient has two components a constant rotation and the actual distortion or strain.
-There are two ways to deal with this:
-\begin{enumerate}
-  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-Lagrangian strain tensor with quadratic components:
-    \begin{equation}\label{eq:green}
-      E _ { i j } = \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right).
-    \end{equation}
-  \item \label{it:polar} Polar decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric positive definite) distortional $\boldsymbol{S}$ parts $\boldsymbol{F}=\boldsymbol{R}\boldsymbol{S}$.
-\end{enumerate}
-That equation \ref{eq:green} cancels out the rotational part can be shown by item \ref{it:polar}:
-$$ \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( S _ {m i}R_{k m} R _ { k n} S_{n j} - \delta _ { i j } \right)$$
-$$= \frac { 1 } { 2 } \left( S _ {m i} \delta_{mn} S_{n j} - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( S _ {i m} S_{m j} - \delta _ { i j } \right)$$
-$$= \frac{1}{2} \left(\boldsymbol{S}^2 - I \right).$$
-Regarding item \ref{it:polar}: Assuming a singular value decomposition of
-\begin{equation}\label{eq:svd}
-  \boldsymbol{F} = \boldsymbol{U\Sigma V}^T
-\end{equation}
-is already computed where $\boldsymbol{U}$,$\boldsymbol{V}$ are orthogonal matrices and $\boldsymbol{\Sigma}$ is a diagonal matrix containing the singular values $\sigma_1 \geq \sigma_2 \geq ...
-\geq \sigma_r$ of $\boldsymbol{F}$.
-$(r-n)$ forms the dimension of the null space.
-The polar decomposition is computable as:
-\begin{equation}
-  \boldsymbol{R} = \boldsymbol{V} \boldsymbol{W} ^ { T }, \quad \boldsymbol{S} = \boldsymbol{W} \boldsymbol{\Sigma} \boldsymbol{W} ^ { T }.
-\end{equation}
-Since singular values are positive, it is straightforward to see that the properties for $\boldsymbol{R}$ and $\boldsymbol{S}$ hold.
-The components of the singular value decomposition are important to gain an intuition for its usefulness: The columns of $\boldsymbol{U}$ and $\boldsymbol{V}$ span bases for the row and column spaces of $\boldsymbol{F}$ using the left and right singular vectors $\boldsymbol{u}_i,\boldsymbol{v}_i$, respectively \cite{MIT:SVD}.
-For illustrating purposes imagine the manipulation of $\boldsymbol{v}_1$ due to $\boldsymbol{U\Sigma V}^T$:
-\begin{enumerate}
-  \item Transform from the right singular vector space to standard basis space: $\boldsymbol{V}^T\boldsymbol{v}_1$ $ = \boldsymbol{e}_1$.
-  \item Scale by singular values to transform to principal stretch space: $\boldsymbol{\Sigma} \boldsymbol{e}_1 = \sigma_1 \boldsymbol{e}_1$.
-  \item \label{it:tr_sing_vect} Transform to left singular vector space: $\boldsymbol{U}\sigma_1\boldsymbol{e}_1 = \sigma_1 \boldsymbol{u}_{1}$.
-\end{enumerate}
-For a positive definite matrix the singular value decomposition becomes even easier as $\boldsymbol{U}=\boldsymbol{V}$. Item \ref{it:tr_sing_vect} then effectively just becomes a transform 'back'.
-
-In the following a summary of the $3 \times 3$ singular value decomposition as in \cite{SVD:3x3} is given.
-The proposed singular value decomposition is also called the 'Polar SVD' and follows a specific convention.
-\begin{enumerate}\label{ref:itemize_conv}
-  \item\label{item:refl_free} $\boldsymbol{U}$,$\boldsymbol{V}$ are reflection-free corresponding to true rotation matrices, i.e.
-both $\text{det}(\boldsymbol{U}), \text{det}(\boldsymbol{V}) = 1$ hold.
-  \item If $\text{det}(\boldsymbol{F}) = -1$ the negative sign needs to move on to $\boldsymbol{\Sigma}$ as a result of item \ref{item:refl_free}.
-The lowest singular value in magnitude will get a negative sign attached.
-  \end{enumerate}
-This convention does not change the existence or uniqueness of the singular value or polar decomposition although strictly speaking $\boldsymbol{S}$ is not positive definite anymore.
-\\
-The algorithm proceeds as follows:
-\begin{enumerate}
-  \item \label{it:eig} \textbf{Symmetric eigenanalysis:} A Jacobi eigenvalue algorithm begins with the symmetric positive definite matrix $\boldsymbol { S } ^{(0)}= \boldsymbol { A } ^ { T } \boldsymbol { A } = \boldsymbol { V } \boldsymbol { \Sigma } ^ { 2 } \boldsymbol { V } ^ { T }$.
-    \begin{enumerate}[label*=\arabic*.]
-      \item \label{item:sk} Iteratively compute (an also symmetric, positive definite) $\boldsymbol { S } ^ { ( k + 1 ) } = [ \boldsymbol { Q } ^ { ( k )}] ^ { T } \boldsymbol { S } ^ { ( k ) } \boldsymbol { Q } ^ { ( k ) }$ where $\boldsymbol{Q}$ is a Givens-Rotation aiming to eliminate off $S_{12}$.
-Store $\boldsymbol{V}^{(k+1)} = \boldsymbol{V}^{(k)}  \boldsymbol{Q}^{(k)}$.
-      \item \label{item:sii} Do \ref{item:sk} again for the other off-diagonal entries $S_{13},S_{23}$.
-      \item \label{item:redo_sii} Redo \ref{item:sk} - \ref{item:sii} a fixed amount of steps $m$.
-    \end{enumerate}
-  \item \textbf{Sorting singular values:} Compute $\boldsymbol{B} := \boldsymbol{AV}$, where $\boldsymbol{V} = \boldsymbol{V}^{(3m)}$.
-Acquire $\boldsymbol{\Sigma}$ by $\|\boldsymbol{b}_i \|_2 = \|\boldsymbol{u}_i\sigma_i\|_2 = |\sigma_i|$, where also $\boldsymbol{B=U\Sigma}$ holds.
-Permute the singular values by sorting them in decreasing order.
-Apply the same permutation to the columns of $\boldsymbol{B}$ and $\boldsymbol{V}$, where switches in $\boldsymbol{V}$ also cause a sign change.
-Enforce the convention mentioned above.
-  \item \label{item:qr} \textbf{QR-factorization:} Compute $\boldsymbol{U}$ using a $\boldsymbol{QR}$-factorization with Givens-Rotations where $\boldsymbol{B}=\boldsymbol{QR}=\boldsymbol{U\Sigma}$.
-The $\boldsymbol{QR}$-factorization is done once in the same fashion as in item \ref{item:sk} - \ref{item:sii}.
-\end{enumerate}
-Due to inherent normalization, fast multiplication and storage efficiency quaternions are preferred over actual rotation matrices.
-In item \ref{item:qr} a $\boldsymbol{QR}$-factorization is preferred over a column normalization of $\boldsymbol{\Sigma U}$ due to its inaccuracy at near-zero singular values.
-In general $\boldsymbol{R}$ is an upper triangular matrix.
-In item \ref{item:qr} it reduces to being diagonal.
-\cite{MPM:COURSE}\cite{SVD:3x3}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%                         Constitutive Models                         %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Constitutive models}\label{seq:constitutive_models}
-It is common practice in elastoplascity theory to decompose the deformation gradient in its elastic and plastic parts $\boldsymbol{F} = \boldsymbol{F}_E \boldsymbol{F}_P$.
-The elastic part is recoverable.
-The plastic part is irreversibly lost.
-Plastic models are not covered here.
-They extend the energy densities discussed here with additional terms $\Psi(\boldsymbol{F}_E,\boldsymbol{F}_P)$.
-For more information refer to the literature \cite{ochsner2014elasto}.
-\subsubsection{Linear elasticity}\label{sec:linear_elasticty}
-The first aim will be to find the strain energy density $\Psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
-The most general linear stress-strain relationship is given by Hooke's Law in three dimensions:
-\begin{equation}\label{eq:hook}
-  \sigma_{ij} = C_{ijkl} \epsilon_{kl},
-\end{equation}
-where $\boldsymbol{\sigma}$ and $\bm{\epsilon}$ are second-order tensors with (3x3) = 9 elements. $C_{ijkl}$ is a fourth-order stiffness-tensor with (3x3)x(3x3) = 81 elements.
-Assuming following symmetries reduces the tensors to 6 and 21 unique elements respectively:
-\begin{enumerate}
-  \item Conservation of angular momentum: $\sigma_{ij} = \sigma_{ji} \Rightarrow C_{ijkl} = C_{jikl}$.
-  \item No-net-rotation: $\epsilon_{kl} = \epsilon_{lk} \Rightarrow C_{ijkl} = C_{ijlk}$.
-  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\Psi$:
-    \begin{equation}
-      C _ { i j k l } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },
-    \end{equation}
-\end{enumerate}
-where item \ref{it:mixed_partials} holds for the strain energy density functional of an (hyper-)elastic material.
-This potential $\Psi$ is thus similar to potentials for example involved in gravitation, electrodynamics or fluids.
-The stress may then also be calculated by
-\begin{equation}\label{eq:partial_energy}
-  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \Psi } } { \partial \epsilon _ { i j } },
-\end{equation}
-if such a $\Psi$ is given.
-An isotropic (direction-independent) linear elastic material further only has three unique elements $C _ {i j k l}$.
-Using Voigt-Notation, which collapses indices $i=j$ and $k=l$, equation \ref{eq:hook} can be rewritten as:
-\begin{equation} \label{eq:voigt}
-  \left[ \begin{array} {c}
-      {\sigma_{11}}\\
-      {\sigma_{22}}\\
-      {\sigma_{33}}\\
-      {\sigma_{23}}\\
-      {\sigma_{13}}\\
-      {\sigma_{12}}\\
-  \end{array} \right]
-  = \left[ \begin{array} { c c c c c c }
-      { C _ { 11 }   } & { C _ { 12 } } & { C _ { 12 } } & 0                & 0                & 0 \\
-      {}               & { C _ { 11 } } & { C _ { 12 } } & 0                & 0                & 0 \\
-      {}               & {}             & { C _ { 11 } } & 0                & 0                & 0 \\
-      {}               & {}             & {}             & { C _ { 22 } }   & 0                & 0 \\
-      {}               & {}             & {}             & {}               & { C _ { 22 } }   & 0 \\
-      {sym}            & {}             & {}             & {}               & {}               & { C _ { 22 } } \\
-  \end{array} \right]
-  \left[ \begin{array} {c}
-      {\epsilon_{11}}\\
-      {\epsilon_{22}}\\
-      {\epsilon_{33}}\\
-      {\gamma_{23}}\\
-      {\gamma_{13}}\\
-      {\gamma_{12}}\\
-  \end{array} \right].
-\end{equation}
-Experimental results of Hooke's law commonly give
-\begin{equation}
-  \boldsymbol{\epsilon} =
-  \epsilon _ { i j } = \frac { 1 } { E } \left[ ( 1 + \nu ) \sigma _ { i j } - \nu \sigma _ { k k } \delta _ { i j } \right]
-  = \boldsymbol{C}^{-1} \bm{\sigma}
-\end{equation}
-using engineering constants Young's modulus $E$ and Poisson ratio $\nu$. Inverting $\boldsymbol{C}^{-1}$ and switching to Lamé parameters $\lambda$ and $\mu$ results in the equation:
-\begin{equation} \label{eq:stress_strain}
-  \sigma_{ij} = 2\mu \epsilon_{ij} + \lambda \text{tr}(\boldsymbol{\epsilon}) \delta_{ij}.
-\end{equation}
-Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$, and $C_{22} = \mu$.
-$\boldsymbol{\gamma}$ is also referred to as the engineering strain.
-Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\Psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:
-\begin{equation}\label{eq:linear_elas}
-  \Psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).
-\end{equation}
-\begin{flushright}\cite{MIT:LINEAR_ELASTICITY}\end{flushright}
-\subsubsection{Corotated hyperelasticity}\label{sec:cor_hyper}
-The simplest tensor assumed by infinitesimal strain theory is the small strain tensor: \begin{equation}
-  \boldsymbol{\epsilon}_l = \frac { 1 } { 2 } \left( \boldsymbol { F }_E + \boldsymbol { F }_E ^ { T } \right) - \boldsymbol { I }.
-\end{equation}
-The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant w.r.t.
-to $\boldsymbol{F}_E$: $\Psi_{LE}(\boldsymbol{\epsilon}_l(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \Psi_{LE}(\boldsymbol{\epsilon}_l(\boldsymbol{F}_E))$.
-Rigid body motions however don't result in strain and consequently don't need to be recovered from.
-So energy should not change.
-Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$ an alternate strain measure may be defined as:
-\begin{equation}
- \boldsymbol { \epsilon } \left( \boldsymbol { F }_E \right) = \boldsymbol { \epsilon }_l \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) = \frac { 1 } { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E + \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E \right) ^ { T } \right) - \boldsymbol { I } = \boldsymbol { S }_E - \boldsymbol { I }.
-\end{equation}
-Substituting $ {\boldsymbol{\epsilon}}$ into equation \ref{eq:linear_elas} leads to the energy definition of corotational hyperelasticity:
-\begin{equation}\label{eq:corot_S}
-  \Psi_{CH} = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } ).
-\end{equation}
-Using rotation-invariance of the Frobenius norm:
-\begin{equation}
-  \Psi_{CH} = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)
-\end{equation}
-Even more insight yields the relationship to their singular values $\sigma _ i$, also called the principal stretches:
-$$\operatorname {tr}\left(\boldsymbol{S}\right) = \sum _ { i = 1 } \sigma _ { i } = \operatorname{tr}\left(\boldsymbol{\Sigma}_E\right)$$
-$$\| \boldsymbol{S}_E \| _ { F }^2 = \sum _ { i = 1 } \sigma _ { i } ^ { 2 } = \| \boldsymbol{\Sigma}_E\|_{F}^2$$
-$$\Rightarrow \| \boldsymbol{S}_E\boldsymbol{-I} \| _ { F }^2 = \|\boldsymbol{S}_E \| _ { F }^2 - 2 \operatorname{tr}(\boldsymbol{S}_E)  + \|\boldsymbol{I}\|_{F}^2 = \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 }$$
-\begin{equation} \label{eq:corot_sing}
-  \Psi_{CH}(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )
-\end{equation}
-Equation \ref{eq:corot_sing} uses the diagonal matrix $\boldsymbol{\Sigma}_E$ containing the principal stretches typically acquired by the singular value decomposition $\boldsymbol{F}_E \boldsymbol{= U \Sigma}_E\boldsymbol{ V ^T }.$ Since the energy density in \ref{eq:corot_sing} is a function of only three (singular) values, which describe stretch/compression of the material, isotropy of the material is underlined.
-\begin{flushright}\cite{ADAMS:ELASTICITY}\end{flushright}
-\subsubsection{Fixed corotated hyperelasticity}
-Numerical implicit stepping algorithms mostly rely on using the first and second derivative of $\Psi$, chapter \ref{sec:newton}.
-Furthermore one may describe an isotropic elastic model due to their principal stretches $\sigma_1,\sigma_2,\sigma_3$, chapter \ref{sec:svd}.
-The rest configuration is reached, when $\sigma_1,\sigma_2,\sigma_3 = 1$.
-I.e.
-no elastic forces will be exerted because the material is completely relaxed (again).
-
-Under material relaxation these stepping algorithms are attracted to a material-dependent primary contour.
-For instance for a model that aims for high volume preservation (Poisson ratio $\nu =0.5$) the primary contour is primarily enforcing volume preservation ($ J_E = \sigma_1\sigma_2\sigma_3 \approx 1$).
-Reaching the rest configuration is only the secondary goal.
-
-The primary contour can be formalized as $\boldsymbol{v}\cdot \boldsymbol{g} = 0$ of the gradient $g_i = \frac{\partial \Psi}{\partial \sigma_i}$ and the eigenvector with the largest-magnitude eigenvalue $\boldsymbol{v}$ of the Hessian $H_{ij} = \frac{\partial \Psi}{\partial \sigma_i \sigma_j}$.
-Problems arise in $\Psi_{CH}$ as the primary contour can easily cross into the inverted region ($\sigma_3 < 0$) leading to inverted configurations while relaxing.
-This is undesired behavior.
-Furthermore, the primary contour can lead into energy kinks at extreme stretches leading to oscillations.
-
-Therefore, they propose to use the fixed corotated hyperelasticity energy density:
-\begin{equation}
-  \Psi_{FCH} = \mu \|\boldsymbol{\Sigma}_E - \boldsymbol{I}\|^2_F + \frac{\lambda}{2}(J_E-1)^2.
-\end{equation}
-This leads to the primary contour $J=1$ which does not cross the inverted region as only one singular value may be negative.
-\begin{flushright}\cite{MPM:INVERT}\end{flushright}
-\subsection{Governing equations}
-Starting from the governing equations this chapter will lead to the weak formulation core of any finite element and material point method.
-\subsubsection{Conservation of mass}
-Let the Eulerian mass density be $_t\rho(_t\boldsymbol{x},t)$. Similarly, let its (Lagrangian) pull back be $_0\rho(_0\boldsymbol{x},t)$.
-After \ref{eq:volume_integral} they are related as:
-\begin{equation} \label{eq:density_pull_back}
-  \int _ {_tB _ { \epsilon }}_t\rho(_t\boldsymbol{x},t)d_t\boldsymbol{x} =   \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}.
-\end{equation}
-An open ball $ _ { 0 }B _ { \epsilon } $ in the reference configuration will have the same mass as its respective open ball $B_{\epsilon}^t$ in the current configuration.
-Keep in mind that both refer to a deformed state.
-
-Conservation of mass dictates that mass does not depend on motion or time. Only the space occupied by this mass may be more or less.
-\begin{equation}
-  \frac{d}{dt} \int _ {_tB _ { \epsilon }} _t\rho(_t\boldsymbol{x},t) d_t\boldsymbol{x} = 0
-\end{equation}
-Equivalently this can be formulated with the constant undeformed initial mass in Lagrangian view:
-\begin{equation}\label{eq:lagr_mass}
-  \left(\int _ {_tB _ { \epsilon }} _t\rho(_t\boldsymbol{x},t) d_t\boldsymbol{x} \stackrel{\text{\ref{eq:density_pull_back}}}{=} \right)
-  \int _ {_0B _ { \epsilon }} {^t_0J}_0\rho(_0\boldsymbol{x},t) d_0\boldsymbol{x}=
-  \int _ {_0B _ { \epsilon }} _0\rho(_0\boldsymbol{x},0)d_0\boldsymbol{x}.
-\end{equation}
-In Eulerian view the conservation of mass is more difficult to develop and starts with the Lagrangian view. Since the integrals do account for arbitrary volumes, they are left out in the following:
-\begin{equation}\label{eq:euler_density_evol}
-  \frac{\partial}{\partial t}(_0\rho^t_0J) = \frac{\partial _0\rho}{\partial t}{^t_0J} + \frac{\partial{^t_0J}}{\partial t} {_0\rho} = 0.
-\end{equation}
-The left side could be immediately pushed forward, the right side is harder:
-\begin{equation}\label{eq:evol_jacobian}
-  \frac{\partial{J}}{\partial t}
-  = \frac{\partial{J}}{\partial {F_{ij}}} \frac{\partial {F_{ij}}}{\partial t}
-  \stackrel{\text{\ref{eq:det_deriv},\ref{eq:evol_def_grad}}}{=} {J}{F}^{-1}_{ji}\frac{\partial v_i}{\partial x_k}F_{kj}
-  = J \delta_{ik}\frac{\partial v_i}{\partial x_k} =  J \frac{\partial v_i}{\partial x_i}.
-\end{equation}
-The determinant differentiation rule can be shown by expressing the determinant with Laplace's expansion and applying the derivative on it:
-\begin{equation}\label{eq:det_deriv}
-  \frac{\partial J}{\partial F_{ij}} = \frac{\partial (F_{ik}\text{adj}(F)_{ki})}{\partial F_{ij}} =  \text{adj}(F)_{ji} = JF^{-1}_{ji}.
-\end{equation}
-The time-evolution of the deformation gradient is:
-\begin{equation}\label{eq:evol_def_grad}
-  \frac{\partial {{^t_0F_{ij}}}}{\partial t} =
-  \frac{\partial}{\partial t}\frac{\partial {^t_0\phi_i}}{\partial _0x_j}(_0\boldsymbol{x},t) = \frac{\partial {_0v_i}}{\partial _0x_j}(_0\boldsymbol{x},t) =
-  \frac{\partial {_tv_i}}{\partial _tx_k}(^t_0\phi(_0\boldsymbol{x},t),t)\frac{\partial {_tx_k}}{\partial {_0x_j}}(_0\boldsymbol{x},t).
-\end{equation}
-Pushing forward \ref{eq:euler_density_evol} with the result of \ref{eq:evol_jacobian} using the material derivative formulation \ref{eq:material_derivative} leads to the Eulerian conservation of mass:
-\begin{equation}
-  \frac{D}{Dt}\rho(_t\boldsymbol{x},t) + \rho(_t\boldsymbol{x},t) \vec{\nabla} \cdot \boldsymbol{v}(_t\boldsymbol{x},t)=0.
-\end{equation}
-Commonly used is the Nabla operator: $\vec { \nabla } = \left( \frac { \partial } { \partial x _ { 1 } } , \dots , \frac { \partial } { \partial x _ { n } } \right)$.
-\cite{MPM:COURSE}
-\subsubsection{Conservation of momentum}
-Continuum forces are divided up into body and surface forces.
-A surface force acts upon the surface of the material $\partial _tB_\epsilon$.
-While a body force scales upon the volume of the material $_tB_\epsilon$.
-Conservation of momentum may then be expressed in a similar way to the conservation of mass as:
-\begin{equation}
-  \frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
-\end{equation}
-Already assumed is angular momentum conservation $\boldsymbol{\sigma}^T  = \boldsymbol{\sigma} $ out of the second part of this section. Beginning with a mix out of Lagrangian and Eulerian view
-$$\left(\frac { d } { d t } \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{v} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} \stackrel{\text{\ref{eq:eulerian_general}}}{=} \right)
-\frac { d } { d t } \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{v} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}$$
-\begin{equation}
-  \stackrel{\text{\ref{eq:euler_density_evol}}}{=}
-  \int _ {_0B _ { \epsilon}  } {^t_0J} \rho ( _0\boldsymbol{x} , t ) \boldsymbol{a} ( _0\boldsymbol{x} , t ) d _0\boldsymbol{x}
-  = \int _ { \partial _tB _ {\epsilon  } } \boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}
-\end{equation}
-where conservation of mass (equation \ref{eq:euler_density_evol}) was applied.
-The Eulerian push-forward of the left side combined with the divergence theorem becomes:
-\begin{equation}\label{eq:eul_transport}
-  \int _ {_tB _ { \epsilon}  } \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} =
-  \int _ { \partial _tB _ {\epsilon } } \vec{\nabla} \cdot \boldsymbol{\sigma} d_t\boldsymbol{x}  + \int _ { _tB_\epsilon } \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
-\end{equation}
-The acceleration $_t\boldsymbol{a}$ is again defined due to the material derivative \ref{eq:material_derivative}. Thus the Eulerian momentum balance equation becomes:
-\begin{equation}
-  _t\rho \frac{D_t\boldsymbol{v}}{Dt} = \vec{\nabla} \cdot \boldsymbol{\sigma} +\boldsymbol{f} ^ {\text{body} }.
-\end{equation}
-There is a quantity left to be defined for the Lagrangian view.
-The Cauchy stress $\boldsymbol{\sigma}$ is defined in the current configuration.
-Pulling back the Cauchy stress leads to a stress measure named the first Piola Kirchhoff stress
-\begin{equation}
-  \int _ { \partial _tB _ {\epsilon  } } _t\boldsymbol{\sigma} d_t\boldsymbol{A} (_t\boldsymbol{x} )
-  \stackrel{\text{\ref{eq:surface_integral}}}{=}
-  \int _ { \partial _0B _ {\epsilon  } }  _0\boldsymbol{\sigma}\boldsymbol{F}^{-T} Jd_0\boldsymbol{A} (_0\boldsymbol{x} )=
-  \int _ { \partial _0B _ {\epsilon  } }  \boldsymbol{P}d_0\boldsymbol{A} (_0\boldsymbol{x} ),
-\end{equation}
-denoted in the literature as:
-\begin{equation}\label{eq:piola}
-  \boldsymbol{P} =
-  \frac{\partial \Psi}{\partial \boldsymbol{F}}=
-  \boldsymbol{\sigma}\boldsymbol{F}^{-T}J.
-\end{equation}
-Summarized, the Lagrangian view of the momentum equation with an initial momentum is ($^0_0J=1$):
-\begin{equation}\label{eq:lagr_mom}
-  _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
-= \vec{\nabla} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
-\end{equation}
-For the stress strain relationship $\boldsymbol{C}$ in chapter \ref{sec:linear_elasticty} as well as the start of this chapter conservation of angular momentum caused $\sigma_{ij} = \sigma_{ji}$.
-This can be shown as follows.
-The description of angular momentum follows that of linear momentum ($ _t\boldsymbol{x}\times \ref{eq:eul_transport}$) where product rule generally applies $\left(\frac{D_t\boldsymbol{x}}{Dt} \times {_t\boldsymbol{v}} = {_t\boldsymbol{v}} \times {_t\boldsymbol{v}} = 0\right)$ :
-\begin{equation}
-\int _ {_tB _ { \epsilon}  } _t\boldsymbol{x} \times \rho ( _t\boldsymbol{x} , t ) \boldsymbol{a} ( _t\boldsymbol{x} , t ) d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } }_t\boldsymbol{x} \times  \boldsymbol{\sigma}^T d_t\boldsymbol{A} (_t\boldsymbol{x} ) + \int _ { _tB_\epsilon }_t\boldsymbol{x} \times  \boldsymbol{f} ^ {\text{body} } d _t\boldsymbol{x}.
-\end{equation}
-In component form this becomes:
-\begin{equation}\label{eq:gen_ang_mom}
-  \int _ {_tB _ { \epsilon}  } \varepsilon_{ijk} \rho x_j a_k d _t\boldsymbol{x} = \int _ { \partial _tB _ {\epsilon  } } \varepsilon_{ijk} x_j \sigma_{mk} dA_m(_t\boldsymbol{x} ) + \int _ { _tB_\epsilon } \varepsilon_{ijk} x_j f_k ^ {\text{body} } d _t\boldsymbol{x}.
-\end{equation}
-The divergence theorem is again applied to the surface forces $\left(\vec{\nabla} = \frac{\partial}{\partial x_m}\right)$:
-$$
-\int _ { \partial _tB _ {\epsilon  } } \varepsilon_{ijk} x_j \sigma_{mk} dA_m (_t\boldsymbol{x} ) =
-\int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \frac{\partial(x_j \sigma_{mk})}{\partial x_m} d_t\boldsymbol{x}
-$$
-\begin{equation}\label{eq:surface_ang_mom}
-  = \int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \left(\delta_{jm} \sigma_{mk}+ x_j\frac{ \partial \sigma_{mk}}{\partial x_m}\right) d_t\boldsymbol{x}.
-\end{equation}
-The conservation of momentum (equation \ref{eq:eul_transport}) can then be applied to the result of plugging \ref{eq:surface_ang_mom} back into \ref{eq:gen_ang_mom} leaving only:
-\begin{equation}
-  \int _ {_tB _ {\epsilon  } } \varepsilon_{ijk} \sigma_{jk} d_t\boldsymbol{x} = 0.
-\end{equation}
-Leaving out the integral and multiplying by ${\varepsilon_{irs}}$ enables equation \ref{eq:compact_levi}. The Cauchy stress is constrained to
-\begin{equation}
-  \sigma_{ij} =\sigma_{ji}
-\end{equation}
-as assumed before. Note that $\boldsymbol{P}$ however is not constrained to be symmetric.
-\begin{flushright}\cite{MPM:COURSE}\cite{MIT:CONTINUUM_MECHANICS}\end{flushright}
-\subsubsection{Weak formulation}\label{sec:weak}
-Before deriving the weak form an explanation of what it achieves is in order.
-The previous presented governing equations are written in the strong form: a solution to the equation needs to be exact on the whole domain and is as such influenced by the whole domain in general.
-Such a solution can be found for simplified models analytically.
-These act as a ground truth to numerical methods like the material point method.
-Analytical methods of today can't handle complex problems: Numerical solutions try to overcome that hurdle.
-
-The complete mathematical description of the weak formulation is beyond this thesis.
-Numeric books that discuss finite element methods will provide one (\cite{bathe2006finite}, \cite{dahmen2008numerik}).
-For simplicity: The weak formulation combined with Galerkin discretization restricts the globality of the strong method due to so called 'test functions': $_0\boldsymbol{q}_h$ .
-A 'test function' generally only has limited support, i.e.
-$ _0\boldsymbol{q}_h\neq 0$ on a very small subset of the whole domain $\Omega^0$.
-This is mostly used to gather information on a local neighborhood and not on the whole domain.
-
-The weak formulation requires an observation of the dot product. Consider the conservation of momentum in the strong form as formulated before \ref{eq:lagr_mom} as:
-\begin{equation}\label{eq:strong}
-  _0\rho(_0\boldsymbol{x},0)_0\boldsymbol{a}(_0\boldsymbol{x},t)
-  = \vec{\nabla} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
-\end{equation}
-Multiply both sides with the dot product of an arbitrary function $_0\boldsymbol{q}(\cdot,t):\Omega^0 \rightarrow \mathbb{R}^d$ and integrate over $\Omega^0$.
-If a solution solves the balance of \ref{eq:strong} then it also solves:
-$$
-\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \left(_0\rho (_0\boldsymbol{x} , 0 ) _0a _{ i } (_0\boldsymbol{x} , t ) - {_0\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}\right) d _0\boldsymbol{x}
-$$
-\begin{equation}
-  =\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \frac{\partial P _ {ij}}{\partial x_j} ( _0\boldsymbol{x} , t ) d_0\boldsymbol{x}.
-\end{equation}
-With the help of partial integration in multiple dimension, the derivative moves over to the test-function.
-Balancing out the order of derivatives, by moving a derivative to the test function, is the main motive of the weak form:
-$$
-\int _ { \Omega ^ { 0 } } \frac{\partial (_0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t ))}{\partial _0x_j}
-- \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
-d_0\boldsymbol{x}.
-$$
-The divergence theorem allows to convert the first term to a boundary integral.
-\begin{equation}
-  \int _ { \partial \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t)
-  d_0\boldsymbol{A}(_0\boldsymbol{x})
-  -\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
-  d_0\boldsymbol{x}.
-\end{equation}
-Mathematically the boundary integral serves as a boundary condition which is set by the specific problem (e.g.
-context of the simulation).
-Finally putting together the previous results gives the weak form of force balance in the Lagrangian view:
-$$
-\int _ { \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) \left(_0\rho (_0\boldsymbol{x} , 0 ) _0a _{ i } (_0\boldsymbol{x} , t ) - {_0f_i} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}\right) d _0\boldsymbol{x}
-$$
-\begin{equation} \label{eq:lagr_force_bal}
-  = \int _ { \partial \Omega ^ { 0 } } _0q _ { i } ( _0\boldsymbol { x } , t ) P _ {ij} ( _0\boldsymbol{x} , t)
-  d_0\boldsymbol{A}(_0\boldsymbol{x})
-  -\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij}( _0\boldsymbol{x} , t )
-  d_0\boldsymbol{x}.
-\end{equation}
-In the material point method stress computations are more naturally done in the current configuration or equally in terms of the Cauchy stress as seen in chapter \ref{sec:linear_elasticty}.
-Pushing the equation forward to Eulerian view with the push forward $_t\boldsymbol{q}:\Omega^t \rightarrow \mathbb{R}^d$ of $_0\boldsymbol{q}$ is only a problem for the last term:
-$$
-\int _ { \Omega ^ { 0 } }  \frac{\partial _0q _ { i }}{\partial _0x_j} ( _0\boldsymbol { x } , t ) P _ {ij} d_0\boldsymbol{x}
-\stackrel{\text{\ref{eq:piola}, \ref{eq:volume_integral}}}{=}
-\int _ { \Omega ^ { 0 } }  \left({^t_0}F_{kj}\frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\right) \left({_t^0}F_{kj} \sigma _ {ik} {_0^tJ}\right) {_t^0J} d_t\boldsymbol{x}
-$$
-\begin{equation}  =  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
-\end{equation}
-This completes the Eulerian view to be:
-$$
-\int _ { \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \left(_t\rho (_t\boldsymbol{x} , t) _ta _{ i } (_t\boldsymbol{x} , t ) - {_tf_i} ^ {\text{body}}(_t\boldsymbol{x},t)\right) d _t\boldsymbol{x}
-$$
-\begin{equation}\label{eq:eul_force_bal}
-  = \int _ { \partial \Omega ^ { t } } _tq _ { i } ( _t\boldsymbol { x } , t ) \sigma _ {ij} d_tA_j(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t } } \frac{\partial _tq _ { i }}{\partial _tx_k} ( _t\boldsymbol { x } , t )\sigma _ {ik} d_t\boldsymbol{x}.
-\end{equation}
-\begin{flushright}\cite{MPM:COURSE}\cite{strang2007computational}\cite{bathe2006finite}\end{flushright}
-\subsection{Material point method}\label{sec:mat_point}
-The key idea of the material point method is to use (Lagrangian) particles as a consistent storage of material properties.
-All stress based forces are computed on a Eulerian grid however.
-This grid does not store any material properties and is therefore often referred to as a scratch pad.
-
-As a corollary there needs to be a way to transfer information from a particle to the neighboring grid cells.
-This also induces switching from Lagrangian to Eulerian view.
-After stresses are computed there also needs to be a way to get back the relevant information from the grid to the particles.
-Advection is typically hard to do in Eulerian/FE-like methods and cause of a lot of problems down the development pipeline.
-In a Lagrangian view particle advection is trivial.
-
-It is very important that the two transfers as well as the grid solver are in compliance with all governing equations.
-While the grid solver will be derived from the weak form of the governing equation the transfers also need to be chosen in a way that conserve the properties defined in them.
-\cite{MPM:COURSE}
-\subsubsection{Interpolation weights}
-The choice of interpolation weights is flexible. Nevertheless a kernel $w(\boldsymbol{x})$ requires some important properties to be qualified for MPM:
-\begin{enumerate}
-  \item Partition of unity:
-    \begin{equation}\label{eq:partition_unity}
-      \sum _ { i } w \left(  { \boldsymbol{x} } -  { \boldsymbol{x} } _ { i } ^ { n } \right) & = 1.
-    \end{equation}
-  \item Identity relation:
-    \begin{equation}
-      \sum _ { i }  { \boldsymbol{x} } _ { i } ^ { n } w \left(  { \boldsymbol{x} } -  {\boldsymbol{ x} } _ { i } ^ { n } \right) & =  { \boldsymbol{x} }.
-    \end{equation}
-  \item Non-negativity: $w \geq 0$. Negative weights can cause severe instability or non-physical behavior, unlike in FEM where they are used. \cite{gao2017adaptive}
-  \item Limited local support to reduce the number of discretizations, chapter \ref{sec:discretization}.
-  \item $C^1$-continuity s.t. $\nabla w$ is continuous.
-\end{enumerate}
-As a reminder of chapter \ref{sec:notation}: $\boldsymbol{x}_p$ refers to a particle's position. $\boldsymbol{x}_i$ to a grid cell's position.
-For a more general discussion refer to \cite{gao2017adaptive}. Often dyadic products of one-dimensional interpolation functions suffice
-\begin{equation}
-  w(\boldsymbol{x}-\boldsymbol{x}_i^n) = w_i^n(\boldsymbol{x}) = w\left(\frac{1}{h}\left(x-x_i^n\right)\right)w\left(\frac{1}{h}\left(y-y_i^n\right)\right)w\left(\frac{1}{h}\left(z-z_i^n\right)\right)
-\end{equation}
-\begin{equation}
-  \nabla w_{i}^n(\boldsymbol{x})
-  =\frac { 1 } {  { h } }
-  \left(
-    \begin{array} {l}
-      w^{\prime} (\frac { 1 } {  { h } } ( x  - x _ { i }^n ))  { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
-      w(\frac{1} {  { h } } ( x  - x _ { i }^n ))               { w }^{\prime} ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }         ( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
-      w(\frac{1} {  { h } } ( x  - x _ { i }^n ))               { w }          ( \frac { 1 } {  { h } } (  { y }  -  { y } _ { i}^n  ) )   { w }^{\prime}( \frac { 1 } {  { h } } ( z  - z _ { i }^n ) \\
-    \end{array}
-  \right)
-\end{equation}
-where $h$ is the grid spacing (of a uniform grid). An interpolation function often employed is a cubic B-spline ($C^2$):
-\begin{equation}\label{eq:cubic_weights}
-  w(x) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array} \right.
-.
-\end{equation}
-Since these function are used to weight (or filter) particles a shortening in notation may be employed as $ w_{ip}^n = w(\boldsymbol{x}_p^n-\boldsymbol{x}_i^n)$.
-\cite{MPM:APIC}\cite{MPM:COURSE}\cite{steffen2008analysis}\cite{gao2017adaptive}
-\subsubsection{Mass transfer}\label{sec:mass}
-Each material point will be assigned an initial volume $_0V_p$ as well as an initial mass $_0m_p$.
-The volume the material point occupies may change in time due to equation \ref{eq:j}: ${^t_0}J _0V_p = {_tV_p}$.
-But, due to conservation of mass \ref{eq:lagr_mass} it will have a constant, initial mass associated with it.
-I.e.
-there will only be a transfer to the grid and no transfer back.
-A transfer of mass to the grid may then be expressed as:
-\begin{equation}\label{eq:mpm:mass}
-  m_i = \sum_p w_{ip}m_p.
-\end{equation}
-$\sum_i m_i =\sum_p m_p$ is a complete prove this fulfills the conservation of mass.
-Remember there is no transfer back.
-In doing so no information can be lost on the particles.
-Proving that the mass transfer to the grid is conserving is enough.
-The stress based-solver may manipulate this information further.
-Mass-lumping strategies typically will use this information directly, equation \ref{eq:mass_discr}.
-\begin{equation}
-  \sum_i m_i
-  \stackrel{\text{\ref{eq:mpm:mass}}}{=}
-  \sum _i \sum_p w_{ip} m_p =
-  \sum _p m_p \sum_i w_{ip}
-  \stackrel{\text{\ref{eq:partition_unity}}}{=}
-  \sum_p m_p
-\end{equation}
-\begin{flushright}\cite{MPM:APIC}\cite{MPM:COURSE}\end{flushright}
-\subsubsection{APIC-transfers}\label{sec:apic}
-The momentum transfer round trip could be defined as:
-\begin{enumerate}
-  \item Particle to grid momentum transfer:
-    \begin{equation}\label{eq:lin_p2g}
-      (m\boldsymbol{v})_i^n = \sum_p w_{ip}m_p\boldsymbol{v}^n_p.
-    \end{equation}
-  \item Factoring out mass:
-    \begin{equation}
-      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
-    \end{equation}
-  \item Coupled with either ($\alpha \in \{0,1\}$) or a combination ($\alpha \in ]0;1[$) of:
-    \begin{equation}
-      \boldsymbol{v}_p^{n+1} = \alpha\boldsymbol{v}_{p,{PIC}}^{n+1} + (1-\alpha)\boldsymbol{v}_{p,{FLIP}}^{n+1}
-    \end{equation}
-    \begin{equation}\label{eq:PIC}
-      \boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}.
-    \end{equation}
-    \begin{equation}\label{eq:FLIP}
-      \boldsymbol{v}_{p,{FLIP}}^{n+1} = \boldsymbol{v}_p^{n} + \sum_i w_{ip}(\boldsymbol{v}_i^{n+1}-\boldsymbol{v}_i^{n+1}).
-    \end{equation}
-\end{enumerate}
-While $PIC$-transfers are very stable, they suffer from excessive (energy) dissipation due to double interpolating on the whole quantity \ref{eq:lin_p2g},\ref{eq:PIC}.
-This causes a heavy loss in angular momentum and velocity modes.
-$FLIP$-Transfers avoid dissipation and loss of angular momentum by only updating the velocity with a difference \ref{eq:FLIP}.
-However, some velocity modes are also not recognized on the grid and may cause unpredictable and unstable behavior in following steps.
-Therefore, often a combination of both is taken.
-
-$APIC$ builds on top of the very stable $PIC$-transfers and effectively only adds an extra term of the Taylor series to increase accuracy.
-This extra term $\boldsymbol{C}_p$ may be in short just referred to as the velocity derivative.
-The local velocity field around a particle may then be characterized by the affine function $\boldsymbol{v}(\boldsymbol{x}) = \boldsymbol{C}_p(\boldsymbol{x}-\boldsymbol{x}_p)$.
-
-Motivated by the theory of angular momentum and moment of inertia one can define a quantity
-\begin{equation}
-\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T }
-\end{equation}
-which is similar to the classically known inertia tensor:
-\begin{equation}
-  \boldsymbol{I} _ { p } = - \sum _ { i } m _ { i } [\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}][\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}]^T
-\end{equation}
-$$
-=
-\sum _ { i } m _ { i }((\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})\boldsymbol{I} -(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T)
-$$
-Bearing in mind, that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
-In classical mechanics the angular velocity $\boldsymbol{\omega_p}$ can be then described using the inertia tensor $\boldsymbol{I}_p$ with the help of the angular momentum $\boldsymbol{L}_p$:
-\begin{equation}
-  \boldsymbol{\omega}_p = \boldsymbol{I}_p^{-1} \boldsymbol{L}_p.
-\end{equation}
-This motivates the velocity derivative to be equally defined by a similar relationship where $\boldsymbol{B}^n_p$ holds momentum information.
-\begin{equation}
-  \boldsymbol{C}^n_p = (\boldsymbol{D}_p^n)^{-1} \boldsymbol{B}^n_p.
-\end{equation}
-The transfers of the $APIC$ scheme are then summarized:
-\begin{enumerate}
-  \item Particle to grid:
-    \begin{equation}\label{eq:apic_mom_p2g}
-    (m\boldsymbol{v})^n_i = \sum_p w^n_{ip}m_p (\boldsymbol { v } _ { p } ^ { n } + \boldsymbol { B } _ { p } ^ { n } ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) )
-\end{equation}
-  \item Factoring out mass:
-    \begin{equation}\label{eq:mass_divide}
-      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
-    \end{equation}
-  \item Grid to particle transfer (in a $PIC$ manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
-    \begin{equation}
-      \boldsymbol{x}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{x}_i^{n+1}
-    \end{equation}
-    \begin{equation}
-      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}
-    \end{equation}
-    $$
-    {^-}\Delta \boldsymbol{x} = \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } +   \boldsymbol { x }_ { i } ^ { n + 1 } - \boldsymbol { x } _ { p } ^ { n + 1 },
-    \quad {^+}\Delta \boldsymbol{x} = \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 }
-    $$
-\begin{equation}\label{eq:apic_mom_g2p}
-  \boldsymbol { B } _ { p } ^ { n + 1 } = \frac { 1 } { 2 } \sum _ { i } w _ { i p } ^ { n } (\boldsymbol { v }_ { i } ^ { n + 1 } ({^-}\Delta \boldsymbol{x}) ^ { T }) {+}  {^+}\Delta \boldsymbol{x} ( { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T }).
-\end{equation}
-
-\end{enumerate}
-For a full proof, that these transfers preserve linear and angular momentum, consult \cite{MPM:APIC}.
-For the choice of dyadic products of cubic b-splines (\ref{eq:cubic_weights}) $\boldsymbol{D}_p$ takes on the simple form:
-\begin{equation}
-\boldsymbol{D}_p^n = \frac { 1 } { 3 } h ^ { 2 } \boldsymbol { I }.
-\end{equation}
-A simple proof (Appendices: \ref{app:dp_proof}) cancelling out the numerous polynomials can be done for instance using SymPy (\cite{Sympy}).\cite{MPM:APIC}\cite{MPM:OLD_APIC}
-\subsubsection{CFL condition} \label{sec:cfl}
-The CFL condition is prominent for FEM: For a stable integration a particle should not travel father than the grid spacing $h$ in a discrete time-step $\Delta t$. $\Delta t$ is thus limited by:
-\begin{equation}
-  \Delta t \leq \frac{h}{\|\boldsymbol{v}^n_i\|_2}.
-\end{equation}
-Assuming $\|\boldsymbol{x}_i^n -\boldsymbol{x}_p^n\| \leq \kappa h$, where $\kappa$ is determined by the interpolation stencil support (cubic 3D: $\kappa = 2\sqrt{3}$) and additionally assuming $\boldsymbol{D}_p^n = k\boldsymbol{I} \Rightarrow (\boldsymbol{D}_p^n)^{-1} = \frac{1}{k}\boldsymbol{I}$ (cubic 3D: $k=\frac{1}{3}h^2$),
-$\|\boldsymbol{v}^n_i\|_2$ can be estimated on the particles. Typically the number of particles is lower. Given Eq. \ref{eq:apic_mom_p2g} this leads to the following estimate:
-$$
-\|\boldsymbol{v}^n_i\|_2 \leq \frac{1}{m_i^n}\left(\sum _ { p } w _ { i p } ^ { n } m _ { p } \| \boldsymbol { v } _ { p } ^ { n } \|_2 + \sum _ { p } w _ { i p } ^ { n } m _ { p } \| \boldsymbol { B } _ { p } ^ { n } \| _ { F } \| ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) \|_2\right)
-$$
-\begin{equation}
-\leq \max _ { p } ( \| \boldsymbol { v } _ { p } ^ { n } \|_2 + \frac { \kappa } { k } \Delta x \| \boldsymbol { B } _ { p } ^ { n } \| _ { F } )
-\end{equation}
-\begin{flushright}\cite{MPM:APIC}\end{flushright}
-\subsection{Discretization}\label{sec:discretization}
-The weak form of the force-balance in (\ref{eq:lagr_force_bal},\ref{eq:eul_force_bal}) implies the following, for MPM preferable, description:
-\begin{equation} \label{eq:weak_mpm}
-  \int _ { \Omega ^ { 0 } } (_0q _ \alpha) (_0\rho_0) (_0a _\alpha)  d _0\boldsymbol{x}= \int _ { \partial \Omega ^ { t^n } } _tq _ \alpha \sigma _ {\alpha\beta} d_tA_\beta(_t\boldsymbol{x}) -  \int _ { \Omega ^ { t^n} } \frac{\partial _tq _ \alpha}{\partial _tx_\beta} \sigma _ {\alpha\beta} d_t\boldsymbol{x}.
-\end{equation}
-The boundary integral is mostly due to collisions and will be ignored.
-\cite{MPM:OPTIMI_INTEGR} discuss level set collisions due to constraint collisions, object penalty collision and penalty self-collisions.
-Collision treatment would need to be involved in the solving of the equation.
-A simple though less accurate method is to process particle collisions separately in a typical computer graphics manner which is assumed for now.
-\subsubsection{Discretize time}
-Any integrator conserving linear and angular momentum could be used to discretize time. The class of time integrators used here are characterized by \begin{equation}\label{eq:midpoint}
-  \frac{y ^ { n + 1 } - y ^ { n }}{ \Delta t} = f^{n+\lambda} = f \left(t^n+\lambda\Delta t, (1-\lambda) y ^ { n } +\lambda y^{n+1}\right)
-\end{equation}
-for a differential equation of order one:
-\begin{equation}
-  \frac{\partial y}{\partial t}( t ) = f (t,y(t)) , \quad y (0) = y _ { 0 }.
-\end{equation}
-A prominent member of this class is the implicit midpoint rule ($\lambda =\frac{1}{2}$).
-Replacing the Lagrangian acceleration $_0a_\alpha$ in equation \ref{eq:weak_mpm} with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:
-\begin{equation}\label{eq:time_disc}
-  \frac{1}{\Delta t} \int _{ \Omega^{t^n}}(_tq_\alpha)(_t\rho)({_tv_\alpha^{n+1}}-{_tv_\alpha^{n}}) d_t\boldsymbol{x} =
-  -  \int _ { \Omega ^ { t^n} } \frac{\partial _tq _ \alpha}{\partial _tx_\beta} \sigma _ {\alpha\beta}^{n+\lambda} d_t\boldsymbol{x}.
-\end{equation}
-\subsubsection{Discretize space}
-A Galerkin discretization brings all spatial terms of equation \ref{eq:time_disc} to a finite-dimen\-sion\-al space: $\boldsymbol{q} \rightarrow \boldsymbol{q}_h$.
-To not further clutter up the notation, the $h$ will be omitted.
-This will replace $q_\alpha, v^n_\alpha , v_\alpha^{n+1}$ with their finite-dimensional grid-based interpolants:
-\begin{equation}
-  {_t q } _ { \alpha } ^ {  { n } } =  ({_t q } _ {  { i } \alpha } ^ {  { n } })  ({_t w } _ { i }) , \quad _tv _ { \alpha } ^ {  { n } } = ({_tv _ {  { j\alpha } } ^ {  { n } }} )( {_t w } _ {  { j } }) , \quad {_tv _ { \alpha } ^ {  { n } + 1 }} = ({_tv _ {  { j } \alpha } ^ {  { n } + 1 }}) ({_t w } _ {  { j } }).
-\end{equation}
-Further, chapter \ref{sec:weak} mentions $\boldsymbol{q}$ can be chosen arbitrarily.
-The Galerkin discretization of a $d$-dimensional space with $m$ grid nodes therefore uses the standard basis functions $e_1,e_2, ...
-,e_{d \times m}$.
-Due to the scalar-product $d\times m$ equations would need to be solved:
-\begin{equation}\label{eq:galerkin}
-  \frac{1}{\Delta t} \int _{ \Omega^{t^n}} ({ _tw } _ {i  })(_t\rho)({_t w } _ { j  })({_tv_{j\alpha}^{n+1}}-{_tv_{j\alpha}^{n}}) d_t\boldsymbol{x} =
- - \int _ { \Omega ^ {  t^ n } }   \frac{\partial{_t w } _ { {i  }}}{\partial_tx_\beta }  \sigma _ {{ \alpha } \beta }  { d } \boldsymbol { x }.
-\end{equation}
-A mass matrix can be factored out as:
-\begin{equation}
-  m_{ij}^n = \int _{ \Omega^{t^n}} {_t w } _ {i  }(_t\rho){_t w } _ { j  } d_t\boldsymbol{x}.
-\end{equation}
-The Lagrangian pull-back relates this to the initial density in the Lagrangian view and discretizing the integral with the initial time-invariant particle mass $m_p \approx V_p^0\rho(_0x_p,0)$:
-\begin{equation}
-m_{ij}^n = \int _{ \Omega^{t^0}} ({_tw_i})(_0\rho_0)({_tw} _ { j  }) d_0 \boldsymbol{x} \approx \sum _ {  { p } }  { m } _ {  { p } }  { w } _ {  { i } } (\boldsymbol { x } _ {  { p } } )  { w } _ {  { j } } ( \boldsymbol { x } _ {  { p } } ).
-\end{equation}
-This matrix is symmetric positive semi-definite (since mass is positive).
-Numerically this matrix is mostly not used as is due to possibility of it being singular.
-This is solved commonly due to a mass-lumping strategy.
-Replacing $m^n_{ii}$ with the $i$-th row sum and clearing all other elements leads to a diagonalization
-\begin{equation}\label{eq:mass_discr}
-\sum _ {  { p } }  { m } _ {  { p } }  { w } _ {ip}   w_{jp}
-\stackrel{\text{\ref{eq:partition_unity}}}{\approx}
-\sum_p m_p w_{ip}
-\stackrel{\text{\ref{eq:mpm:mass}}}{=}
-m_i^n,
-\end{equation}
-where partition of unity $\sum_j w_{jp} = 1$ is used. This is exactly the mass transfer as in equation \ref{eq:mpm:mass}, s.t. no further assembling of a mass matrix is needed.
-The discretization of the right side of \ref{eq:galerkin} happens with an estimated per particle stress $\boldsymbol{\sigma}_p^{n+\lambda}$:
-\begin{equation}\label{eq:stress_discr}
-\int _ { \Omega ^ {  t^ n } }   \frac{\partial{_t w } _ { {i  }}}{\partial_tx_\beta }  \sigma _ {{ \alpha } \beta }  { d } \boldsymbol { x } \approx
-\sum _ { p } (\sigma _ { p }^{n+\lambda}) _ { \alpha \beta } \frac{{ \partial w } _ { {ip}}^n}{ \partial x_\beta }  V _ { p } ^ { n }.
-\end{equation}
-Setting in equation \ref{eq:mass_discr} and \ref{eq:stress_discr} into \ref{eq:galerkin} summarizes the space discretization as:
-\begin{equation}\label{eq:space_discr}
-  \frac{1}{\Delta t} (({m^n\boldsymbol{v}^{n+1}})_i-({m^n\boldsymbol{v}^n})_i) =
-  -\sum _ { p } \boldsymbol{\sigma} _ { p }^{n+\lambda} \nabla w_{ip}^n V _ { p } ^ { n }
-  = \boldsymbol{f}_i^{n+\lambda}.
-\end{equation}
-The momentum change of the left side is by construction equal to a (grid node) force.
-
-Given, that the material point method keeps track of the deformation by a deformation gradient, each particle will have one associated with it for the deformation of its local neighborhood $\boldsymbol{F}^n_p$.
-Based on this one may also gain a volume change measure around the particle as $J^n_p = \text{det}(\boldsymbol{F}^n_p)$.
-Starting with an initial volume of a particle $V_p^0$ the volume may be tracked in time by:
-\begin{equation}\label{eq:volume_discr}
-  V^n_p \stackrel{\text{\ref{eq:j}}}{\approx} V_p^0J^n_p.
-\end{equation}
-In equation \ref{eq:piola} an alternate measure for the stress by the first Piola-Kirchhoff stress is given. The results of \ref{eq:space_discr} may therefore equally expressed by it:
-$$
-\boldsymbol{f}^{n+\lambda}_i \stackrel{\text{\ref{eq:piola},\ref{eq:volume_discr}}}{=}
-- \sum_p \frac{1}{J^n_p}\boldsymbol{P}^{n+\lambda}_{p}(\boldsymbol{F}^{n}_{p})^T \nabla w_{ip}V^0_pJ^n_p
-$$
-\begin{equation}\label{eq:force_disc}
-  = - \sum_p \boldsymbol{P}^{n+\lambda}_{p}(\boldsymbol{F}^{n}_{p})^T \nabla w_{ip}V^0_p.
-\end{equation}
-\begin{flushright}\cite{MPM:COURSE} \cite{MPM:APIC} \cite{bathe2006finite}\end{flushright}
-\subsubsection{Deformation gradient evolution}
-In equation \ref{eq:evol_def_grad} the evolution of the deformation gradient is shown to be:
-\begin{equation}
-  \frac{\partial ^t_0\boldsymbol{F}}{\partial t} =  \nabla{_0\boldsymbol{v}}(_0\boldsymbol{x},t).
-\end{equation}
-Discretizing the Lagrangian deformation gradient in time with equation \ref{eq:midpoint} results in:
-\begin{equation}
-  \frac{\boldsymbol{F}^{n+1}_p + \boldsymbol{F}^{n}_p}{\Delta t} =  \nabla_0\boldsymbol{v}^{n+\lambda}(_0\boldsymbol{x}).
-\end{equation}
-Pushing the right side forward to Eulerian view
-\begin{equation}
-  \frac{\boldsymbol{F}^{n+1}_p + \boldsymbol{F}^{n}_p}{\Delta t} =  \nabla_t\boldsymbol{v}^{n+\lambda}(_t\boldsymbol{x})^t_0\boldsymbol{F} = \nabla_t\boldsymbol{v}^{n+\lambda}(_t\boldsymbol{x})\boldsymbol{F}^n_p
-\end{equation}
-and further applying the Galerkin discretization
-\begin{equation}
-  (_tv^{n+\lambda})_\alpha = (v_{i}^{n+\lambda})_\alpha w_i \Rightarrow \frac{\partial (_tv^{n+\lambda})_\alpha}{\partial x_\beta} =(v_i^{n+\lambda})_\alpha\frac{\partial w_i}{\partial x_\beta}
-\end{equation}
-leads to the final update rule for the deformation gradient:
-\begin{equation}\label{eq:evol_def_grad_disc}
-\boldsymbol{F}^{n+1}_p = \left( \boldsymbol{I} + \Delta t\sum_i \boldsymbol{v}_i^{n+\lambda}({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
-\end{equation}
-The discretization of the position will also be necessary to advance the particles and weight them back:
-\begin{equation}  \frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1}
-  - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda}.
-\end{equation}
-The grid position $\boldsymbol{\hat{x}}_i^{n+1}$ does not correspond to an actual deformation.
-The grid never actually gets deformed (unlike in FEM-methods).
-The discretized evolution of the deformation gradient  \ref{eq:evol_def_grad_disc} is directly a function of $\boldsymbol{\hat{x}}$.
-For the point $\boldsymbol{\hat{x}}_i^{n+1}$ this becomes:
-\begin{equation}\label{eq:evol_def_grad_disc_x}
-  \boldsymbol{\hat{F}}^{n+1}_p(\boldsymbol{\hat{x}}_i^{n+1}) =  \left( \boldsymbol{I} + \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
-\end{equation}
-As part of the class of time integrators in use (\ref{eq:midpoint}) a function of $\boldsymbol{x}_i$ gets evaluated at an in-between point given by:
-\begin{equation}\label{eq:midpoint_x}
-  \boldsymbol{x}_i^{n+\lambda} = \lambda \boldsymbol{x}_i^{n+1} + (1-\lambda)\boldsymbol{x}_i^n.
-\end{equation}
-Plugging this point into \ref{eq:evol_def_grad_disc_x} leads to the following generalization:
-$$\boldsymbol{\hat{F}}^{n+\lambda}_p(\boldsymbol{x}_i^{n+\lambda}) = \left( \boldsymbol{I} + \lambda \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p
-  $$
-  \begin{equation} \label{eq:def_grad_lambda}
-  = (1-\lambda)\boldsymbol{F}^n_p + \lambda \boldsymbol{F}^{n+1}_p.
-\end{equation}
-\begin{flushright}\cite{MPM:COURSE}\end{flushright}
-\subsubsection{Grid nodal forces}
-The notion of a total elastic potential energy function $\Psi$ was introduced in chapter \ref{sec:cor_hyper}. The MPM approximation to this function can be defined by:
-\begin{equation}\label{eq:energy_discr}
-  e(\hat{\boldsymbol{x}}) = \sum_pV^0_p \Psi(\hat{\boldsymbol{F}}_{Ep}(\hat{\boldsymbol{x}})).
-\end{equation}
-The evolution of the deformation gradient \ref{eq:evol_def_grad_disc_x} for a general $\boldsymbol{\hat{x}}_i$ is necessary
-\begin{equation}\label{eq:partial_def_grad}
-\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
-= \delta_{\omega \alpha}
-\frac{\partial w_{ip}}{\partial x_\gamma}F_{\gamma\beta} = \delta_{\omega \alpha}F_{\gamma\beta} \frac{\partial{w_{ip}}}{\partial x_\gamma}
-\end{equation}
-for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of equation \ref{eq:force_disc}:
-$$
-  \frac{\partial e}{\partial \hat{x}_{i\alpha}}(\hat{\boldsymbol{x}})
-  = \sum_p V^0_p \frac{\partial \Psi}{\partial \hat{F}_{\omega\beta}}(\hat{\boldsymbol{F}}_{Ep})\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
-  $$
-  \begin{equation}\label{eq:nodal_force}
-  = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}.
-\end{equation}
-The nodal force can be also described in terms of the Cauchy stress:
-\begin{equation}
-  \boldsymbol{\hat{f}}_i = - \sum_p V^n_p \boldsymbol{\sigma}_p(\boldsymbol{\hat{F}}_{Ep})\nabla{w^n_{ip}}.
- \end{equation}
-Due to equation \ref{eq:def_grad_lambda} the stress computation is summarized for the class of functions in use as:
-\begin{equation}
-  \boldsymbol{P}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{P}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}),
-  \quad
-  \boldsymbol{\sigma}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{\sigma}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}).
-\end{equation}
-\begin{flushright}\cite{MPM:COURSE}\cite{MPM:APIC}\end{flushright}
-\subsubsection{Symplectic midpoint scheme}
-Starting from the last update on the grid which is typically the position with equation \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
-\begin{equation}\label{eq:vel_midpoint}
-  \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \boldsymbol{v}_i^{n+\frac{1}{2}}.
-\end{equation}
-Due to \ref{eq:space_discr} a velocity update can be put together as:
-\begin{equation}\label{eq:force_midpoint}
-\boldsymbol{\hat{v}}^{n+1}_i
-= \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\frac{1}{2}}.
-\end{equation}
-A variable of $n+\frac{1}{2}$ then gets evaluated at the midpoint (\ref{eq:midpoint_x}):
-\begin{equation}
-  \boldsymbol{f}^{n+\frac{1}{2}}_i =  \boldsymbol{f}_i\left(\frac{\boldsymbol{x}_i^n + \boldsymbol{\hat{x}}_i^{n+1}}{2}\right).
-\end{equation}
-The modified energy conserving implicit midpoint scheme from \cite{GONZALEZ} for the material point method differs by the use of a trapezoidal approximation of $\boldsymbol{v}_i^{n+\frac{1}{2}}$:
-\begin{equation}\label{eq:trapezoidal}
-  \boldsymbol{v}_i^{n+\frac{1}{2}} \approx \frac{\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n}}{2}.
-\end{equation}
-The trapezoidal rule has the same order of error $O(\Delta t^2)$ as the implicit midpoint scheme.
-This modification allows for a more direct one-step scheme as shown in the following.
-Plugging $\boldsymbol{x}_i^{n+1}$ of equation \ref{eq:vel_midpoint} into \ref{eq:force_midpoint} leads to the discretized momentum equation:
-\begin{equation}
-  \boldsymbol{h}(\boldsymbol{\hat{v}}_i^{n+1}) = m_i^n \frac{\boldsymbol{\hat{v}_i}^{n+1}-\boldsymbol{v}_i^n}{\Delta t} - \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{\hat{v}}_i^{n+1} + \boldsymbol{v}_i^{n})\right) = 0.
-\end{equation}
-One can recast this method back to an energy function by integrating
-\begin{equation}
-  E(\boldsymbol{v}_i) = \sum_i \frac{m_i^n}{8} \|\boldsymbol{v}_i - \boldsymbol{v}_i^n\|^2_2 + e(\boldsymbol{x}_i^n + \frac{\Delta t}{4}(\boldsymbol{v}_i + \boldsymbol{v}_i^n))
-\end{equation}
-where $e$ is just the discretized elastic potential energy out of \ref{eq:energy_discr}.
-In general this is an optimization objective that needs minimizing to solve for the updated velocities.
-Recasting allows to solve the energy objective in a general manner by an optimization integrator.
-
-In general starting off with an energy description is a more physical approach of the problem.
-The first term can be identified clearly as the kinetic energy.
-Moreover, it easily allows adding potential terms for gravity or collisions \cite{MPM:OPTIMI_INTEGR}, or modifying the energy function altogether based off for instance the fundamental state of matter: as an example the phase transition to and physics of a liquid \cite{MPM:PHASE_CHANGE}.
-The analysis of minimizing the objective is equivalent to finding the zero crossing of the derivative:
-\begin{equation}
-  \argmin_{\forall \boldsymbol{v}_i}(E(\boldsymbol{v}_i)) \Leftrightarrow g(\boldsymbol{v}_i) = \frac{\partial E}{\partial \boldsymbol{v}_i} = 0.
-\end{equation}
-Since the objective is minimizing, the scale of $\boldsymbol{E}$ can be chosen arbitrarily. I.e. a zero crossing does not scale. The scaling here can be identified as:
-\begin{equation}
-  \boldsymbol{g}(\boldsymbol{v}_i) = \frac{\Delta t}{4} h(\boldsymbol{v}_i) = m_i^n \frac{\boldsymbol{v_i}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4} \boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
-\end{equation}
-\begin{flushright}\cite{MPM:COURSE}\cite{MPM:APIC}\end{flushright}
-\subsubsection{Newton's Method}\label{sec:newton}
-\newcommand{\var}[1]{{\ttfamily#1}}% variable
-\begin{algorithm}[t]
-  \caption{Conjugate gradient}\label{alg:conj_grad}
-  \begin{algorithmic}[1]
-    \Procedure{CONJUGATE-GRADIENT}{$H,x,f$}
-    \State $\boldsymbol{x} \gets \Call{InitialGuess}$
-    \State $\boldsymbol{Hx} \gets \Call{ComputeHp}{x}$
-    \State $\boldsymbol{r}\gets \boldsymbol{f} - \boldsymbol{Hx}$
-    \State $\boldsymbol{p}\gets \boldsymbol{r}$
-    \State $\gamma \gets \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
-    \Repeat
-      \State $\boldsymbol{Hp} \gets \Call{ComputeHp}{p}$
-      \State $\boldsymbol { s } \gets \boldsymbol { H } \boldsymbol { p }$
-      \State $\alpha \gets \frac { \gamma } { \langle \boldsymbol { p } , \boldsymbol { s } \rangle }$
-      \State $\boldsymbol { x } \gets \boldsymbol { x } + \alpha \boldsymbol { p }$
-      \State $\boldsymbol { r } \leftarrow \boldsymbol { r } - \alpha \boldsymbol { s }$
-      \State $\kappa \leftarrow \langle \boldsymbol { r } , \boldsymbol { r } \rangle$
-      \If{$ \kappa < \epsilon$} \Comment{alt. fixed amount of steps}
-        \State \Return \label{alg:conj_grad:exit}
-      \EndIf
-      \State $\beta \gets \frac{\kappa}{\gamma}$
-      \State $\boldsymbol { p } \leftarrow \boldsymbol { r } + \beta \boldsymbol { p }$
-      \State $\gamma \gets \kappa$
-      \Until{false} \Comment{exit at \ref{alg:conj_grad:exit}}
-    \EndProcedure
-  \end{algorithmic}
-\end{algorithm}
-Many minimization and root finding algorithms are available. One of such is Newton's Method, which allows for rapid quadratic convergence in a near local neighborhood.
-\begin{equation}
-  \boldsymbol{v}_i^{(i+1)} = \boldsymbol{v}^{(i)}_i + \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right]^{-1}\boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right)
-\end{equation}
-Computing the inverse is numerically irresponsible. Instead the following linear system is solved $\left(\Delta \boldsymbol{v} = \boldsymbol{v}_i^{(i+1)} - \boldsymbol{v}^{(i)}_i\right)$:
-\begin{equation}\label{eq:linear_system}
-  \left[\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}\left(\boldsymbol{v}^{(i)}_i\right)\right] \Delta \boldsymbol{v} =  \boldsymbol{g}\left(\boldsymbol{v}^{(i)}_i\right).
-\end{equation}
-Using the Newton's Method however requires a computation of the Hessian of $E(\boldsymbol{v}_i)$:
-\begin{equation}
-  \frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j} = \frac{m_i^n}{4} - \frac{\Delta t^2}{16}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right).
-\end{equation}
-Computing $\frac{\partial f_i}{\partial x_j}$ for every combination $ij$ would be quite memory-intensive.
-Instead immediately the matrix-vector product on an increment $\delta \boldsymbol{u}_j$ is solved:
-\begin{equation}\label{eq:hessian-matrix-vector}
-  \sum_j\frac{\partial \boldsymbol{g}_i}{\partial \boldsymbol{v}_j}\delta \boldsymbol{u}_j = \frac{m_i^n}{4}\delta \boldsymbol{u}_j - \frac{\Delta t^2}{16}\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}_i + \boldsymbol{v}_i^{n})\right) \delta \boldsymbol{u}_j.
-\end{equation}
-$\frac{\partial \boldsymbol{g}}{\partial \boldsymbol{v}}$ is symmetric, positive definite due it being the Hessian of a convex energy function $E$ of a hyper-elastic material.
-The linear system of equations \ref{eq:linear_system} can be solved with the conjugate gradient method (Algorithm \ref{alg:conj_grad}). There is two important things to notice:
-\begin{enumerate}
-  \item  One may want to use a preconditioner $\boldsymbol{P}^{-1}$ to reduce iteration times.
-However, conservation of momentum on incomplete convergence is fulfilled if \ref{eq:linear_system} is premultiplied by the inverse diagonal mass matrix $\boldsymbol{M}^{-1}$ \cite{MPM:APIC}.
-Thus the equation becomes:
-    $$
-      \left(\frac{1}{4} - \frac{\Delta t^2}{16}\frac{1}{m_i^n}\frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j}\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right)\right) \Delta v
-      $$
-    \begin{equation}
-      = \frac{\boldsymbol{v}_i^{(i)}-\boldsymbol{v}_i^n}{4} -\frac{\Delta t}{4}\frac{1}{m_i^n}\boldsymbol{f}_i\left(\boldsymbol{x}_i^n + \frac{\Delta t}{4} (\boldsymbol{v}^{(i)}_i + \boldsymbol{v}_i^{n})\right).
-    \end{equation}
-  \item Due to \ref{eq:hessian-matrix-vector} the product of the Hessian with an increment ($\sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j$) needs to be computed anew every step since a pure matrix store is too costly.
-\end{enumerate}
-The Hessian is derived following \ref{eq:partial_def_grad} and \ref{eq:nodal_force}:
-\begin{equation}
-  \frac{\partial e}{\partial {x}_{i\alpha} \partial x_{j\tau}}
-= \sum_p V^0_p\frac { \partial ^ { 2 } \Psi } { \partial  { F}_{\alpha \beta} \partial { F } _{\tau \sigma}}  (F^n_p)_{\omega \sigma}\frac{\partial w_{jp}}{\partial x_\omega} (F^n_p)_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\frac{\partial {f}_{i\alpha}}{\partial x_{j\tau}}.
-\end{equation}
-The Hessian increment can be computed with a two-stage process:
-\begin{enumerate}
-  \item Compute a particle quantity $\boldsymbol{A}_p$ with a grid-to-particle-transfer:
-    \begin{equation}\label{eq:Ap}
-    \boldsymbol{A}_p = \frac { \partial ^ { 2 } \Psi } { \partial \boldsymbol { F } \partial \boldsymbol { F } } ( \hat { \boldsymbol { F } } _ { p } ( \hat { \boldsymbol { x } } )) : \left(\sum_j \delta \boldsymbol{u}_j (\nabla w_{jp}^n)^T \boldsymbol{F}^n_p\right).
-    \end{equation}
-  \item Compute the Hessian increment ($\delta \boldsymbol{f}_i$) with a particle-to-grid-transfer:
-    \begin{equation}\label{eq:delta_f}
-  - \delta \boldsymbol{f}_i = - \sum_j \frac{\partial \boldsymbol{f}_i}{\partial \boldsymbol{x}_j} \delta \boldsymbol{u}_j = \sum_p V_p^0 \boldsymbol{A}_p(\boldsymbol{F}^n_p)^T \nabla w_{ip}^n.
-\end{equation}
-\end{enumerate}
-\clearpage
-\subsection{General-purpose computing on graphics processing units}\label{sec:gpgpu}
-This chapter is split into two parts.
-\ref{sec:optim} tends to the memory hierarchy of the GPU and typical optimizations done in GPGPU work.
-Furthermore, chapter \ref{sec:metrics} will elaborate metrics to quantify optimizations.
-\subsubsection {GPGPU Optimization opportunities}\label{sec:optim}
-The speed of these memory types is given in the following starting with the fastest:
-Register > Shared Memory > Local, Global. Table \ref{table:gpu_mem} gives an overview over their capabilities which is referred to in the following.
-
-Therefore, one should prefer register memory for intra-thread caching.
-Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block.
-Global memory is the persistent storage on the GPU.
-
-\textbf{Host communication:} A common performance bottleneck is stalling the GPU with CPU-GPU transfers.
-The attainable bandwidth for PCIe x16 3.0 between host and device is 16GB/s while the communication between device and GPU is 224GB/s (GTX 970) \cite{NVIDIA:GTX970}.
-This includes doing calculations on the GPU that may run faster on the CPU.
-The API may also stall the GPU if the CPU is not issuing instructions fast enough to the GPU.
-This may be due to calculations on the CPU or, even worse, calculations depending on GPU results.
-This thesis tries to avoid PCI-transfers wherever possible.
-
-\begin{table}
-  \begin{tabular}{ | l | l | l | l | l |}    \hline
-    Memory       &Location on/off chip   &Cached   &Scope                &Lifetime   \\\hline
-    Register     &On                     &n/a      &1 thread(no shuffle) &thread     \\\hline
-    Local        &Off                    &L2       &1 thread             &thread     \\\hline
-    Shared       &On                     &n/a      &block                &block      \\\hline
-    Global       &Off                    &L2       &all threads+host     &host alloc \\
-    \hline
-  \end{tabular}
-  \caption{GPU Memory \cite{NVIDIA:BEST:PRACTICE}}\label{table:gpu_mem}
-\end{table}
-\textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists.
-NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront.
-In general they are called subgroups.
-All operations are issued per subgroup.
-A memory access uses the L2-caching behavior.
-A L2-cache-line is 32 bytes long.
-Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.
-If all threads of a warp access memory within this cache-line, only one memory request of 32 bytes needs to be made.
-In contrast, if a warp accesses N different cache-lines, N*32 bytes need to be loaded.
-
-The unified L1/texture-cache acts as a coalescing buffer for the thread of a subgroup (Maxwell architecture \cite{NVIDIA:MAXWELL}).
-It does not matter in which order the cache-line is accessed by the warp.
-I.e.
-random or reversed access on a single cache-line are also coalesced.
-On a global level the easiest to control access pattern thus becomes consecutive or reversed memory accesses.
-L2 is a write back cache.
-Thus this assessment also qualifies for global memory writes.
-\cite{NVIDIA:MEMORY_BAND}
-
-\textbf{Shared Memory bank conflicts:} Shared memory is split into 32 banks of 4 bytes.
-The hardware groups together unique memory accesses on these banks.
-E.g., if two threads within a subgroup access bank 17, a bank conflict occurs, and the access needs to be split into two groups.
-The one exception is a broadcast: All threads within a subgroup access the same bank.
-The number of banks is equal to the warp size on the GTX970 s.t.
-throughput for 32 floats is maximized if every thread does a unique bank access.
-\cite{NVIDIA:BEST:PRACTICE}
-
-\textbf{Register spilling:} Local memory should be avoided.
-Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached.
-The available number of registers per thread is calculable with the CUDA Occupancy Calculator \cite{NVIDIA:OCCUPANCY} on NVIDIA hardware.
-If the occupancy falls off to 0\%, register spilling occurs.
-The occupancy is either given absolute due the amount of active warp or relative divided by the possible maximum amount of warps on a Streaming Multiprocessor(SM).
-Occupancy allows the SM to schedule between warps: If the current warp cannot continue (memory latency, barrier etc.), the warp scheduler will switch to another.
-This is called a warp stall, table \ref{table:metric}.
-Occupancy is mostly affected due to resource limits on the SM (shared memory, register), as can be seen in the Occupancy Calculator.
-Non optimal occupancy might not have any performance impact at all when the GPU has enough warps to switch to at all times.
-\cite{NVIDIA:ACHIEVED_OCCUPANCY}
-
-There is no exact way to find out the exact number of registers for any OpenGL shader.
-The program object assembly output of \textit{nvemulate} \cite{NVIDIA:EMULATE} may give an indicator.
-It will output an intermediate language of one of the families of \textit{NV\_gpu\_\-programX} (where $X$ stands for the version).
-TEMP variables are registers.
-Since \textit{NV\_gpu\_programX} is a family of intermediate languages, there is however no guarentee these are not further optimized.
-Otherwise, close monitoring of the occupancy is recommended.
-All other metrics discussed are readily available for OpenGL.
-
-\textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits.
-One should start however with a block size of at least 64, as can be seen in the Occupancy Calculator.
-Increasing the block size is useful in combination with shared memory usage.
-Shared memory accesses arinstructionspically faster than global memory accesses.
-The more threads of a block reuse shared memory data, the faster the memory throughput will be.
-Therefore, operators which involve filtering (also MPM-transfers) can use this behavior.
-
-\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it.
-Otherwise, merging sequential and parallel work improves I/O latency hiding.
-Since the block size is limited by the GPU, some algorithms need to be split in tree structures to be processed.
-An indirection cannot be avoided for writing back and loading again from global memory.
-Therefore, algorithms often improve much further from more sequential work to avoid this indirection.
-
-\textbf{Loop unrolling:} Foremost, loop unrolling reduces instructions and can thus speed up compute-bound work.
-Moreover, loop unrolling might help the compiler reorder instructions and thus execute instructions faster.
-Instructions benefiting greatly from this are global loads.
-\cite{NVIDIA:PEAK_PERFORMANCE} shows at the example of motion blur and ray-marching that this greatly can increase throughput.
-Whereas the motion blur example uses a fixed amount of samples and therefore is easily unrollable, ray-marching is a dynamic process.
-Dynamic loops can however be partially unrolled.
-In the case of ray-marching this would for instance fetch two samples at once although it might process only one.
-In short, this merges two consecutive loop iterations.
-The latency of the second instruction might be partly hidden by the first one and thus speed up the work load.
-Mind however, that loop unrolling can increase register pressure.
-More registers need to be stored and process the relevant code where a dynamic loop reuses the same registers for any iteration.
-
-\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\cite{NVIDIA:PEAK_PERFORMANCE}\end{flushright}
-\subsubsection{GPU Metrics}\label{sec:metrics}
-The GPU metric topping all others is the raw computing time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
-\begin{enumerate}
-  \item For the GPU most interesting is the scalability with increasing input data sizes. All algorithms here however scale linear with the input size.
-  \item \label{it:speedup}Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$.
-Therefore, one might be interested how long the GPU needs to calculate one physical second:
-    \begin{equation}
-    \Delta t^c_{1s} = \frac{\Delta t_c}{\Delta t_p}.
-    \end{equation}
-    Otherwise, one would calculate the speedup of one procedure over a another procedure for a direct comparison:
-
-    \begin{equation}
-      \text{Speedup of Procedure 1} = \frac{\Delta t_{c1}}{\Delta t_{p1}} \frac{\Delta t_{p2}}{\Delta t_{c2}}.
-    \end{equation}
-\end{enumerate}
-\begin{table}
-  \begin{tabular}{ | l | l | l |}    \hline
-    \textbf{Metric}   & \textbf{Max}(GTX 970) & \textbf{Description} \\ \hline
-    Speedup           & -                     & item \ref{it:speedup}\\\hline
-    $\Delta t_c$      & -                     & computing time\\\hline
-    VRAM SOL\%        & 100                   & memory througput w.r.t. to hw.-limit\\\hline
-    SM SOL\%          & 100                   & instruction throughput\\\hline
-    L2 SOL\%          & 100                   & L2-cache throughput\\\hline
-    Tex SOL\%         & 100                   & L1-cache throughput\\\hline
-    L2-Hitrate\%      & 100                   & L2 hit-miss ratio\\\hline
-    SM Active\%       & 100                   & one warp active avg. over SMs\\\hline
-    SM Issue Util.\%  & 100                   & amount of cycles an instr. was issued\\\hline
-    SM Occupancy      & 64.0                  & active warps avg. over SMs\\\hline
-    Read Active\%     & 100                   & \% of time spend on VRAM reads\\\hline
-    Write Active\%    & 100                   & \% of time spend on VRAM writes\\\hline
-    \multicolumn{3}{|l|}{\textbf{Warp Stall Reasons:}}\\\hline
-    Long Scoreboard\% & 100                   & VRAM latency (mostly reads)\\\hline
-    Short Scoreboard\%& 100                   & L2/Shared Mem. latency\\\hline
-    Barrier\%         & 100                   & block synchronization \\\hline
-    No Instruction\%  & 100                   & instruction fetching or\\
-		      &                       & waiting on instruction cache miss\\\hline
-  \end{tabular}
-\caption{GPU Performance Metrics}\label{table:metric}
-\end{table}
-The other GPU metrics in table \ref{table:metric} are acquired using NVIDIA Nsight with OpenGL Performance Markers \cite{NVIDIA:METRICS}.
-
-None of these metrics necessarily have a correlation to computing time.
-They are however good indicators on which operations/hardware units the GPU spends its time on.
-The top speed of light(SOL)\% metric measures fraction of the throughput to the bandwidth of that hardware unit.
-NVIDIA aims to get the top SOL\% metric between 60\% and 80\%.
-Occupancy was discussed in \ref{sec:optim}.
-\cite{NVIDIA:PEAK_PERFORMANCE}
-
-Aiming to get shaders memory-bandwidth limited, and minimizing the amount of needed accesses for any given workload, is the top priority for simple tasks.
-Achieving memory-bandwidth limit is equal to bringing the VRAM SOL\% between 60\% and 80\%.
-The metrics Read Active\%, Write Active\% and Long Scoreboard\% will simultaneously increase by following that aim.
-
-The metric SM Issue Utilization (per active cycle) is important for compute-bound work, as memory-bandwidth limited work will spend most of its time outside of the SM fetching or writing data.
-This metric tells if the SM can issue instructions fast.
-NVIDIA also aims between 60\% to 80\% for this metric if the shader is compute-bound.
-Furthermore, for compute-bound work often the top warp stall reason is waiting on instruction fetches.
-
-\clearpage
-\section{Implementation}\label{sec:implementation}
-The computing time $\Delta t_c$ of shaders is gained by OpenGL timer queries \cite{KHRONOS:TIMER_QUERY}.
-A small benchmarker takes care of queuing timer queries and resolving them a frame later.
-This is done to not stall the GPU pipeline waiting on those queries.
-At the end of testing data is gathered and statistical quantities are written into a log file.
-\subsection{GPU memory layout}\label{sec:soa_aos}
-\noindent\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[caption={AoS-Layout},language={GLSL},style={GL}]
-struct Particle {
-  vec4 position;
-  vec4 velocity;
-} Particles[n];
-\end{lstlisting}
-\end{minipage}
-\hfill\noindent\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[caption={SoA-Layout},label={llst:soa},language={GLSL},style={GL}]
-struct Container{
-  vec4 positions[n];
-  vec4 velocities[n];
-} Particles;
-\end{lstlisting}
-\end{minipage}
-
-Coalescing leads to the motivation of using structure of arrays(SoA) over arrays of structures(AoS) memory layouts.
-The latter is the more commonly taught approach while high performance CPU code may also prefer SoA.
-This opens up the field of data-oriented design which is too big too cover and not main part of this thesis.
-
-In short, one might not access every variable of a particle in a shader.
-I.e.
-in most compute work one has a pipeline of different passes each requiring another subset of the variables.
-Assume for instance, one shader only requires the position.
-Then it is more favorable for consecutive threads to load consecutive position vectors.
-The cache-line is then filled with position vectors which all will get used.
-In contrast, in an AoS layout the cache-line will have a position and a velocity loaded but the shader never uses the velocity.
-This halves the throughput.
-This process is shown in Figure \ref{fig:soa} and \ref{fig:aos}.
-
-\noindent\begin{wrapfigure}{R}{0.5\textwidth}
-\hfill\begin{minipage}{0.45\textwidth}
-\begin{lstlisting}[caption={OpenGL Layout},label={llst:opengl},language={GLSL},style={GL}]
-//if SoA
-#define Particle PREC_VEC_TYPE
-#define Particle_position 0
-#define Particle_velocity 1
-#define Particle_size 2
-//if AoS
-struct Particle{
-  PREC_VEC_TYPE position;
-  PREC_VEC_TYPE velocity;
-};
-buffer Container{
-  Particle variables [ ];
-};
-\end{lstlisting}
-\end{minipage}
-\end{wrapfigure}
-
-Due to this we motivate the C++-library \textit{magic\_get} \cite{APOLUKHIN:MAGIC_GET} for basic reflection.
-Our code manipulates AoS layouts into SoA layouts and back using reflection.
-The MSVC-Compiler has a hard coded stack limit for these kind of operations.
-A struct may be not bigger than 256 Bytes before throwing a compiler error.
-With the optional use of double-precision this limit is quite restrictive towards that compiler.
-
-Furthermore, for sizes starting with 256 KB the NVIDIA-Compiler could not link the OpenGL program using the SoA-Layout as in Listing \ref{llst:soa}.
-As an alternative meta information is provided in the style of reflection macros.
-Therefore, the container is collapsed as in Listing \ref{llst:opengl}.
-Memory is then accessible by SoA by memory offset and by AoS by variable or by memory offset for a more abstract view of the memory.
-The AoS setup is shared between C++ and OpenGL code.
-One could also build a shader using the abstract layout to convert from SoA to AoS.
-For more coherency converting the whole process to reflection macros is sensible.
-
-\begin{figure}[htbp]
-  \centering
-  \includesvg[svgpath =images/,width=\textwidth]{soa}
-  \caption{SoA-Layout in practice. The first instruction is shown in orange, the second instruction in green. If green is left out throughput is still maximized}
-\label{fig:soa}
-\end{figure}
-\begin{figure}[htbp]  \centering
-  \includesvg[svgpath =images/,width=\textwidth]{aos}
-  \caption{AoS-Layout in practice.
-The first instruction is shown in orange, the second instruction in green.
-Orange already loads both cache-lines.
-If green is left out throughput is halved.
-Also green is an L2-cache lookup requiring additional overhead over the direct SoA way.}
-\label{fig:aos}
-\end{figure}
-
-To name a few more cases that macros handle for us:
-\begin{enumerate}
-  \item Multiple buffering, e.g. double buffer for unsorted and sorted particles.
-  \item Single and double precision (PREC\_VEC\_TYPE in Listing \ref{llst:opengl}).
-  \item A decorator pattern allows to add buffer information to existing buffer information: For instance index sort needs an extra buffer containing the indices.
-Full sort requires a double buffer.
-Thus a sorted buffer gets an attachment(decoration) to contain these information as well as signaling which sort was used in the first place for further operations down the pipeline.
-  \item Generalize methods to use variants of the same shader with different operations:
-
-    \texttt{\#define unary\_op(x) length(x)}
-
-    \texttt{\#define binary\_op(x,y) x+y}
-\end{enumerate}
-For debugging purposes the shader is written to the hard drive with all processed includes.
-The gcc-pre\-pro\-ces\-sor(\cite{GCC:PREP}) can preprocess the shader with options: \texttt{gcc -E -x c -P}.
-Except for the omission of the glsl version or extension macros no problems were encountered in this process.
-
-\subsection{Parallel Reduction \& Scan} \label{sec:red_scan}
-The schemes presented in the following are a result of the optimization opportunities discussed in \ref{sec:optim}.
-
-\textbf{Parallel Reduction}: Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ and an array of values $\left[a_0,a_1, ... ,a_n\right]$ reduction computes:
-\begin{equation}
-  r = a_0 \circ a_1 \circ ... \circ a_n.
-\end{equation}
-This OpenGL parallel reduction implementation uses \cite{NVIDIA:PARALLEL_REDUCTION} as a reference.
-To avoid memory padding in shared memory the addressing scheme is given by the sequential addressing scheme in Figure \ref{fig:sequential}.
-\begin{figure}[htbp]
-    \centering
-  \includesvg[svgpath =images/,width=\textwidth]{sequential}
-  \caption{Sequential addressing for \texttt{vec4}s.
-Even threads are blue.
-Odd threads are yellow.
-Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers.
-The second instruction is only green.
-The first and all following instructions can be handled fully concurrently.
-No memory padding is required.}
-\label{fig:sequential}
-\end{figure}
-
-Parallel reduction schemes using warp shuffle-operations require interleaved addressing due to memory locality restrictions, Figure \ref{fig:interleaved}. \cite{NVIDIA:SHUFFLE}
-
-\cite{NVIDIA:PARALLEL_REDUCTION} also make use of combining parallel and sequential work.
-Every thread adds a multiple of two sequential adds into shared memory.
-After this the thread does its parallel work.
-Therefore, the reduction factor one invocation does is multiple times higher and allows higher input data sizes without losing any performance, refer to chapter \ref{sec:eval}.
-
-Parallel reduction solves, assuming a \texttt{unary\_op(x)} directly after memory load, the following problems in MPM:
-\begin{itemize}
-  \item CFL condition of chapter \ref{sec:cfl} with operators,
-  \begin{itemize}
-  \item \texttt{unary\_op(x) := length(x)} and
-  \item \texttt{binary\_op(x,y) := max(x,y)}.
-  \end{itemize}
-\item discretized conservation of governing equations between time steps or transfers in chapter \ref{sec:mass} and \ref{sec:apic} with
-\begin{itemize}
-  \item \texttt{unary\_op(x) := x} and
-  \item \texttt{binary\_op(x,y) := x + y}.
-  \end{itemize}
-\item calculate the maximum count of particles of all bins within a block (\textit{per-block parallel reduction} \ref{sec:active_blocks}):
-\begin{itemize}
-  \item \texttt{unary\_op(x) := x} and
-  \item \texttt{binary\_op(x,y) := max(x,y)}.
-\end{itemize}
-\end{itemize}
-
-\textbf{Scan:}  Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ...
-,a_n\right]$ an exclusive scan computes the array:
-\begin{equation}
-  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
-\end{equation}
-\begin{wrapfigure}[13]{r}{0.20\textwidth}
-\vspace{-11}
- \includesvg[svgpath =images/,width=0.20\textwidth]{scan}
- \caption{Scan (add)}
-  \label{fig:scan}
-\end{wrapfigure}
-The scan uses an interleaved addressing pattern, Figure \ref{fig:interleaved}.
-The interleaved addressing pattern has the benefit over the sequential one of already computing the partial scans for all odd indices (although misaligned).
-Note, that the reduction needs to reduce to the right achieved by writing into the right element instead of the left in Figure \ref{fig:scan}.
-A full scan is then computed as \cite{NVIDIA:SCAN}:
-\begin{enumerate}
-  \item \label{it:up-sweep} Upsweep-Phase: Top-to-Bottom reduction to the right of the array.
-  \item Inserting the neutral element into the last element of the array.
-  \item Downsweep-Phase: Reverse execution order of item \ref{it:up-sweep} (inverted tree), and in addition overwrite the left element with the right one's.
-\end{enumerate}
-Again, sequential work can be combined with parallel work.
-\cite{NVIDIA:SCAN_MODERN} name the process: Raking.
-The general technique is also called register blocking.
-Instead of loading one element per thread, one thread loads N elements and computes a sequential partial scan in register memory.
-The reduced last value of the array is written into shared memory and the parallel scan begins.
-Afterwards the parallel scan result is spread out(raking) with the sequential partial scan.\cite{NVIDIA:SCAN_MODERN}
-\begin{figure}[htbp]
-  \includesvg[svgpath =images/,width=\textwidth]{interleaved}
-  \caption{Interleaved addressing \texttt{vec4}s.
-Even threads are blue, odd are yellow.
-Shared-memory banks per instruction are colored: The banks of the first instruction are green and red numbers.
-The second instruction is only green.
-Memory padding needs to be done for concurrent shared-memory access.
-The first and all following instructions can be handled fully concurrently.}
-\label{fig:interleaved}
-\end{figure}
-As mentioned, the interleaved addressing requires a memory padding for conflict-free addresses. For block sizes of 32 - 512 this becomes:
-\begin{equation}\label{eq:bank}
-  p = u + (u << \log_2(n)), \text{ or }   p = u + \frac{u}{n}
-\end{equation}
-where $p$ is the padded memory offset, $u$ is the unpadded memory offset and $n$ is the number of banks.
-For a block size of 1024 additional padding is needed for a full conflict-free access:
-\begin{equation}\label{eq:bank_1024}
-  p = u + (u << \log_2(n)) + (u << (2 * \log_2(n))), \text{ or }   p = u + \frac{u}{n} + \frac{u}{n*n}.
-\end{equation}
-However, the added computation time does not outweigh the bank-conflicts.
-So the process is reverted to equation \ref{eq:bank}.
-Note, that this is a correction to \cite{NVIDIA:SCAN} where C++-Operator-Precedence of $+$ over $<<$ is not recognized, and additionally incorrectly $ u << n$ is used in the first part of equation \ref{eq:bank_1024}.
-
-A scan is used to calculate memory offsets for stream compaction or sorting algorithms.
-
-One can also calculate a per block scan (with halo) without doing a full scan for every block (Appendices \ref{sec:block_scan}).
-This implementation did not end up using this more closely related approach to \cite{NVIDIA:NNSEARCH} as it would commit shared memory bank-conflicts defying the architecture of the GPU.
-It also puts restriction on the maximum allowed particles per block.
-Instead preferred is batching a fixed amount of particles, more in chapter \ref{sec:transfers}.
-
-The input size for reduction and scan is limited.
-It's calculable by the $block\text{ }size$ $ \times sequential\text{ }loads$.
-To support bigger input sizes pyramid structures are in use to process the elements the blocks reduced.
-In this implementation the input size is mostly reduced s.t.
-the first level above in the pyramid can be computed within a single block.
-For the scan this result is then written back.
-This was enough for the input sizes employed and can be similarly extended by pyramid structures.
-
-\subsection{Counting sort \& stream compaction}\label{sec:sort}
-Sorting on the GPU can dramatically increase workload performance of subsequent steps if they can profit from following things:
-\begin{enumerate}
-  \item Accesses can now be handled in a coalesced fashion. \label{it:sort_1}
-  \item Data can now be reused due to L2-Cache and/or shared memory. \label{it:sort_2}
-\end{enumerate}
-This is the case for algorithms which require neighbor communication.
-Supported is a full deep copy of all particle variables.
-A deep copy profits from item \ref{it:sort_1} and \ref{it:sort_2} but takes a hit in performance of uncoalesced accesses for all variables.
-Alternatively, an indexing to access an existing particle buffer in the specified sorted manner is although supported.
-Indexing only profits from item \ref{it:sort_2} and subsequent shaders will be uncoalesced.
-
-\textbf{Stream compaction} reduces applicable data for subsequent operations.
-The applicable data is checked by a condition.
-There are a lot of cases where stream compaction can reduce running times: In chapter \ref{sec:rel_work} particle activation is mentioned as one of such.
-Chapter \ref{sec:active_blocks} will reduce active blocks with an indexed stream compaction.
-Collision treatment could be implemented as an operation following a collision detection etc.
-Again indexing and deep copies can be executed for the same reasons mentioned in \ref{it:sort_1} and \ref{it:sort_2}.
-
-\textbf{Counting sort}: As the sorting algorithm of choice counting sort is used.
-Counting sort is split into three parts:
-\begin{enumerate}
-  \item
-\parbox[t]{\dimexpr\textwidth-\leftmargin}{%
-      \vspace{-7.5}
-      \begin{wrapfigure}[10]{r}{0.425\textwidth}
-	\vspace{-33}
- \includesvg[svgpath =images/,width=0.425\textwidth]{binning}
-  \caption{Binning: Particle adds one to $c_i$ after storing it as $o_p$.}
-  \label{fig:binning}
-\end{wrapfigure}
-\textbf{Binning}: Transform the particle position into the grid space s.t.
-every integer in range corresponds to a grid node.
-Bin due to flooring the transformed particle: Store the particle count $c_i$ of this grid node on the particle as an offset $o_p$ into this grid node.$$o_p := c_i.$$Following that, up the counter on the grid by one:$$c_i := c_i +1.$$
-}
-This process is visualized in Figure \ref{fig:binning}.
-  \item \textbf{Scan}: Compute a per grid node scan $s_i$ on the counter $c_i$ following \ref{sec:red_scan}.
-
-  \item \textbf{Reordering}: $i_p := s_i + o_p$ will now lead to a new ordering of the particle indices.
-Sorting is now similar to methods \texttt{sort\_by\_key(uint i)} where ordering of particles within the same bin is undefined.
-\end{enumerate}
-\begin{flushright}\cite{NVIDIA:NNSEARCH}\cite{MPM:GPU}\end{flushright}
-
-The one-dimensional indexing of the grid nodes can be chosen as preferred. Possible are a per-block indexing or a global indexing which are both supported.
-
-Particles could potentially access the same bin at the same time resulting in a race condition.
-A naive solution to the problem are atomic accesses.
-As long as the (atomic) writes can be handled coalesced, this process can be done fast by the GPU, chapter \ref{sec:eval}.
-This is however only the case for a full deep sort.
-Therefore, binning in the case of a full sort acts upon the last sorted state.
-Double buffering particles allows to do so.
-However, binning is done after a position update that changes the ordering.
-Full coalescing will therefore generally not occur.
-
-Alternatively, one could work again with the CFL condition. Since particles move at most one bin in the 27-neighborhood, one could mimic a particle-to-grid-transfer, chapter \ref{sec:p2g}:
-\begin{enumerate}
-  \item \textbf{Labeling}: Label the particles by the bin they want to change to.
-  \item \textbf{Neighborhood lookup}: A bin then assigns offsets to the particles in the neighborhood and increases its counter.
-\end{enumerate}
-Similarly to the particle-to-grid-transfer one could choose from the different algorithms presented there to do so.
-However without additional endeavors the order within a bin is not preserved.
-One would need to rely on the L1-cache as a coalescing buffer as does the atomic global solution.
-
-Alternatively, one could choose a closer approach to \cite{NVIDIA:NNSEARCH}'s neighborhood search which greedily loads all particles. Ordering can easily be preserved then.
-
-Note that the support in their approach is smaller (27 instead of 64 nodes) and the data required per particle is only the label.
-Chapter \ref{sec:active_blocks} shows how to do this only for active blocks.
-Due to the difference in support the same pipeline cannot be used.
-
-In summary, there is little gains to do so yet as binning is only executed once per step and is not the dominant performance cost.
-
-\subsection{MPM operations}\label{sec:transfers}
-Easy to parallelize are all operations that are executed entirely on the grid or entirely on the particles:
-$$\square_p = \square_p \circ \square_p \circ ... \circ \square_p,$$
-$$\square_i = \square_i \circ \square_i \circ ... \circ \square_i.$$
-One thread will just equal one node or particle, respectively.
-For one variable assignments a map shader is used that applies a unary operation to any input element.
-This can be used for instance for the mass divide of momentum on the grid (equation \ref{eq:mass_divide}) or resetting buffers that are target of global atomic operations.
-
-In contrast much harder are transfers between the two structures that are essential to the material point method:
-
-Particle-to-grid(P2G)-transfers are the mass (equation \ref{eq:mpm:mass}) and momentum (equation \ref{eq:apic_mom_p2g}) transfers to the grid, as well as the computation of the Hessian of equation \ref{eq:delta_f}.
-In general these can be summarized as:
-$$\square_i = \sum_p \square_p \circ \square_{ip}.$$
-
-The transfers back to the particles are called grid-to-particle(G2P)-Transfers.
-Apart from the APIC-transfers (equation \ref{eq:apic_mom_g2p}) and the intermediate Matrix $\boldsymbol{A}_p$ for the Hessian (equation \ref{eq:Ap}), the deformation gradient update $\boldsymbol{F}^{n+1}_p$ is also a G2P-transfer (equation \ref{eq:evol_def_grad_disc_x}).
-In short, all transfers of the form:
-$$\square_p = \sum_i \square_i \circ \square_{ip}.$$
-This concludes all necessary operations for the MPM as derived.
-
-The number of transfers per physical time step is much higher than in simple PIC- or SPH-code which are mostly interested in pressure contributions($float$).
-This motivates making the transfers a very highly optimized operation.
-The material point method requires for an elastic material as derived:
-\begin{enumerate}
-  \item One P2G-transfer of mass and momentum $m_i, \boldsymbol{v}_i$(combined $float4$).
-  \item Per conjugate gradient iteration:
-    \begin{enumerate}
-      \item One G2P-transfer for the $3 \times 3$ Matrix $\boldsymbol{A}_p$
-      \item One P2G-transfer of $\delta \boldsymbol{f}_i$ ($float4$)
-    \end{enumerate}
-  \item Per Newton iteration: One G2P-transfer for $\boldsymbol{F}^{n+1}_p$
-  \item One G2P-transfer of the $3 \times 3$ Matrix $\boldsymbol{B}_p^{n+1}$, the position $\boldsymbol{x}_p^{n+1}$ and the velocity $\boldsymbol{v}_p^{n+1}$
-\end{enumerate}
-In summary, a lot of data is to be transferred between grid and particles which will only increase with more complex models.
-In contrast, any preprocessing like sorting (chapter \ref{sec:sort}) or filtering active blocks (chapter \ref{sec:active_blocks}) need to be executed only once every physical times step.
-Each transfer can profit from it.
-Therefore, the transfers are the main focus of this implementation.
-
-The MPM-transfers are different from a typical stencil or filter operation in two main factors:
-\begin{enumerate}
-  \item \textbf{Dynamic weights}: The stencil is not static.
-In MPM the weighting function $w_{ip}$ (or $\nabla w_{ip}$ etc.) is dependent on the particle position $w_i(\boldsymbol{x}_p)$.
-This also means the position needs to be temporarily stored to apply the transfer.
-  \item \label{it:part_nodes} \textbf{Participating nodes}: A stencil operation reads from a fixed amount of nodes in its neighborhood given by the support of the stencil.
-This is a node to node relationship which all threads typically participate in from start to finish for the full range of the support.
-
-In a MPM-transfer the number of participating nodes can vary.
-A node may have one, zero or any positive integer of particles associated with it.
-Thus in a transfer some nodes may not participate at all or have more load than any of the nodes in its neighborhood.
-\end{enumerate}
-Item \ref{it:part_nodes} does sound more problematic from a parallelization perspective than it is in practice.
-In chapter \ref{sec:rel_work} particle resampling is mentioned of a means to fill material gaps.
-This is an inherent mechanic to control the number of particles within a cell.
-Although it is designed to augment the numerics, the merge method can be easily used to control the upper boundary of particles within a cell.
-Nevertheless the number of particles within a cell is still variable.
-All methods presented in the following are not restricted to bins of a fixed size of particles.
-
-The sorting and binning employed in chapter \ref{sec:sort} allows to split the grid into a partitioning of blocks.
-Blocks map well to the thread group nature of GPUs.
-A single thread within a thread group thus corresponds to a grid node within that block, respectively.
-
-\subsubsection{Grid-to-particle transfers}\label{sec:g2p}
-The straightforward solution to this problem is to handle all particles independently from each other.
-One particle adds up the contributions from each of the grid nodes in range of the interpolation function.
-In three dimensions this can be done for instance with a two times nested for-loop within a thread.
-Luckily, grid nodes in a uniform grid are inherently sorted which takes care of coalescing.
-
-Furthermore, sorting particles due to chapter \ref{sec:sort} increases caching behavior as variables of grid nodes are directly reused by particles in the same group.
-This results in direct L2-cache hits instead of going the long way of reloading from global memory.
-
-With the mentioned partitioning of the grid into blocks, the shared memory architecture of the GPU can be exploited.
-This requires two level binning of particles to nodes within blocks \cite{MPM:GPU}.
-Figure \ref{fig:g2p_transfer} shows how the G2P-transfer using shared memory is realized.
-Algorithm \ref{alg:g2p} shows this process briefly.
-Due to binning a particle is associated with its lower left node.
-In Figure \ref{fig:g2p_transfer} four threads are active and form the thread group.
-All visible grid nodes need to be loaded by the threads.
-This includes the halo of the block which is dependent upon the support of the interpolation function.
-For uneven polynomials of the interpolation function the floored position is very favorable.
-It allows the left support to be one smaller than the right support.
-As a result, the weight of a most outer-edge particle (of the active particle region) just reaches zero for any neighborhood-node outside the halo.
-
-\begin{figure}[htbp]
-    \centering
-  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_inv_par}
-  \caption{G2P-transfer: The transfer of a single particle is split into 64 sequential runs (using cubic weights). Shown are the first two runs.}
-  \label{fig:g2p_transfer}
-\end{figure}
-The bigger the block the less halo nodes need to be loaded in total.
-Given an interpolation function with left support $l$ and right support $r$, total grid size $g$, and block sizes $b_x,b_y,b_z$ the general function to calculate the number of elements loaded is:
-\begin{equation}
-  m(b_x,b_y,b_z) = \frac{g}{b_x*b_y*b_z}(l+b_x+r)(l+b_y+r)(l+b_z+r).
-\end{equation}
- As an example, take a 128*128*128 grid and the cubic interpolation function which has a left support of 1, and a right support of 2. Subdivide the grid into 4*4*4 or 8*4*4 elements respectively:
- $$m(4,4,4): 11,239,424\text{ Elements, }m(8,4,4):  8,830,976\text{ Elements}$$
-Note however, that in practice there is a good chance, that a majority of the additional elements results in L2-cache hits.
-
-Next up is the transfer: Each thread corresponds to one particle within the blue nodes in Figure \ref{fig:g2p_transfer}.
-The threads run in unison over their relative neighbors within the support and collect the weighted contribution to this particle.
-In this process the particle's position needs to be loaded to the weighted corresponding contribution.
-The transfer needs to be repeated for each particle in the bin.
-Since the number of particles is variable, this loop is dynamic and will lead to branch divergence within a warp.
-One should rely on external mechanics like particle resampling to reduce branching.
-
-\algblockdefx{FORALLP}{ENDFAP}[1]%
-  {\textbf{for all }#1 \textbf{do in parallel}}%
-  {\textbf{end for}}
-\begin{algorithm}[htpb]
-  \caption{G2P-transfer}\label{alg:g2p}
-  \begin{algorithmic}[1]
-    \FORALLP{$blockNode \in grid$}
-    \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
-    \ForAll{$(node,local\_id) \in \text{assignedNodes}(blockNode, H\_FLAT)$}
-      \State $blockAndHalo[local\_id] \gets node$\Comment{global load}
-      \EndFor
-      \State \text{barrier()}
-      \ForAll{$particle \in \text{bin}(blockNode)$}
-      \State $sum \gets$ vec4(.0)
-      \ForAll{$neighbor\_id \in \text{support}(blockNode)$}\Comment{transfer}
-      \State $w_{ip} \gets weight(particle,neighbor\_id)$
-      \State $sum $ += $ blockAndHalo[neighbor\_id]*w_{ip}$
-	\EndFor
-	\State $particle \gets sum$ \Comment{global write}
-      \EndFor
-    \ENDFAP
-  \end{algorithmic}
-\end{algorithm}
-Alternatively, one could come up with a per warp load balancing scheme: One thread would then not directly correspond to one active node.
-They would need to act as another node in the same warp when their load is done.
-This will inevitably lead to shared memory bank conflicts and other difficulties.
-For the particle push transfers this would necessarily introduce atomic operations on shared memory again.
-
-In Appendices \ref{sec:shared_transfer} the accesses of one warp on shared memory banks is shown.
-The interpolation function is assumed to be cubic and the first block dimension is $x=4$ or $x=8$.
-The shared memory can thus be handled as concurrently as possible.
-
-As an extension to this process batching is introduced.
-Instead of conservatively loading and writing one particle at a time, each thread handles multiple particles together which it stores in register memory.
-Important to this process is to unroll the corresponding loops that load, write, as well as process these particles.
-Instead of needing to rerun the transfer for each particle, the number of runs now is the number of non-batched runs divided by the number of particles.
-This process adds to the amount of registers allocated and can thus also negatively impact performance.
-\subsubsection{Particle-to-grid-transfers}\label{sec:p2g}
-Straightforward to implement are various variations on global memory.
-Consider first binning was not introduced.
-A grid node would not know how to lookup the particles in its neighborhood.
-This would require the particle to find and write to the grid node in its support.
-Since any particle in the neighborhood of a grid node contributes to it, all writes have to be atomic.
-A typical approach is thus simply parallelizing for each particle and atomically write to the neighbor nodes in a loop.
-In the unsorted case this stalls the warps immensely.
-In \cite{Meyer2015} the static loop over the support is also parallelized resulting in higher throughput.
-The benefits of sorting in chapter \ref{sec:sort} apply to the P2G-transfers very significantly.
-The results of above get reversed and looping performs better.
-Stalling is much less of an issue.
-
-\begin{figure}[t]
-    \centering
-  \includesvg[svgpath =images/,width=\textwidth]{block_transfer_par}
-  \caption{P2G-transfer: The first run is shown.
-On the left the P2G-sync-transfer is shown that writes into the neighborhood.
-On the right the P2G-pull approach is shown that instead reads out of the neighborhood.}
-  \label{fig:p2g_transfer}
-\end{figure}
-
-\textbf{P2G-pull-transfer}: Sorting and binning however open up the possibility of shared memory implementations.
-The equivalent of the G2P-transfer for the P2G-transfer is pulling the relevant data from the neighborhood.
-The process is shown in Figure \ref{fig:p2g_transfer}.
-
-It is important to notice that the supports in the P2G-pull approach are reversed: Left becomes right, right becomes left.
-Other than that the transfer works in the same way as the G2P-transfer.
-Instead of reading from grid nodes, reading from particles is done.
-Consequently, one particle of each bin is loaded into shared memory.
-This typically puts a lot more strain onto shared memory.
-For a simple PIC-transfer (equation \ref{eq:lin_p2g})
-$$
-(m\boldsymbol{v})_i^n = \sum_p w_{i}(\boldsymbol{x}_p)m_p\boldsymbol{v}^n_p
-$$
-the position $\boldsymbol{x}_p$ and the velocity $\boldsymbol{v}_p$ need to be stored in shared memory for every node in the halo.
-Additionally the count $c_i$ and scan $s_i$ need to be available to load and access particles.
-For more complex transfers this will get even worse.
-
-By construction the material point method tries to avoid matrices on the grid due to storage cost.
-Conversely, the particles have matrices defined over them, e.g.
-$\boldsymbol{B}_p$,$\boldsymbol{F}_p$.
-In conclusion one should motivate approaches where grid nodes correspond to shared memory due to the shared memory size limitation.
-
-There is one implementation detail left out for the P2G-pull-transfer which is solved in chapter \ref{sec:active_blocks}.
-That is grid nodes do not know if there are any particles in the neighborhood left processing.
-\ref{sec:active_blocks} solves this by finding the maximum particle count in the block (with halo).
-Thus all nodes within the block need to stay active during transfers.
-
-It is possible to delegate this decision down from the block to its subgroups (respecting their respective haloes).
-Delegating this from the subgroup to the threads is possible as well.
-Each thread would calculate the summed up count in its neighborhood.
-But this hardly matters for compute-bound shaders as all other threads need to wait on the thread with the highest count in the subgroup anyway.
-
-\begin{algorithm}[t]
-  \caption{P2G-sync-transfer}\label{alg:p2g}
-  \begin{algorithmic}[1]
-    \FORALLP{$blockNode \in grid$}
-    \State \text{shared vec4 } $blockAndHalo$ [$H\_FLAT$]
-    \ForAll{$local\_id \in \text{assignedNodesID}(blockNode, H\_FLAT)$}
-      \State $blockAndHalo[local\_id] \gets $vec4(.0)
-      \EndFor
-      \State \text{barrier()}
-      \ForAll{$particle \in \text{bin}(blockNode)$} \Comment{global load} \label{alg:p2g_dyn_loop}
-      \ForAll{$neighbor\_id \in \text{support}(blockNode)$} \Comment{transfer}
-	\State $w_{ip} \gets weight(particle,neighbor\_id)$
-	\State $blockAndHalo[neighbor\_id] $ += $ particle*w_{ip}$
-	\State \text{barrier()} \Comment{uniform flow control needed!} \label{alg:p2g_uniform_barrier}
-	\EndFor
-      \EndFor
-      \ForAll{$(node,local\_id) \in \text{assignedNodes}(blockNode, H\_FLAT)$}
-      \State atomicAdd($node$,$blockAndHalo[local\_id]$) \Comment{global write}\label{alg:atomicAdd}
-      \EndFor
-    \ENDFAP
-  \end{algorithmic}
-\end{algorithm}
-
-In the following the two Particle-Push-Transfers are examined as they put less strain on shared memory:
-
-\textbf{P2G-atomic-transfer}: One can invert the process of pulling into pushing.
-Instead of reading from shared memory corresponding to particles, this means now writing to shared memory corresponding grid nodes.
-Accesses now need to be either atomic or explicitly scheduled if blocks are larger than a subgroup.
-
-As can be seen in Figure \ref{sec:p2g}, push approaches have the big advantage of only needing to load particles within the block region.
-The transfers then spread the particle's contribution to the grid nodes in the neighborhood.
-If in algorithm \ref{alg:p2g} line \ref{alg:p2g_uniform_barrier} the \texttt{barrier()} is left out, and \texttt{atomicAdd()}s on shared memory are in use, this process is straightforward to implement.
-One has to make sure to reset the shared memory as their contents are otherwise undefined.
-
-The write back to memory however now also needs to feature atomics.
-This process will be the same in the following synchronized approach.
-For a node that is part of any halo, \texttt{atomicAdd()}s are a strict requirement.
-Their results are only partially computed and need to be complemented by neighbor blocks.
-For such a node these are two writes for a side node, four for an edge node, and eight for a corner node.
-
-\textbf{P2G-sync-transfer}: Warps/Subgroups are inherently synchronized.
-No race conditions are present between their threads as they all access different memory positions (as can be seen in Figure \ref{fig:p2g_transfer}).
-The problem arises between different warps of the same thread group.
-In a synchronized approach this is solved by a \texttt{barrier()}.
-Care needs to be taken as the particle loop of line \ref{alg:p2g_dyn_loop} in algorithm \ref{alg:p2g} is dynamic.
-
-The OpenGL language states the following requirements for external synchronization:
-\textit{"\texttt{barrier()} can be called from flow-control, but it can only be called from uniform flow control.
-(...) In short, (...) every execution must hit the exact same set of \texttt{barrier()} calls in the exact same order."} \cite{KHRONOS:BARRIER}
-
-The dynamic loop has to become static within a thread group.
-The problem is solved in same fashion as in the P2G-pull method with chapter \ref{sec:active_blocks}.
-The maximum particle of each bin within the block without the halo is the minimum amount of times the loop needs to run across the threads to hit uniform flow control.
-
-Warps are still accelerated if all threads within that warp have lower count. They still however have to commit to wait at the same \texttt{barrier()}.
-
-In summary all transfers have the same optimal shared-memory access patterns as shown already by Appendices \ref{sec:shared_transfer}.
-Batching is again a valid option that is executed in the same manner by unrolling the corresponding loops.
-The block size as well as the level of batching can be chosen arbitrarily to optimize for the workload.
-
-\begin{figure}[!b]
-  \includesvg[svgpath =images/,width=\textwidth]{indirect_dispatch}
-  \caption{Pipeline to filter active blocks.
-Returned is the number of active blocks for the launch of a dispatch.
-A scan and an indexed stream compaction give back the filtered indices to access the active blocks in that dispatch.}
-  \label{fig:active_block}
-\end{figure}
-
-\subsubsection{Active Blocks \& Maximum Block Count:}\label{sec:active_blocks}
-The P2G-pull- and P2G-sync-transfer of chapter \ref{sec:transfers} required the maximum count of particles within a region.
-The active particle region is the region that needs to get reduced.
-Therefore, it is sufficient for the P2G-sync-transfer to reduce the counter within the block, left side of Figure \ref{fig:p2g_transfer}.
-The P2G-pull-transfer however also need to take the halo into account, right side of Figure \ref{fig:p2g_transfer}.
-Both are achievable with a single level of parallel reduction (in contrast to a pyramid).
-This is the \textit{per block bin count} from now on that uses as the functional input for the parallel reduction:
-\begin{enumerate}
-  \item \texttt{unary\_op(count) = count} .
-  \item \texttt{binary\_op(lc,rc) = max(lc,rc)}.
-\end{enumerate}
-
-For a P2G-sync-transfer one may just start with a parallelization level equal to the blocks employed in \ref{sec:transfers}.
-Then the shader can be narrowed down to do more sequential work to improve throughput.
-For the P2G-pull-transfer one may also do the first but additionally load all rest elements (by the halo).
-From there on out more sequential work might be favorable.
-
-If a per block indexing mode is employed, all blocks are directly behind each other in the buffer.
-Reduction then happens in the typical fashion.
-A global indexing needs to be permuted such that the indexing puts blocks first.
-
-The process of filtering active blocks can be easily appended.
-Figure \ref{fig:active_block} shows the process.
-An active block is one that has a non-zero per-block counter.
-The scan that will refer to the active block indices takes thus functional input:
-\begin{enumerate}
-  \item \texttt{unary\_op(count) = (count>0)?1:0}, which uses the ternary operator that increases count by one for any active block.
-  \item \texttt{binary\_op(lc,rc) = lc + rc}.
-\end{enumerate}
-
-Starting a dispatch for every block only the active blocks will now write their index at the scan's position.
-It results in an indexed stream compaction.
-This process is not coalesced.
-Coalescing will still be guarenteed for all nodes within the block for the P2G-transfers and G2P-transfer.
-The only uncoalesced access within these shaders is loading the maximum per block count from VRAM which there are few of in comparison.
-
-\clearpage
-\section{Evaluation}\label{sec:eval}
-An important part of any operation is to verify them.
-Chapter \ref{sec:verify} does that.
-Following up, chapter \ref{sec:perf} will use the metrics out of chapter \ref{sec:metrics} to evaluate the performance of the implementation.
-\subsection{Verifying Results}\label{sec:verify}
-\textbf{Parallel reduction}, \textbf{map}, and \textbf{scan} verify against \texttt{std::transform\_reduce} as it supplies the same functionality as a fully reduced parallel reduction.
-It is standard since C++17.
-These shaders are exact on integers.
-Floating point precision can be seen in table \ref{tab:parallel_red}.
-\begin{table}[!b]
-  \begin{tabular}{|l|r|r|r|}\hline
-    Shader       & \texttt{unary\_op} & Total abs. error & Total rel. error \\\hline
-    Map (float)   & x          & 0.0    & 0.0      \\\hline
-    Map (vec4,dvec4)   & length(x)  &0.0625 & 7.94e-08 \\\hline\hline
-    MapReduce (float)& x  &0.0183 & 3.51e-08 \\\hline
-    MapReduce (double)& x  &1.78e-08 & 3.42e-14 \\\hline
-    MapReduce (vec4)& length(x)  &212.813 & 2.71e-04 \\\hline
-    MapReduce (dvec4)& length(x)  &1.58e-05 & 2.02e-11 \\\hline
-  \end{tabular}
-  \caption{Error of various shaders on 1024*1024 random elements. Scalars are just between [0.0-1.0]. The vectors have a length between [0.0-1.0].}
-  \label{tab:parallel_red}
-\end{table}
-
-\textbf{Counting sort} is tested as a whole by giving back the bin keys of the particles. The bin keys are then tested on the CPU against \texttt{std::is\_sorted}, which is standard since C++11.
-
-The Eigen-library \cite{EIGEN:LIBRARY} is a well regarded linear algebra library.
-The \textbf{Singular value decomposition} is tested against \texttt{Eigen::JacobiSVD}.
-As a test one million random matrices are created by Eigen.
-Tests account for the 'Polar SVD'.
-There is not a single sign error.
-The average error between the singular values of Eigen and the implementation of chapter \ref{sec:svd} is 4.91782e-07.
-$det(\boldsymbol{U})$ and $det(\boldsymbol{V})$ should be as close as possible to 1.
-The errors are 3.55707e-07 and 3.26429e-07, respectively.
-
-The \textbf{MPM-Transfers} need to be mass and momentum conserving. The total mass on the grid $m_g^{tot}$ and the particles $m_p^{tot}$ has to be the same after any transfer:
-\begin{equation}
-  m_{p}^{tot}= \sum_p m_p =  \sum_i m_i = m_{g}^{tot}.
-\end{equation}
-Following this procedure, the same should be true for the total linear momentum $\boldsymbol{p}^{tot}$:
-\begin{equation}
-  \boldsymbol{p}_{p}^{tot}= \sum_p m_p \boldsymbol{v}_p=  \sum_i m_i \boldsymbol{v}_i = \boldsymbol{p}_{g}^{tot}
-\end{equation}
-and also for angular momentum $\boldsymbol{L}^{tot}$ (out of convention upper-case):
-\begin{equation}
-  \boldsymbol{L}_{p}^{tot}= \sum_p \boldsymbol{x}_p \times m_p \boldsymbol{v}_p =  \sum_i \boldsymbol{x}_i \times m_i \boldsymbol{v}_i = \boldsymbol{L}_{g}^{tot}.
-\end{equation}
-The test case spreads one million particles uniformly with random velocities on a $128\times128\times128$ grid, table \ref{tab:roundtrip}.
-A roundtrip consists of a P2G-transfer and a G2P-transfer for relevant variables.
-In this process the (random) velocities will get filtered.
-In PIC-Transfers this regionally averages out velocities and dissipates angular momentum $\boldsymbol{L}_p$.
-APIC-Transfers can conserve the angular momentum better as they define a local velocity field around the particle.
-Thus table \ref{tab:roundtrip} shows the effects on the total error after 1,000 roundtrips.
-
-The total errors vary little between the methods presented in chapter \ref{sec:transfers}.
-They are caused by floating-point inaccuracy.
-The order of execution on the GPU can vary.
-Thus the floating-point error will vary from execution to execution.
-\begin{table}[t]
-\centering
-  \begin{tabular}{|l|r|r|}\hline
-    Quantity& Abs. error  & Rel. error \\ \hline
-    \multicolumn{3}{|c|}{PIC-Transfer}\\\hline
-    $m^{tot}_i$ & 6.438     & 6.437e-06     \\\hline
-    $\boldsymbol{p}^{tot}_i$  & (0.882,0.977,0.018)&(1.417,1.722,0.003)e-04 \\\hline
-    $\boldsymbol{L}^{tot}_i$  & \textbf{(27.81,43.55,24.35)}&(0.024,0.036,0.273)\\\hline
-    \multicolumn{3}{|c|}{APIC-Transfer}\\\hline
-    $m^{tot}_i$ & 7.188     & 7.188e-06     \\\hline
-    $\boldsymbol{p}^{tot}_i$  & (1.230,0.990,0.855)&(1.975,1.745,1.371)e-04 \\\hline
-    $\boldsymbol{L}^{tot}_i$  & \textbf{(0.081,0.184,5.6e-04)}&(0.7043,1.510,0.063)e-04\\\hline
-   \end{tabular}
-   \caption{Error of 1,000 PIC/APIC-roundtrips of 1M uniformly placed particles with random velocity in a $128\times128\times128$ grid.
-Used this instance is a P2G-sync-transfer and G2P-transfer with batching = 4 and block size = (4,4,4).
-Particles are fixed.}
-  \label{tab:roundtrip}
-\end{table}
-\clearpage
-\subsection{Performance}\label{sec:perf}
-The given metrics in \ref{sec:metrics} give valuable insight into how well the GPU can utilize a given shader or program.
-The GPU was a black box for a long time in this regard.
-The metrics provided by NVIDIA are a first step to open up the GPUs inner workings.
-Theory about optimization for the architecture such as \cite{NVIDIA:BEST:PRACTICE} can be more readily applied.
-However for abstraction purposes theory as well as descriptions about the architecture might be deliberately vague.
-These metrics give the programmer more power of identifying the right optimizations for a given code.
-
-All metrics in the following are gathered from a NVIDIA GTX970.
-All timings are averaged over at least 1000 runs with OpenGL Timer Queries \cite{KHRONOS:TIMER_QUERY}.
-As more runs are done, the GPU stays very consistent on bringing the average runtime closer to the minimum recorded time.
-This indicates that the first few runs can take comparably longer.
-
-\textbf{SoA vs.
-AoS}: In chapter \ref{sec:soa_aos} is already hinted that throughput could be halved in an AoS-Layout.
-These effects can be hideous without knowledge of the architecture.
-Take for instance a map operation (or equivalently a copy operation).
-Figure \ref{fig:soa} and \ref{fig:aos} in chapter \ref{sec:soa_aos} already gave an overview of the procedure for \texttt{vec4}.
-Table \ref{tab:map_1} compares differences between an AoS- and a SoA-Layout with a buffer of two attributes.
-\begin{table}[htbp]
-  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
-    AoS         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
-    SoA         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\
-    \hline
-  \end{tabular}
-\caption{Metrics between different buffer layouts.
-Map operation with one instruction \texttt{float x = length(vel)} on $1024 \times 1024$ \texttt{vec4} buffer storing position and velocity (\texttt{vel}).}
-\label{tab:map_1}
-\end{table}
-
-Immediately apparent becomes that the SoA-Layout is 2.26 times faster.
-However, both memory subunits(VRAM,L2) have the same utilization as in an AoS-Layout.
-Though the big difference is in the memory access, the wrong access may be conversely spotted in the non-memory subunits.
-Since an AoS-Layout loads twice as many \texttt{vec4}, every non-memory related metric shrinks by half.
-An AoS-Layout however issues the same instructions.
-The SM should do equal work and be active for the same amount of cycles which is not the case.
-\begin{table}[htpb]
-  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
-    Layout &  $\Delta t_c (\mu s)$ & Speedup       &VRAM             & L2              & L2-Hit & Read Active \\\hline
-    AoS         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
-    SoA         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
-    \hline
-  \end{tabular}
-\caption{Metrics between different buffer layouts.
-Map operation with two instructions \texttt{float x = length(vel)}, \texttt{float y = length}\texttt{(pos)} on 1024 $\times$ 1024 \texttt{vec4} buffer storing position(\texttt{pos}) and velocity (\texttt{vel}).}
-\label{tab:map_2}
-\end{table}
-
-The case of using both attributes is also visualized in Figure \ref{fig:soa} and \ref{fig:aos}.
-One map instruction for each attribute is done within the same shader.
-Table \ref{tab:map_2} shows the results.
-Now AoS and SoA load the same amount of elements from VRAM.
-The SoA-Layout is still faster by 35 $\mu s$.
-
-In the SoA-Layout the second attribute lays consecutively in memory and is directly fetched.
-The SoA timing from table \ref{tab:map_1} is doubled.
-As expected the two operations are independent of each other and could thus be separated into two shaders with minimal performance impact; Assuming invocation cost is minimal.
-
-In the AoS-Layout the second attribute lays in the L2-Cache from fetching the first attribute.
-Thus using the second attribute will yield to cache hits.
-But going the extra way of checking for the element in the L2-cache requires an additional lookup.
-Thus in table \ref{tab:map_2} L2-SOL and L2 Hit rate go up.
-
-The view given was restricted to \texttt{vec4} as its the most commonly occurring type.
-For float/uint data structures the impacts of AoS vs.
-SoA can be even more dramatic.
-Instead of two \texttt{vec4}s eight scalars fit into a cache-line.
-In contrast only one \texttt{dvec4} (double-precision) fits into a cache-line.
-Performance of AoS and SoA in this case is equal.
-
-As a closure the reader is again referred to the subject of data-oriented design.
-While for non-random access SoA is (as shown) the more efficient layout, random access algorithms can improve performance by using an AoS or mixed setup.
-In a random access pattern the nearby elements of an SoA-Layout will be left unused in the cache-line.
-
-\textbf{Parallel Reduction \& Scan:} Chapter \ref{sec:red_scan} presents a shared memory implementation of parallel reduction.
-To further speed up the work load a sequential pattern is preferred to avoid bank conflicts.
-Table \ref{tab:map_reduce} compares this to an interleaved addressing without padding to illustrate the effects of bank conflicts on performance.
-Similarly the same can be done for a scan where however the interleaved addressing is preferred anyway.
-Table \ref{tab:scan} shows the comparison for a scan.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
-    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
-    Interleaved    & 305                   & -       &23.0\%          & \textbf{60.9}\%  & S. Scoreb.(17.2\%)\\\hline
-    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0\%)} \\\hline
-    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%         & L. Scoreb.(80.1\%)\\\hline
-    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%         & L. Scoreb.(84.4\%)\\\hline
-    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%         & L. Scoreb.(76.9\%)\\
-    \hline
-  \end{tabular}
-  \caption{Optimization of one parallel reduction dispatch on $1024\times1024$ \texttt{vec4} with \texttt{unary\_op(vec4 x)=length(x)} with block\_size = 1024.
-$\Delta t_c$ is in $\mu s$.
-Note that, the interleaved addressing in this example intentionally commits bank conflicts to show performance impacts.
-The number in parentheses states the number of additional added sequential elements.
-Reduced is by block\_size$\times$ sequential elements.
-Thus they do more work, while also performing better!}\label{tab:map_reduce}
-\end{table}
-
-The metrics between the parallel reduction and scan differ quite a lot when it comes to the resolve of bank-conflicts.
-Where the scans issue utilization goes drastically up from 34.2\% to 51.3\%, the same metric goes down for parallel reduction from 62.2\% to 33.0\%.
-SM throughput decreases as VRAM throughput increases although little for the scan.
-The top warp-stall reason for both is the barrier synchronization between thread groups.
-
-The warp-stall reason however showing the increased throughput due to respecting the shared memory architecture is the Short Scoreboard.
-The shaders in their original state have no reason to stall for the Short Scoreboard as any memory access is directly fetched.
-In the parallel reduction case the change is quite visible as this metric drops from 17.2\% to only 2.0\%.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
-    Interleaved       &  $\Delta t_c$         & Speedup &VRAM              & SM      & L2     & Sel. Stall Reas.\\\hline
-    Conflicts           & 748                   & -       &17.5\%            & 51.7\%  & 8.2\%  & S. Scoreb.(6.3\%)\\\hline
-    No confl.           & 571                   & 1.31x   &25.7\%            & 50.1\%  & 5.4\%  & \textbf{S. Scoreb.(2.6\%)}\\\hline
-    part. confl.        & 546                   & 1.36x   &23.5\%            & 50.8\%  & 6.0\%  & \textbf{S. Scoreb.(3.2\%)} \\\hline\hline
-    Seq. (2x)           & 323                   & 2.31x   &41.7\%            & \textbf{43.0}\%  & 17.2\% & S. Scoreb.(32.0\%)\\\hline
-    unrolled            & 258                   & 2.90x   &\textbf{53.8}\%   & 42.2\%  & \textbf{22.0}\% & S. Scoreb.(32.0\%)\\\hline\hline
-    Seq. (4x)           & 311                   & 2.40x   &56.8\%            & 22.7\%  & 37.9\% & L. Scoreb.(43.3\%)\\\hline
-    unrolled            & 297                   & 2.52x   &61.1\%            & 21.2\%  & 40.6\% & L. Scoreb.(49.9\%)\\
-    \hline
-  \end{tabular}
-  \caption{Optimization of one scan dispatch on $4 \times 1024\times1024$ integers with \texttt{binary\_op(x,y) := x+y}.
-$\Delta t_c$ in $\mu s$, block size is 1024.
-The number in parentheses states the number of additionally loaded sequential elements.
-Segment of each scan is block\_size$\times$ sequential elements.
-Higher reduction factor corresponds to more work being done.}\label{tab:scan}
-\end{table}
-The second optimization is giving these algorithms more sequential work.
-This will lighten the SM throughput and increase VRAM throughput as can be seen in both cases.
-Note, that each thread group in table \ref{tab:map_reduce} and \ref{tab:scan} does the same amount of parallel work; the sequential work is additionally being done by each thread.
-However, due to the additional sequential work the amount of thread groups go down by the same factor.
-As mentioned in \cite{NVIDIA:PARALLEL_REDUCTION} this saturates I/O-Latency and thus improves performance.
-
-Doing sequential work in parallel reduction adds almost no strain.
-Thus high amounts sequential work can be done by each thread.
-The SM work will shrink while the shader get more and more memory-bound.
-For a reduction factor of 256 first signs of performance decrease can be seen as only four groups are involved in the dispatch.
-The GTX970 has 13 SMs.
-Therefore, work is not parallelized enough on a block level anymore.
-The metric SM active\% will go rapidly down at this point.
-For four groups its 27.8\% which is roughly $\frac{4}{13}$.
-
-For the scan this is a little bit harder as elements need to be stored in registers.
-The effects can be seen in table \ref{tab:scan}.
-The segment each scan can handle is by far not as extensive as in parallel reduction.
-But the speedup achieved from a partially conflict free to an implementation with sequential work, is in contrast much higher.
-Since each thread requires the nearest by elements, L2-cache hits increase rapidly as does the L2 utilization.
-Thus the scoreboard stall reasons rise again.
-Unrolling the static loop which loads the sequential elements additionally hides memory latency.
-
-\textbf{Counting Sort:} The aim of sorting is coalescing as well as increased caching behavior for all following shaders.
-As a motivation the binning's similarity to the MPM-Operations might already be highlighted.
-Table \ref{table:bin} shows how a sorted particle position increases coalescing and L2-Cache hits.
-The VRAM-throughput increases to 75.3\% due to coalescing.
-Nearby particles hit the same bins and thus L2 rises up to 35.0\%.
-The 6.95x speed up shows how important the sorting as a preprocess is.
-\begin{table}[htbp]
-  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
-    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
-    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
-    Deep sort           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
-    \hline
- \end{tabular}
-\caption{Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
-\end{table}
-
-Table \ref{table:sort} shows the performance of reordering with index and deep sorting as is.
-The index sorting will have constant timing across cases.
-Their writing access will stay uncoalesced.
-Moreover shaders down the pipeline will also not benefit from coalescing.
-The binning in table \ref{table:bin} thus for instance always runs on random ordering.
-Index sorting is not affecting it.
-
-Again differences similarly to the binning can be seen when already working with a sorted state.
-Coalescing highly increases VRAM throughput.
-So much that a deep sort on eight \texttt{vec4}s is almost on equal footing with a random index sort.
-The speedup from an uncoalesced to a fully coalesced state is 5.68x.
-The full sort suffers for each (particle) variable from an uncoalesced access.
-However if sorting is not done the MPM-Transfers will then suffer from those.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r| }    \hline
-    Best Case    &  $\Delta t_c (\mu s)$&VRAM  & Top Stall Reas.    & Read   & Write  \\\hline
-    Index        &  1,516       &25.6\%& L. Scoreb.(78.7\%) & 16.30\%& 9.30\% \\\hline
-    Full(8 vec4) &  1,731       &\textbf{75.3}\%& L. Scoreb.(92.2\%) & 38.60\%& 34.60\%\\
-    \hline
-    \hline
-    Worst Case             &  $\Delta t_c(\mu s)$         &VRAM              & Top Stall Reas.    & Read   & Write\\\hline
-    Index                  &  1,516                &26.0\%            & L. Scoreb.(73.5\%) & 16.10\%& 9.40\%\\\hline
-    Full(8 vec4)           &  9,847                &24.0\%            & L. Scoreb.(82.2\%) & 9.90\% & 14.0\%\\
-    \hline
- \end{tabular}
-\caption{Order dependency of reordering of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.
-Best case is already sorted, worst case is unsorted and still fully random.}\label{table:sort}
-\end{table}
-
-Binning as well as reordering will rarely run fully sorted as mentioned in \ref{sec:sort}.
-Sorting works on the last sorted state.
-The CFL condition encourages limited reach.
-Most physical processes will move uniformly.
-These two properties in general will guarantee higher coalescing as well as L2-cache hit rates.
-It is however up to the dynamics of the problem how much time is spend sorting.
-\\
-
-The following metrics are gathered for PIC-Transfers of momentum as they form a reduced problem to based on only one \texttt{vec4} transfer.
-The MPM-Operations are evaluated with two easily reproducible examples.
-
-For the \textbf{G2P-transfers} the different datasets can be seen in table \ref{table:g2p_random} and \ref{table:g2p_uniform4}.
-The first one uses random particle positions and therefore is overly pessimistic for the material point method.
-Random particle positions as well as random velocities are a configuration that in practice will never occur.
-Even for a fundamental state of matter of gas this is unrealistic.
-The second one is optimistic as warp branch divergence is minimized.
-
-The global G2P-transfers does not reach desirable throughput in any of the GPU subunits.
-The top warp-stall reason is Long Scoreboard: Nearly all information needs to be reloaded uncoalesced from global memory.
-The data in \ref{table:g2p_uniform4} has an ordering that encourages data reuse.
-But is not aligned with the grid.
-However, the SM already runs on three times its throughput compared to the random data.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |r|}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         & L2    &SM     &Top Warp-Stall Reas. \\\hline
-    G2P-global             &   38,318             &-               &34.4\% & 9.6\% &L. Scoreb.(95.1\%)  \\\hline
-    sorted                 &    \textbf{2,991}    &12.81x          &\textbf{52.3\%}&\textbf{67.1}\% &No Instr.(14.8\%)   \\\hline
-    G2P-shared             &    9,665             & 3.96x          & 2.5\% &65.0\% &No Instr.(14.2\%)\\\hline
- \end{tabular}
- \caption{G2P-transfers of a $128\times 128\times128$ grid back to one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
-Block size is (8,4,4).}\label{table:g2p_random}
-\end{table}
-
-As elucidated in chapter \ref{sec:sort} deep sorting increases coalescing and caching behavior.
-The effects of sorting impact performance by a 12.81x speedup in the randomized case.
-The density of particles per bin is 0.5.
-Heavy use is made of the L2-Cache rising up to 52.3\% reading from many distinctive, successive memory positions.
-Meanwhile the density of four particles per bin in table \ref{table:g2p_uniform4} puts comparably low strain on the L2-Cache with 9.4\%.
-The runtime on random data is only 12.2\% slower than the uniform data.
-
-The G2P-shared method lies miles beyond that as its execution time is 9,665 $\mu s$.
-It can however get an edge on the more favored uniform data where 2,232 is the best recorded time.
-The G2P-shared method relies on shared memory instead of the L2-Cache to read data.
-Thus this method can be faster when little branch divergence between warps is the case.
-The random data however has highly divergent branching and is cause for the high execution times.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         & L2   &SM             & Top Warp-Stall Reas.\\\hline
-    G2P-global             &   10,316             &-               &34.4\%&\textbf{32.4}\%         & L. Scoreb.(79.9\%) \\\hline
-    sorted                 &    2,629		  &3.92x           &\textbf{9.4}\%&66.9\%          & No Instr.(14.1\%) \\\hline
-    G2P-shared             &    \textbf{2,232}             &4.62x           &4.1\% &\textbf{63.6}\%& No Instr.(22.7\%) \\\hline
- \end{tabular}
- \caption{G2P-transfers of a $128\times 128\times128$ grid back to one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).}\label{table:g2p_uniform4}
-\end{table}
-
-In NVIDIA-Nsight run times recorded for the G2P-transfers were comparably lower while making use of batching.
-However, OpenGL timer queries recorded little difference in trade-off for the added register pressure.
-Thus G2P-transfers should stay away of batching.
-
-For the sorted G2P-global as well as the G2P-shared method following statements are also of interest.
-The top-warp stall reason is waiting on the instruction fetches which is typical for compute-bound shaders with high SM usage.
-The SM Issue Utilization per Active Cycles is consistent around 65\%.
-This indicates that neither increasing occupancy nor reducing warp-stalls would increase performance.
-
-As a summary, relying on the comparably fast L2-Cache memory transactions over shared memory is consistent across the reduced examples for the sorted G2P-global method.
-The main benefit of the global solution is that warps do not suffer from branch divergence in that method.
-However, the G2P-shared method may still give an edge in performance.
-
-The \textbf{P2G-transfer} is tested on the same data in \ref{table:p2g_random} and \ref{table:p2g_uniform4}, respectively.
-The global atomic approaches suffer from low throughput across all units.
-Cause are uncoalesced accesses and low cache utilization.
-
-The top stall reason for all global operations is the short scoreboard.
-This is caused by the high number of atomic writes which are accumulated in L2-Cache.
-The sorting brings forth this problem more acutely as L2-Caches throughput rises to the top SOL with 44.0\% in the uniform case of table \ref{table:p2g_uniform4}.
-The random data example in table \ref{table:p2g_random} relies even more heavily on the L2-Cache with 65.8 SOL\% similarly to what has been seen in the G2P-transfer.
-
-Nevertheless, this already speeds up the random case by a factor of 11.88x as it followed no particular ordering. The sorting makes sure that the incoming data is preprocessed for optimal access.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM    \\\hline
-    global                 &  254,410             &-               &17.8\%&9.6\% &1.7\%  \\\hline
-    \cite{Meyer2015}       &  217,997             &1.17x           &21.7\%&3.7\% &13.2\% \\\hline
-    \cite{Meyer2015} sorted&   30,477		  &8.35x           &64.2\%&64.2\%&22.1\% \\\hline
-    global sorted          &   21,413		  &11.88x          &7.2\% &\textbf{65.8}\%&19.4\% \\\hline
-    \hline
-    P2G-pull               &   26,822             &9.46x           &1.6\% &2.3\% &35.2\%\\\hline
-    P2G-atomic*		   &   15,552             &16.36x          &3.7\% &7.1\% &\textbf{65.4}\%\\\hline
-    P2G-sync*              &   \textbf{12,801}    &19.87x          &3.6\% &7.3\% &\textbf{61.3}\%\\\hline
- \end{tabular}
- \caption{P2G-transfers of one million randomly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-Block size is (8,4,4).
-Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_random}
-\end{table}
-
-Using the \textbf{P2G-pull method} on the random case performs worse than the global method.
-It can significantly save time however in the uniform case.
-The P2G-pull-transfer is 4.31 times faster than the global P2G-transfer on the sorted uniform data.
-
-Top SOL\% is now the SM but it lacks throughput with 39.4\%.
-Its SM Issue Utilization per active cycle is 40\% which is below the desired threshold of 60\%.
-The top warp stall reason is instruction based.
-So as long as the algorithm does not change no increase can be achieved here.
-As explained in chapter \ref{sec:p2g} the pull method requires high amount of shared memory.
-The occupancy calculator as well as NVIDIA-Nsight report an occupancy of 16.0.
-This is $\frac{1}{4}$-th of the maximum occupancy of the GTX970.
-This is however already optimized for thread group size.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM     \\\hline
-    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
-    \cite{Meyer2015}       &   45,342             &0.98x           &25.1\%&42.4\%&11.7\% \\\hline
-    \cite{Meyer2015} sorted&   23,007		  &1.97x           &43.8\%&59.0\%&23.9\% \\\hline
-    global sorted          &   20,484		  &2.21x           &7.0 \%&\textbf{44.0}\%&16.1\% \\\hline
-    \hline
-    P2G-pull               &    4,747             &9.55x           &3.7\% &6.7\% &39.4\%\\\hline
-    P2G-atomic*		   &    3,148             &14.40x          &5.3\% &6.7\% &\textbf{65.0}\%\\\hline
-    P2G-sync*              &    \textbf{2,595}    &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
- \end{tabular}
- \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).
-Methods marked with a star(*) are executed with batching = 4.}\label{table:p2g_uniform4}
-\end{table}
-
-A closer look is taken of the P2G-atomic- and P2G-sync-transfers to illustrate the performance improvements of batching and filtering for active blocks in table \ref{table:p2g_optim}.
-
-The \textbf{P2G-atomic-Transfer} in its initial state uses one atomic transfer per particle on shared memory.
-Its effects are quite visible on the SM SOL\% with 82.9\% to 90.9\% in the call for active blocks.
-The top stall reason is Short Scoreboard which is indicative of the high number of stalls due to atomic writes on shared memory.
-
-Batching reduces the atomic writes.
-Particles get handled together.
-In succession one atomic write is done.
-This changes the top stall reason to the more favorable instruction fetches as the main stall reason with 20.8\%.
-Short scoreboard goes down to 11.6\%.
-As previously the SM was strained, throughput goes down to 65.0\%.
-The greatest time improvement can be seen when batching and indirect call lead to a combined speedup of 2.2x.
-This is much higher than the multiplicative speedup of 1.51x one could expect.
-
-The \textbf{P2G-sync-transfer} also may use batching and save time.
-The speedups are much smaller with 1.21x in the uniform case of table \ref{table:p2g_optim}.
-Batching in this case reduces the amount of thread group barriers needed for synchronization.
-The warp-stall reason barrier goes down from 10.4\% to 5.8\% in this process.
-The external synchronization is however much less invasive than atomic writes.
-The P2G-sync method without batching has roughly the same timing as the P2G-atomic-Transfer with batching of four particles.
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ & Speedup &SM    & Top Stall Reas.  & Occup.\\\hline
-    P2G-atomic		   &    6,933             & -       &82.9\%&S. Scoreb.(24.7\%)& 43.3\\\hline
-    indirect		   &    6,276             & 1.10x   &90.8\%&\textbf{S. Scoreb.(31.7\%)} & 43.4\\\hline
-    batching=4           &    4,929             & 1.41x   &65.1\%&No Instr.(20.8\%)& \textbf{34.9}\\\hline
-    both                   &    3,148             & \textbf{2.20x}   &\textbf{65.0}\%&No Instr.(23.2\%) & 35.4\\\hline
-    \hline
-    P2G-sync     &    3,161             & -       &72.4\%&No Instr.(17.7\%) & 43.3\\\hline
-    batching=4             &    2,595             & 1.21x   &67.0\%&No Instr.(23.7\%) & \textbf{35.4}\\\hline
- \end{tabular}
- \caption{P2G-transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).
-Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:p2g_optim}
-\end{table}
-
-Batching is a process that adds register pressure as can be seen by the reduced occupancy of 34.9-35.4.
-Using batching every new particle variable involved in the transfer takes the batching amount of that variables size more registers.
-
-The P2G-atomic-Transfer is very reliant on batching where the P2G-sync-transfer is not.
-This is further highlighted in table \ref{table:p2g_area}.
-With one particle per cell(area 47.6\%) batching does not help at all.
-The P2G-sync-transfer however can still work comparably fast to either the high density of four particles per cell losing about 7.2\% speed or atomic operations because batching is not available.
-This also means that the P2G-sync-transfer only scales slightly with the area covered which is due to the overhead of loading and writing more blocks.
-
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r |}    \hline
-    Method                 &  Area=47.6\% & Area=23.8\% &Area=11.9\%\\\hline
-    P2G-sync		   &  3,375$\mu s$& 3,266$\mu s$ & 3,129$\mu s$\\\hline
-    batching = 2           & -&2,942$\mu s$ & 2,745$\mu s$\\\hline
-    batching = 4           & -&- &2,622$\mu s$\\\hline
- \end{tabular}
- \caption{P2G-sync-transfer of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with one/two/four particles per cell corresponding to the area of the grid that is covered with 47.6\%/23.8\%/11.9\%.
-Block size is (4,4,4).}
- \label{table:p2g_area}
-\end{table}
-To underline the extensibility of all shaders table \ref{table:apic} implements the APIC-Transfers of chapter \ref{sec:apic}.
-The G2P-shared algorithm continues to lose ground on the global sorted G2P method as its execution time now only measures $300\mu s$ instead of the $400\mu s$ in table \ref{table:g2p_uniform4}.
-Its SM throughput also overtakes the global method with 81.8\%.
-\begin{table}[htpb]
-  \begin{tabular}{ | l | r | r | r | r |}    \hline
-    Method                 &  $\Delta t_c(\mu s)$ &SM    & Top Stall Reas.  & Occup.\\\hline
-    G2P(sorted)		   &    3,647             &74.9\%&No Instr.(14.2\%)& 57.7\\\hline
-    G2P-shared             &    3,384             &\textbf{81.8}\%&No Instr.(18.7\%) & 43.3\\\hline
-    \hline
-    P2G-sync               &    3,843             &68.9\%&No Instr.(24.9\%) & 43.3\\\hline
-    batching=2             &    3,634             &67.3\%&No Instr.(26.5\%) & 31.6\\\hline
- \end{tabular}
- \caption{APIC-Transfers of one million uniformly positioned particles with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid.
-They form a rotated (unsorted) cube with four particles per cell.
-Block size is (8,4,4).
-Shown are the effects of filtering active blocks(indirect) and batching.}\label{table:apic}
-\end{table}
-
-The P2G-atomic- and the even more favorable P2G-sync-transfer give great speedups over the global atomic methods.
-Even in a random case (table \ref{table:p2g_random}) they can outdo the global sorted method despite the high branch divergence.
-They however shine in the uniform case where speedups of 6.5x and 7.8x respectively can be achieved over the global sorted method.
-\clearpage
-\section{Conclusion \& Future Work}\label{sec:conclusion}
-The derivation of the material point method for a classic elastic material with an implicit midpoint scheme and the APIC scheme ensures that a wide coverage of the method is shown.
-Although the MPM consists of many mathematical tasks four operations on the GPU can be identified.
-Exclusive grid and particle operations are parallelized with a 1:1 correspondence to threads.
-The optional use of stream compaction reduces applicable particles/nodes where necessary.
-
-The other two operations are particle and grid transfers which are the main focus of this thesis.
-Different particle and grid transfers utilizing the hierarchical memory structure of the GPU are compared.
-Presented is a pipeline to efficiently execute these transfers on the GPU.
-All transfers do not rely on fixed bin sizes in any stages of the pipeline.
-Since these operations are executed multiple times per physical time step these operations are augmented with two preprocessing steps.
-The preprocessing steps need to be done once and are outweighed by the transfers.
-
-Deep sorting particles with counting sort in a double-buffered manner allows coalesced accesses and high cache hit rates on transfers.
-Although possible, this process does not rely on the CFL condition.
-The sorting further guarantees the independence of the ordering of the incoming data and may long-term benefit other operations relying on neighborhood information.
-
-The second preprocessing step is filtering the active blocks of grid and particle transfers with a stream compaction pipeline where no particles are present.
-It also supplies the P2G-pull- and P2G-sync-transfer with a maximum per block bin counter.
-
-While the shared memory architecture can give an edge in performance it produces branch divergence and may perform worse over a simpler algorithm relying on the L2-cache.
-The P2G-sync-transfer is shown to be an easily extendable and even in edge cases well performing operation.
-Particle push transfers could be shown to profit from batching as multiple particles are handled in unison.
-The sorted global G2P-transfer's consistency of utilizing the L2-Cache is a surprise further to be monitored.
-
-To show the weaknesses and strengths of the transfers tests on an overly pessimistic and an optimistic case are performed and examined.
-While this strongly furthers understanding it leaves real simulation performance comparisons to be desired.
-
-However, before this process can be concluded merge and split techniques need to be taken into consideration.
-They are designed to augment the numerics of the material point method but will inherently reduce branch divergence as particle numbers even out between cells in the same warp.
-As such they will stray less from the optimistic example.
-The extent to which they will stray from it dependends however on the dynamics of the problem.
-\clearpage
-\begin{appendices}
-\section{$D_p$ for Cubic Splines}\label{app:dp_proof}
-  \begin{code}
-\caption{$\boldsymbol{D}_p$ proof}
-  \begin{minted}{python}
-import numpy as np
-from sympy import *
-
-def round_expr(expr, num_digits):
-  return expr.xreplace(
-  {n : round(n, num_digits) for n in expr.atoms(Number)}
-  )
-
-# Limit a,b,c,x to interval [0,1] for simplyfing
-a, b, c = symbols('a b c', nonnegative=true)
-a = a/(1+a)
-b = b/(1+b)
-c = c/(1+c)
-
-x = symbols('x', nonnegative=true)
-x = x/(1+x)
-
-# Define cubic interpolation function
-def N1(x):
-  return 0.5*pow(abs(x),3)-pow(x,2)+2/3
-def N2(x):
-  return -1/6*pow(abs(x),3)+pow(x,2)-2*abs(x)+4/3
-def wip(i,x):
-  if(i==1 or i==2):
-    return N1(x)
-  else:
-    return N2(x)
-
-# Define the parametrized position in the grid
-def grid_points(x):
-  return np.array([-x-1,-x,1-x,2-x])
-
-alphas = grid_points(a)
-betas  = grid_points(b)
-gammas = grid_points(c)
-
-D_temp = np.array([[0,0,0],[0,0,0],[0,0,0]])
-
-# Loop over all grid nodes in the vicinity
-for i,ai in enumerate(alphas):
-  for j,bj in enumerate(betas):
-    for k,ck in enumerate(gammas):
-      # xi_xp is the distance from parametrized
-      # position to grid node [i,j,k]
-      xi_xp = np.array([ai,bj,ck])
-      # each outer product weighted by interpolation
-      # functions
-      this_outer = wip(i,ai)*wip(j,bj)*wip(k,ck)*
-		    np.outer(xi_xp,xi_xp)
-      # summed up over all grid nodes
-      D_temp = np.add(D_temp,this_outer)
-
-D = np.array([[0.0,0.0,0.0],
-	      [0.0,0.0,0.0],
-	      [0.0,0.0,0.0]])
-
-for i,D_row in enumerate(D_temp):
-  for j,D_ij in enumerate(D_row):
-    # simplify to cancel polynoms
-    # round_expr because of numerical cancellation
-    D[i][j] = round_expr(simplify(D_ij),14)
-    print(D)
-\end{minted}
-\end{code}
-  Prints out:
-  $$
-  \left[
-    \begin{array}{ccc}
-      0.33333333 &0.         &0.        \\
-      0.         &0.33333333 &0.	\\
-      0.         &0.         &0.33333333\\
-  \end{array}\right]= \frac{\boldsymbol{D}_p}{h^2}
-  $$
-\section{Block Scan}\label{sec:block_scan}
-
-\textbf{A Block Scan:} Assume a global scan is already computed on a n-dimensional uniform grid and now want to compute the local scan of an n-dimensional tile/block in that grid.
-This again forms a uniform grid.
-For simplicity assume two dimensions:
-
-\noindent Given the initial two-dimensional pre-scan array:
-\begin{equation}
-  \boldsymbol{G}=
-  \left[
-  \begin{array} {c c c c }
-    a_{00} & a_{01} &  ... & a_{0n}\\
-
-    a_{10} & a_{11} &  ... & a_{1n}\\
-    ...    & ...    &  ... & ...   \\
-    a_{m0} & a_{m1} &  ... & a_{mn}\\
-\end{array}
-\right].
-\end{equation}
-We can identify a block with $i,j,k,l\geq 0$, $i+k<m$ and $j+l<n$ as:
-\begin{equation}
-  \boldsymbol{L}=
-  \left[
-  \begin{array} {c c c c }
-    a_{ij} & a_{(i)(j+1)} &  ... & a_{(i)(n-l)}\\
-
-    a_{(i+1)(j)} & a_{(i+1)(j+1)} &  ... & a_{(i+1)(n-l)}\\
-    ...    & ...    &  ... & ...   \\
-    a_{(m-k)(j)} & a_{(m-k)(j+1)} &  ... & a_{(m-k)(n-l)}\\
-\end{array}
-\right].
-\end{equation}
-We use the $\Sigma$-Notation for sums although any associative binary operator would hold. The global scan is computed as all previous rows + the current row to the current element:
-\begin{equation}
-  S^G_{st} = \sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(t-1)} a_{sr}
-\end{equation}
-with $0 \leq s,t \leq m,n$.
-The local scan is computed similar:
-\begin{equation}
-  S^L_{bc} = \sum_{q,r=i,j}^{(b-1),(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br}.
-\end{equation}
-This however requires a full scan of $\boldsymbol{S}^L$, where $i,j \leq b,c \leq (m-k),(n-l)$.
-We can alternatively compute the local scan from the global scan with the following five steps:
-\begin{enumerate}
-\item Subtract the global scan's first element of every row-element and limit your view to indices $b,c$ then:
-  $$
-  P_{st} = S^G_{st} - \left(\sum_{q,r=0}^{s-1,n} a_{qr} + \sum_{r=0}^{(j-1)} a_{sr}\right)
-  $$
-\begin{equation}
-  \Rightarrow P_{bc} = \sum_{r=j}^{(c-1)} a_{br}
-\end{equation}
-\item Add the last value of $\boldsymbol{L}$ of the previous row (if it exists) to the current row.
-  \begin{equation}
-    \Rightarrow T_{bc} = a_{(b-1)(n-l)} + \sum_{r=j}^{(c-1)} a_{br}
-  \end{equation}
-\item Extract the last column $c =(n-l)$:
-  \begin{equation}
-    t_b = a_{(b-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{br}.
-  \end{equation}
-\item Compute the in dimensionality-by-one reduced exclusive scan (of the last column):
-  \begin{equation}
-    s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q-1)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right).
-\end{equation}
-\item Row-wise add back:
-  $$T_{bc} + s_{b} = \sum_{q=i}^{(b-1)}\left(a_{(q)(n-l)} + \sum_{r=j}^{(n-l-1)} a_{qr}\right) + \sum_{r=j}^{(c-1)} a_{br}$$
-  \begin{equation}
-    = \sum_{q=i}^{(b-1)}\sum_{r=j}^{(n-l)} a_{qr} + \sum_{r=j}^{(c-1)} a_{br} = S^L_{bc}.
-  \end{equation}
-\end{enumerate}
-For a simpler understanding of this process Figure \ref{fig:tile_scan} shows the process with the same numbering:
-\begin{figure}[htbp]
-  \includesvg[svgpath =images/,width=0.53\textwidth]{tile_scan}
-  \caption{A per block scan following the above enumeration.}
-  \label{fig:tile_scan}
-\end{figure}
-
-\clearpage
-\section{Shared memory accesses on MPM-Transfers}\label{sec:shared_transfer}
-\newcommand{\mc}[2]{\multicolumn{#1}{c}{#2}}
-\definecolor{Gray}{gray}{0.85}
-
-\newcolumntype{a}{>{\columncolor{Gray}}c}
-\newcolumntype{b}{>{\columncolor{white}}c}
-\begin{table}[htpb]
-  \begin{tabular}{ | l | a | a | a | a | l | l | l |}\hline
-    \rowcolor{white}
-    BlockAndHalo(vec4) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
-    Banks (1.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
-    Banks (2.) & 7 & 0 & 1 & 2 & 3 & 4 & 5 \\\hline
-    Banks (3.) & 6 & 7 & 0 & 1 & 2 & 3 & 4 \\\hline
-    Banks (4.) & 5 & 6 & 7 & 0 & 1 & 2 & 3 \\\hline
-    Banks (5.) & 4 & 5 & 6 & 7 & 0 & 1 & 2 \\\hline
-    Banks (6.) & 3 & 4 & 5 & 6 & 7 & 0 & 1 \\\hline
-    Banks (7.) & 2 & 3 & 4 & 5 & 6 & 7 & 0 \\\hline
-    Banks (8.) & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\\hline
-    \rowcolor{white}
-    Banks (9.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
-  \end{tabular}
-  \caption{First run of transfer: Bank accesses by warp with block size $x=4$.
-Bank accesses are colored grey.
-The halo ranges due to the support from $0-7$.
-Every bank gets accessed four times.
-It is easy to see that a window shift would lead to the same amount of bank accesses.}
-\end{table}
-\begin{table}[htpb]
-  \begin{tabular}{ | l | a | a | a | a | a | a | a |a | l | l | l |}\hline
-    \rowcolor{white}
-    BlockAndHalo(vec4) & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\\hline
-    Banks (1.) & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2  \\\hline
-    Banks (2.) & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5  \\\hline
-    Banks (3.) & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0  \\\hline
-    Banks (4.) & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3  \\\hline
-    \rowcolor{white}
-    Banks (5.) & 4 & 5 & 6 & 7 & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\\hline
-  \end{tabular}
-  \caption{First run of transfer: Bank accesses by warp with block size $x=8$.
-Bank accesses are colored grey.
-The halo ranges due to the support from $0-10$.
-Every bank gets accessed four times.
-It is easy to see that a window shift would lead to the same amount of bank accesses.}
-\end{table}
-\end{appendices}
-\clearpage
-\listoffigures
-\listoftables
-\lstlistoflistings
-\clearpage
+\end{frame}
 \selectlanguage{english}

### 2019-02-07 pres/pres.tex
video embedded

@@
 \documentclass{beamer}
+\usetheme{Malmoe}
+\usecolortheme{seahorse}
+\usepackage{csquotes}
 \usepackage{listings}
@@
 \usepackage{xcolor,colortbl}
+\usepackage{media9}
+\usepackage[font=Times,
+timeinterval=30,
+timeduration=20,
+timewarningfirst=75,
+timewarningsecond=90,
+colorwarningfirst=blue,
+colorwarningsecond=red]{tdclock}
+\setbeamertemplate{footline}{%
+    \hspace*{2ex}%
+    \scriptsize{%
+        \cronominutes~-~\tdhours\hspace{0.1ex}:\hspace*{0.5ex}\tdminutes%
+    }%
+    \hfill\insertframenumber/\inserttotalframenumber
+    \hspace*{2ex}%
+}
 
@@
 
+%Information to be included in the title page:
+
+\title{GPU Acceleration of the Material Point Method}
+
+\author % (optional, for multiple authors)
+{Fabian Meyer}
+
+\institute % (optional)
+{University of Koblenz}
+
+\date{Kolloquium Computergrafik, 14 February 2019}
+
 \begin{document}
 \begin{frame}
+\titlepage
+\end{frame}
+
+\section{Motivation on a GPU MPM Approach}
+
+\subsection{a}
+\begin{minipage}{.4\pdfpagewidth}
+\begin{frame}
+\includemedia[
+  width=160,
+  height=90,
+  activate=pageopen,
+  addresource=snowh264.mp4,
+  flashvars={source=snowh264.mp4&autoPlay=true&loop=true&repeat=always}
+  ]{\includegraphics{snow.jpg}}{VPlayer.swf}
+Videoi
+\end{minipage}
+\end{frame}
+
+\section{A Gentle Introduction to the MPM}
+
+\subsection{b}
+
+\begin{frame}
+\frametitle{hi}
   \begin{enumerate}
@@
 \end{frame}
+
+\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
+
+\subsection{c}
+
+\begin{frame}
+\frametitle{hi}
+  \begin{enumerate}
+    \item \cite{MPM:COURSE}
+  \end{enumerate}
+\end{frame}
+
+\section{Delving Deeper: Further Opportunities}
+
+\subsection{d}
+
+\begin{frame}
+\frametitle{hi}
+  \begin{enumerate}
+    \item \cite{MPM:COURSE}
+  \end{enumerate}
+\end{frame}
+
 \selectlanguage{english}

### 2019-02-10 pres/pres.tex
motiv., begin math

--- 2019-02-07 pres/pres.tex
@@
 \documentclass{beamer}
+\usetheme{Malmoe}
+\usecolortheme{seahorse}
+
+\usepackage{textpos}
+\usepackage{csquotes}
 \usepackage{listings}
@@
 \usepackage{xcolor,colortbl}
+\usepackage{media9}
 
+\let\oldcite=\cite
+\renewcommand{\cite}[1]{\textcolor[rgb]{.55,.55,.89}{\oldcite{#1}}}
+
+\setbeamerfont{frametitle}{size=\small}
 \addbibresource{../mscthesis.bib}
+\addbibresource{pres.bib}
 \captionsetup{justification=raggedright,singlelinecheck=false}
@@
 
+%Information to be included in the title page:
+
+\title{GPU Acceleration of the Material Point Method}
+
+\author % (optional, for multiple authors)
+{Fabian Meyer}
+
+\institute % (optional)
+{University of Koblenz}
+
+\date{Kolloquium Computergrafik, 14 February 2019}
+
 \begin{document}
 \begin{frame}
+\titlepage
+\end{frame}
+
+\section{Motivation on a GPU MPM Approach}
+
+\subsection{A Brief MPM Overview: Do You Want to Build a Snowman?}
+
+\begin{frame}
+
+\frametitle{A Brief MPM Overview: Do You Want to Build a Snowman?}
+A short historical summary of MPM:
+\vfill
+\begin{minipage}{0.43\paperwidth}
+  \begin{itemize}
+    \item Belongs to family of particle-in-cell(PIC) techniques \cite{evans1957particle}.
+    \item Initial application to solids \cite{sulsky1995application} $\rightarrow$ MPM
+    \item From research to production in \textit{Disney's} animation film \textit{Frozen} \cite{MPM:SNOW}.
+    \item Avalanche research \cite{MPM:AVALANCHE}
+  \end{itemize}
+\end{minipage}
+\begin{minipage}{0.35\paperwidth}
+\includemedia[
+  width=144,
+  height=81,
+  activate=pageopen,
+  addresource=media/snowh264.mp4,
+  flashvars={source=media/snowh264.mp4&autoPlay=true&loop=true&repeat=always}
+  ]{\includegraphics{media/snow.jpg}}{VPlayer.swf}
+  \tiny{Video result of my bachelor thesis on the simulation of snow \cite{Meyer2015}.}
+\end{minipage}
+\end{frame}
+
+\begin{frame}
+PIC ideas:
+\vfill
+\begin{minipage}{0.45\paperwidth}
+\begin{itemize}
+    \item Combine Lagrangian particles \& Eulerian grid    \item Particles store all information
+\end{itemize}
+\vfill
+Typical PIC/MPM roundtrip:
+\vfill
+\begin{enumerate}
+  \item Particle-to-grid(P2G) transfer to an unmoving grid
+    \item Solve discretized governing equations on grid
+    \item Grid-to-particle(G2P) transfer back to particles \& move them
+    \item[$\Rightarrow$] \textbf{meshfree, non-empirical}
+\end{enumerate}
+\end{minipage}
+\begin{minipage}{0.35\paperwidth}
+	\includesvg[width=0.35\paperwidth]{media/pic.svg}
+	\tiny{Transfers: Interpolation functions are defined over grid nodes.}
+\end{minipage}
+\end{frame}
+\subsection{GPGPU for performance enthusiasts}
+\begin{frame}
+  \frametitle{GPGPU for performance enthusiasts}
+  Why would('nt) you?
+  \vfill
+  \begin{columns}[T]
+    \begin{column}{0.45\textwidth}
+      \textbf{Drawbacks:}
+    \begin{itemize}
+      \item Interactivity much easier on CPU, but slow PCI-Bus communication
+      \item Code is mostly written against GPU architecture
+      \item A lot of strain on the programmer
+    \end{itemize}
+    \end{column}
+    \begin{column}{0.45\textwidth}
+      \textbf{Benefits:}
+    \begin{itemize}
+      \item Data is already on the GPU for rendering
+      \item \textbf{Higher parallelization acceleration}
+    \end{itemize}
+    \end{column}
+  \end{columns}
+\end{frame}
+\section{A Gentle Introduction to the MPM}
+
+\subsection{Governing Equations: Conservation of Mass \& Momentum}
+
+\begin{frame}
+  \frametitle{Governing Equations: Conservation of Mass \& Momentum}
+  \textbf{Conservation of mass}, continuum assumption holds.\\
+  Lagrangian (moving with a particle ${}_0\boldsymbol{x}$):
+  \begin{equation}
+    {^t_0J}\rho(_0\boldsymbol{x},t) = {}\rho(_0\boldsymbol{x},0).
+  \end{equation}
+Eulerian (outside observer ${}_t\boldsymbol{x}$):
+\begin{equation}
+  \frac{\partial}{\partial t}\rho(_t\boldsymbol{x},t) = -  \vec{\nabla} \cdot (\rho(_t\boldsymbol{x},t) \boldsymbol{v}(_t\boldsymbol{x},t)).
+\end{equation}
+Lagrangian and Eulerian view measure differently but give same results. Equations are given in the strong form! \cite{MPM:COURSE}\cite{MIT:CONTINUUM_MECHANICS}
+\end{frame}
+
+\begin{frame}
+  \textbf{Conservation of momentum}:\\
+  Lagrangian (moving with a particle ${}_0\boldsymbol{x}$):
+  \begin{equation}
+  \rho(_0\boldsymbol{x},0)\boldsymbol{a}(_0\boldsymbol{x},t)
+= \vec{\nabla} \cdot \boldsymbol{P}(_0\boldsymbol{x},t) + {\boldsymbol{f}} ^ {\text{body}}(_0\boldsymbol{x},t) {^t_0J}.
+  \end{equation}
+Eulerian (outside observer ${}_t\boldsymbol{x}$):
+\begin{equation}
+  \rho(_t\boldsymbol{x},t) \boldsymbol{a}(_t\boldsymbol{x},t) = \vec{\nabla} \cdot \boldsymbol{\sigma}(_t\boldsymbol{x},t) + \boldsymbol{f} ^ {\text{body}}(_t\boldsymbol{x},t)
+\end{equation}
+Solving this equation will tell us how the velocity fields $\boldsymbol{v}(_t\boldsymbol{x}),\boldsymbol{v}(_0\boldsymbol{x})$ change on the whole domain due to acceleration $\boldsymbol{a}$.
+This is important to advect particles accounting for all forces.
+\cite{MPM:COURSE}\cite{MIT:CONTINUUM_MECHANICS}
+\end{frame}
+
+\subsection{The Pretty Strong but Mathematically Weak Formulation}
+\begin{frame}
+\frametitle{The Pretty Strong but Mathematically Weak Formulation}
+\begin{align}
+  \int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot  \Bigl[(_0\rho_0)(_0\boldsymbol{a}) - & {}_t\boldsymbol{f}^{\text{body}}\Bigr] d _0\boldsymbol{x}= \nonumber \\
+										     &  \int _ { \partial \Omega ^ { t^n } } _t\boldsymbol{q} \cdot \boldsymbol{\sigma} d_t\boldsymbol{A} - \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}
+\end{align}
+\end{frame}
+\section{A MPM Guide on GPGPU Pitfalls and Optimizations}
+
+\subsection{c}
+
+\begin{frame}
+\frametitle{hi}
   \begin{enumerate}
@@
 \end{frame}
+
+\section{Delving Deeper: Further Opportunities}
+
+\subsection{d}
+
+\begin{frame}
+\frametitle{hi}
+  \begin{enumerate}
+    \item \cite{MPM:COURSE}
+  \end{enumerate}
+\end{frame}
+
 \selectlanguage{english}

### 2019-02-11 pres/pres.tex
more math

@@
 \usecolortheme{seahorse}
+\usepackage{pgfplots}
+\usepackage{tikz}
+\usetikzlibrary{datavisualization}
+\usetikzlibrary{datavisualization.formats.functions}
 
@@
 \frametitle{The Pretty Strong but Mathematically Weak Formulation}
+
+\textbf{Weak Formulation} (or Principle of Virtual Work):\\
+Dot product equations with arbitrarily 'test functions' $\boldsymbol{q}$ and apply divergence theorem:
 \begin{align}
-  \int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot  \Bigl[(_0\rho_0)(_0\boldsymbol{a}) - & {}_t\boldsymbol{f}^{\text{body}}\Bigr] d _0\boldsymbol{x}= \nonumber \\
-										     &  \int _ { \partial \Omega ^ { t^n } } _t\boldsymbol{q} \cdot \boldsymbol{\sigma} d_t\boldsymbol{A} - \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}
+\int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot  \Bigl[(_0\rho_0)(_0\boldsymbol{a}) - & {}_0\boldsymbol{f}^{\text{body}}{}_0^tJ\Bigr] d _0\boldsymbol{x}= \nonumber \\
+										     &  \int _ { \partial \Omega ^ { t^n } } _t\boldsymbol{q} \cdot \boldsymbol{\sigma} d_t\boldsymbol{A} - \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
 \end{align}
+A strong solution is also a solution to the weak formulation. Leave out body forces(like gravity) and boundary condition (e.g. collisions) for now:
+\begin{equation}
+  \int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot (_0\rho_0)(_0\boldsymbol{a}) d _0\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
+\end{equation}
+\end{frame}
+\begin{frame}[fragile]
+\textbf{Weak Derivative}:
+\vfill
+\begin{minipage}{0.49\textwidth}
+  $y=|t|$ has weak derivative:
+  \\\\
+$ v = \left\{ \begin{array} { l l }
+{ - 1 , } & { \text { if } t < 0 } \\
+{ c , } & { \text { if } t = 0 } \\
+{ 1 , } & { \text { if } t > 0 }
+\end{array} \right.$
+\\
+
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{tikzpicture}
+\datavisualization [school book axes,
+		    visualize as line/.list={fx},
+		    fx={style={color=blue}}
+]
+
+data [set=fx,format=function] {
+      var x : interval [-1.5:1.5] samples 5;
+      func y = abs(\value x);
+      };
+\end{tikzpicture}
+\end{minipage}
+
+\vfill
+\begin{minipage}{0.49\textwidth}
+  Heaviside step function has no weak derivate: \\\\
+  $y= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+
+\begin{tikzpicture}
+
+  \datavisualization [school book axes,
+		      visualize as line/.list={fx},
+		      fx={style={color=blue}}]
+data[set=fx] {
+  x, y
+  -1.5, 0
+  0, 0
+  0, 1
+  1.5, 1
+};
+\end{tikzpicture}
+\end{minipage}
+\vfill
+Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
+\end{frame}
+\subsection{Discretization of Space and Time}
+\begin{frame}
+\frametitle{Discretization of Space and Time}
+\textbf{Time discretization} with implicit midpoint scheme:
+  \begin{equation}
+    \frac { y ^ { n + 1 } - y ^ { n } } { \Delta t } = f ^ { n + \frac{1}{2} } = f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)
+  \end{equation}
+\begin{itemize}
+  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
+  \item \textbf{midpoint} as it conserves gov. equations
+\end{itemize}
+
+  \begin{equation}
+    \Rightarrow \int _ { \Omega ^ { t } } _t\boldsymbol{q} \cdot {}_t\rho(_t\boldsymbol{v}^{n+1}-{}_t\boldsymbol{v}^{n}) d _t\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}^{n+\frac{1}{2}}  d_t\boldsymbol{x}.
+  \end{equation}
+\end{frame}
+
+\begin{frame}[fragile]
+\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$ with limited support. Here dyadic products
+\begin{equation}
+  w_i(\boldsymbol{x}) =w(\boldsymbol{x}-\boldsymbol{x}_i) = w(\frac{1}{h}(x-x_i))w(\frac{1}{h}(y-y_i)w(\frac{1}{h}(z-z_i))
+\end{equation}
+of cubic b-splines suffice:
+\begin{equation}
+w ( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array}.
+\right
+\end{equation}
+\end{frame}
+\begin{frame}
+
+\begin{tikzpicture}[domain=0:1]
+\begin{axis}
+\addplot {x*x};
+\end{axis}
+\end{tikzpicture}
+
+\end{frame}
+\begin{frame}
+Thus set in:
+$$
+{}_t\boldsymbol{q}(\boldsymbol{x},t^n) = \sum_i \boldsymbol{e}_iw_i(\boldsymbol{x}),
+{}_t\boldsymbol{v}^{n(+1)}(\boldsymbol{x}) = \sum_j \boldsymbol{v}^{n(+1)}_jw_j(\boldsymbol{x}).
+$$
+Combine it with Numerical Integration where the particles function as quadrature points:
 \end{frame}
 \section{A MPM Guide on GPGPU Pitfalls and Optimizations}
-
-\subsection{c}
+\subsection{d}
 

### 2019-02-11 pres/pres.tex
impl. begin

@@
 \usepackage{wrapfig}
-\usepackage[english,ngerman]{babel}
+\usepackage[english]{babel}
 \usepackage[backend=biber,style=alphabetic]{biblatex}
@@
 \captionsetup{justification=raggedright,singlelinecheck=false}
-\DeclareMathOperator*{\argmin}{arg min}
+
 \makeatletter
-\def\ext@algorithm{lol}% algorithm captions will be written to the .lol file
-% share the list making commands and redefine the heading
 \AtBeginDocument{%
@@
 \makeatother
+
 \lstdefinelanguage{GLSL}
@@
 \begin{tikzpicture}
+  \small
 \datavisualization [school book axes,
+		    y axis = {attribute = y, label},
 		    visualize as line/.list={fx},
@@
   Heaviside step function has no weak derivate: \\\\
-  $y= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
+  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
 \end{minipage}
 \begin{minipage}{0.4\textwidth}
-
 \begin{tikzpicture}
-
-  \datavisualization [school book axes,
-		      visualize as line/.list={fx},
-		      fx={style={color=blue}}]
-data[set=fx] {
-  x, y
-  -1.5, 0
-  0, 0
-  0, 1
-  1.5, 1
-};
+  \small
+  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
+                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
+                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
+            \end{axis}
 \end{tikzpicture}
@@
 \end{equation}
-\end{frame}
-\begin{frame}
+\begin{tikzpicture}[
+  declare function={
+    func(\x)= and(abs(\x) >= 0 , abs(\x)<1) * (pow(abs(\x),3)/2 - pow(abs(\x),2) +2/3)  +
+	      and(abs(\x) >= 1 , abs(\x)<2) * (1/6* pow((2-abs(\x)),3))
+   ;
+  }
+]
+\begin{axis}[
+  width=10cm,height=4cm,
+  axis x line=middle, axis y line=middle,
+  ymin=-0.1, ymax=1, ytick={0,1}, ylabel=$w$,
+  xmin=-2.5, xmax=2.5, xtick={-2,...,2}, xlabel=$x$,
+  domain=-2.5:2.5,samples=101, % added
+]
 
-\begin{tikzpicture}[domain=0:1]
-\begin{axis}
-\addplot {x*x};
+\addplot [blue,thick] {func(x)};
 \end{axis}
 \end{tikzpicture}
-
 \end{frame}
@@
 $$
-Combine it with Numerical Integration where the particles function as quadrature points:
+Combine it with numerical integration where the particles function as quadrature points \cite{Steffen}:
+
+\begin{minipage}{0.6\textwidth}
+
+$$
+g _ { i } = \int _ { \Omega } g ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \boldsymbol{x}
+$$
+\begin{equation}
+  \approx \sum _ { p } g _ { p }  w_i(\boldsymbol{x}_p) V _ { p }.
+\end{equation}
+due to integration by midpoint rule:
+\begin{equation}
+\int _ { \Omega } f ( x ) d x \approx \sum _ { i = 1 } ^ { N } f \left( x _ { i } \right) h _ { i }.
+\end{equation}
+
+\end{minipage}
+\begin{minipage}{0.3\textwidth}
+  \begin{tikzpicture}[/pgf/declare function={f=-.1*(x)*(x-8);}]
+\pgfplotsset{
+    integral segments/.code={\pgfmathsetmacro\integralsegments{#1}},
+    integral segments=3,
+    integral/.style args={#1:#2}{
+        ybar interval,
+        domain=#1+((#2-#1)/\integralsegments)/2:#2+((#2-#1)/\integralsegments)/2,
+        samples=\integralsegments+1,
+        x filter/.code=\pgfmathparse{\pgfmathresult-((#2-#1)/\integralsegments)/2}
+    }
+}
+\begin{axis}[
+    ylabel=$f$,
+    width =5cm,height=5cm,
+    domain=0:6,
+    samples=100,
+    axis lines=middle
+]
+\addplot [thick] {f};
+\addplot [
+    red,
+    integral=0:6
+] {f};
+\end{axis}
+\end{tikzpicture}
+\end{minipage}
+\end{frame}
+  \subsection{Velocity Fields: APIC-Transfers}
+\begin{frame}
+  \frametitle{Velocity Fields: APIC-Transfers}
+  \begin{minipage}{0.64\textwidth}
+  PIC-transfer
+  \begin{enumerate}
+    \item $(m\boldsymbol{v})_i^n = \sum_p w_{ip}^nm_p\boldsymbol{v}^n_p$
+    \item $\boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}$
+    \item $\boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w^n_{ip}\boldsymbol{v}_i^{n+1}$
+  \end{enumerate}
+\vspace*{15}
+APIC-transfers add a local velocity field $\boldsymbol{C}^n_p$ around $\boldsymbol{v}^n_p$:
+\\\vspace*{-10}\\
+$ ( m \boldsymbol{v} ) _ { i } ^ { n } = \sum _ { p } w _ { i p } ^ { n } m _ { p } \left( \boldsymbol{v} _ { p } ^ { n } + \boldsymbol{C} _ { p } ^ { n } ( \boldsymbol{x} _ { i } ^ { n } - \boldsymbol{x} _ { p } ^ { n }))$
+\end{minipage}
+\begin{minipage}{0.30\textwidth}
+\begin{minipage}{1.0\textwidth}
+  \small
+  \includesvg[width=1.0\textwidth]{media/pic.svg}
+  P2G-transfer $\sum_p$
+\\
+\end{minipage}
+\begin{minipage}{1.0\textwidth}
+  \small
+  \includesvg[width=1.0\textwidth]{media/g2p.svg}
+  G2P-transfer $\sum_i$
+\end{minipage}
+\end{minipage}
 \end{frame}
 \section{A MPM Guide on GPGPU Pitfalls and Optimizations}
-\subsection{d}
+\subsection{Layout of the data: SoA vs. AoS}
+\begin{frame}[fragile]
+\frametitle{Layout of the data: SoA vs. AoS}
 
+\begin{minipage}{0.31\textwidth}
+  \tiny
+\begin{minipage}{0.95\textwidth}
+  SoA-Layout
+\begin{lstlisting}[language={GLSL},style={GL},basicstyle=\tiny]
+struct Container{
+  vec4 positions[n];
+  vec4 velocities[n];
+} Particles;
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.95\textwidth}
+  AoS-Layout
+\begin{lstlisting}[language={GLSL},style={GL},basicstyle=\tiny]
+struct Particle {
+  vec4 position;
+  vec4 velocity;
+} Particles[n];
+\end{lstlisting}
+\end{minipage}
+
+\end{minipage}
+\begin{minipage}{0.64\textwidth}
+  \tiny
+  \includesvg[svgpath =../images/,width=\textwidth]{soa}
+  Structures of arrays(SoA)
+  \\
+  \includesvg[svgpath =../images/,width=\textwidth]{aos}
+  Arrays of structures(AoS)
+
+\end{minipage}
+\end{frame}
 \begin{frame}
-\frametitle{hi}
-  \begin{enumerate}
-    \item \cite{MPM:COURSE}
-  \end{enumerate}
+\begin{table}[htbp]
+  \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
+    Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
+    AoS         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
+    SoA         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\
+    \hline
+\end{tabular}
+\end{table}
+
   \end{frame}
-
 \section{Delving Deeper: Further Opportunities}

### 2019-02-12 pres/pres.tex
layouting

@@
 \begin{minipage}{0.31\textwidth}
+\begin{minipage}{0.95\textwidth}
+Coalescing:
+\vspace*{23}
+\end{minipage}
 \tiny
@@
 \end{lstlisting}
+\vspace*{29}
 \end{minipage}
@@
 \end{minipage}
-
 \end{minipage}
@@
 \begin{frame}
-\begin{table}[htbp]
+Nvidia Nsight\cite{NVIDIA:METRICS} now offers metrics to identify bottlenecks:
+\vspace{-5}
+\begin{table}
+  \footnotesize
+  \begin{tabular}{ | l | l |}    \hline
+    \textbf{Metric}   & \textbf{Description} \\ \hline
+    VRAM SOL\%        & memory througput w.r.t. to hardware limit\\\hline
+    SM SOL\%          & instruction throughput\\\hline
+    L2 SOL\%          & L2-cache throughput\\\hline
+    Tex SOL\%         & L1-cache throughput\\\hline
+    SM Issue Util.\%  & amount of cycles an instr. was issued\\\hline
+  \end{tabular}
+\end{table}
+\\
+\vspace{5}
+A simple map or copy shader with one instruction shows SoA vs. AoS differences:
+\vspace{-15}
+\begin{table}
+  \footnotesize
   \begin{tabular}{ | l | l | l | l | l | l | l |}    \hline
@@
 \end{table}
-
+$\Rightarrow$ SoA increases coalescing for non-random access.
+\end{frame}
+\subsection{Parallel Reduction \& Scan}
+\begin{frame}
+  \frametitle{Parallel Reduction \& Scan}
 \end{frame}

### 2019-02-12 pres/pres.tex
red., scan

@@
 \vspace{5}
-A simple map or copy shader with one instruction shows SoA vs. AoS differences:
+A simple map(\texttt{y=length(x)}) shader on 1024$\times$1024 Elements SoA vs. AoS differences:
 \vspace{-15}
@@
     Layout &  $\Delta t_c (\mu s)$	& Speedup &VRAM     & SM     & L2     & SM Issue Util. \\\hline
-    AoS         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
-    SoA         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\
+    AoS(1 instr.)         & 243                   & -       &77.7\%   & 7.3\%  & 30.3\% & 6.8\% \\\hline
+    SoA(1 instr.)         & \textbf{120}                   & 2.26x   &75.4\%   & \textbf{14.3}\% & 29.4\% & \textbf{14.0\%} \\\hline
+    AoS(2 instr.)         & 275                   & -        &61.3\%           & \textbf{41.8}\% & \textbf{53.8}\% & 48.9\% \\\hline
+    SoA(2 instr.)         & 240                   & 1.16x    &75.4\%  & 29.4\%          & 20.0\% & 62.3\% \\
     \hline
@@
 \frametitle{Parallel Reduction \& Scan}
+Assuming an associative \texttt{binary\_op(x,y)}:= $x \circ y$ , a neutral element $e$ of the \texttt{binary\_op}, and an array of values $\left[a_0,a_1, ... ,a_n\right]$.\\\vspace{10}
+\begin{itemize}
+  \item \textbf{Parallel reduction} computes the scalar:
+\begin{equation}
+  r = a_0 \circ a_1 \circ ... \circ a_n.
+\end{equation}
+\item (Exclusive) \textbf{scan} computes the array:
+\begin{equation}
+  \left[e , a_0 , \left(a_0 \circ a_1\right), \left(a_0 \circ a_1 \circ a_2\right), \ ... \ ,\left(a_0\circ a_1 \circ a_2 \circ ...\circ a_{n-1}\right)\right].
+\end{equation}
+\end{itemize}
+\\\vspace{5}
+Here, only shared memory approaches without (NVIDIA exclusive) warp shuffle operations.
+\end{frame}
+\begin{frame}
+\textbf{Shared Memory Bank Conflicts}:
+\\\vspace{5}
+\begin{minipage}{0.94\textwidth}
+\begin{minipage}{0.5\textwidth}
+  \tiny
+  \includesvg[svgpath =../images/,width=1.0\textwidth]{sequential}
+  Sequential Addressing
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \tiny
+  \includesvg[svgpath =../images/,width=1.0\textwidth]{interleaved}
+  Interleaved Addressing
+\end{minipage}
+\end{minipage}
+\\\vspace{10}
+Interleaved Addressing causes bank conflicts $\Rightarrow$ padding needed.
+\textbf{Sequential work}:
 \end{frame}

### 2019-02-13 pres/pres.tex
red,sorting

@@
 \begin{itemize}
-  \item \textbf{Parallel reduction} computes the scalar:
+  \item \textbf{Parallel reduction} computes the value:
 \begin{equation}
@@
 \textbf{Shared Memory Bank Conflicts}:
-\\\vspace{5}
+\\\vspace{10}
 \begin{minipage}{0.94\textwidth}
@@
 \\\vspace{10}
-Interleaved Addressing causes bank conflicts $\Rightarrow$ padding needed.
-\textbf{Sequential work}:
+\\
+Interleaved addressing causes bank conflicts (Short Scoreboard activity) $\Rightarrow$ padding needed.
+\vspace{-12}
+\begin{table}
+    \footnotesize
+  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
+    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
+    Interl. no padd.    & 305                   & -       &23.0\%          & \textbf{60.9}\%  & S. Scoreb.(17.2\%)\\\hline
+    Sequential     & 141                   & 2.16x   &\textbf{49.8}\%   & 37.1\%         & \textbf{S. Scoreb.(2.0\%)} \\\hline
+  \end{tabular}
+    \caption{\footnotesize Parallel reduction on $1024\times1024$ vectors with \texttt{y=length(x)} as input.}
+\end{table}
+\vspace{-10}
+More elements than thread group size require pyramid schemes.
+\end{frame}
+
+\begin{frame}
+
+\textbf{Sequential work}: multiple elements per thread.\begin{itemize}
+  \item Memory latency hiding (Long Scoreboard up)
+  \item Higher reduction factor each dispatch $\Rightarrow$ Less global memory indirections
+  \item Unrolling loops can help but adds register pressure.
+\end{itemize}
+\begin{table}
+  \footnotesize
+  \begin{tabular}{ | l | r | r | r | r | r | r | r |}    \hline
+    Method         &  $\Delta t_c$ & Speedup &VRAM              & SM              & Sel. Warp-Stall Reas.\\\hline
+    Sequential     & 141                   & 2.16x   &49.8\%   & 37.1\%         & S. Scoreb.(2.0\%) \\\hline\hline
+    Seq. (2x)      & 100                   & 3.05x   &69.5\%            & 26.2\%         & L. Scoreb.(80.1\%)\\\hline
+    Seq. (128x)    & 98                    & 3.1x    &\textbf{72.9}\%   & 16.9\%         & \textbf{L. Scoreb.(84.4\%)}\\\hline
+    Seq. (256x)    & 101                   & 3.0x    &66.4\%            & 14.6\%         & L. Scoreb.(76.9\%)\\
+    \hline
+  \end{tabular}
+  \caption{\footnotesize Parallel reduction on $1024\times1024$ vectors with \texttt{y=length(x)} as input. Methods have 504, 8, 4 thread groups, respectively. A GTX970 has 13 SMs.}
+\end{table}
+\vspace{-10}
+Scan is similar but cannot profit as much from sequential work having to keep multiple elements in register memory.
+\end{frame}
+\subsection{Binning \& Counting Sort: Where Are You?}
+\begin{frame}
+\frametitle{Binning \& Counting Sort: Where Are You?}
+
+Grid node does not know its neighboring particles $\Rightarrow$ Binning.
+\\\vspace{10}
+Binning combines nicely with \textbf{Counting Sort:}\\
+\begin{minipage}{0.55\textwidth}
+\begin{enumerate}
+  \item \textbf{Binning}: Per node counting.
+  \item \textbf{Scan}: Computes new memory offset for particles.
+  \item \textbf{Reordering}: Give back indexing list or do deep copy.
+\end{enumerate}
+\end{minipage}
+\begin{minipage}{0.40\textwidth}
+  \footnotesize
+ \includesvg[svgpath =../images/,width=\textwidth]{binning}
+ \caption{\footnotesize Binning: Particle adds one
+to counter after storing it as offset.}
+ \end{minipage}
+\end{frame}
+
+\begin{frame}
+Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
+\begin{enumerate}
+  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
+  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
+\end{enumerate}
+\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
+\begin{table}
+  \footnotesize
+  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
+    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
+    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
+    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
+    \hline
+ \end{tabular}
+\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
+\end{table}
+\end{frame}
+\subsection{The MPM Specific Transfers}
+\begin{frame}
+\frametitle{The MPM Specific Transfers}
+All MPM operations belong to one of those parallelization schemes:
+\begin{itemize}
+ \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
+ \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
+ \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
+ \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
+\end{itemize}
+\vfill
+MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
+$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
 \end{frame}

### 2019-02-13 pres/pres.tex
pres. finish

@@
 \usepackage{media9}
+\usepackage{setspace}
 
@@
  \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
+ \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
  \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
- \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \end{itemize}
@@
 \end{frame}
+\begin{frame}
+\begin{minipage}{\textwidth}
+  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
+  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
+  Pull-Operations read!
+\end{minipage}
+\end{frame}
+\begin{frame}
+Similar to filter or stencil operations on the GPU:
+\begin{itemize}
+  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
+  \item Interpolation function however dependent on particle position.
+  \item Needs to be rerun for every particle in the cell.
+\end{itemize}
+\vfill
+Typical setup of transfers:
+\begin{enumerate}
+  \item Initialize shared memory (pull: with nodes from global memory).
+  \item Perform transfers.
+  \item Write back to global memory (push: global atomics since writes on halo).
+\end{enumerate}
+\end{frame}
+\begin{frame}
+\begin{minipage}{\textwidth}
+  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
+  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
+  Pull-Operations read, Push-operations write!\\
+  $\Rightarrow$ Race conditions.
+\end{minipage}
+\end{frame}
+\begin{frame}
+  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
+\begin{itemize}
+  \item Block is active if any cell counter is active.
+  \item Cell counter is active if it has at least one particle.
+\end{itemize}
+\vfill
+\begin{minipage}{0.9\textwidth}
+  \footnotesize
+  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
+\end{minipage}
+\end{frame}
+\begin{frame}
+\begin{itemize}
+  \item Block and its halo are always target of shared memory operations $\Rightarrow$ P2G-Pull low occupancy.
+  \item Batching multiple particles can increase performance due to hiding synchronization, unroll!
+  \item Transfers respect shared memory bank conflicts fully.
+  \item Warp divergence for varying cell counts.
+\end{itemize}
+
+{\footnotesize
+\begin{table}
+  \begin{tabular}{ | l | r | r | r | r | r |}    \hline
+    Method                 &  $\Delta t_c(\mu s)$ &Speedup         &VRAM  & L2   &SM     \\\hline
+    global                 &   44,442             &-               &4.6 \%&34.4\%&7.7\%  \\\hline
+    global sorted          &   20,484		  &2.21x           &7.0 \%&\textbf{44.0}\%&16.1\% \\\hline
+    P2G-sync              &    \textbf{2,595}    &17.47x          &5.9\%&7.6\%  &\textbf{67.0}\%\\\hline
+ \end{tabular}
+\end{table}
+P2G-transfers of one uniformly million particles with 4 particles per cell with random velocities between $v_x,v_y,v_z \in [-1.0;1.0]$ in a $128\times 128\times128$ grid. Block size is (8,4,4).
+}
+
+\end{frame}
 \section{Delving Deeper: Further Opportunities}
+\subsection{A comparison to \cite{Gao:2018}}
+\begin{frame}
+  \frametitle{A comparison to \cite{Gao:2018}}
 
-\subsection{d}
+Largely same decision making:\\
+\begin{table}
+  \begin{tabular}{|l|c|c|}
+    \hline
+           & Me              & \cite{Gao:2018} \\\hline
+    Sort   & Count/Histogram & Count/Histogram \\
+	   & for each var.   & sel. variables  \\\hline
+    Filtering domain& Filter-op.& Sparse Grid structure \\\hline
+    Transfers        & Shared mem. only & Warp-shuffle op.\\\hline
+  \end{tabular}
+\end{table}
+Warp shuffle allows for fast parallel segmented reduction of cells of varying counts.\\
+$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster in these instances but NVIDIA only.
 
-\begin{frame}
-\frametitle{hi}
-  \begin{enumerate}
-    \item \cite{MPM:COURSE}
-  \end{enumerate}
 \end{frame}

### 2019-02-13 pres/pres.tex
last apge

@@
 \begin{itemize}
-  \item Block and its halo are always target of shared memory operations $\Rightarrow$ P2G-Pull low occupancy.
-  \item Batching multiple particles can increase performance due to hiding synchronization, unroll!
+  \item Block and halo are always target of shared memory operations $\Rightarrow$ P2G-pull low occupancy.
+  \item \textbf{Batching} multiple particles can increase performance due to hiding synchronization, unroll!
   \item Transfers respect shared memory bank conflicts fully.
-  \item Warp divergence for varying cell counts.
+  \item \textbf{Warp divergence} for varying cell counts.
 \end{itemize}
@@
 
-Largely same decision making:\\
+Simultaneously being worked on. Largely same decision making:\\
 \begin{table}
-  \begin{tabular}{|l|c|c|}
+\begin{tabular}{|c|c|c|}
     \hline
-           & Me              & \cite{Gao:2018} \\\hline
-    Sort   & Count/Histogram & Count/Histogram \\
+	   & \textbf{Me}              & \textbf{\cite{Gao:2018}} \\\hline
+  \textbf{Sort}   & Count/Histogram & Count/Histogram \\
 	   & for each var.   & sel. variables  \\\hline
-    Filtering domain& Filter-op.& Sparse Grid structure \\\hline
-    Transfers        & Shared mem. only & Warp-shuffle op.\\\hline
+  \textbf{Filtering domain}& Filter-op.& Sparse Grid structure \\\hline
+  \textbf{Transfers}        & Shared mem. only & Warp-shuffle op.\\\hline
   \end{tabular}
 \end{table}
-Warp shuffle allows for fast parallel segmented reduction of cells of varying counts.\\
-$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster in these instances but NVIDIA only.
-
+\textbf{Warp shuffle} allows for fast parallel segmented reduction of cells of varying counts.\\
+$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster for varying counts but \textbf{NVIDIA only}.
+\end{frame}
+\begin{frame}
+  \centering{\textbf{Thank you for your attention!}}
+  \vfill
+\begin{itemize}
+  \item \textbf{Bachelor thesis}: \url{https://github.com/MeyerFabian/snow}
+  \item \textbf{Master thesis}: \url{https://github.com/MeyerFabian/msc}
+  \item \textbf{Presentation}: \url{https://github.com/MeyerFabian/msc/pres}
+      \item \textbf{Code} (right now): \url{https://github.com/mpm-msc/snow}
+\end{itemize}
+\vfill
+\centering
+\textbf{Questions?}
 \end{frame}

### 2019-09-23 pres/pres.tex
redid some things

@@
 \section{Motivation on a GPU MPM Approach}
+\subsection{GPGPU for performance enthusiasts}
+\begin{frame}
+\frametitle{GPGPU for performance enthusiasts}
+  Why would('nt) you?
+  \vfill
+  \begin{columns}[T]
+    \begin{column}{0.45\textwidth}
+      \textbf{Drawbacks:}
+    \begin{itemize}
+      \item Interactivity much easier on CPU, but slow PCI-Bus communication
+      \item Code is mostly written against GPU architecture
+      \item A lot of strain on the programmer
+    \end{itemize}
+    \end{column}
+    \begin{column}{0.45\textwidth}
+      \textbf{Benefits:}
+    \begin{itemize}
+      \item Data is already on the GPU for rendering
+      \item \textbf{Higher parallelization acceleration}
+    \end{itemize}
+    \end{column}
+  \end{columns}
+\end{frame}
 
@@
 \begin{frame}
+\begin{minipage}{0.45\paperwidth}
 PIC ideas:
-\vfill
-\begin{minipage}{0.45\paperwidth}
 \begin{itemize}
-    \item Combine Lagrangian particles \& Eulerian grid    \item Particles store all information
+    \item Particles store all information
+    \item Eulerian grid as a non-deformable scratchpad
+    \item[$\Rightarrow$] \textbf{meshfree}
 \end{itemize}
@@
 Typical PIC/MPM roundtrip:
-\vfill
 \begin{enumerate}
-  \item Particle-to-grid(P2G) transfer to an unmoving grid
+  \item Transfer particle quantities to the grid (\textit{P2G})
   \item Solve discretized governing equations on grid
-    \item Grid-to-particle(G2P) transfer back to particles \& move them
-    \item[$\Rightarrow$] \textbf{meshfree, non-empirical}
+  \item Transfer back (\textit{G2P})
 \end{enumerate}
 \end{minipage}
-\begin{minipage}{0.35\paperwidth}
-	\includesvg[width=0.35\paperwidth]{media/pic.svg}
-	\tiny{Transfers: Interpolation functions are defined over grid nodes.}
+\begin{minipage}{0.30\paperwidth}
+\begin{minipage}{1.0\textwidth}
+  \small
+  \includesvg[width=1.0\textwidth]{media/pic.svg}
+  P2G-transfer $\sum_p$
+\\
 \end{minipage}
-\end{frame}
-\subsection{GPGPU for performance enthusiasts}
-\begin{frame}
-  \frametitle{GPGPU for performance enthusiasts}
-  Why would('nt) you?
-  \vfill
-  \begin{columns}[T]
-    \begin{column}{0.45\textwidth}
-      \textbf{Drawbacks:}
-    \begin{itemize}
-      \item Interactivity much easier on CPU, but slow PCI-Bus communication
-      \item Code is mostly written against GPU architecture
-      \item A lot of strain on the programmer
-    \end{itemize}
-    \end{column}
-    \begin{column}{0.45\textwidth}
-      \textbf{Benefits:}
-    \begin{itemize}
-      \item Data is already on the GPU for rendering
-      \item \textbf{Higher parallelization acceleration}
-    \end{itemize}
-    \end{column}
-  \end{columns}
+\begin{minipage}{1.0\textwidth}
+  \small
+  \includesvg[width=1.0\textwidth]{media/g2p.svg}
+  G2P-transfer $\sum_i$
+\end{minipage}
+\end{minipage}
 \end{frame}
@@
 
-\subsection{The Pretty Strong but Mathematically Weak Formulation}
-\begin{frame}
-\frametitle{The Pretty Strong but Mathematically Weak Formulation}
+\subsection{Discretization of Space and Time}
 
-\textbf{Weak Formulation} (or Principle of Virtual Work):\\
-Dot product equations with arbitrarily 'test functions' $\boldsymbol{q}$ and apply divergence theorem:
-\begin{align}
-\int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot  \Bigl[(_0\rho_0)(_0\boldsymbol{a}) - & {}_0\boldsymbol{f}^{\text{body}}{}_0^tJ\Bigr] d _0\boldsymbol{x}= \nonumber \\
-										     &  \int _ { \partial \Omega ^ { t^n } } _t\boldsymbol{q} \cdot \boldsymbol{\sigma} d_t\boldsymbol{A} - \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
-\end{align}
-A strong solution is also a solution to the weak formulation. Leave out body forces(like gravity) and boundary condition (e.g. collisions) for now:
-\begin{equation}
-  \int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot (_0\rho_0)(_0\boldsymbol{a}) d _0\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
-\end{equation}
-\end{frame}
-\begin{frame}[fragile]
-\textbf{Weak Derivative}:
-\vfill
-\begin{minipage}{0.49\textwidth}
-  $y=|t|$ has weak derivative:
-  \\\\
-$ v = \left\{ \begin{array} { l l }
-{ - 1 , } & { \text { if } t < 0 } \\
-{ c , } & { \text { if } t = 0 } \\
-{ 1 , } & { \text { if } t > 0 }
-\end{array} \right.$
-\\
-
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\begin{tikzpicture}
-  \small
-\datavisualization [school book axes,
-		    y axis = {attribute = y, label},
-		    visualize as line/.list={fx},
-		    fx={style={color=blue}}
-]
-
-data [set=fx,format=function] {
-      var x : interval [-1.5:1.5] samples 5;
-      func y = abs(\value x);
-      };
-\end{tikzpicture}
-\end{minipage}
-
-\vfill
-\begin{minipage}{0.49\textwidth}
-  Heaviside step function has no weak derivate: \\\\
-  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\begin{tikzpicture}
-  \small
-  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
-                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
-                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
-            \end{axis}
-\end{tikzpicture}
-\end{minipage}
-\vfill
-Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
-\end{frame}
-\subsection{Discretization of Space and Time}
 \begin{frame}
 \frametitle{Discretization of Space and Time}
-\textbf{Time discretization} with implicit midpoint scheme:
-  \begin{equation}
-    \frac { y ^ { n + 1 } - y ^ { n } } { \Delta t } = f ^ { n + \frac{1}{2} } = f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)
-  \end{equation}
+\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$.
+Here dyadic products of cubic b-splines suffice.
+$w_i$ should satisfy at least \cite{gao2017adaptive}:
 \begin{itemize}
-  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
-  \item \textbf{midpoint} as it conserves gov. equations
+  \item Partition of unity: $\sum_i w (\boldsymbol{x}-\boldsymbol{x}^n_i)=1$
+  \item Identity relation: $\sum_i \boldsymbol{x}_i w (\boldsymbol{x}-\boldsymbol{x}^n_i)=\boldsymbol{x}$ $(\boldsymbol{x}=\boldsymbol{x}_p)$
+  \item Non-negativity: $w \geq 0$.
+  \item Limited support.
+  \item $C^1$-continuity.
 \end{itemize}
-
-  \begin{equation}
-    \Rightarrow \int _ { \Omega ^ { t } } _t\boldsymbol{q} \cdot {}_t\rho(_t\boldsymbol{v}^{n+1}-{}_t\boldsymbol{v}^{n}) d _t\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}^{n+\frac{1}{2}}  d_t\boldsymbol{x}.
-  \end{equation}
-\end{frame}
-
-\begin{frame}[fragile]
-\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$ with limited support. Here dyadic products
-\begin{equation}
-  w_i(\boldsymbol{x}) =w(\boldsymbol{x}-\boldsymbol{x}_i) = w(\frac{1}{h}(x-x_i))w(\frac{1}{h}(y-y_i)w(\frac{1}{h}(z-z_i))
-\end{equation}
-of cubic b-splines suffice:
-\begin{equation}
-w ( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array}.
-\right
-\end{equation}
-\begin{tikzpicture}[
+Shortening $w^n_{ip} = w(\boldsymbol{x}_p^n - \boldsymbol{x}_i^n).
+\begin{tikzpicture}[remember picture,overlay,
   declare function={
@@
 \begin{axis}[
-  width=10cm,height=4cm,
+  width=7.5cm,height=3.5cm,
   axis x line=middle, axis y line=middle,
@@
 \end{frame}
+
 \begin{frame}
-Thus set in:
-$$
-{}_t\boldsymbol{q}(\boldsymbol{x},t^n) = \sum_i \boldsymbol{e}_iw_i(\boldsymbol{x}),
-{}_t\boldsymbol{v}^{n(+1)}(\boldsymbol{x}) = \sum_j \boldsymbol{v}^{n(+1)}_jw_j(\boldsymbol{x}).
-$$
-Combine it with numerical integration where the particles function as quadrature points \cite{Steffen}:
+Transfer quantities from particles to grid. Numerical integration where the particles function as quadrature points \cite{Steffen}:
 
-\begin{minipage}{0.6\textwidth}
-
-$$
-g _ { i } = \int _ { \Omega } g ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \boldsymbol{x}
-$$
-\begin{equation}
-  \approx \sum _ { p } g _ { p }  w_i(\boldsymbol{x}_p) V _ { p }.
-\end{equation}
-due to integration by midpoint rule:
-\begin{equation}
-\int _ { \Omega } f ( x ) d x \approx \sum _ { i = 1 } ^ { N } f \left( x _ { i } \right) h _ { i }.
-\end{equation}
-
+$$m_i = \int_\Omega \rho(\boldsymbol{x})w_i(\boldsymbol{x}) d\Omega \approx \sum_p \rho_p w_{ip} V_p \approx \sum_p m_p w_{ip}$$
+$$A _ { i } = \int _ { \Omega }  A ( \boldsymbol{x} ) w _ { i } ( \boldsymbol{x} ) d \Omega \approx \sum _ { p } A _ { p }  w_{ip} V _ { p }.$$ \\
+\begin{minipage}{0.62\textwidth}
+APIC-transfers add a local velocity field $\boldsymbol{C}_p$ around $\boldsymbol{v}_p$:
+\\\vspace*{-10}\\
+$ ( m \boldsymbol{v} ) _ { i } = \sum _ { p } w _ { i p } m _ { p } \left( \boldsymbol{v} _ { p } + \boldsymbol{C} _ { p } ( \boldsymbol{x} _ { i } - \boldsymbol{x} _ { p } ))$
 \end{minipage}
-\begin{minipage}{0.3\textwidth}
-  \begin{tikzpicture}[/pgf/declare function={f=-.1*(x)*(x-8);}]
-\pgfplotsset{
-    integral segments/.code={\pgfmathsetmacro\integralsegments{#1}},
-    integral segments=3,
-    integral/.style args={#1:#2}{
-        ybar interval,
-        domain=#1+((#2-#1)/\integralsegments)/2:#2+((#2-#1)/\integralsegments)/2,
-        samples=\integralsegments+1,
-        x filter/.code=\pgfmathparse{\pgfmathresult-((#2-#1)/\integralsegments)/2}
-    }
-}
-\begin{axis}[
-    ylabel=$f$,
-    width =5cm,height=5cm,
-    domain=0:6,
-    samples=100,
-    axis lines=middle
-]
-\addplot [thick] {f};
-\addplot [
-    red,
-    integral=0:6
-] {f};
-\end{axis}
-\end{tikzpicture}
-\end{minipage}
-\end{frame}
-  \subsection{Velocity Fields: APIC-Transfers}
-\begin{frame}
-  \frametitle{Velocity Fields: APIC-Transfers}
-  \begin{minipage}{0.64\textwidth}
-  PIC-transfer
-  \begin{enumerate}
-    \item $(m\boldsymbol{v})_i^n = \sum_p w_{ip}^nm_p\boldsymbol{v}^n_p$
-    \item $\boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}$
-    \item $\boldsymbol{v}_{p,{PIC}}^{n+1} = \sum_i w^n_{ip}\boldsymbol{v}_i^{n+1}$
-  \end{enumerate}
-\vspace*{15}
-APIC-transfers add a local velocity field $\boldsymbol{C}^n_p$ around $\boldsymbol{v}^n_p$:
-\\\vspace*{-10}\\
-$ ( m \boldsymbol{v} ) _ { i } ^ { n } = \sum _ { p } w _ { i p } ^ { n } m _ { p } \left( \boldsymbol{v} _ { p } ^ { n } + \boldsymbol{C} _ { p } ^ { n } ( \boldsymbol{x} _ { i } ^ { n } - \boldsymbol{x} _ { p } ^ { n }))$
-\end{minipage}
-\begin{minipage}{0.30\textwidth}
+\begin{minipage}{0.33\textwidth}
 \begin{minipage}{1.0\textwidth}
@@
 \end{minipage}
+
+\end{minipage}
+\end{frame}
+
+\begin{frame}
+\begin{minipage}{0.64\textwidth}
+The MPM can be described in the weak formulation (FEM: virtual quantities). This allows us to 'shift' the derivative:\begin{itemize}
+\item $\boldsymbol{A}_p = \sum_i \boldsymbol{A}_i w_{ip} $
+\item $\nabla \boldsymbol{A}_p = \sum_i \boldsymbol{A}_i \nabla w_{ip}^T $
+\item$\nabla \cdot \boldsymbol{A}_p = \sum_i \boldsymbol{A}^T_i \nabla w_{ip} $
+\end{itemize}
+\end{minipage}
+\begin{minipage}{0.30\textwidth}
 \begin{minipage}{1.0\textwidth}
@@
 \end{minipage}
+\vfill
+\textbf{Time discretization} with implicit midpoint scheme:\\
+$v_i ^ { n + 1 } - v_i ^ { n } = \frac{\Delta t}{m_i^n}f_i ^ { n + \frac{1}{2} } =\frac{\Delta t}{m_i^n} f \left( t ^ { n } + \frac{\Delta t}{2}, \frac{1}{2} y ^ { n } + \frac{1}{2} y ^ { n + 1 } \right)$
+\begin{itemize}
+  \item \textbf{implicit} requires linear system solve $\Rightarrow$ more stable, larger time steps
+  \item \textbf{midpoint} as it conserves angular momentum
+\end{itemize}
 \end{frame}
@@
  \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
- \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
+ \item 1 thread : 1 node: $\square_i = \square_i \circ \square_i \circ ... \circ \square_i$.
  \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
@@
 \printbibliography
+
+\begin{frame}
+\frametitle{The Pretty Strong but Mathematically Weak Formulation}
+
+\textbf{Weak Formulation} (or Principle of Virtual Work):\\
+Dot product equations with arbitrarily 'test functions' $\boldsymbol{q}$ and apply divergence theorem:
+\begin{align}
+\int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot  \Bigl[(_0\rho_0)(_0\boldsymbol{a}) - & {}_0\boldsymbol{f}^{\text{body}}{}_0^tJ\Bigr] d _0\boldsymbol{x}= \nonumber \\
+										     &  \int _ { \partial \Omega ^ { t^n } } _t\boldsymbol{q} \cdot \boldsymbol{\sigma} d_t\boldsymbol{A} - \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
+\end{align}
+A strong solution is also a solution to the weak formulation. Leave out body forces(like gravity) and boundary condition (e.g. collisions) for now:
+\begin{equation}
+  \int _ { \Omega ^ { 0 } } _0\boldsymbol{q} \cdot (_0\rho_0)(_0\boldsymbol{a}) d _0\boldsymbol{x}= \int _ { \Omega ^ { t^n} } \nabla {}_t\boldsymbol{q} : \boldsymbol{\sigma}  d_t\boldsymbol{x}.
+\end{equation}
+\end{frame}
+\begin{frame}[fragile]
+\textbf{Weak Derivative}:
+\vfill
+\begin{minipage}{0.49\textwidth}
+  $y=|t|$ has weak derivative:
+  \\\\
+$ v = \left\{ \begin{array} { l l }
+{ - 1 , } & { \text { if } t < 0 } \\
+{ c , } & { \text { if } t = 0 } \\
+{ 1 , } & { \text { if } t > 0 }
+\end{array} \right.$
+\\
+
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{tikzpicture}
+  \small
+\datavisualization [school book axes,
+		    y axis = {attribute = y, label},
+		    visualize as line/.list={fx},
+		    fx={style={color=blue}}
+]
+
+data [set=fx,format=function] {
+      var x : interval [-1.5:1.5] samples 5;
+      func y = abs(\value x);
+      };
+\end{tikzpicture}
+\end{minipage}
+
+\vfill
+\begin{minipage}{0.49\textwidth}
+  Heaviside step function has no weak derivate: \\\\
+  $s= \left\{ \begin{array} { l l } { 0 , } & { \text { if } t < 0 } \\ { 1 , } & { \text { if } t \geq 0 } \end{array} \right.$
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{tikzpicture}
+  \small
+  \begin{axis}[axis lines = middle, width=5.1cm,height=3.5cm,xtick={-1,0,1},ytick={0,1},ymin=-0.3,ymax=1.25,xmin=-1.8,xmax=1.8,ylabel=$s$]
+                \addplot[blue,mark=*,mark options={fill=white},samples at={-2.1,0}] {0};
+                \addplot[blue,mark=*,samples at={0,2.1}] {1.0};
+            \end{axis}
+\end{tikzpicture}
+\end{minipage}
+\vfill
+Allows for point loads, material discontinuities and more. \cite{bathe2006finite}
+\end{frame}
+
+\begin{frame}[fragile]
+\textbf{Space Discretization} is done in a Galerkin/FEM fashion with grid based interpolants $w_i$ with limited support. Here dyadic products
+\begin{equation}
+  w_i(\boldsymbol{x}) =w(\boldsymbol{x}-\boldsymbol{x}_i) = w(\frac{1}{h}(x-x_i))w(\frac{1}{h}(y-y_i)w(\frac{1}{h}(z-z_i))
+\end{equation}
+of cubic b-splines suffice:
+\begin{equation}
+w ( x ) = \left\{ \begin{array} { l l } { \frac { 1 } { 2 } | x | ^ { 3 } - | x | ^ { 2 } + \frac { 2 } { 3 } } & { 0 \leqslant | x | < 1 } \\ { \frac { 1 } { 6 } ( 2 - | x | ) ^ { 3 } } & { 1 \leqslant | x | < 2 } \\ { 0 } & { 2 \leqslant | x | } \end{array}.
+\right
+\end{equation}
+\begin{tikzpicture}[
+  declare function={
+    func(\x)= and(abs(\x) >= 0 , abs(\x)<1) * (pow(abs(\x),3)/2 - pow(abs(\x),2) +2/3)  +
+	      and(abs(\x) >= 1 , abs(\x)<2) * (1/6* pow((2-abs(\x)),3))
+   ;
+  }
+]
+\begin{axis}[
+  width=10cm,height=4cm,
+  axis x line=middle, axis y line=middle,
+  ymin=-0.1, ymax=1, ytick={0,1}, ylabel=$w$,
+  xmin=-2.5, xmax=2.5, xtick={-2,...,2}, xlabel=$x$,
+  domain=-2.5:2.5,samples=101, % added
+]
+
+\addplot [blue,thick] {func(x)};
+\end{axis}
+\end{tikzpicture}
+\end{frame}
+\section*{Rest}
 \end{document}
